### 2020-04-27 manuscript/main.tex
manuscript before preprint upload, including result data, figure code and figure files

\documentclass[fleqn,12pt,inline]{wlscirep}

\usepackage{xr-hyper}

\newcommand{\Prob}{{\mathbf{P}}}
\newcommand{\cR}{{\mathfrak R}}
\newcommand{\cB}{{\mathbf{B}}}

\usepackage{euscript}

%\usepackage[left]{lineno}
%\linenumbers

\usepackage{setspace}
\doublespacing

% xr helper code to generate dependency definition command \myexternaldocument
\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

\myexternaldocument{supplement}

\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data}
\author[1,2,3,4]{David Lähnemann}
\author[4,5]{Johannes Köster}
\author[3]{Ute Fischer}
\author[3]{Arndt Borkhardt}
\author[1,2,+,*]{Alice C. McHardy}
\author[5,+,*]{Alexander Schönhuth}
\affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
\affil[2]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[3]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
\affil[4]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
\affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
\affil[+]{shared last authors}
\affil[*]{To whom correspondence should be addressed: \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de}, \href{as@cwi.nl}{as@cwi.nl}.}

\begin{abstract}

Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which when applied to single cells generate significant numbers of false positives and false negatives.
Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.

We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified single cell DNA sequencing data.
It introduces a mechanistically motivated empirical model of amplification bias that improves the quantification of genotyping uncertainty.
To account for amplification errors, it jointly models the single cell sample with a bulk sequencing sample from the same cell population---also enabling a biologically relevant imputation of missing genotypes for the single cell.
Through these innovations, ProSolo achieves substantially higher performance in calling and genotyping single nucleotide variants in single cells in comparison to all state-of-the-art tools. 
Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.

ProSolo's model is implemented into a flexible framework, encouraging extensions. The source code and usage instructions are available at: \url{https://github.com/prosolo/prosolo}
\end{abstract}

\begin{document}

\flushbottom
\maketitle

\thispagestyle{empty}

\section{Introduction}

Originally, genome sequences have been queried for genetic germline variation or for highly abundant somatic variation, for example in cancer.
The advent of high-throughput single cell sequencing has recently turned the spotlight on a type of so far understudied variation: 
the often less abundant somatic or post-zygotic variation that constantly accumulates with every mitotic cell division throughout the lifetime of an organism, turning every individual into a complicated genomic mosaic\cite{forsberg_mosaicism_2017}.
Estimates for somatic single nucleotide variants (SNVs) range from around $0.6 \cdot 10^{-9}$  up to $60 \cdot 10^{-9}$  mutations per genome position per cell division\cite{lynch_rate_2010,lynch_evolution_2010,tomasetti_half_2013,li_somatic_2013}, with a recent estimate\cite{milholland_differences_2017} based on single-cell sequencing at $2.66 \cdot 10^{-9}$.
With the size of the human (reference) genome at approximately $3.2 \cdot 10^9$ base pairs, these numbers indicate that even during healthy development, most cells harbour cell-specific point mutations.
This enables retrospective monitoring of lineages involved in normal organism development, merely by sampling some cells\cite{kester_single-cell_2018}, without having to interfere with its general development, or kill the individual.
In other words, this establishes a universally applicable methodology for \textit{in vivo} lineage tracing.

In cancer development, this variation can be used to trace the cellular ancestry of tumour subclones and metastases\cite{ross_onconem:_2016,zafar_sifit:_2017,malikic_integrative_2019,kuipers_advances_2017}, and to characterise the evolutionary dynamics of cancer progression\cite{skums_inference_2019,lahnemann_eleven_2020}.
In the long run, methods that account for the dynamics of mutational signatures in cellular evolution will improve diagnosis, treatment and prognosis of diseases for which somatic alterations are a key factor.
To this end, obtaining accurate profiles of the genetic variation affecting single cells is essential.

In sequencing libraries prepared directly from single cells, only a small fraction of the genome is sampled.
To obtain coverage levels that allow for the consistent identification of SNVs across larger parts of the genome, \textit{in vitro} whole genome amplification is crucial. 
Among whole genome amplification methods, multiple displacement amplification (MDA\cite{dean_comprehensive_2002}) has proven the least error-prone and is therefore considered the state-of-the-art in single cell SNV profiling\cite{de_bourcy_quantitative_2014,hou_comparison_2015,huang_single-cell_2015,estevez-gomez_comparison_2018}.
But, although the type of polymerase used in MDA ($\Phi29$) has the highest fidelity currently attainable (due to its proof-reading functionality), amplification errors still occur at a rate of $1.24 \cdot 10^{-6}$ to $9.5 \cdot 10^{-6}$ per copied base\cite{paez_genome_2004,wang_genome-wide_2012,wang_clonal_2014,de_bourcy_quantitative_2014,lodato_somatic_2015}---three orders of magnitude higher than the estimates for the somatic mutation rate.
Further, the efficiency and fidelity of $\Phi29$ polymerase depends on the template sequence context\cite{de_paz_high-resolution_2018}, implying that the amplification error rate systematically varies around this average.
Moreover, the degree of amplification depends on the quality of the template DNA extracted from the single cell\cite{baumer_exploring_2018} and how accessible each stretch of DNA is to amplification initiation via priming\cite{picher_trueprime_2016}.
As a result, sequencing coverage after amplification differs both between sites along the genome and between the two alleles at a particular site, up to the entire dropout of alleles\cite{spits_optimization_2006}.
Because standard variant callers assume that alleles are uniformly covered, they do not perform well on the resulting data and are substantially outperformed by  single-cell  variant callers\cite{zafar_monovar:_2016,dong_accurate_2017}.
Clearly, when calling SNVs for single cells, the statistical uncertainties introduced by the amplification need to be dealt with at the largest possible accuracy.

Thus, variant callers for whole-genome-amplified single-cell data need to account for both amplification errors and allelic bias, in addition to accounting for the site-specific variation.
However, state of the art single cell SNV callers routinely assume fixed global rates when modeling uneven allelic coverage (up to dropout) and amplification errors. 
For example, to reflect the amplification error rate, both MonoVar\cite{zafar_monovar:_2016} and SCcaller\cite{dong_accurate_2017} work with global false positive error rates for calling the presence of an alternative allele at a particular site.
This assumes that $\Phi29$ polymerase is agnostic to local template sequence context, although it is not\cite{de_paz_high-resolution_2018}.
Similarly, for modelling allele dropout, MonoVar\cite{zafar_monovar:_2016} and SCIPhI\cite{singer_single-cell_2018} assume that one rate applies globally (and across all cells).
This neglects that allele dropout, as the extreme case of uneven allele coverage, varies greatly along the genome and in particular also between cells, because their DNA is amplified separately.
Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
Improving on that point, tools exist that account for local variation in allelic amplification bias. SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019} both estimate the minor allele frequency from nearby germline heterozygous sites. 
However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
Thus, to the best of our knowledge, there is no statistical model that allows for local variation of bias and errors due to amplification, and for statistically sound false discovery control when calling and genotyping SNVs in single cells.

We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
Importantly, our model further enables a computationally efficient implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_enhancing_2019}.
ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.

\section{Results}

\subsection{Single Cell Sequencing Model}

We describe a novel probabilistic model that addresses the genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
In the following, we introduce the central innovations of our model and demonstrate its advantages in comparison to existing approaches.
More details and a detailed derivation of all model elements can be found in the Supplement, including a summary of the core model.

Our model addresses the two major issues of MDA: (i) the differential amplification of the two alleles present in a diploid cell ("amplification bias" in the following); (ii) MDA induced errors ("amplification errors" in the following) which are copy errors introduced by the $\Phi29$ polymerase used in MDA. To address amplification bias, we leverage a mechanistically motivated, empirically derived model of differential amplification of alleles.
To assess amplification errors, we evaluate single cell samples together with a bulk sample from which the single cell is supposed to stem.
Regarding the latter, we argue that a bulk sample should be added to single cell sequencing experiments wherever possible: it samples from the same cell population without requiring amplification, and is therefore unaffected by amplification bias and errors and thus makes a particularly useful background sample to address the statistical uncertainties and biases induced by MDA.
At the same time, one of the major features of the core model and its implementation is that it can easily be adapted to flexibly deal with other sampling setups, so it could be extended to further scenarios.
For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_enhancing_2019}.

\begin{figure}[!tpb]%figure1
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
    \end{minipage}
  \end{minipage}
  \vspace{2ex}
  \hspace{1ex}
  \begin{minipage}[t]{.99\linewidth}
    \begin{minipage}{.47\linewidth}
      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
    \end{minipage}
    \hspace{2ex}
    \begin{minipage}{.49\linewidth}
      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
    \end{minipage}
  \end{minipage}
  \caption{
   \textbf{A,B} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
   Homozygous reference sites in \textbf{(A)} are assumed to follow a beta-binomial distribution;
   sites heterozygous for the alternative allele in \textbf{(B)} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
   \textbf{(C)} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs. $\theta_s = 0.5$).
   Alternative nucleotide T (orange) is an amplification error.
   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
   This is formalised with the model in D.
   \textbf{(D)} Definition of single cell events based on true underlying alternative allele frequencies in the single cell ($\theta_s$) and the bulk ($\theta_b$) (assuming that the bulk sample has a deep coverage that captures somatic variants).
   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).\small\newline
   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
  \label{fig:prosolo_alt-calling}
\end{figure}

\paragraph{A mechanistically motivated model of amplification bias, trained on data, gives realistic coverage-specific single-cell (genotype) likelihoods.}
To account for MDA amplification bias up to the complete dropout of individual alleles, we distinguish between two alternative allele frequencies:
\begin{enumerate*}[label={(\roman*)}]
  \item The theoretical underlying allele frequency at a site in a single cell: $\theta_s$.
   This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
   However, the ratio of reads harboring the different alleles from a single cell sequencing experiment does not reflect the true allele frequency, because of the biases induced by amplification. 
  Instead, the ratio of reads reflects
  \item the allele frequency after its distortion through amplification bias.
   For a site with total coverage of $l$ reads, of which $k$ reads bear the alternative allele, the formal definition of this measurable frequency is $\rho_s=\frac{k}{l}$.
\end{enumerate*}
To accurately quantify the uncertainty introduced by amplification bias, we consider

\begin{equation}
 \mathbf{P}(\rho_s\mid\theta_s)\qquad \text{for}  \quad\theta_s\in\{0,0.5,1\},
\end{equation}

the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele.

To do this, we follow the considerations of Lodato et al.\cite{lodato_somatic_2015} (see Figure S5 and the section ``Modeling MDA-derived alternative read counts'' of the respective supplement for the details), who fitted well-studied probability distributions to empirical distributions they obtained for $\mathbf{P}(\rho_s\mid\theta_s)$ of $\theta_s\in\{0,0.5\}$.
For sites that are homozygous for the reference allele, amplification bias cannot initially happen.
However, once an amplification error creates an alternative allele, this can be amplified to large frequencies due to amplification bias.
Lodato et al.\cite{lodato_somatic_2015} thus consider the empirical distribution $\mathbf{P}(\rho_s\mid\theta_s=0)$ to follow a beta-binomial.
Effectively, this means that the probability of a non-zero alternative read count  ($\rho_s>0$, which is any alternative read count above $0$ in Figure~\ref{fig:prosolo_alt-calling}A) will be non-zero ($\mathbf{P}(\rho_s\mid\theta_s=0)>0$), merely because of sequencing and amplification errors.
In contrast, at heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B), the distribution is dominated by amplification bias.
Thus, for $\mathbf{P}(\rho_s\mid\theta_s=0.5)$ they found a mixture of two beta-binomial distributions to appropriately fit the empirically observed distributions (Figure~\ref{fig:prosolo_alt-calling}B).
We further motivate the choice of the beta-binomial distribution mechanistically by an analogy to its generative urn model named after Pólya\cite{eggenberger_uber_1923}:
Take an urn with two white and two black balls, where each ball represents one strand of each (double-stranded) allele at a heterozygous site.
In the Pólya urn model, drawing a ball leads to replacement of that ball with two balls of the same color, analogous to a strand copy by the $\Phi29$ polymerase (additional discussion in the Supplement, Section~\ref{sec:polya-urn}).
Finally, the necessity for a mixture of beta-binomials (Figure~\ref{fig:prosolo_alt-calling}B) becomes evident when contrasting it with the binomial distribution observed at heterozygous sites in bulk experiments.
Namely, the bulk distribution for reads supporting the alternative allele would narrowly peak at a count of $k=\frac{l}{2}$, which is not the case for the corresponding $k=10$ in Figure~\ref{fig:prosolo_alt-calling}B.
Instead, the mixture of beta-binomials peaks at the extreme read counts of $k=0$ and $k=l$, highlighting that the dropout of the alternative or reference allele is quite likely to occur.
Most likely, this mixture arises from differences in the accessibility of different pieces of the genomic DNA for amplification\cite{picher_trueprime_2016} (Supplement, Section~\ref{sec:polya-urn}).
Finally, for homozygous alternative sites (i.e. $\mathbf{P}(\rho_s\mid\theta_s=1)$), we rely on the symmetry of the cases $\theta_s=0$ and $\theta_s=1$.
In summary, we obtain the following equations for $\mathbf{P}(\rho_s\mid\theta_s)$:

\begin{equation}
% TODO: center equations
	\label{eq:BBLodato}
	\begin{split}
		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 0 ) 			=~ &\cB\cB(~ k,~ l;~ \alpha(l),~ \beta(l)~ ),\\
		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = \frac12) =~ &w(l) \times \cB\cB(~ k,~ l;~ \alpha_1(l),~ \alpha_1(l)~ )~ +\\
                                                       ~&~(~ 1-w(l)~ ) \times \cB\cB(~ k,~ l;~ \alpha_2(l),~ \alpha_2(l)~ ),\\
		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 1 ) 			=~ &\cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ).
	\end{split}
\end{equation}

Here, $\cB\cB$ represents the beta-binomial probability mass function where $k\in\{0,...,l\}$ with $l$ being the total read coverage of the site that is considered. All of parameters $\alpha, \alpha_1,\alpha_2$, $\beta$, and $w$ scale linearly in $l$.
Allowing to vary distributions through these parameters allows amplification bias to depend on the total coverage of a site, and thereby to vary locally.
Symmetry of distributions for the homozygous sites ($\theta_s = 0$ in Figure~\ref{fig:prosolo_alt-calling}A, and $\theta_s = 1$) is established by swapping the shape parameters $\alpha(l)$ and $\beta(l)$.
The distribution for heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B) corresponds to a mixture of two beta-binomials with shape parameters $\alpha_1(l)=\beta_2(l)$ and $\alpha_2(l)=\beta_2(l)$, where equality of $\alpha(l)$ and $\beta(l)$ yields symmetry of the distributions relative to $k = \frac{l}{2}$ (the expected value for alternative read counts at heterozygous sites).
Slopes and intersects for the scaling of all parameter values across different choices of $l$ are given in Table~\ref{tab:Lodato}.

\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
This establishes a formal, statistical argument to why one should consider bulk experiments in single cell sequencing whenever possible and why the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.

For our model, we derive likelihoods for all possible alternative allele frequencies in the background bulk sample.
Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:

\begin{equation}
 \label{eq:bulk-prob}
 L\left(\theta_b = \frac{m}{n} \mid \boldsymbol{Z}^b\right) \propto
 P\left(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n}\right) =
 \prod_{j=1}^n P\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)
\end{equation}

Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_enhancing_2019}.

\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihoods works for calling genotypes in the single cell.
However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:

\begin{equation}
  \label{eq:event-space}
  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\theta_s} \times [0, 1]_{\theta_b}.
\end{equation}

Within this space, we define single cell events as mutually exclusive subspaces, for example an error-free homozygous alternative site is defined by allele frequencies $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\theta_s} \times (\frac{1}{2}, 1]_{\theta_b}$  or the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):

\begin{equation}
  \label{eq:event-set}
  \begin{split}
    \mathfrak{E} &= \{ \boldsymbol{E}_{\text{hom ref}}, \boldsymbol{E}_{\text{err alt}}, \boldsymbol{E}_{\text{ADO to alt}}, \boldsymbol{E}_{\text{het}}, \boldsymbol{E}_{\text{ADO to ref}}, \boldsymbol{E}_{\text{err ref}}, \boldsymbol{E}_{\text{hom alt}} \}
  \end{split}
\end{equation}

Assuming a flat prior across the possible underlying allele frequencies for both the bulk and the single cell, we can compute likelihoods for all those single cell events (e.g.~Equation~\ref{eq:ado-to-ref-likelihood-data}).
The sum of the likelihoods of all these (mutually exclusive) events yields the marginal probability (Equation \ref{eq:marginal-prob}).
Using the marginal probability, we can calculate the posterior probability for any of these events.
For example, the posterior probability of event $\boldsymbol{E}_{\text{hom alt}}$ (Figure~\ref{fig:prosolo_alt-calling}D) can be calculated with:

\begin{equation}
 \label{eq:hom-alt-posterior-samples}
 \begin{split}
  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
      \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{hom alt}}) \\
    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
      \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) \times \Prob(\boldsymbol{Z}^b \mid \theta_b \in (\frac12,1])\\
  \end{split}    
\end{equation}

Accounting for the sample likelihoods based on Equation \ref{eq:theta-likelihood} (assuming $\rho_b = \theta_b$ for the bulk that has no amplification step, Equation \ref{eq:rho-b}), and evaluating only point estimates of these likelihoods at possible alternative allele frequencies, this gives:

\begin{equation}
 \label{eq:hom-alt-posterior-likelihoods}
 \begin{split}   
 \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
     &\sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \Prob(\rho_s = \frac{k}{l} \mid \theta_s =  1)\right\} \times \\
     &\sum_{m=\frac{n}{2}}^{n} \Prob(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n})\\
  \end{split}    
\end{equation}

Accounting for amplification bias with Equation \ref{eq:BBLodato} and computing the likelihood of the sample-specific allele frequency ranges with Equations \ref{eq:read-product} and \ref{eq:bulk-prob}, this becomes:

\begin{equation}
 \label{eq:hom-alt-posterior-reads}
 \begin{split}   
 \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
      &\sum_{k=0}^{l} \left\{ \prod_{i=1}^l \Prob(Z_i^s \mid \rho_s = \frac{k}{l}) \times \cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ) \right\} \times\\
      &\sum_{m=\frac{n}{2}}^{n} \prod_{j=1}^l \Prob\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)\\
  \end{split}    
\end{equation}

With read probabilities calculated with Equation \ref{eq:read_probability}, we can calculate the posterior probability of the $\boldsymbol{E}_{\text{hom alt}}$ event (for an analogous and more detailed derivation for $\boldsymbol{E}_{\text{ADO to ref}}$, see Supplement, Section~\ref{sec:sc-model}).
To obtain posterior probabilities for compound events, we sum up the posterior probabilities of the events it comprises, effectively joining up their respective allele frequency range combinations into compound range combinations.
For the site-specific single cell probability for the presence of an alternative allele in the single cell, we thus get the compound event (blue events in Figure~\ref{fig:prosolo_alt-calling}D; Table~\ref{tab:Events}):

\begin{equation}
 \label{eq:compound-event-alt-presence}
 \begin{split}
  \boldsymbol{E}_{\text{alt presence}_s}
  =~&\boldsymbol{E}_\text{ADO to ref}~~\cup~~
    \boldsymbol{E}_\text{err ref}~~\cup~~
    \boldsymbol{E}_\text{het}~~\cup~~
    \boldsymbol{E}_\text{ADO to alt}~~\cup~~
    \boldsymbol{E}_\text{hom alt} \\
  =~&\{0\}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}~~\cup~~
    \{0, \frac{1}{2}\}_{\theta_s} \times [1]_{\theta_b}~~\cup~~
    \{\frac{1}{2}\}_{\theta_s} \times (0,1)_{\theta_b}~~\cup~~\\
    &\{1\}_{\theta_s} \times (0,\frac{1}{2}]_{\theta_b}~~\cup~~
    \{1\}_{\theta_s} \times (\frac{1}{2},1]_{\theta_b} \\
  =~&\{0\}_{\theta_s} \times [\frac{1}{2},1]_{\theta_b}~~\cup~~
    \{\frac{1}{2},1\}_{\theta_s} \times (0,1]_{\theta_b}
 \end{split}
\end{equation}

whose posterior probability we can obtain from this sum:

\begin{equation}
 \label{eq:alt-posterior-prob}
 \begin{split}
  \Prob(\text{alt presence}_s \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =~&
    \Prob(\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~ 
    \Prob(\boldsymbol{E}_{\text{err ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
  &\Prob(\boldsymbol{E}_{\text{het}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
  \Prob(\boldsymbol{E}_{\text{ADO to alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
 \end{split}
\end{equation}

To genotype, we calculate the posterior probability of all three possible single-cell genotypes and choose the genotype with the maximum posterior probability. 
In Figure~\ref{fig:prosolo_alt-calling}D, events are colored purple when implying the homozygous reference genotype, light blue when implying the heterozygous genotype and dark blue when implying the homozygous alternative genotype (Table \ref{tab:Events}).
Similarly, to calculate the posterior probability of an allele dropout at a particular site, we sum up the posterior probabilities of the two {\ttfamily ADO} events.

For any such compound event, we can estimate a threshold on the posterior probabilities that controls for a specified false discovery rate. 
This is based on the approach described by Müller et al.\cite{muller_optimal_2004,muller_fdr_2006}---for further details see the Supplement (Section~\ref{sec:fdr}) and Köster et al\cite{koster_enhancing_2019}.

\paragraph{Biologically relevant imputation based on the bulk sample.}
As argued above, a large enough sampling of the bulk cell population that the single cell comes from should contain the single cell's genotype at a particular site, unless this cell is genuinely the first cell to harbour a mutation at that site.
This bulk background sample can thereby render credibility to single cell variants with low coverage, while at the same time eliminating amplification errors in the single cell sample, as these will not exist in the bulk sample.
Interestingly, the bulk sample also provides a mechanism of biologically meaningful imputation at sites that have no read coverage in the single cell.
If imputation is desired for sites with no read coverage in the single cell, we set $\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) = \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac{1}{2}) = \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) = 1$, rendering all (unknown underlying) single cell genotypes equally likely.
Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will usually favor germline genotypes over any existing (lower frequency) somatic genotypes at a site.

\paragraph{ProSolo is an easy-to-use command-line tool, based on a modular framework.}

ProSolo is an easy-to-use command-line tool---following usability standards\cite{taschuk_ten_2017}---and its source code is available at \url{https://github.com/prosolo/prosolo}, including instructions for an easy installation via Bioconda\cite{gruning_bioconda:_2018}.
Its main contribution in terms of software is the implementation of its comprehensive statistical model into the Rust variant calling library of Varlociraptor\cite{koster_enhancing_2019}.
See the Supplement (Section~\ref{sec:implementation}) for further implementation details.

\subsection{Benchmarking}

We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}. We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 5.4.0}) to implement the benchmarking workflows.
For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo}.

\subsubsection{Datasets and Ground Truths}

\begin{figure}[!tpb]
  \centering
  \begin{minipage}{.26\linewidth}
    \includegraphics[height=32ex]{figs/Dong2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{4ex}
  \begin{minipage}{.4\linewidth}
    \includegraphics[height=32ex]{figs/Laehnemann2017_data_overview.pdf} \\
  \end{minipage}
  \hspace{4ex}
  \begin{minipage}{.22\linewidth}
    \includegraphics[height=32ex]{figs/data_overview_color_legend.pdf} \\
  \end{minipage}
  \caption{
   Benchmark Datasets.
   For single cells, DNA was multiple displacement amplified (MDA).
   \textbf{A} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
   \textbf{B} Newly generated whole exome sequencing (WXS) dataset of blood cells.
   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
   }
  \label{fig:datasets}
\end{figure}

We benchmarked ProSolo on two experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\

\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\

\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
% refs for MACS and FACS? Ute said, this is usually not required.
Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
% ref for q-rt-PCR? Ute said, this is usually not required.
For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.

To generate the ground truth of this dataset, we could leverage previously sequenced bulk whole exome data from the same person, their parents and three siblings \cite{hoell_constitutional_2014} (Figure~\ref{fig:datasets}B).
To create ground truth germline alternative allele calls, we ran three pedigree-aware variant callers (BEAGLE4.0\cite{browning_improving_2013}, polymutt\cite{li_likelihood-based_2012} and FamSeq\cite{peng_rare_2013,peng_famseq:_2014}) and created a consensus by including only calls where all callers agree at a site and where a maximum of one caller not calling the site was allowed (Figure~\ref{fig:granulocytes-ground-truth}).

\subsubsection{ProSolo achieves highest alternative allele calling accuracy}

The most precise single cell variant callers to date, SCcaller and SCIPhI, only call the presence vs. the absence of an alternative allele (i.e. the heterozygous and the homozygous alternative genotypes called jointly, the joint probability of the blue event areas in Figure~\ref{fig:prosolo_alt-calling}D).
We thus focused on this for the main benchmarking.

\begin{figure}[!tpb]
 \begin{minipage}{.40\linewidth}
  \includegraphics[height=45ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
 \end{minipage}
 \begin{minipage}{.55\linewidth}
  \includegraphics[height=45ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
 \end{minipage}

 \caption{
  Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
  Note that both panels are strong zoom-ins, focusing on (different) areas of interest.
  Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\\
  Threshold parameters:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
\label{fig:alt-calling_prec-rec}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}A and \ref{fig:alt-calling_prec-rec_global}A), at recall rates of 45\% and higher. 
In comparison to all other tools, ProSolo shows striking increases in recall.
For example, an increase of nearly 10\% for a precision above 99\%, where its maximum recall is 76.6\%, compared to 70.5\% for MonoVar, 68.7\% for SCIPhI and 61.0\% for SCcaller.
The only exception with a recall of 0.01\% is SCAN-SNV (at a precision of 99.2\%).
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}B).
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above 99\%, with ProSolo reaching a 20\% increase of recall to 17.6\%, compared to SCIPhI's 14.6\%, and SCcaller with 7.2\% (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
In comparison, MonoVar achieved a maximum precision of only 96.29\%. However, this was at a much higher recall (14.13\%) than for example SCcaller (9.55\% at a precision of 97.23\%).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.

On this dataset, SCAN-SNV's recall increased to 0.16\% at a decreased maximum precision of 89.7\%.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the homozygous reference germline genotype and will incorrectly classify (existing) alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate.
However, as the other tools also provide alternative allele calls for all sites where the single cells retained this germline genotype, the relative effect on their precision will be smaller.

At the same time, this germline ground truth caveat also indicates that the recall of SCIPhI's sensitive mode and ProSolo's imputation mode will be an overestimate.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
Both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation.
However, at single cell sites where a somatic mutation has created a true alternative allele, but no coverage is provided, we expect that both methods are most likely to call the homozygous reference germline genotype.
In then comparing this to the germline genotype as the ground truth, these calls will be classified as true negatives even though they really constitute false negatives, thus artificially increasing recall.
While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, \ref{fig:F_1-scores}B and \ref{fig:FDR-ground-truth-vs-theoretical}B for reference).

Finally, a feature where ProSolo clearly stands out is the control over the false discovery rate.
As can be seen in Figure~\ref{fig:alt-calling_prec-rec} (and Figure~\ref{fig:alt-calling_prec-rec_global}), ProSolo provides flexible control over precision vs. recall via specifying a false discovery rate of interest.
While no other tool provides a formal control over the easily interpretable false discovery rate, several of the tools provide other types of thresholds that we varied in attempts to achieve higher precision or recall.
However, none of them provide control over similar ranges of precision and recall.
The only limit to that range with ProSolo's current model, is that it becomes less accurate when controlling for very small false discovery rates (below 0.01\% for alternative allele calling in the whole genome dataset, see Supplement Section~\ref{sec:fdr-of-alt-calling}).
But this still leaves ProSolo as the only tool that provides the user with the choice of either aiming for more discoveries at the cost of a higher rate of false discoveries, or at aiming for a more limited number of discoveries with higher confidence in each of them.

\subsubsection{Estimates of allele dropout rate validate ProSolo's model}

Leveraging our ground truths and using three different ways to calculate the allele dropout rate, we can confirm the general validity of our single-cell event definitions and also explore limitations of the current model.
For the allele dropout rate, we will focus on the set of sites where the respective ground truth call is heterozygous, as these are the sites where the dropout of one of the alleles can be identified in a non-ambiguous manner.

More details for the three ways in which we calculate allele dropout rates are given in the Supplement (Section~\ref{sec:suppl-adr-calc}).
Here, we give a short explanation:

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/prosolo_ado-rate_expected-vs-ground-truth.pdf} \\
  \caption{
    Concordance of three differently calculated per-cell (IL-1 and PAG cells) allele dropout rates across ground truth heterozygous sites, in the context of allele dropout rates from the literature\cite{wang_clonal_2014,hou_single-cell_2012,xu_single-cell_2012,ling_evaluation_2009,spits_whole-genome_2006,spits_optimization_2006,renwick_proof_2006,lodato_somatic_2015}.
    The expected value of the allele dropout events in ProSolo (dark gray) is concordant with the number of false homozygous genotype calls made by ProSolo on those sites (black) and both values are well within the range of published allele dropout rates for single cell MDA sequencing data ("published", to the right, very light gray).
    The naive allele dropout rate (light gray)---calculated as ground truth heterozygous sites with a minimum coverage of seven and either no read with the reference allele or no read with the alternative allele---shows discrepancies with ProSolo's estimates of allele dropout for samples with a more uniform coverage (IL-11, PAG1, PAG10, Figure~\ref{fig:coverage-dists}).
  }
  \label{fig:allele-dropout}
\end{figure}

\begin{enumerate*}
 \item At each ground truth heterozygous site, we sum the posterior probabilities of the two allele dropout events defined in ProSolo ("ADO to alt" and "ADO to ref" in Figure~\ref{fig:prosolo_alt-calling}D) to obtain a total allele dropout probability (Equation \ref{eq:ado-posterior-prob}) and use these to compute the expected value of allele dropout across all sites.
 This gives us an expected allele dropout rate ("expected value" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.expected-allele-dropout-rate}).
 \item We genotype all the ground truth heterozygous sites with ProSolo, take the most likely genotype and then compare against the ground truth: heterozygous sites that ProSolo calls as homozygous are counted as dropout sites. The number of such sites is divided by the total number of ground truth heterozygous sites where the respective single cell had coverage ("hom at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.hom-at-ground-truth-het-allele-dropout-rate}).
 \item We identify all heterozygous ground truth sites with a coverage of at least 7, where without amplification bias we could be reasonably sure to sample both alleles (for the reasoning see Supplement, Section~\ref{sec:suppl-adr-calc}).
 We then count a site as an allele dropout if there is one allele (reference or alternative) with no read coverage at all, and again divide by the total number of ground truth heterozygous sites where the respective single cell had coverage ("no alt/ref read at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.no-alt-ref-read-at-ground-truth-het-adr}).
\end{enumerate*}

The expected allele dropout rates based on the ProSolo probabilities for allele dropout clearly fall into the range of previously published allele dropout rates\cite{wang_clonal_2014,hou_single-cell_2012,xu_single-cell_2012,ling_evaluation_2009,spits_whole-genome_2006,spits_optimization_2006,renwick_proof_2006,lodato_somatic_2015} ("published" in Figure~\ref{fig:allele-dropout}).
This analysis also clearly shows that the ProSolo expected allele dropout rates, based on the model's probabilities, correspond to those determined by comparing ProSolo genotypes with the ground truth (Figure~\ref{fig:allele-dropout}).
This demonstrates that the explicit modelling of allele dropout events works and is useful for genotyping.
However, in that comparison, the expected allele dropout rate was consistently underestimated on our own whole exome data ("granulocytes", Figure~\ref{fig:allele-dropout}), and slightly overestimated for the data from Dong et al.\cite{dong_accurate_2017} ("Dong 2017", Figure~\ref{fig:allele-dropout}).
The comparison to a naively calculated allele dropout rate consistently shows an overestimation of the allele dropout probability, which is strongest for the samples with a higher overall coverage (IL-11, PAG1, PAG10, Figures~\ref{fig:allele-dropout} and \ref{fig:coverage-dists}).
However, this overestimation of the allele dropout probability does not seem to impact the genotyping resolution (Figures~\ref{fig:gt-calling-true-false-colouring} and \ref{fig:gt-calling-gt-colouring}).

\section{Discussion}

ProSolo is the first method for SNV calling from MDA single cell sequencing data to comprehensively model both amplification bias and amplification errors in a way that allows for site-specific variation (Figure \ref{fig:prosolo_alt-calling}).
We achieve this by combining a data-driven model of amplification errors that incorporates a bulk background sample alongside the single cell sample with an empirical model of amplification bias, based on a statistical understanding of the MDA process that is mechanistically motivated and dependent upon a site's coverage.
The underlying model calculates posterior probabilities for fine-grained single cell event definitions, whose false discovery rate can be controlled for and that can pass more information about data uncertainties to probabilistic models in downstream analyses.
Using a whole genome and a whole exome dataset---each with a different type of ground truth (Figure~\ref{fig:datasets})---we demonstrate that these two innovations of ProSolo combined to achieve better accuracy (Figure~\ref{fig:alt-calling_prec-rec}).
Moreover, the model allows to accurately control the false discovery rate of variant calls, another novelty of great practical value, and both the model and its modular implementation are easy to adapt and extend.

Especially the joint modeling of a single cell and a bulk sample is favorable from a statistical point of view: in terms of MDA induced bias, the bulk acts as an unbiased sample of the population from which the single cell was drawn.
This provides a drastically more sensitive approach compared to, for example, the consensus rule of calling alternative alleles only when there is evidence in at least two (or even three) single cells (as implemented in MonoVar\cite{zafar_monovar:_2016}).
A more systematic and biologically relevant model of sharing information is implemented in SCIPhI, where the phylogenetic relationship of cells is part of the estimated parameters.
Intuitively, if two cells are closely related, they have a higher likelihood of sharing a genotype at a particular site.
However, both the consensus rule and the sharing of information via an inferred phylogeny requires that more than one single cell exhibit sufficient coverage of an alternative allele to call it reliably.
This will rarely be achieved in single cell MDA experiments, which are often limited to a few dozen cells.
In contrast, bulk sequencing can easily be scaled to a much larger sampling of a cell population---representing hundreds or thousands of cells of a cell population---simply by adding a single higher coverage sample.
Our benchmarking demonstrated that ProSolo achieves a substantially higher recall than any other tool for a precision above 99\% in identifying the presence of an alternative allele.
To summarize our benchmarking of alternative allele calling: SCcaller seems optimized for precision, MonoVar achieved higher recall than SCcaller, and SCAN-SNV does not seem suitable for general variant calling on MDA single cell sequencing data, but is only applicable when restricting interest to somatic variants.
ProSolo clearly achieves the best accuracy, with particularly striking increases in recall. 
The only other tool that shows consistent performance across both datasets, but with a recall 10 to 20\% below ProSolo, is SCIPhI.
However, SCIPhI takes several weeks on a single core without any possibility of parallelization, compared to a runtime of only a few days on a cluster for all the other tools.

Jointly calling single cell variants with a bulk background sample also enables a biologically relevant imputation of genotypes at sites where single cells lack coverage.
If genotype profiles without missing values are for example required by downstream tools, ProSolo can provide a cell-population-specific imputation instead of resorting to a majority assignment (limited by the number of single cells sequenced) or even imputation based on external databases (e.g.~dbsnp).
As our results further demonstrate, ProSolo’s empirical model of amplification bias is robust also for whole exome sequencing data. 
In contrast, SCcaller---that estimates local amplification bias based on minor allele frequencies at known germline genotypes in the vicinity of an examined site---suffers from a scarcity of neighbouring sites in whole exome data.

Finally, ProSolo is the only method that provides a clearly interpretable false discovery rate parameter that can actually control the trade-off between precision and recall.
This can be used on alternative allele calls (Figure~\ref{fig:alt-calling_prec-rec}, Supplement Section~\ref{sec:fdr-of-alt-calling}), the main focus of our benchmarking, but also on any of the events pertinent to single cell analysis (such as allele dropout or amplification errors) or combinations thereof (see Figure~\ref{fig:prosolo_alt-calling}D for all events).
Thereby, beyond just calling alternative allele presence in a statistically reliable way, ProSolo can, for example, also compute the expected allele dropout rate across the entire genome of a particular cell in a robust manner (Equation \ref{eq:ado-posterior-prob}, Figure~\ref{fig:allele-dropout}).

In addition to such global rates, ProSolo determines reliable site-specific posterior probabilities for all single cell events that might be of interest, including allele-resolved genotypes, allele dropout or amplification errors. 
We anticipate that such fine-grained probabilities---and the uncertainties they capture---will be informative for improving probabilistic modelling in downstream analyses.
They could e.g.~prove useful in models for phylogenetic reconstruction of the lineage relationship of sequenced single cells\cite{singer_single-cell_2018,zafar_siclonefit:_2018,koptagel_scuphr:_2018}, while keeping them computationally tractable\cite{lahnemann_eleven_2020}.

An in-depth look at one of the above-mentioned fine-grained events, allele dropout, showed that ProSolo's allele dropout rate estimates were within the range of published estimates.
This confirms that our modeling of allele dropout events is realistic and can be useful in alternative allele calling and genotyping.
However, the allele dropout rate estimations were slightly off in different directions for the different benchmarking datasets, suggesting that the empirical distributions we currently make use of (based on Lodato et al.\cite{lodato_somatic_2015}) may not suit all datasets.
For example, whole genome amplification may introduce further variability so far not captured by our model.  
This conclusion is further supported by the naively calculated allele dropout rate, which is much lower in the high coverage cells of the two datasets (IL-11 for the Dong et al. data\cite{dong_accurate_2017}, PAG1 and PAG10 in our granulocytes; see Figure~\ref{fig:coverage-dists}).
Interestingly, the sample IL-11 from Dong et al.\cite{dong_accurate_2017} has been suggested to be a doublet\cite{luquette_identification_2019}, which might explain the higher overall coverage and points to a possible source for the discrepancy between the naively calculated and the estimated allele dropout rates.
If samples PAG1 and PAG10 were doublets as well, this would indicate that our use of empirical distributions in ProSolo provides for more robust event probabilities in the presence of doublets, while heuristic thresholding (as in our naive allele dropout estimate) is very sensitive to such perturbations.
In general, the use of a fixed empirical model for MDA allelic bias does not seem to impede ProSolo's performance in alternative allele calling compared to the other tools, but has a noticeable effect on genotyping (Supplement, Section~\ref{sec:suppl-gt-calling}) and might be responsible for slight imprecisions when controlling for very small false discovery rates (Supplement Section~\ref{sec:fdr-of-alt-calling}).
When future datasets are generated based on improved MDA protocols\cite{lahnemann_eleven_2020}, these effects might be exacerbated.
We therefore consider it important future work to improve on the fitting of the empirical read count distributions observed when applying MDA to single cells (Supplement, Section~\ref{sec:better_beta-binom}).
For example, the parameters of the mechanistically motivated combination of beta-binomial distributions for modelling heterozygous genotypes could be learned per single cell sample at germline heterozygous sites---similar to the approaches of SCcaller and SCAN-SNV, but globally per cell with their local variation modelled by their dependence on a site's coverage.
Further room for improvement also remains for the modeling of the homozygous genotypes: while the current distributions account for both amplification and sequencing errors, sequencing errors are already safely accounted for elsewhere in our latent variable model.
An amplification error profile that is not compounded with sequencing errors, would need to be based on the $\Phi$29 polymerase error rate and a better understanding of the statistical distribution that these errors generate (Supplement, Section~\ref{sec:better_beta-binom}).
Studying and implementing the corresponding changes in the future has the potential to further improve the accurate site-specific event probabilities that ProSolo already provides through the joint modelling with a bulk background sample.

Finally, the modular implementation of ProSolo within the context of the Varlociraptor library\cite{koster_enhancing_2019} facilitates the implementation of further features, such as the ones mentioned above, read-backed phasing\cite{bohrson_linked-read_2019,hard_conbase:_2019} or even more variable models of amplification bias\cite{luquette_identification_2019} that could be integrated to exhaustively leverage MDA data information content.
Since the Varlociraptor library also provides advanced functionality for the calling of insertions, deletions and multiple nucleotide variants (MNVs), one of the immediately following next steps will be to adapt ProSolo to calling those in single cells.\\

Overall, ProSolo provides an accurate and easy-to-use variant caller for single cell MDA sequencing data, which will empower more research using single cell sequencing data.

\bibliography{Benchmarking_ProSolo}

\section*{Acknowledgements}

\section*{Funding}
This work has been supported by the Helmholtz Association, in particular through a Helmholtz Incubator grant (Sparse2Big ZT-I-0007), by the compute cluster at the Helmholtz Institute for Infection Research, and by the Katharina Hardt Stiftung. Alexander Schönhuth was supported by the Netherlands Organisation for Scientific Research (NWO: Vidi grant 639.072.309). Arndt Borkhardt and Ute Fischer were further supported by the German Federal Office for Radiation Protection (BfS) grant nos. 3618S32274 and 3618S32275.

\section{Online Methods}
See Supplement PDF.

\end{document}

### 2020-04-27 manuscript/supplement.tex
manuscript before preprint upload, including result data, figure code and figure files

\documentclass[authoryear,preprint,11pt]{scrartcl}
\addtokomafont{disposition}{\rmfamily}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{url}
\usepackage{epsfig}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{natbib}

%\usepackage[left]{lineno}
%\linenumbers

\usepackage{setspace}
\doublespacing

\usepackage{rotating}
\usepackage{xtab}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{multirow} % enable table cells spanning multiple rows (\multirow{}{}{})
\usepackage{bigdelim} % enable brackets in tables that span multiple rows (e.g.: `\multirow{}{}{\ldelim}`)
\usepackage{colortbl} % enable background coloring of table rows (\rowcolor{})
\usepackage{tikz}

\usepackage{listings} % enable source code listings, e.g.~\begin{lstlisting}
\usepackage[inline]{enumitem} % enable enumerate* environment for inline enumeration

\usepackage{hyperref}
\usepackage{tocstyle}
% Define Supplement prefix for numbered references
\newcommand{\supplpref}{S~}
% Set Supplement prefix for numbered references
\renewcommand{\thetable}{\supplpref\arabic{table}}
\renewcommand{\thesection}{\supplpref\arabic{section}}
\renewcommand{\thesubsection}{\supplpref\arabic{section}.\arabic{subsection}}
\renewcommand{\thesubsubsection}{\supplpref\arabic{section}.\arabic{subsection}.\arabic{subsubsection}}
\renewcommand{\thefigure}{\supplpref\arabic{figure}}
\renewcommand{\theequation}{\supplpref\arabic{equation}}

\newcommand{\Prob}{{P}}

\newcommand{\cR}{{\mathfrak R}}
\newcommand{\cB}{{\mathfrak B}}

\usepackage{xr-hyper}

% xr helper code to generate dependency definition command \myexternaldocument
\makeatletter
\newcommand*{\addFileDependency}[1]{% argument=file name and extension
  \typeout{(#1)}
  \@addtofilelist{#1}
  \IfFileExists{#1}{}{\typeout{No file #1.}}
}
\makeatother

\newcommand*{\myexternaldocument}[1]{%
    \externaldocument{#1}%
    \addFileDependency{#1.tex}%
    \addFileDependency{#1.aux}%
}

\myexternaldocument{main}

\begin{document}

\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data}
\subtitle{Supplementary Methods and Results}

\author{David Lähnemann, Johannes Köster, Ute Fischer, \\ Arndt Borkhardt, Alice McHardy, Alexander Schönhuth}

\maketitle

\tableofcontents

\section{Probabilistic model of ProSolo}

\subsection{Underlying True Alternative Nucleotide Frequencies: $\theta$}
\label{sec.afs}

The eventual goal in variant calling is to infer the true frequency of a possible alternative, and thus variant, nucleotide in the sample you sequenced.
For such an alternative nucleotide $x\in\{A,C,G,T\}$ at a particular site in a genome, we therefor define\footnote{We omit an explicit notation for a site, but every $\theta$ is always specific to one genome position.}:
\begin{equation}
 \theta_{b}(x), \theta_{s}(x)
\end{equation}
as the (true, but usually unknown) frequency of alternative nucleotide $x$ in the bulk and single cell sample, respectively.
These frequencies translate into probabilities of sampling a DNA fragment containing the respective nucleotide at a site:
either directly, as with the sequencing of DNA from a bulk of cells $b$;
or indirectly, as with the sequencing of whole genome amplified DNA of a single cell $s$.
It is our eventual goal to work back from bulk and single cell data to derive likelihoods for possible values of $\theta_{b}(x)$ and $\theta_{s}(x)$, respectively, allowing us to place likelihoods on possible underlying genotypes and events like allele dropouts in single cells.\\

For a single diploid cell $s$, the possible alternative nucleotide frequencies are
\begin{equation}
 \theta_s(x) =
 \begin{cases}
  0       \\
  \frac12 \\
  1
 \end{cases}
\end{equation}
reflecting the cases that:
(i) $x$ is not present at the position---if no other allele is present, this constitutes a homozygous reference site;
(ii) $x$ reflects an alternative nucleotide that is present in only one of the two genome copies of the single cell---this constitutes a heterozgyous site;
(iii) $x$ is an alternative nucleotide present in both genome copies of the single cell---this constitutes a homozygous alternative site.

Bulk experiments contain subpopulations that can differ in their genotype regarding a particular site, through a somatic mutation that one of the subpopulations has acquired in one allele.
They can thus exhibit any alternative nucleotide frequency ($\theta_{b}(x) \in [0,1]$), depending on the genotypes present and their relative abundance (Figure~\ref{fig:event-space}).

\subsection{Non-reference Nucleotide Frequencies After Biased Whole Genome Amplification: $\rho$}
\label{sec.abundance}

Let further $\rho_{s}(x),\rho_{b}(x)$ be the abundance of genome copies showing alternative nucleotide $x$ at a particular site in the single or bulk DNA sample, respectively, right before the sequencing---for the single cell, this means after the whole genome amplification that is necessary to obtain enough DNA for sequencing with technologies like Illumina.
These abundances translate into probabilities to draw a read containing $x$ from the respective sample (in the sequencing experiment), given that such a read covers the genome position.

Under the assumption of uniform abundance of all parts of the genome and uniform abundance of each haplotype, which is reasonable in bulk sequencing (i.e.~without amplification), it holds that $\rho = \theta$, so we henceforth assume that
\begin{equation}
 \label{eq:rho-b}
 \rho_{b}(x) = \theta_{b}(x).
\end{equation}

In contrast, in single cell sequencing---due to the necessary whole genome amplification process---the assumption of uniform abundance of all parts of the genome and of each haplotype does not hold.
Instead, each allele at a particular site can be amplified to a different degree, ranging from no amplification at all (i.e.~a complete allele dropout) to several hundred or thousand copies.
Thus

\begin{equation}
 \label{eq.nonUniDependency}
 \rho_s(x) \ne \theta_s(x)\quad\text{but instead}\quad \rho_s(x) \sim \Prob(\rho_s \mid \theta_s(x))
\end{equation}

that is, $\rho_s$ is a random variable governed by a probability distribution ${\Prob(\rho_s \mid \theta_s(x) )}$ that depends on the original alternative nucleotide frequency $\theta_s$.
$\Prob(\rho_s \mid \theta_s(x))$ reflects the differential amplification of the two alleles, which is only detectable for heterozygous sites ($\theta_s = \frac12$), thus

\begin{equation}
 \label{eq.rho-given-theta}
 \Prob(\rho_s \mid \theta_s(x) ) =
 \begin{cases}
  \Prob(\rho_s \mid \theta_s(x) = 0)        \\
  \Prob(\rho_s \mid \theta_s(x) = \frac12 ) \\
  \Prob(\rho_s \mid \theta_s(x) = 1 )
 \end{cases}
\end{equation}

Understanding these probability densities is key to understanding the statistical issues single cell whole genome amplification methods come along with\footnote{
 Note, in addition to the whole genome amplification bias, these distributions can implicitly also model the amplification errors that the polymerases used for whole genome amplification introduce in the copy process.
}.
From the Supplementary Material of \cite{lodato_somatic_2015} (see Figure S5 and the section ``Modeling MDA-derived alternative read counts'' of the respective supplement for the details), we get an empirical model that characterizes them for the whole genome amplification method multiple displacement amplification (MDA)\footnote{
 For other whole genome amplification methods with different amplification bias properties, we can simply implement a different instance of the general model for $\Prob(\rho_s \mid \theta_s(x))$ accordingly.
 And for MDA, different empirical models learned from other or more published datasets could be implemented as well as strategies of inferring parameters from the very dataset analysed in a certain run of the software.
}:

\begin{equation}
 \begin{split}
  \Prob(\rho_s \mid \theta_s = 0 ) 			&= \cB\cB_{l,\alpha, \beta}(\rho_s)\\
  \Prob(\rho_s \mid \theta_s = \frac12) &= w(l) \times \cB\cB_{l,\alpha_1,\alpha_1}(\rho_s) + (1-w(l)) \times \cB\cB_{l,\alpha_2,\alpha_2}(\rho_s)\\
  \Prob(\rho_s \mid \theta_s = 1 ) 			&= \cB\cB_{l,\beta,\alpha}(\rho_s)
 \end{split}
\end{equation}

where $\cB\cB$ represents beta-binomial functions with the parameters linearly scaled by $l$, the total observations at a site (i.e.~the total read coverage).
For homozygous reference sites ($\theta_s = 0$), the parameters of the beta-binomial function are $\alpha(l)$ and $\beta(l)$.
For the heterozygous sites ($\theta_s = \frac12$), both beta-binomial functions are symmetrical with $\alpha_1(l),\alpha_2(l)$ dependent on the site-specific coverage $l$ and combined by factor $w(l)$.
$\rho_s \in \frac{k}{l}, k \in {0, 1, \dots, l}$ are the possible cases of alternative nucleotide frequencies.
In their Supplementary Figure S5 in panels C and D, \cite{lodato_somatic_2015} give estimates of the parameters $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ obtained by fitting to empirical data.
We list the values we extracted in Table~\ref{tab:Lodato}.
For the third case (homozygous alternative nucleotide, i.e.~$\Prob(\rho_s \mid \theta_s = 1)$) we mirror the homozygous reference function\footnote{
Note that the amount of sites that are homozygous for the reference allele is much larger, which also means that statistics for $\theta_s=1.0$ are more robust when being derived from the case $\theta_s=0.0$.
}.

Using one of the possible expressions of the respective probabilities (Equation 6.28 in Section 6.2.4 of \cite{johnson_univariate_2005})\footnote{
 Here, the parameters $b,w,c$ from the formulation in Equation 6.28 in Section 6.2.4 of \cite{johnson_univariate_2005} are summarised into $\alpha_1,\alpha_2$ from \cite{lodato_somatic_2015} with:
 \begin{equation}
 \label{eq.johnson_relation}
 \begin{split}
  \alpha_1 =~ & \frac{w_1}{c} = w_1 = b_1\\
  \alpha_2 =~ & \frac{w_2}{c} = w_2 = b_2
 \end{split}
 \end{equation}
 I.e., the number of starting reference and alternative alleles is identical within each beta-binomial, yielding symmetric distributions.
 In terms of the Pólya urn model used as motivation in \cite{johnson_univariate_2005}, this equates to the number of $w$ white and $b$ black balls being identical.
 $c$, the number of balls of identical color that are added whenever a ball of particular color is added, is set to $1$.
 Further considerations for obtaining the values of $b,w$ are discussed below in relation to the Pólya urn model.
}, with regards to the beta function $\mathfrak{B}$, we can compute the probability of observing $k$ reads with the alternative nucleotide at a site of coverage $l$ for the heterozygous case as follows:

\begin{equation}
 \label{eq.beta-binomial}
 \begin{split}
  \Prob(K = k \mid \theta_s = \frac{1}{2}) =~ &w(l) \times {l \choose k} \frac{ \mathfrak{B}_{k+\alpha_1,l-k+\alpha_1} }{ \mathfrak{B}_{\alpha_1,\alpha_1} } +\\
  &(1-w(l)) \times {l \choose k} \frac{ \mathfrak{B}_{k+\alpha_2,l-k+\alpha_2} }{ \mathfrak{B}_{\alpha_2,\alpha_2} }\\
 \end{split}
\end{equation}

In the future, these distributions could be learned from the dataset at hand, whenever large enough sets of sites that have a high probability of being homozygous reference or heterozygous are available.
In addition, the exact distribution used could be further refined with the below considerations regarding the suitability of beta-binomial distributions in this case.\\

\begin{table}[tbp]
 \caption{Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}. Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.}
 \label{tab:Lodato}
 \renewcommand{\arraystretch}{1.15}

 \begin{center}
  \begin{tabular}{lrr}
   parameter  & slope        & intersect   \\
   \toprule
   $\alpha$   & -0.000027183 & 0.068567471 \\
   $\beta$    & 0.007454388  & 2.367486659 \\
   \midrule
   $w$        & 0.000548761  & 0.540396786 \\
   $\alpha_1$ & 0.057378844  & 0.669733191 \\
   $\alpha_2$ & 0.003233912  & 0.399261625 \\
   \bottomrule
  \end{tabular}
 \end{center}
\end{table}

\subsubsection{The Pólya urn model: motivating the use of beta-binomial distributions}\label{sec:polya-urn}

For the heterozygous case, the applicability of the beta-binomial distribution \citep{irwin_distribution_1954} should be immediately evident from the formulation of the Pólya urn model \citep{eggenberger_uber_1923} that generates beta-binomial distributions \citep{irwin_distribution_1954,johnson_univariate_2005}.
It works directly analogous to the amplification process\footnote{The notation used here is analogous to Equation 6.28 in Section 6.2.4 of \cite{johnson_univariate_2005}. Also see Equation \ref{eq.johnson_relation} in the above footnote for its relation to $\alpha_1, \alpha_2$ in \cite{lodato_somatic_2015}.}:

At the start, the urn contains $w$ $w$hite and $b$ $b$lack balls, analogous to the two alleles present at a heterozygous site---let us say white balls represent the reference allele and black balls the alternative allele.
Whenever a ball is drawn, it is replaced to the urn and another $c$ balls of the same $c$olor are added, analogous to the polymerase copying an allele.

\begin{figure}[!tpb]
 \begin{minipage}[t]{.47\linewidth}
  \textbf{A}\\
  \includegraphics[width=.95\linewidth]{figs/Beta_binom_cov-20_w-2_b-2_f-1.pdf} \\
  \textbf{C}\\
  \includegraphics[width=.95\linewidth]{figs/Beta_binom_cov-20_w-2_b-2_f-0-25.pdf} \\
 \end{minipage}
 \begin{minipage}[t]{.47\linewidth}
  \textbf{B}\\
  \includegraphics[width=.95\linewidth]{figs/Beta_binom_cov-20_w-2_b-2_f-0-5.pdf}
  \textbf{D}\\
  \includegraphics[width=.95\linewidth]{figs/Beta_binom_cov-20_w-4_b-5-9e-05_f-0-5.pdf}
  \llap{\raisebox{4ex}{%  move next graphics to top right corner
        \includegraphics[width=.7\linewidth]{figs/Beta_binom_cov-20_w-4_b-5-9e-05_f-0-5_focus_tail.pdf}
        \hspace{1ex}
    }}
 \end{minipage}
 \caption{
  Beta-binomial distributions with different shape parameters, that correspond to our theoretical numbers of starting allele strands ($w_t = b_t = 2$) scaled to different amplification accessibility by $f$.
  \textbf{A} Perfect accessibility of both strands of both alleles for amplification ($f = 1$) and the resulting alternative allele frequencies.
  \textbf{B} Half that accessibility ($f = 0.5$), resulting in the uniform distribution.
  \textbf{C} For any accessibility scaling value below $\frac{1}{2}$ (e.g.~$f = 0.25$), the beta-binomial distribution will give maxima at alternative allele frequencies $0$ and $1$.
  \textbf{D} The heavily non-symmetrical case ($w_t \gg b_t$), analogous to homozgyous reference genotypes.
  Instead of the distribution currently used in ProSolo, which was fit to data and compounds sequencing errors with amplification errors (Figure~\ref{fig:prosolo_alt-calling}A in the main manuscript), this distribution makes the following theoretical assumptions:
  It sets $w_t = 4$ for the $4$ reference strands available for amplification and $b_t$ to four times the known $\Phi$29 error rate for drawing / introducing an error alternative allele instead of the reference allele.
  The inset zooms into the y-axis to show details of the tail of the distribution.}
 \label{fig:beta-binomials}
\end{figure}

As we have two strands each for both alleles, a natural choice of starting alleles would be $w_t = b_t = 2$.
We here subscript the values with $_t$ to denote the theoretical nature of the parameter choice, as opposed to practical values of $w_p,b_p$ introduced below.
And as one strand is added with every copy operation, we set $c = 1$ (Figure~\ref{fig:beta-binomials}A).
This gives us a slight peak around the alternative allele frequency of $0.5$ (Figure~\ref{fig:beta-binomials}A), which we can also make out in the real data (Figure~\ref{fig:prosolo_alt-calling}B of the main manuscript).

However, it does not capture the allele dropout peaks at the allele frequencies of $0$ and $1$ seen there.
A likely reason for these dropout peaks is, that not every position on every allele is equally accessible to amplification initiation \citep{picher_trueprime_2016}.
I.e.~mechanistically, some close-by site to the one under consideration must be single-stranded at some point during the continuous amplification reaction (at a constant temperature), to be accessible for an initial priming that will start the amplification of the respective allele for that region.
To account for this lower initiation accessibility, we can introduce a scaling factor $f$ for the starting frequencies of the two alleles, with $0 <= f <= 1$, to scale down the initially available allele count:
if strands are already separated for this allele and amplification initiation works perfectly for both strands ($f = 1$), then we would get $w_p = w_t * f = 2 * 1 = 2$;
whereas at the other extreme of $f = 0$, amplification initialisation of the allele would be impossible.
Then, for $f = \frac{1}{2}$, we get the uniform distribution ($w_p = w_t * f = \frac{1}{2} * 2 = 1$, Figure~\ref{fig:beta-binomials}B), and for all $f < \frac{1}{2}$, we will get distributions with peaks at $0$ and $1$ (e.g.~Figure~\ref{fig:beta-binomials}C).

If we then consider a mixture of sites with different $f$, we would expect distributions as observed e.g.~in Figure~\ref{fig:prosolo_alt-calling}B of the main manuscript; and a reasonable approximation of this is the mixture of the two symmetrical beta-binomials as fitted by \cite{lodato_somatic_2015}, one with $f \approx 0.40 < \frac{1}{2}$ and one with $f \approx 0.67 > \frac{1}{2}$ (i.e.~the intersects of $\alpha_2$ and $\alpha_1$ plus their slight slope-adjustments, as given in Table~\ref{tab:Lodato}).

\subsubsection{Outlook: more realistic beta-binomial distributions} \label{sec:better_beta-binom}

However, something that this does not capture is possible asymmetry of the distributions, i.e.~the case of $w \neq b$.
While the theoretical starting amounts of strands of both alleles in the heterozygous case should remain the same ($w_t = b_t = 2$), we would have to introduce allele-specific scaling factors $f_w \neq f_b$ that capture the different availability of the two alleles for amplification (e.g.~because one of the alleles is more densely wrapped around Histones than the other).
As a result we would get $w_p = w_t * f_w \neq f_b * b_t = b_p$, and instead of the one shape parameter in the symmetrical case, we would then have to fit the distributions with a second shape parameter.
I.e.~instead of Equation \ref{eq.johnson_relation}, we would get:
\begin{equation}
 \label{eq.assym_het_beta-binom}
 \begin{split}
  \alpha_1 =~ & w_p = w_t * f_w = 2 f_w\\
  \beta_1 =~ & b_p = b_t * f_b = 2 f_b
 \end{split}
\end{equation}\\

However, assuming that the dropout of both alleles---reference and alternative---is equally likely, this asymmetry will average out over all heterozygous sites, and the symmetry assumption remains a justified reduction in parameters to fit when fitting a mixture of two beta-binomials.

This is somewhat different for the homozygous case.
Here, \cite{lodato_somatic_2015} already modelled the allele frequency deviations ($>0$ for homozygous reference) as an asymmetric beta-binomial distribution.
This gives a reasonable fit to the data and makes sense to use, as we would expect these deviations to be the result of amplified polymerase errors from the MDA.
While we just use the distributions fit by \cite{lodato_somatic_2015} in our initial implementation we present here, this could in the future be replaced by a theoretically derived beta-binomial distribution.
For example, for the homozygous reference case, a very low $b_t$ (the probability of ``choosing'' a previously non-existing error allele) could be determined by (four times) the known $\Phi$29 polymerase error rate, whereas $w_t$ would be set to $4$ to account for four reference strands available for copying (Figure~\ref{fig:beta-binomials}D).
As $f$ would in this case affect both the reference ($w$) and the error allele ($b$) alike, it could be set to some identical value for both, e.g.~to an average accessibility of $0.5$ (Figure~\ref{fig:beta-binomials}D).
This would avoid compounding the uncertainty due to MDA with sequencing errors, that we already account for separately in our latent variable model through base calling quality scores and is one very likely source of the current mis-classifications of ground truth heterozygous sites as homozygous (Figure~\ref{fig:gt-calling-gt-colouring}).\\

\subsection{Reads: Sampling Alternative Nucleotide Frequencies in the Sequencing Process}
\label{sec.reads}

Let $\cR$ (=$\cR(g)$) denote all reads covering a site $g$, with $R_i\in\cR$ denoting an individual read out of the $l$ reads aligning to that site (${l\in{\mathbb N}, i \in \{1,2,\dots,l\}}$).
Formally, we index positions in aligned reads by the respective genome indices $g$, but we omit the index $g$ for readability.
Working with a given $\rho$, one can assume independence among the reads in the respective sample (while, of course, given $\theta$, one can assume independence only in bulk, due to the necessary amplification and the resulting coverage nonuniformity in single cell samples).
That is

\begin{equation}
 \label{eq.independence}
 L(\rho\mid\cR) \propto \Prob(\cR\mid\rho) = \prod_{R\in\cR}\Prob(R\mid\rho) \propto \prod_{R\in\cR}L(\rho\mid R)
\end{equation}

both for bulk and single cell sequencing. 
Let $S_{R_i}[g]$ be the nucleotide determined by the sequencer for position $g$ of read $R_i$ and $Q_{R_i}[g]$ be the Phred quality score for the nucleotide at that position of read $R_i$\footnote{
 As above, we now omit $g$ for $S_{R_i}$ and $Q_{R_i}$ for ease of notation and reading, as we will for $M_{R_i}$.
 Also note that the value of $Q_{R_i}[g]$ can be adjusted by a base quality score recalibration that accounts for systematic base calling errors, before using it as input to ProSolo.
}.
Let further $M_{R_i}[g]$ be the Phred alignment quality score\footnote{
 Also called mapping quality; represented by the MAPQ tag in the sequence alignment (SAM) format specifications v1 (2a802cd) and by the MQ tag in the variant call format (VCF) specification v4.3 (2a802cd).
 As before, we will omit the position $g$ from notation from this point on.
} provided by the aligner.
We now define $Z_i[g]$ ($i\in\{1,\dots,l\}$) to be the collection of the observable variables for the alignment of one of $l$ reads $R_i \in \cR$ to a genome position $g$.
Omitting index $g$, this is the vector ${Z_i = \langle S_{R_i}, Q_{R_i}, M_{R_i} \rangle}$ and drawing on this information, we now introduce hyperparameters analogously to the model for tumor-normal variant calling as outlined by \cite{koster_enhancing_2019} for Varlociraptor.
These will be the latent variables $\omega_i\in\{0,1\}$ to model correct placement of an alignment ($\omega_i=1$ if correct and $\omega_i=0$ otherwise) and $\xi_i\in\{0,1\}$ to specify whether an alignment $Z_i$ is associated with the alternative nucleotide ($\xi_i=1$) or not ($\xi_i=0$).
Taken together, they give three cases

\begin{equation}
 \label{eq.read-cases}
 (\omega_i,\xi_i) =
 \begin{cases}
  (1,1): R_i \text{ mapped correctly, alternative nucleotide}  \\
  (1,0): R_i \text{ mapped correctly, reference nucleotide} \\
  (0,\{0,1\}): R_i \text{ not mapped correctly, nucleotide irrelevant}
 \end{cases}
\end{equation}\\

We thus model the two major sources of uncertainty on the level of the individual alignment $Z_i$, 1) \emph{alignment uncertainty} and 2) \emph{typing uncertainty}, by associating every alignment $Z_i$ with these two binary-valued, latent variables, reflecting uncertainty hyperparameters, $\omega_i$ and $\xi_i$:\\

\emph{First},
\begin{equation}
 \label{eq.map-prob}
 \omega_i \sim \text{Bernoulli}\left(\pi_i \right) \text{ for }i=1,\dots,l
\end{equation}

where $\pi_i$ is the (Phred scaled) posterior probability proportional to $M_{R_i}$ provided by the aligner.
We can compute it with

\begin{equation}
 \label{eq.MQ.phred}
 \pi_i = 10^{-M_{R_i}/10} \\
\end{equation}\\

\emph{Second},
\begin{equation}
 \label{eq.var-prob}
 \xi_i \sim \text{Bernoulli}\left(\rho \right) \text{ for }i=1,\dots,l
\end{equation}
which reflects that sampling a fragment from the locus that bears the alternative nucleotide agrees with the probability to sample a genome copy with the alternative nucleotide from the pool of amplified DNA.

Whether $\xi_i$ is $1$ or $0$ is generally not evident from the observed $Z_i$, due to typing uncertainty.
We formally define\footnote{
 If $\omega_i=0$, that is the alignment is incorrect, $Z_i \mid \omega_i = 0 \sim 1$ reflects that $Z_i$ has no influence on the posterior probability distribution of $\rho$.
}
\begin{equation}
 \label{eq.abs-pres}
 Z_i \mid \omega_i = 1, \xi_i = 0 \sim a(Z_i) \quad\text{ and }\quad
 Z_i \mid \omega_i = 1, \xi_i = 1 \sim p(Z_i)
\end{equation}

where $a,p$ are probability distributions over correct $S_{R_i}$ when the alternative nucleotide is either \emph{a}bsent or \emph{p}resent in read $R_i$ and $a(\cdot)$ and $p(\cdot)$ are the respective functions to sample from this distribution for a particular read.
With these probability distributions, we can account for any number of factors for which we know how they affect these probability distributions---but if we lack any further information for an alternative nucleotide, we can assume $Q_{R_i}$ to be a specification of these distributions for $R_i$, accounting for the uncertainties of the sequencing process.
With ${x \in \{A,C,G,T\}}$ the alternative nucleotide under inspection, we can thus compute:

\begin{align}
 a(Z_i) & =
 \label{eq:absent-phred}
 \begin{cases}
  10^{-Q_{R_i}/10}     & \text{ if } S_{R_i} = x    \\
  1 - 10^{-Q_{R_i}/10} & \text{ if } S_{R_i} \neq x
 \end{cases} \\
 p(Z_i) & = 1 - a(R_i)
 \label{eq:present-phred}
\end{align}

\subsection{Single cell latent variable model} \label{sec:sc-model}

By applying the Chapman-Kolmogorov equation with our binary hyperparameters to the single cell case\footnote{
 We will only be looking at single cell samples in the following, as the case of $\rho_b = \theta_b$ is the simpler one and analogous without the whole genome amplification considerations. Please also refer to the tumor and normal sample derivations in \cite{koster_enhancing_2019}, but note that we need no purity parameter, here.
}, we get

\begin{equation}
 L(\rho_s \mid Z_i^s) \propto~ \Prob(Z_i^s \mid \rho_s)= \int_{\omega_i^s,\xi_i^s}\Prob(Z_i^s \mid \omega_i^s,\xi_i^s)\times \Prob(\omega_i^s,\xi_i^s \mid \rho_s)\;d(\omega_i^s,\xi_i^s)
\end{equation}

which fans out to the following cases from equation \ref{eq.read-cases}:

\begin{equation}
 \begin{split}
  \Prob(Z_i^s \mid \rho_s) =~ &\Prob(Z_i^s \mid \omega_i^s=0) \times \Prob(\omega_i^s=0 \mid \rho_s) \\
  &+ \Prob(Z_i^s \mid \omega_i^s=1,\xi_i^s=0) \times \Prob(\omega_i^s=1,\xi_i^s=0 \mid \rho_s) \\
  &+ \Prob(Z_i^s \mid \omega_i^s=1,\xi_i^s=1) \times \Prob(\omega_i^s=1,\xi_i^s=1 \mid \rho_s) \\
 \end{split}
\end{equation}

Knowing that the mapping probability of reads is independent from the given (distorted) allele frequency $\rho_s$, and assuming a sampling probability of 1 for any mismapped read, this becomes:

\begin{equation}
 \begin{split}
  \Prob(Z_i^s \mid \rho_s) =~ &\Prob(\omega_i^s=0) \\
  &+ \Prob(Z_i^s \mid \omega_i^s=1,\xi_i^s=0) \times \Prob(\omega_i^s=1) \times \Prob(\xi_i^s=0 \mid \rho_s) \\
  &+ \Prob(Z_i^s \mid \omega_i^s=1,\xi_i^s=1) \times \Prob(\omega_i^s=1) \times \Prob(\xi_i^s=1 \mid \rho_s) \\
 \end{split}
\end{equation}

With Equation~\ref{eq.map-prob} for $\omega_i^s$, Equation~\ref{eq.var-prob} for $\xi_i^s$ and Equations~\ref{eq:absent-phred} and \ref{eq:present-phred} for the conditional probabilities of the data given an absence or presence of the observed nucleotide, we get:

\begin{equation}
 \label{eq:read_probability}
 \begin{split}
  \Prob(Z_i^s \mid \rho_s) &= (1 - \pi_i^s) \\
  &+~ a^s(Z_i^s)\times \pi_i^s \times (1 - \rho_s) \\
  &+~ p^s(Z_i^s) \times \pi_i^s \times \rho_s \\
 \end{split}
\end{equation}

As the $l$ reads are conditionally independent, given a particular (distorted) alternative nucleotide frequency $\rho_s$ (also see Equation \ref{eq.independence}), we can calculate the likelihood of such a particular $\rho_s$ with:

\begin{equation}
 \label{eq:read-product}
 L(\rho_s \mid\boldsymbol{Z}^s) \propto \Prob(\boldsymbol{Z}^s \mid \rho_s) = \prod_{i=1}^l \Prob(Z_i^s \mid \rho_s)
\end{equation}

With the resulting likelihoods for $\rho_s \in [0,1]$, we can calculate the likelihood of a particular $\theta_s$:

\begin{equation}
 \label{eq:theta-likelihood}
 L(\theta_s \mid\boldsymbol{Z}^s) \propto \Prob(\boldsymbol{Z}^s \mid \theta_s) = \sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \Prob(\rho_s = \frac{k}{l} \mid \theta_s) \right\}
\end{equation}

For $\theta_s \in \{0,\frac12,1\}$, with equation \ref{eq.rho-given-theta} specifying the cases of $\theta_s$ and Equation \ref{eq:BBLodato} further specifying $\Prob(\rho_s \mid \theta_s
 )$ for MDA, this gives us:

\begin{equation}
 \label{eq:thetas-likelihood}
 \begin{split}
  \Prob(\boldsymbol{Z}^s \mid \theta_s = 0) &=  \sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \cB\cB_{l,\alpha, \beta} (\rho_s = \frac{k}{l}) \right\}\\
  \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac12) &= \sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \left[ w(l) \times \cB\cB_{l,\alpha_1,\alpha_1}(\rho_s = \frac{k}{l}) + \right.\right.\\
   &\qquad \qquad \qquad \left.\left.(1-w(l)) \times \cB\cB_{l,\alpha_2,\alpha_2}(\rho_s = \frac{k}{l}) \right] \right\} \\
  \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) &=  \sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \cB\cB_{l,\beta, \alpha} (\rho_s = \frac{k}{l}) \right\}\\
 \end{split}
\end{equation}

For the current implementation, we use the empirical values for $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ as deduced by \cite{lodato_somatic_2015} (Table~\ref{tab:Lodato}).
But as mentioned above, these could be learned from a dataset at hand whenever high-confidence sets of heterozygous and homozygous sites are available (e.g.~by calling variants on a background bulk and cross-referencing with dbSNP).\\

\subsection{Bulk background: Accounting for Amplification Errors and Allele Dropout}

\begin{figure}[tbp]
 \includegraphics[width=.7\linewidth]{figs/Event_space.pdf}\\
 \caption{
  It depicts single cell events at a particular genomic site, defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
  The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
  A tabular representation of the same event definitions is given in Table~\ref{tab:Events} below (this is a larger copy of Figure~\ref{fig:prosolo_alt-calling}D in the main manuscript, reproduced  here for easier cross-reference with that table).\footnotesize\newline
  ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference
  }
 \label{fig:event-space}
\end{figure}

\begin{center}
 \begin{table}[tbp]
  \caption{
   Single cell events at a particular genomic site are defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
   A graphical representation of the same event definitions is given in Figure~\ref{fig:event-space}.
  }
  \label{tab:Events}
  \renewcommand{\arraystretch}{0.9}

  \begin{tabular}{p{22ex}p{12ex}p{12ex}p{33ex}}
   \toprule
   single cell event & $\theta_s(x)$   & $\theta_b(x)$ & implicit single cell genotype \\
   \toprule
   $\boldsymbol{E}_{\text{hom ref}}$      & $0$             & $[0,\frac12)$ & homozygous ref                \\
       \midrule
   $\boldsymbol{E}_{\text{ADO to ref}}$   & $0$             & $[\frac12,1)$ & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{err ref}}$      & $\{0,\frac12\}$ & $1$           & homozygous alt                \\
         \midrule
   $\boldsymbol{E}_{\text{err alt}}$      & $\{\frac12,1\}$ & $0$           & homozygous ref                \\
         \midrule
   $\boldsymbol{E}_{\text{het}}$          & $\frac12$       & $(0,1)$       & heterozygous                  \\
         \midrule
   $\boldsymbol{E}_{\text{ADO to alt}}$   & $1$             & $(0,\frac12]$ & heterozygous                  \\
       \midrule
   $\boldsymbol{E}_{\text{hom alt}}$      & $1$             & $(\frac12,1]$ & homozygous alt                \\
   \bottomrule
   \multicolumn{4}{l}{\footnotesize ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous,}\\
   \multicolumn{4}{l}{\footnotesize hom -- homozygous, ref -- reference} \\
   \bottomrule
  \end{tabular}
 \end{table}
\end{center}

Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequencies $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
We then define mutually exclusive subspaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$, providing a set of events $\mathfrak{E}$.
Assuming that there is no informative prior information about different allele frequencies (i.e. assuming a flat prior across all possible allele frequencies in both samples), we can then calculate the likelihood of such an event as the product of the likelihoods of the allele frequency ranges in the two samples.
For $\boldsymbol{E}_{\text{ADO to ref}}$---and using Equation \ref{eq:thetas-likelihood} to calculate the likelihoods of the allele frequency ranges---this for example becomes:

\begin{equation}
  \label{eq:ado-to-ref-likelihood-samples}
  \begin{split}
    \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{ADO to ref}})
      =~&\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) \times \Prob(\boldsymbol{Z}^b \mid \theta_b \in [\frac12,1))\\
      =~&\sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \cB\cB_{l,\alpha, \beta} (\rho_s = \frac{k}{l}) \right\} \times\\
        &\sum_{m=\frac{n}{2}}^{n-1} \Prob(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n})\\
  \end{split}
\end{equation}

Here, we use the fact that for the bulk sample without amplification, we can assume that $\theta_b = \rho_b$ and thus use a simpler form of Equation \ref{eq:theta-likelihood}.
Also, we take $n$ to be the total number of bulk reads and $m$ to be the possible numbers of reads with the alternative allele for this event---analogous to $k$ and $l$ for the single cell sample, and amounting to only taking point estimates of the likelihoods at possible alternative allele counts given the total coverage of the examined genomic site.

As all the defined events (Equation~\ref{eq:event-set}) are mutually exclusive (Figure~\ref{fig:event-space}, Table~\ref{tab:Events}), the sum of the likelihoods of all events yields the marginal probability:

\begin{equation}
  \label{eq:marginal-prob}
  \begin{split}
    \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b)
      =~&\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E})     =~\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e)}\\
      =~&\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{hom ref}})~+~ \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{hom alt}})~+\\
        &\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{het}})~+\\
        &\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{err alt}})~+~ \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{ADO to alt}})~+\\
        &\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{err ref}})~+~ \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{ADO to ref}})
  \end{split}
\end{equation}

Then, using Equation \ref{eq:read-product} for the likelihood of $\rho_s$ in the single cell sample and the equivalent relationship for the likelihood of $\theta_b$ in the bulk sample (with $j$ as an index running through all bulk sample reads), we get:

\begin{equation}
  \label{eq:ado-to-ref-likelihood-reads}
  \begin{split}
    \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{ADO to ref}})
      =~&\sum_{k=0}^{l} \left\{ \prod_{i=1}^{l} \Prob(Z_i^s \mid \rho_s = \frac{k}{l}) \times \cB\cB_{l,\alpha, \beta} (\rho_s = \frac{k}{l}) \right\} \times\\
        &\sum_{m=\frac{n}{2}}^{n-1} \prod_{j=1}^{n} \Prob(Z_j^b \mid \theta_b = \frac{m}{n})\\
  \end{split}    
\end{equation}

Using Equation \ref{eq:read_probability} for the likelihood of an individual read in the single cell sample given a $\rho_s$ and the respective formulation for an individual read in the bulk sample given a $\theta_b$, this will calculate the likelihood of the {\ttfamily ADO to ref} event as:

\begin{equation}
  \begin{split}
  \label{eq:ado-to-ref-likelihood-data}
    \Prob(\boldsymbol{Z}^s,&\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{ADO to ref}}) =\\
    &\sum_{k=0}^{l} \left\{\vphantom{\prod_{i}{l}} \prod_{i=1}^{l} \right.\left( (1 - \pi_i^s) + a^s(Z_i^s)\times \pi_i^s \times (1 - \frac{k}{l}) + p^s(Z_i^s) \times \pi_i^s \times \frac{k}{l} \right) \times \\
    &\qquad~~\left.\cB\cB_{l,\alpha, \beta} (\rho_s = \frac{k}{l}) \right\} \times\\
    &\sum_{m=\frac{n}{2}}^{n-1} \left\{ \prod_{j=1}^{n} \left( (1 - \pi_j^b) + a^b(Z_j^b)\times \pi_j^b \times (1 - \frac{m}{n}) + p^b(Z_j^b) \times \pi_j^b \times \frac{m}{n} \right) \right\}\\
  \end{split}
\end{equation}

Analogously, the likelihoods for all of the events as specified in Figure~\ref{fig:event-space} can be computed and their sum according to Equation \ref{eq:marginal-prob} is the marginal probability.
We can use the marginal probability to calculate the posterior probability of the {\ttfamily ADO to ref} event with:

\begin{equation}
  \begin{split}
  \label{eq:ado-to-ref-posterior-prob}
    \Prob(&\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =\\
    &~~\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times\\
    &~~\sum_{k=0}^{l} \left\{\vphantom{\prod_{i}{l}} \prod_{i=1}^{l} \right.\left( (1 - \pi_i^s) + a^s(Z_i^s)\times \pi_i^s \times (1 - \frac{k}{l}) + p^s(Z_i^s) \times \pi_i^s \times \frac{k}{l} \right) \times \\
    &~~\qquad~~\left.\cB\cB_{l,\alpha, \beta} (\rho_s = \frac{k}{l}) \right\} \times\\
    &~~\sum_{m=\frac{n}{2}}^{n-1} \left\{ \prod_{j=1}^{n} \left( (1 - \pi_j^b) + a^b(Z_j^b)\times \pi_j^b \times (1 - \frac{m}{n}) + p^b(Z_j^b) \times \pi_j^b \times \frac{m}{n} \right) \right\}\\
  \end{split}
\end{equation}

With analogous equations, we can compute the posterior probability for all of the events defined by ProSolo (Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
Each of them is reported in the output VCF or BCF file.
Based on the implicit single cell genotypes as given in Table~\ref{tab:Events}, we can then compute posterior probabilities for genotypes in the single cell by summing the posterior probabilities of all the events that imply each genotype.
For the heterozygous case (light blue fields in Figure~\ref{fig:event-space}), this would be:

\begin{equation}
 \label{eq:het-posterior-prob}
 \begin{split}
  \Prob(\text{heterozygous}_s \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =~&
    \Prob(\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
  &\Prob(\boldsymbol{E}_{\text{het}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
  &\Prob(\boldsymbol{E}_{\text{ADO to alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
 \end{split}
\end{equation}

The genotype with the maximum probability among the three possible genotypes is then chosen.
Similarly, we can also compute posterior probabilities for any compound event.
For example, the overall posterior probability of an allele dropout at a particular site ({\ttfamily ADO} events in Figure~\ref{fig:event-space} and Table~\ref{tab:Events}) is:

\begin{equation}
 \label{eq:ado-posterior-prob}
 \begin{split}
  \Prob(\text{allele dropout} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =~&
    \Prob(\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
  &\Prob(\boldsymbol{E}_{\text{ADO to alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
 \end{split}
\end{equation}

Another example is the probability of the presence of any alternative allele at a site (all blue fields in Figure~\ref{fig:event-space}, Equation \ref{eq:alt-posterior-prob} in the main manuscript).
This posterior probability is the one used for the main benchmarking of ProSolo against the other existing single cell specific variant callers (MonoVar, SCAN-SNV, SCcaller, and SCIPhI), as SCcaller and SCIPhI only call the presence (vs. absence) of an alternative nucleotide and this is thus the only measure we can compare all tools on.

\subsection{Implementation} \label{sec:implementation}

ProSolo is an easy-to-use command-line tool (following usability standards \citep{taschuk_ten_2017}) and its source code is available at \url{https://github.com/prosolo/prosolo}, including instructions for an easy installation (via Bioconda \citep{gruning_bioconda:_2018}). 
The main contribution in terms of software that ProSolo's development provided, is the implementation of its comprehensive statistical model that jointly looks at a single-cell and a bulk sample.
This was implemented as a part of Varlociraptor, a variant calling library implemented in Rust \citep{koster_enhancing_2019}.
Varlociraptor already makes use of all available sources of information regarding biases of normal sequencing data, induced by ambiguous alignments and sequencing errors, but had to be extended to accommodate this type of sample combination and to allow for accounting for amplification bias.
Because Varlociraptor had originally been specializing in indels and structural variants, implementing ProSolo also meant further investment in expanding Varlociraptor's capabilities for basic SNV analysis and required some improvements in the handling of sequencing alignment data in the upstream library rust-htslib \citep{rust-htslib_htslib_nodate}, a Rust wrapper around htslib \citep{htslib_c_nodate}. 
Further, we optimized Varlociraptor's calculations wherever possible, for example by reducing the number of redundant likelihood calculations and by caching frequently reused values.
And finally, numerical issues required an addition to the upstream library rust-bio \citep{koster_rust-bio:_2016}.

\paragraph{False Discovery Rate (FDR) Control of Event Probabilities} \label{sec:fdr}

Last---but not least---ProSolo can control the false discovery rate of any set of single cell events that a user defines.
At any scale of high throughput sequencing data (from targeted panels over whole exome sequencing to whole genome sequencing), we generate a large amount of event calls and thus have to control for false discoveries, e.g.~by controlling the false discovery rate \citep{benjamini_controlling_1995}.
Using Varlociraptor's functionality \citep{koster_enhancing_2019}, we can estimate thresholds on our posterior probabilities based on the approach described by \cite{muller_optimal_2004,muller_fdr_2006}.
For ProSolo, we extended Varlociraptor's capabilities from estimating such thresholds for the posterior probability of a single event to estimating thresholds for any combination of the above defined events.
For example, to control the false discovery rate for allele dropout events, we jointly obtain a threshold for the combination of the two ADO events in Figure~\ref{fig:prosolo_alt-calling}D.
Or, to control the false discovery rate in identifying the presence of an alternative allele, we jointly obtain a threshold for the combination of all the respective events in Equation \ref{eq:alt-posterior-prob}. 
This functionality, like the improvements mentioned in Section~\ref{sec:implementation}, is now also available in Varlociraptor's subcommands.

\section{Benchmarking} \label{sec:suppl-benchmarking}

All code used for benchmarking is available at: \url{https://www.github.com/prosolo/benchmarking_prosolo}.

\subsection{Datasets and Ground Truths} \label{sec:suppl-datasets}

We have performed our benchmarking analyses on two datasets, one previously published for the evaluation of SCcaller by \cite{dong_accurate_2017} and one generated especially for this Benchmarking (see Figure~\ref{fig:datasets} in the main manuscript).
For them, we generated different kinds of ground truths:\\

\textbf{Whole genome sequencing of almost identical kin-cells from a cell-line \citep{dong_accurate_2017}.}
The first dataset is from the publication of a previous variant caller for single cell sequencing data generated using MDA, SCcaller (\cite{dong_accurate_2017}; Figure~\ref{fig:datasets}A).
They started with one initial cell from a cell line and grew it in several steps (a complete overview of their setup is given in Figure 1 of \cite{dong_accurate_2017}).
After an initial mini-expansion, they selected one single cell as the founder for their secondary IL expansion into 20-30 cells.
From this they extracted two single cells (IL-11 and IL-12) that were sequenced after multiple displacement amplification (MDA).
Both cells have a minimum coverage of five of at least 80\% of genomic sites (with only about 90\% of the used reference genome represented, as this includes a largely uncovered chromosome Y and alternative haplotype contigs which would only be covered if the particular haplotype is present), with IL-11 having a more even coverage that leads to more sites with a usable coverage (Figure~\ref{fig:coverage-dists}A).
The rest of the kindred cells from that IL clone was then used as a bulk sequencing sample without MDA (IL-1C), showing a coverage profile similar to the whole genome amplified single cells (Figure~\ref{fig:coverage-dists}A).
We generated the ground truth genotypes for this dataset from the kindred clone IL-1C, which should be only very few cell divisions away from the single cells, and thus have almost no difference in the somatic mutations acquired.
The ground truth genotype was generated separately for variant sites and homozygous reference sites:
for variant sites, GATK HaplotypeCaller was run, followed by variant quality score recalibration; for homozygous reference sites, we used bcftools mpileup to identify sites with a coverage above 25 but no alternative nucleotide present.
This bulk sample (IL-1C) was only used as a ground truth and was not provided as input to any of the software compared here.
Three more distant clones (C1, C2, C3 in Figure~\ref{fig:datasets}A), grown from the cells after the initial mini-expansion, were jointly used as a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
With three separate bulk sequencing runs, their joint coverage of about 50 for 80\% of genomic sites is much higher than for the MDA single cells and the unamplified IL-1C bulk (Clones in Figure~\ref{fig:coverage-dists}A).\\

\begin{figure}[!tpb]
 \begin{minipage}{.99\linewidth}
  \includegraphics[width=.99\linewidth]{figs/Dong2017/Dong2017_coverage_dist.pdf}
 \end{minipage} \\
 \begin{minipage}{.99\linewidth}
  \includegraphics[width=.99\linewidth]{figs/Laehnemann2017/Laehnemann2017_coverage_dist.pdf}
 \end{minipage}
 \caption{
 Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples.
 Panel \textbf{A} shows the coverage of the single cells (IL-11, IL-12), the kindred ground truth clone (IL-1c) and the more distant background bulk (Clones C1, C2, and C3 joined into one sample) of the dataset from \cite{dong_accurate_2017}.
 Panel \textbf{B} shows the coverage of the single cells (PAG1, PAG2, PAG5, PAG9, PAG10) and the background bulk (PNG) of the granulocyte data generated for this evaluation.
 }
 \label{fig:coverage-dists}
\end{figure}

\textbf{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figure~\ref{fig:datasets}B).
Single cells from that cell population were isolated using a Fluidigm C1 microfluidics chip and subjected to multiple displacement amplification (MDA) using the GE Healthcare GenomiPhi Kit in its version adapted to that microfluidic chip.
With a panel of 16 custom primer pairs covering different genes across chromosomes used in a quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
For those cells, we performed whole exome capture using the Agilent SureSelectXT Human All Exon V5+UTRs Kit.
The paired-end sequencing library (2x101bp) was then prepared with the TruSeq V3 Kit and sequenced on a HiSeq 2500.
From the remaining sorted cell population, we extracted bulk DNA and submitted it to the same whole exome capture and DNA sequencing without MDA.\\

For the ground truth of this dataset, we could leverage previously published bulk whole exome sequencing data from the same person, their parents and three siblings (\cite{hoell_constitutional_2014}; Figure~\ref{fig:datasets}B).
For ground truth alternative allele calls, we ran three pedigree-aware variant callers on the family whole exome data.
To be able to run them reproducibly, we made all three of them available via bioconda \citep{gruning_bioconda:_2018}:
\begin{itemize}
  \item BEAGLE 4.0 \citep{browning_improving_2013}, bioconda version: {\ttfamily beagle=4.0\_06Jun17=1}
  \item polymutt \citep{li_likelihood-based_2012}, bioconda version: {\ttfamily polymutt=0.18=0}
  \item FamSeq in MCMC ({\ttfamily -method 3}) mode \citep{peng_rare_2013,peng_famseq:_2014}), bioconda version: {\ttfamily famseq=1.0.3=0}
\end{itemize}
From their calls, we created a consensus by only including sites where either all callers agree on a genotype or where a maximum of one caller has a \emph{missing} (as opposed to different genotype) call, while the other two callers agree on the genotype (Figure~\ref{fig:granulocytes-ground-truth}).
The use of the full pedigree ensures high confidence genotypes---however, please bear in mind that these represent the germline genotype.
Single cells will have accumulated somatic genotypes that differ from the germline and we are bound to mis-interpret those as false positive calls in our ground truth comparison.
At the same time, the number of such somatic variants will be small compared to the number of germline variants, and should hence keep our overall ground truth comparison valid.\\

\begin{figure}[!tpb]
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/ground_truth/Hoell2014_pedigree_consensus_calling.pdf}
 \caption{
   Constructing the germline ground truth genotype for our granulocytes dataset from the consensus of three pedigree-aware callers.
   Variants were called on a previously generated whole exome bulk sequencing data set of the person's family (Figure~\ref{fig:datasets}), which is separate from the single cell and bulk whole exome sequencing data generated for this study.
   Variants were called using beagle, FamSeq and polymutt with total counts of individual tools (bottom left bar plot) and tool intersections of sites called (top bar plot) in grey.
   Black fractions of bars represent the sites selected for our consensus: sites where either all callers agree upon a genotype or at least two callers agree upon a particular genotype and the third tool has a missing genotype (i.e.~doesn't contradict the genotype called by the other two).
 }
\label{fig:granulocytes-ground-truth}
\end{figure}

\subsection{Software and Parameters}

ProSolo was compared against the available tools for single nucleotide variant calling in single cell sequencing data from multiple displacement amplified (MDA) DNA, MonoVar \citep{zafar_monovar:_2016}, SCAN-SNV \citep{luquette_identification_2019}, SCcaller \citep{dong_accurate_2017}, and SCIPhI \citep{singer_single-cell_2018}.\\

For ProSolo (version {\ttfamily 0.6.1} installed via bioconda), each single cell was called against the respective cell's bulk (the joined clones 1, 2 and 3 for the \citep{dong_accurate_2017} dataset, PNG for our own dataset, see Figure~\ref{fig:datasets}).
Candidate sites to call were generated using bcftools mpileup, taking all sites where at least one read covered a non-reference nucleotide in the single cell or its respective bulk.
We ran two different modes of the software: Calling sites with a minimum read coverage of 1 in the single cell sample (default mode in Figure~\ref{fig:alt-calling_prec-rec}) or including zero-coverage sites, effectively imputing genotypes at these sites from the bulk sample (imputation).
The different filter thresholds used represent the false discovery rate ({\ttfamily --fdr}) that we can control for (Section~\ref{sec:fdr}).\\

For MonoVar (version {\ttfamily 0.0.1}, which is commit {\ttfamily e0cf2db} cleaned up for the bioconda version we created), parameters were set as recommended in the software repository README \citep{zafar_monovar_nodate}.
We ran two different modes of the software, with consensus filtering between cells of a population ({\ttfamily -c~1}) and without the filtering ({\ttfamily -c~0}).
The different filter thresholds used represent the parameter {\ttfamily -t} being varied.
However, turning on the consensus filtering did not cause any substantial changes, possibly due to the low number of cells sequenced (Figure~\ref{fig:alt-calling_prec-rec_global}).
Varying the parameter {\ttfamily -t} did not cause any substantial changes in MonoVar results, either.\\

SCAN-SNV \citep{luquette_identification_2019} was installed via the install instruction in the repository on 2019-10-30\footnote{
 This are the install instructions at the repository state at \url{https://github.com/parklab/scan-snv/blob/62b7dcbc21e6cac64dc384abdfa708120e33819c/README.md}.
 This includes the following packages and versions from the custom conda channel \url{https://anaconda.org/jluquette}:
 {\ttfamily r-scansnv~0.1}, {\ttfamily scansnv~0.9}, {\ttfamily r-fastghquad~1.0} and {\ttfamily natefoo-slurm-drmaa~1.2.0}.
 Further, it requires manual registration of the java executable for {\ttfamily gatk~3.8}.
} and patched locally to make it run.
The pipeline was then run with parameters set as recommended in the software repository {\ttfamily README.md} and the command line help, with the only changes made for a better comparability with the other tools.
Namely, we tried to call as sensitively as possible by setting: {\ttfamily  --min-sc-alt~1 --min-sc-dp~1 --min-bulk-dp~1}.
However, the default pipeline only provides calls for somatic variants (variants only found in single cells, not in the bulk, as opposed to all the present alternative alleles), heavily filtering candidate sites before calling.
We thus attempted a fairer alternative allele calling comparison by modifying the pipeline to remove the filter that eliminates alternative alleles also seen in the bulk sample and the filter that eliminates known dbsnp variation\footnote{
 I.e.~we edited the snakemake {\ttfamily rule scansnv\_somatic\_sites} to exclude the following requirements:
 {\ttfamily tab[,bulk.alt] == 0 \& tab[,bulk.idx] == '0/0' \& tab\$dbsnp == '.'}.
}.
The different filter thresholds used represent the {\ttfamily --fdr} parameter, which---according to the authors---does not formally control the false discovery rate \citep{luquette_identification_2019}.\\

For SCcaller (version {\ttfamily 1.2}, as available on bioconda), default parameters were set as recommended in the software repository README \citep{dong_sccaller_2018}.
We ran four different modes of the software, varying two things:
\begin{enumerate*}
 \item For the heterozygous candidate sites that SCcaller requires as input, we either used dbSNP \citep{sherry_dbsnp:_2001,ncbi_database_2016} entries or generated them from the bulk background sample that was also used for calling in ProSolo, running GATK HaplotypeCaller.
 As the dbSNP-derived heterozygous candidate sites consistently outperformed those generate from the bulk background sample, only results using dbSNP are reported in the figures, to provide a cleaner comparison.
 \item In addition to the recommended settings that require a minimum coverage of 10 (default, {\ttfamily --min~10 --minvar~4 --RD~20}), we also ran SCcaller with reduced coverage requirements to increase sensitivity for a fairer comparison with ProSolo (sensitive, {\ttfamily --min~4 --minvar~1 --RD~10}), thus including sites with a minimum read coverage of 4 and down to one read covering the alternative nucleotide.
\end{enumerate*}
The three filtering thresholds used represent the filtering levels available for the {\ttfamily -a cutoff} command-line parameter.
The filtering threshold gave certain leverage over the precision-recall trade-off and reducing the coverage requirements improved the sensitivity of results (Figure~\ref{fig:alt-calling_prec-rec_global}).\\

For SCIPhI (version {\ttfamily 0.1.4}, as available on bioconda), we ran two different modes:
\begin{enumerate*}
 \item The default parameters of the command line tool ({\ttfamily default}).
 \item Fewer iterations ({\ttfamily -l~400000}, because of the excessive run time) and more {\ttfamily sensitive} by turning off all heuristic filters ({\ttfamily --cwm~1 --mnp~1 --ms~1 --bns~0 --bnc~0 --ncf~0 --mnc~1}).
\end{enumerate*}
On the whole exome dataset with five cells, SCIPhI ran for three weeks for each of these modes.
On the whole genome dataset with two cells, an initial run of SCIPhI was killed due to a server problem after running for more than eight weeks.
An attempt to restart from intermediate data using the {\ttfamily --il} command line parameter had the model estimations escalating towards unrealistic values, and the crash was thus left unrecoverable.
A rerun on a server with more performant CPUs led to the sensitive mode finishing after 5 weeks and the default mode after 7.5 weeks.
As no option for filtering or thresholding based on sensitivity vs. specificity is available, only two unfiltered data points are given in the respective plots (Figure~\ref{fig:alt-calling_prec-rec_global}).

\subsection{Alternative Allele Calling}

The most precise single cell variant callers to date, SCcaller and SCIPhI, can only call the presence vs. the absence of an alternative allele (i.e.~the heterozygous and the homozygous alternative genotypes called jointly).
We have thus focused on this for the main part of our benchmarking.

For ProSolo, the posterior probability of the presence of an alternative allele is calculated with Equation \ref{eq:alt-posterior-prob}, which represent the blue Event areas in Figures~\ref{fig:prosolo_alt-calling}D and \ref{fig:event-space}.

Calculations of $p$recision and $r$ecall were based on true positives (TP, alternative allele present in the ground truth and called by the respective software), false positives (FP, the respective software calls the presence of an alternative allele that does not exist in the ground truth at that site), and false negatives (FN, the respective software makes no call or calls a homozygous reference genotype at a ground truth site with an alternative allele), using the formulas:

\begin{equation}
  \label{eq:precp-rec}
  \begin{split}
    p &= \frac{TP}{TP + FP} \\
    r &= \frac{TP}{TP + FN}
  \end{split}
\end{equation}

\begin{figure}[!tpb]
 \begin{minipage}{.36\linewidth}
  \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
 \end{minipage}
 \begin{minipage}{.55\linewidth}
  \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
 \end{minipage}
 \caption{
  Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
  }
 \label{fig:alt-calling_prec-rec_global}
\end{figure}

The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher. 
In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
For example, for a precision above 99\%, its maximum recall is 76.6\% compared to 70.5\% for MonoVar, 68.7\% for SCIPhI and 61.0\% for SCcaller.
SCAN-SNV achieves up to 99.2\% precision, but only 0.01\% recall.
This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.

Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
This dataset is considerably more challenging than the first one.
See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above 99\%, with ProSolo reaching a 20\% increase of recall to 17.6\%, compared to SCIPhI's 14.6\%, and SCcaller with 7.2\% (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
In comparison, MonoVar achieved a maximum precision of only 96.29\%. However, this was at a much higher recall (14.13\%) than for example SCcaller (9.55\% at a precision of 97.23\%).
SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
On this dataset, SCAN-SNV's recall increased to 0.16\% at a decreased maximum precision of 89.7\%.
Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.

\begin{figure}[!tpb]
 \begin{minipage}[t]{.35\linewidth}
 \vspace{0pt}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf} \newline
 \end{minipage}
 \begin{minipage}[t]{.64\linewidth}
 \vspace{0pt}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf} \newline
 \end{minipage}
 \caption{
 Cell-specific $F_1$ scores for different filtering threshold levels for \textbf{A} the whole genome dataset and \textbf{B} the whole exome dataset.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:F_1-scores}
\end{figure}

ProSolo's improvements in accuracy can be further formalised using the $F_1$ score, the harmonic mean of precision $p$ and recall $r$, with $\beta = 1$ for the general $F$ score defined as \citep{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}:\\

\begin{equation} \label{eq:F_1_score}
  F_{\beta} = (\beta^2 + 1) \times \frac{p r}{\beta^2 p + r}
\end{equation}

The per-cell $F_1$ score clearly demonstrates ProSolo's performance advantage (Figure~\ref{fig:F_1-scores}), even when restricting our comparison to ProSolo's (and SCIPhI's) default mode for a fairer comparison on the whole exome dataset with a germline ground truth (see caveat discussed above).

\begin{figure}[!tpb]
 \begin{minipage}[t]{.4\linewidth}
 \vspace{0pt}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo_FDR_ground_truth_vs_theoretical.pdf} \newline
 \end{minipage}
 \begin{minipage}[t]{.59\linewidth}
 \vspace{0pt}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo__FDR_ground_truth_vs_theoretical.pdf} \newline
 \end{minipage}
 \caption{
 False discovery rate (FDR, $1 - p$recision) as calculated based on the ground truth, compared to the theoretical FDR we controlled for using ProSolo, for (A) the whole genome data set and (B) the whole exome data set.
 The diagonal indicates identity of the ground truth FDR with the FDR that was controlled for.
 Any points underneath it indicate overly conservative FDR control, any points above it indicate an FDR control that is too lenient.\newline \footnotesize
  Thresholds used were applied via the following command line options:
  MonoVar {\ttfamily --t};
  ProSolo {\ttfamily --fdr};
  SCAN-SNV {\ttfamily --fdr};
  SCcaller {\ttfamily -a cutoff};
  SCIPhI {\ttfamily none available}.
  Software modes:
  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
 }
 \label{fig:FDR-ground-truth-vs-theoretical}
\end{figure}

\subsection{False Discovery Rate Control of Alternative Allele Calling} \label{sec:fdr-of-alt-calling}

Finally, a feature where ProSolo clearly stands out, is the control over the false discovery rate.
As can be seen in Figure~\ref{fig:alt-calling_prec-rec} (and Figure~\ref{fig:alt-calling_prec-rec_global}), ProSolo provides flexible control over precision vs. recall via specifying a false discovery rate of interest.
While no other tool provides a formal control over the easily interpretable false discovery rate, several of the tools provide other types of thresholds which we varied in attempts to achieve higher precision or recall.
However, none of them provide control over similar ranges of precision and recall.
Thus, ProSolo is the only tool that provides the user with the choice of either aiming for more discoveries at the cost of a higher rate of false discoveries, or at aiming for a more limited number of discoveries with higher confidence in each of them.

Further, Figure~\ref{fig:FDR-ground-truth-vs-theoretical}A demonstrates, that ProSolo provides a rigorous control over the false discovery rate on the whole genome dataset.
The only exception---in both default and imputation mode---is when attempting to control for a false discovery rate below 0.0001.
This indicates, that ProSolo's model still needs refinement in its resolution for sites with a very high likelihood of an alternative allele present.
A similar trend can be seen on the whole exome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B), but for any attempt to control for a false discovery rate below 0.02.
This is within the common range of values often chosen as a false discovery rate to control for in practice and not being able to actually achieve the false discovery rate control at such values is usually unacceptable.
However, as also discussed above, we generally expect the precision (of all tools) to be underestimated on this dataset: using the germline genotypes as ground truth results in misclassifying actual somatic variants in single cells as false positives.
This underestimation of precision on this dataset in turn means an overestimation of the false discovery rate ($1 - p$recision), and is thus to be expected.
This artefact of using a germline ground truth also shows in the discrepancy in the ground-truth based false discovery rate between ProSolo's default and imputation mode when controlling for higher false discovery rates.
As expected, imputation turned out less precise in the whole genome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}A).
In contrast, imputation seems to have improved precision for the whole exome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B).
This would suggest that the larger call set from ProSolo's imputation mode (with a higher recall) is also more precise than the smaller call set from the default mode.
However, this increase in precision is most likely explained by an imputation towards the majority genotype in the bulk sample that will usually favor the germline genotype at a site, no matter what the actual single cell genotype would be.
Taken together, this indicates that the false discovery rate that we can calculate with the given germline ground truth for the whole exome dataset is higher than the actual value would be with a more accurate (somatic) ground truth available.

\subsection{Allele dropout rate calculations} \label{sec:suppl-adr-calc}

Leveraging our ground truths and using three different ways to calculate the allele dropout rate, we can confirm the general validity of the respective event definitions (Figure~\ref{fig:prosolo_alt-calling}D) and evaluate limitations.
For this, we will focus on the set of sites where the respective ground truth call is heterozygous, as these are the sites where the dropout of one of the alleles can be identified clearly.\\

The first way of calculating the allele dropout rate leverages the posterior probabilities generated by the ProSolo model (Equation \ref{eq:ado-posterior-prob}).
At each ground truth heterozygous site, we sum up the probabilities of the two allele dropout events defined in ProSolo ("ADO to alt" and "ADO to ref" in Figure~\ref{fig:prosolo_alt-calling}D) to obtain a total allele dropout probability (see Equation \ref{eq:ado-posterior-prob}).
Obtaining the expected value of the allele dropout rate then constitutes adding those totals across all ground truth heterozygous sites with coverage and dividing by their number.
This gives us an expected allele dropout rate based on the data and the ProSolo model ("expected value" in Figure~\ref{fig:allele-dropout}):

\begin{equation}
    \frac{\displaystyle\sum_{\substack{\text{covered ground truth}\\\text{heterozygous sites}}}  \Prob(\text{allele dropout} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b )}{\text{\# covered ground truth heterozygous sites}}
    \label{eq.expected-allele-dropout-rate}
\end{equation}\\

The two ways to validate this expected allele dropout rate both depend on a comparison against the respective ground truth genotypes.
One important difference in the ground truth used here, compared to the ground truth used for alternative allele calling, is that we further filtered the ground truth for the \cite{dong_accurate_2017} dataset.
For this dataset, the basic ground truth was generated using GATK HaplotypeCaller and variant quality score recalibration.
However, this recalibration optimises on the presence (heterozgyous or homozygous alternative) vs. the absence (homozygous reference) of an alternative allele and filtered genotypes will not reliably distinguish between heterozygous genotypes and homozygous alternative genotypes.
We thus excluded all sites where the genotype quality ({\ttfamily GQ} FORMAT field) is below a PHRED score of 30.
This corresponds to a 0.1\% probability of having erroneously called the current genotype over the second most likely genotype, and thus also filters for a clear distinguishability between heterozygous and homozygous alternative genotypes.
This is important both for identifying allele dropout and for identifying exact genotypes as in Section~\ref{sec:suppl-gt-calling}.

The first validation of these rates from within ProSolo is to genotype all the ground truth heterozygous sites with ProSolo.
We take the most likely genotype (i.e.~the genotype whose coloured event area in Figure~\ref{fig:prosolo_alt-calling}D has the highest joint probability).
Then, every ground truth heterozygous site that ProSolo calls as homozygous is counted as a dropout site and the number of dropout sites is divided by the total number of ground truth heterozygous sites where the respective single cell had coverage.
This gives an allele dropout rate based on a comparison against the ground truth ("hom at ground truth het" in Figure~\ref{fig:allele-dropout}):

\begin{equation}
    \frac{\text{\# ground truth heterozygous sites called homozygous}}{\text{\# covered ground truth heterozygous sites}}
    \label{eq.hom-at-ground-truth-het-allele-dropout-rate}
\end{equation}\\

A second validation of these rates is another comparison against the ground truth, but without using ProSolo in any way.
For each single cell, using samtools mpileup, we identify all heterozygous ground truth sites with a coverage of at least 7.
Without amplification bias we could be reasonably sure to sample both alleles of a heterozygous site at coverage 7 (the probability to sample only one allele with a coverage of 7 would be: $2 \cdot 0.5^7 = 0.015625$).
We then count a site as an allele dropout, if there is one allele (reference or alternative) with no read coverage at all.
The allele dropout rate is the fraction of allele dropout sites within the respective cell's heterozygous ground truth sites ("no alt/ref read at ground truth het" in Figure~\ref{fig:allele-dropout}):

\begin{equation}
    \frac{\text{\# ground truth heterozygous site with only one allele covered}}{\text{\# covered ground truth heterozygous sites}}
    \label{eq.no-alt-ref-read-at-ground-truth-het-adr}
\end{equation}\\

The expected allele dropout rates based on the ProSolo probabilities for allele dropout fall into the range of previously published allele dropout rates \citep{wang_clonal_2014,hou_single-cell_2012,xu_single-cell_2012,ling_evaluation_2009,spits_whole-genome_2006,renwick_proof_2006,lodato_somatic_2015} ("published" in Figure~\ref{fig:allele-dropout}).
This analysis also shows that the ProSolo expected allele dropout rates, based on the model's probabilities, correspond with those based on the comparison of ProSolo genotypes with the ground truth (Figure~\ref{fig:allele-dropout}), demonstrating that the explicit modelling of allele dropout events is useful for genotyping.
However, in that comparison, the expected allele dropout rate is consistently underestimated on our own whole exome data ("granulocytes" in Figure~\ref{fig:allele-dropout}), while it is slightly overestimated for the data from \cite{dong_accurate_2017} ("Dong 2017" in Figure~\ref{fig:allele-dropout}).
These deviations in the estimation in different directions suggest that the empirical distributions we currently use do not fit all datasets.
Instead, they most likely depend on the respective whole genome amplification and library preparation procedures, as well as other possible batch effects.
This conclusion is further supported by the much lower naively calculated allele dropout rate in the high coverage cells of the two datasets (IL-11 in \cite{dong_accurate_2017}, PAG1 and PAG10 in our granulocytes, compare \ref{fig:coverage-dists}).
Here, the empirical distributions used seem to overestimate allele dropout and the performance of ProSolo could be further improved with distributions tailored to the specific dropout profile of the experiment on that individual cell.
However, this overestimation of the allele dropout probability does not seem to impact the genotyping resolution (see below, Figures~\ref{fig:gt-calling-true-false-colouring} and \ref{fig:gt-calling-gt-colouring}).

\subsection{Genotype calling} \label{sec:suppl-gt-calling}

As a further performance comparison, we want to include a direct comparison of the actual called genotypes against the ground truth genotypes of the respective data set.
As only ProSolo, MonoVar and SCAN-SNV provide actual genotype calls---a further refinement over the alternative allele calling described above---SCcaller and SCIPhI cannot be included in this comparison.
Further, the small numbers of calls made by SCAN-SNV would not be interpretable in the figures provided, and we have thus also excluded SCAN-SNV from this comparison.

Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
For ProSolo, this amounts to the default mode without imputation (which also ensures a fair comparison against the germline-based ground truth) and controlling for a false discovery rate of $0.2$.
For MonoVar, this means no consensus filtering and no filtering via the {\ttfamily --t} thresholding parameter.\\

\begin{figure}[!tpb]
 \begin{minipage}{.45\linewidth}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_genotyping_recall_prosolo_monovar.pdf}
  \end{minipage}
 \begin{minipage}{.53\linewidth}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_genotyping_recall_prosolo_monovar.pdf}
 \end{minipage}
 \begin{minipage}{.45\linewidth}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_genotyping_precision_prosolo_monovar.pdf}
  \end{minipage}
 \begin{minipage}{.53\linewidth}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_genotyping_precision_prosolo_monovar.pdf}
 \end{minipage}
 \caption{
  Genotype calling precision and recall in comparison against the respective ground truth genotypes (from IL-1c in the left plots; from PNG in the plots on the right).
  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
  Facet labels at the top of plots indicate the single cell.
  Facet labels on the right side of plots indicate the genotype:
  "1/1" for homozgyous alternative, "0/1" for heterozygous.
  \textbf{A} Recall of ground truth genotypes.
  Here, facet labels on the right side of plots indicate the ground truth genotype compared against.
  ProSolo provides a genotype call at more of the ground truth sites, with a slight but noticeable increase of false genotypes provided at ground truth heterozygous sites.
  \textbf{B} Precision of called genotypes.
  Here, facet labels on the right side of plots indicate the genotype of the caller.
  The higher amount of calls by ProSolo (most noticeable for heterozygous calls for the whole genome datasets, and homozygous alternative calls for the whole exome datasets) trades off with a loss in precision.
  }
\label{fig:gt-calling-true-false-colouring}
\end{figure}

\begin{figure}[!tpb]
 \begin{minipage}{.45\linewidth}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_genotyping_calls_per_ground_truth_prosolo_monovar.pdf} \\
  \end{minipage}
 \begin{minipage}{.53\linewidth}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_genotyping_calls_per_ground_truth_prosolo_monovar.pdf} \\
 \end{minipage}
 \begin{minipage}{.45\linewidth}
  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_genotyping_ground_truth_per_calls_prosolo_monovar.pdf} \\
  \end{minipage}
 \begin{minipage}{.53\linewidth}
  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_genotyping_ground_truth_per_calls_prosolo_monovar.pdf} \\
 \end{minipage}
 \caption{
  Genotype calling comparison against the respective ground truth genotypes (from IL-1c in the left plots; from PNG in the plots on the right).
  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
  Panels A and B correspond to the respective panels in Figure~\ref{fig:gt-calling-true-false-colouring}, but with a finer resolution of the false genotypes.
  Facets at the top of plots indicate the single cell.
  \textbf{A} Genotypes called by the respective tool are indicated by the colouring, for the respective ground truth genotypes  as indicated by the facets on the right side of the plots.
  \textbf{B} Ground truth genotypes are indicated by the colouring, for the genotypes called by the respective tool as indicated by the facets on the right side of the plots.
  Both tools show visible amounts of misclassification in most of the possible categories.
  }
\label{fig:gt-calling-gt-colouring}
\end{figure}

Plotting the overall precision and recall of ProSolo and MonoVar genotyping (Figure~\ref{fig:gt-calling-true-false-colouring}), demonstrates that the higher recall of ProSolo is limited to the calling of heterozygous genotypes for the whole genome datasets (IL-11, IL-12 in Figure~\ref{fig:gt-calling-true-false-colouring}A).
For the whole exome dataset, ProSolo achieves a higher recall of both heterozygous and homozygous alternative ground truth genotypes (PAG cells in Figure~\ref{fig:gt-calling-true-false-colouring}A).
However, the recall remains much lower in this dataset, possibly due to the more uneven coverage of the respective cells and the lower coverage of the background bulk sample (PNG in Figure~\ref{fig:coverage-dists}).

Wherever ProSolo generates more genotype calls---be they heterozygous or homozygous alternative calls---it also makes more errors (Figure~\ref{fig:gt-calling-true-false-colouring}B).
In the only case where ProSolo does not generate more calls than MonoVar---homozygous alternative calls in the whole genome datasets---it also does not show a noticeable increase in false genotype calls.

When looking at the exact genotype miscalls (Figure~\ref{fig:gt-calling-gt-colouring}), it becomes apparent that the higher amount of miscalls at higher recall in ProSolo is biased for heterozygous calls made by ProSolo (Figure~\ref{fig:gt-calling-gt-colouring}B):
Heterozygous calls made by ProSolo are more often at ground truth homozygous reference sites, whereas heterozygous calls at ground truth homozygous alternative sites do not occur more often than for MonoVar.
Thus, while MonoVar heterozygous calls are more likely to be miscalled homozygous alternative sites than miscalled homozygous reference sites, the inverse holds for ProSolo.
This indicates, that both models are not yet well-balanced.
For ProSolo, a further improvement could be to adapt the amplification bias distributions currently used (see Section~\ref{sec:better_beta-binom} for some suggestions).
Also, learning distributions from the data at hand could further improve precision and recall wherever data is more uniformly distributed than in the dataset the current distributions are derived from \citep{lodato_somatic_2015}.

Altogether, the genotype-level analysis also confirms that a higher and more uniform coverage of the single cell data leads to higher precision and recall in the variant calling, no matter which caller is used.
E.g.~for the---most evenly covered---whole genome sequencing data (Figure~\ref{fig:coverage-dists}A), both callers achieve a very high recall of ground truth homozygous alternative sites with a very high precision (left panels in Figures~\ref{fig:gt-calling-true-false-colouring} and \ref{fig:gt-calling-gt-colouring}).
Also, for the whole exome data (right panels in Figures~\ref{fig:gt-calling-true-false-colouring} and \ref{fig:gt-calling-gt-colouring}), recall is visibly higher for all genotypes across both tools in the samples with a higher and more uniform coverage (PAG1 and PAG10, see Figure~\ref{fig:coverage-dists}), and precision is noticeably higher in those samples for MonoVar.

\subsection{Further Software and Resources used}

All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo}.
An overview and citations of software and resources beyond those mentioned in the methods section are provided in Table~\ref{tab:software-overview}.

\begin{center}
 \begin{table}[tbp]
  \caption{Software and related resources used in this manuscript. All software versions as installed with conda.}
  \label{tab:software-overview}
  \begin{tabular}{llll}
    \toprule
    software / resource & version   & purpose                       & citation \\
    \toprule
    cutadapt            & 1.14      & adapter trimming              & \citep{martin_cutadapt_2011} \\
    \midrule
    trimmomatic         & 0.36      & quality trimming              & \citep{bolger_trimmomatic_2014} \\
    \midrule
    bwa mem             & 0.7.15    & read alignment / mapping      & \citep{li_aligning_2013} \\
    \midrule
    samtools            & 1.4.1     & handle SAM / BAM files        & \citep{li_sequence_2009} \\
    \midrule
    picardtools         & 2.9.2     & mark duplicates               &  \\
    \midrule
    bcftools            & 1.6       & handle VCF / BCF files        & \citep{danecek_variant_2011} \\
    \midrule
    mosdepth            & 0.2.5     & coverage calculations         & \citep{pedersen_mosdepth_2018} \\
    \midrule
    Okabe Ito palette   &           & colorblind-safe               & \citep{okabe_color_2008} \\
    \midrule
    ggthemes            & 0.2.5     & Okabe Ito palette             & \citep{arnold_ggthemes_2019} \\
    \midrule
    Brewer palettes     &           & colorblind-friendly           & \citep{brewer_colorbrewer_2006} \\
    \midrule
    RColorBrewer        & 1.1\_2     & Brewer palettes              & \citep{neuwirth_rcolorbrewer_2014} \\
    \midrule
    tidyverse           & 1.1.1     & ground truth comparisons      & \citep{wickham_welcome_2019} \\
                        & 1.3.0     & results data handling         &  \\
    \midrule
    ggplot2             & 3.2.1     & result plots                  & \citep{wickham_ggplot2_2016} \\
    \midrule
    Inkscape            & 0.92.4    & info-graphics, improved labels & \\
    \midrule
    VGAM                & 1.1\_1     & distributions in R            & \citep{yee_vgam_2008,yee_vector_2015} \\
    \midrule
    UpsetR              &           & set intersection plot \ref{fig:granulocytes-ground-truth} & \citep{conway_upsetr_2017} \\
    \bottomrule
  \end{tabular}
 \end{table}
\end{center}

\bibliographystyle{abbrvnat}
\bibliography{Benchmarking_ProSolo.bib}

\end{document}

### 2020-04-27 manuscript/figs/Dong2017/Dong2017_calculate_prosolo_ado-rate-against-ground-truth.R

@@ -1,0 +1,45 @@
+library('tidyverse')
+library('stringr')
+
+SCs <- c("IL-11","IL-12")
+
+remove(current)
+remove(counts)
+
+# read in ProSolo data
+
+for ( sc in SCs ) {
+  #for (f in fdrs ) {
+    file <- str_c("prosolo/", sc,".Clones.alt_sites_only.prosolo.min_sc_cov_1.minGQ_30.gt_matching.ML_set.tsv")
+    current <- as_tibble( read_tsv( file ) ) %>%
+                add_column( cell = sc, 
+                            software = "ProSolo"#,
+                            #filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ))
+                            )
+    if (!exists("counts")) {
+      counts <- current
+    } else {
+      counts <- bind_rows(counts,current)
+    }
+  #}
+}
+
+GT_levels <- c("not called", "1/1", "0/1", "0/0")
+
+# order factor levels of GT and PED_GT for consistent bar stacking
+total <- counts %>%
+  filter(PED_GT == "0/1" & !is.na(GT)) %>%
+  group_by(cell) %>%
+  summarise(total = sum(n))
+
+ado <- counts %>%
+  filter(PED_GT == "0/1" & GT != "0/1" & !is.na(GT)) %>%
+  group_by(cell) %>%
+  summarise(ado = sum(n))
+
+ado_rates <- full_join(ado, total, by = "cell") %>%
+  mutate(ado_rate = ado/total) %>%
+  select(cell, ado_rate)
+  
+write_tsv(ado_rates, "prosolo/single-cells.Clones.alt_sites_only.prosolo.min_sc_cov_1.minGQ_30.ado-rates-against-ground-truth.tsv")
+

### 2020-04-27 manuscript/figs/Dong2017/Dong2017_plot_coverage_dist.R

@@ -1,0 +1,54 @@
+library(tidyverse)
+library('RColorBrewer')
+library('extrafont')
+loadfonts(device = "pdf")
+
+samples <- c("IL-11", "IL-12", "IL-1c", "Clones")
+
+# Read in data
+
+current <- tibble()
+coverage_dists <- tibble()
+for (sample in samples) {
+  file <- str_c( sample, ".bps.mosdepth.global.dist.txt")
+      
+  current <- as_tibble( read_tsv( file,
+                                  col_names = c("region", "coverage", "fraction"),
+                                  col_types = "cid"
+                                  ) ) %>%
+              add_column( sample = sample ) %>%
+              mutate( sample = factor(sample, levels = samples, ordered = TRUE) )
+  
+  coverage_dists <- bind_rows(coverage_dists, current)
+}
+
+linetype_values = c(3, 2, 1, 4, 5)
+
+max_power <- 2
+powers <- 10**(0:max_power)
+majors <- c(1,3,5)
+minors <- 1:9
+major_ticks <- rep(majors, times = max_power+1)*rep(powers, each = length(majors))
+minor_ticks <- rep(minors, times = max_power+1)*rep(powers, each = length(minors))
+
+ggplot(coverage_dists %>% filter(region == "total") %>% filter(coverage != 0), aes(x = coverage, y = fraction) ) +
+  coord_cartesian( ylim = c(0,1), xlim = c(1,500) ) +
+  scale_x_continuous( limits = c(1,500),
+                      trans = 'log10',
+                      breaks = major_ticks,
+                      minor_breaks = minor_ticks
+                      ) +
+  scale_y_continuous( limits = c(0,1),
+                      breaks = seq( from = 0, to = 1, by = 1/5),
+                      minor_breaks = seq( from = 0, to = 1, by = 1/10) ) +
+  geom_line( aes( linetype = sample, colour = sample ) ) +
+  scale_linetype_manual(
+    values = linetype_values
+    ) +
+  scale_colour_brewer(palette = "Set1") +
+  theme_bw(base_size=24, base_family="Lato")  +
+  labs(x = "site coverage",
+       y = "fraction of sites",
+       tag = "A",
+       title = "Dong 2017, whole genome")
+ggsave("Dong2017_coverage_dist.pdf", device = cairo_pdf, width=10, height=6)

### 2020-04-27 manuscript/figs/Dong2017/Dong2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -1,0 +1,310 @@
+# packages, colors, palettes and stuff
+source("../common.R")
+
+cells <- c("IL-11", "IL-12")
+
+# Read in MonoVar stats
+
+# TODO: add "0-1" once computed
+ps <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
+cs <- c("0", "1") # indicator whether consensus filtering was on in MonoVar
+
+current <- tibble()
+positives_negatives <- tibble()
+for (cell in cells) {
+  for ( c in cs ) {
+    for ( p in ps ) {
+      file <- str_c("monovar/", cell, ".single_cells.", p, ".c", c, ".monovar.positives_negatives.tsv")
+      
+      if ( c == 1 ) {
+        mode = "MonoVar default"
+      } else {
+        mode  = "MonoVar no consensus"
+      }
+       
+      current <- as_tibble( read_tsv(file) ) %>%
+                  add_column( software = mode,
+                              cell = cell,
+                              filter = if_else(p == "1", 1.0, as.numeric( str_replace(p, '-', '.') )) )
+      
+      positives_negatives <- bind_rows(positives_negatives, current)
+    }
+  }
+}
+
+# Read in SCcaller stats
+
+alphas <- c("0-01", "0-05", "1-0" ) # different alpha values for filtering SCcaller results, 1.0 means no filtering
+dbsnp <- c("", ".dbsnp") # against custom het sites or known dbsnp sites
+low_cov <- c("", ".low-cov") # with repo settings or low-cov settings similar to publication
+
+for (cell in cells) {
+  for ( d in dbsnp ) {
+    for ( m in low_cov ) {
+      for ( a in alphas ) {
+          file <- str_c("sccaller/", cell, d, m, ".sccaller.varcall.cutoff.fil-amp-err_alpha_", a, ".positives_negatives.tsv")
+          
+          current <- as_tibble( read_tsv(file) ) %>%
+                      add_column( software = case_when(
+                                      d != "" & m == "" ~ "SCcaller default dbsnp",
+                                      d != "" & m != "" ~ "SCcaller sensitive dbsnp",
+                                      d == "" & m == "" ~ "SCcaller default bulk",
+                                      d == "" & m != "" ~ "SCcaller sensitive bulk"),
+                                  cell = cell,
[…]

### 2020-04-27 manuscript/figs/Dong2017/Dong2017_plot_prosolo-monovar_genotyping.R

@@ -1,0 +1,180 @@
+library('tidyverse')
+library('stringr')
+library('ggthemes')
+library('scales')
+library('RColorBrewer')
+
+Dong_SCs <- c("IL-11","IL-12")
+
+remove(current)
+remove(counts)
+
+# read in ProSolo data
+
+fdrs <- c( "0-2" )
+
+for ( sc in Dong_SCs ) {
+  for (f in fdrs ) {
+    file <- str_c("prosolo/fdr_alt-presence/", sc,".Clones.alt_sites_only.fdr_", f , "_alt-presence.prosolo.min_sc_cov_1.minGQ_30.gt_matching.ML_set.tsv")
+    current <- as_tibble( read_tsv( file ) ) %>%
+                add_column( cell = sc, 
+                            software = "ProSolo, default",
+                            filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') )) )
+    if (!exists("counts")) {
+      counts <- current
+    } else {
+      counts <- bind_rows(counts,current)
+    }
+  }
+}
+
+# Read in Monovar data
+for ( sc in Dong_SCs ) {
+  file <- str_c("monovar/", sc,".single_cells.1.c0.monovar.minGQ_30.gt_matching.tsv")
+  current <- as_tibble( read_tsv( file ) ) %>%
+              add_column(cell = sc,
+                         software = "MonoVar, no cons",
+                         filter = 1)
+  if (!exists("counts")) {
+    counts <- current
+  } else {
+    counts <- bind_rows(counts,current)
+  }
+}
+
+GT_levels <- c("not called", "1/1", "0/1", "0/0")
+
+# order factor levels of GT and PED_GT for consistent bar stacking
+counts <- counts %>%
+  mutate( match = case_when(
+                          is.na(GT) ~ "not called",
+                          GT == PED_GT ~ "correct GT",
+                          TRUE ~ "false GT"
+                 ),
+          match = factor(match, levels = c("not called", "false GT", "correct GT" ) ),
+          GT = replace_na(GT, "not called"),
+          PED_GT = replace_na(PED_GT, "not called"),
+          PED_GT = factor( PED_GT, levels = GT_levels, exclude = NULL, ordered = TRUE),
+          GT = factor( GT, levels = GT_levels, exclude = NULL, ordered = TRUE )
+  )
+
+# color naming for colorblind palette
[…]

### 2020-04-27 manuscript/figs/Laehnemann2017/Laehnemann2017_calculate_prosolo_ado-rate-against-ground-truth.R

@@ -1,0 +1,43 @@
+library('tidyverse')
+library('stringr')
+
+SCs <- c("PAG1","PAG2", "PAG5", "PAG9", "PAG10")
+
+remove(current)
+remove(counts)
+
+# read in ProSolo data
+
+for ( sc in SCs ) {
+    file <- str_c("prosolo/", sc,".PNG.alt_sites_only.prosolo.min_sc_cov_1.gt_matching.ML_set.tsv")
+    current <- as_tibble( read_tsv( file ) ) %>%
+                add_column( cell = sc, 
+                            software = "ProSolo"#,
+                            #filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ))
+                            )
+    if (!exists("counts")) {
+      counts <- current
+    } else {
+      counts <- bind_rows(counts,current)
+    }
+}
+
+GT_levels <- c("not called", "1/1", "0/1", "0/0")
+
+# order factor levels of GT and PED_GT for consistent bar stacking
+total <- counts %>%
+  filter(PED_GT == "0/1" & !is.na(GT)) %>%
+  group_by(cell) %>%
+  summarise(total = sum(n))
+
+ado <- counts %>%
+  filter(PED_GT == "0/1" & GT != "0/1" & !is.na(GT)) %>%
+  group_by(cell) %>%
+  summarise(ado = sum(n))
+
+ado_rates <- full_join(ado, total, by = "cell") %>%
+  mutate(ado_rate = ado/total) %>%
+  select(cell, ado_rate)
+  
+write_tsv(ado_rates, "prosolo/single-cells.PNG.alt_sites_only.prosolo.min_sc_cov_1.ado-rates-against-ground-truth.tsv")
+

### 2020-04-27 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_coverage_dist.R

@@ -1,0 +1,56 @@
+library(tidyverse)
+library('RColorBrewer')
+library('extrafont')
+loadfonts(device = "pdf")
+
+samples <- c("PAG1", "PAG2", "PAG5", "PAG9", "PAG10", "PNG")
+
+# Read in data
+
+current <- tibble()
+coverage_dists <- tibble()
+for (sample in samples) {
+  file <- str_c( sample, ".bps.SureSelectHumanAllExonV5plusUTR.target.mosdepth.region.dist.txt")
+      
+  current <- as_tibble( read_tsv( file,
+                                  col_names = c("region", "coverage", "fraction"),
+                                  col_types = "cid"
+                                  ) ) %>%
+              add_column( sample = sample ) %>%
+              mutate( sample = factor(sample, levels = samples, ordered = TRUE) )
+  
+  coverage_dists <- bind_rows(coverage_dists, current)
+}
+
+linetype_values = c(3, 2, 1, 4, 5, 6)
+color_palette <- c(brewer.pal(name = "Set1", n = 7)[1:5], brewer.pal(name = "Set1", n = 7)[7])
+
+max_power <- 2
+powers <- 10**(0:max_power)
+majors <- c(1,3,5)
+minors <- 1:9
+major_ticks <- rep(majors, times = max_power+1)*rep(powers, each = length(majors))
+minor_ticks <- rep(minors, times = max_power+1)*rep(powers, each = length(minors))
+
+ggplot(coverage_dists %>% filter(region == "total") %>% filter(coverage != 0), aes(x = coverage, y = fraction) ) +
+  scale_x_continuous( limits = c(1,500),
+                      trans = 'log10',
+                      breaks = major_ticks,
+                      minor_breaks = minor_ticks
+                      ) +
+  scale_y_continuous( limits = c(0,1),
+                      breaks = seq( from = 0, to = 1, by = 1/5),
+                      minor_breaks = seq( from = 0, to = 1, by = 1/10) ) +
+  geom_line( aes( linetype = sample, colour = sample ) ) +
+  scale_linetype_manual(
+    values = linetype_values
+    ) +
+  scale_colour_manual(
+    values = color_palette
+    ) +
+  theme_bw(base_size=24, base_family="Lato")  +
+  labs(x = "site coverage",
+       y = "fraction of exome sites",
+       tag = "B",
+       title = "granulocytes, whole exome")
+ggsave("Laehnemann2017_coverage_dist.pdf", device = cairo_pdf, width=10, height=6)

### 2020-04-27 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -1,0 +1,298 @@
+# packages, colors, palettes and stuff
+source("../common.R")
+
+cells <- c("PAG1", "PAG2", "PAG5", "PAG9", "PAG10")
+
+# Read in MonoVar stats
+
+# TODO: add "0-1" once computed
+ps <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
+cs <- c("0", "1") # indicator whether consensus filtering was on in MonoVar
+
+current <- tibble()
+positives_negatives <- tibble()
+for (cell in cells) {
+  for ( c in cs ) {
+    for ( p in ps ) {
+      file <- str_c("monovar/", cell, ".single_cells.", p, ".c", c, ".monovar.positives_negatives.tsv")
+      
+      if ( c == 1 ) {
+        mode = "MonoVar default"
+      } else {
+        mode  = "MonoVar no consensus"
+      }
+      
+      current <- as_tibble( read_tsv(file) ) %>%
+                  add_column( software = mode,
+                              cell = cell,
+                              filter = if_else(p == "1", 1.0, as.numeric( str_replace(p, '-', '.') )) )
+      
+      positives_negatives <- bind_rows(positives_negatives, current)
+    }
+  }
+}
+
+# Read in SCcaller stats
+
+alphas <- c("0-01", "0-05", "1-0" ) # different alpha values for filtering SCcaller results, 1.0 means no filtering
+dbsnp <- c("", ".dbsnp") # against custom het sites or known dbsnp sites
+low_cov <- c("", ".low-cov") # with repo settings or low-cov settings similar to publication
+
+for (cell in cells) {
+  for ( d in dbsnp ) {
+    for ( m in low_cov ) {
+      for ( a in alphas ) {
+          file <- str_c("sccaller/", cell, d, m, ".sccaller.varcall.cutoff.fil-amp-err_alpha_", a, ".positives_negatives.tsv")
+          
+          current <- as_tibble( read_tsv(file) ) %>%
+                      add_column( software = case_when(
+                                      d != "" & m == "" ~ "SCcaller default dbsnp",
+                                      d != "" & m != "" ~ "SCcaller sensitive dbsnp",
+                                      d == "" & m == "" ~ "SCcaller default bulk",
+                                      d == "" & m != "" ~ "SCcaller sensitive bulk"),
[…]

### 2020-04-27 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar_genotyping.R

@@ -1,0 +1,180 @@
+library('tidyverse')
+library('stringr')
+library('ggthemes')
+library('scales')
+library('RColorBrewer')
+
+SCs <- c("PAG1","PAG2", "PAG5", "PAG9", "PAG10")
+
+remove(current)
+remove(counts)
+
+# read in ProSolo data
+
+fdrs <- c( "0-2" )
+
+for ( sc in SCs ) {
+  for (f in fdrs ) {
+    file <- str_c("prosolo/fdr_alt-presence/", sc,".PNG.alt_sites_only.fdr_", f , "_alt-presence.prosolo.min_sc_cov_1.gt_matching.ML_set.tsv")
+    current <- as_tibble( read_tsv( file ) ) %>%
+                add_column( cell = sc, 
+                            software = "ProSolo, default",
+                            filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') )) )
+    if (!exists("counts")) {
+      counts <- current
+    } else {
+      counts <- bind_rows(counts,current)
+    }
+  }
+}
+
+# Read in Monovar data
+for ( sc in SCs ) {
+  file <- str_c("monovar/", sc,".single_cells.1.c0.monovar.gt_matching.tsv")
+  current <- as_tibble( read_tsv( file ) ) %>%
+              add_column(cell = sc,
+                         software = "MonoVar, no cons",
+                         filter = 1)
+  if (!exists("counts")) {
+    counts <- current
+  } else {
+    counts <- bind_rows(counts,current)
+  }
+}
+
+GT_levels <- c("not called", "1/1", "0/1", "0/0")
+
+# order factor levels of GT and PED_GT for consistent bar stacking
+counts <- counts %>%
+  mutate( match = case_when(
+                          is.na(GT) ~ "not called",
+                          GT == PED_GT ~ "correct GT",
+                          TRUE ~ "false GT"
+                 ),
+          match = factor(match, levels = c("not called", "false GT", "correct GT" ) ),
+          GT = replace_na(GT, "not called"),
+          PED_GT = replace_na(PED_GT, "not called"),
+          PED_GT = factor( PED_GT, levels = GT_levels, exclude = NULL, ordered = TRUE),
+          GT = factor( GT, levels = GT_levels, exclude = NULL, ordered = TRUE )
+  )
+
+# color naming for colorblind palette
[…]

### 2020-04-27 manuscript/figs/Laehnemann2017/ground_truth/Hoell2014_plot_pedigree_caller_consensus.R

@@ -1,0 +1,67 @@
+library(tidyverse)
+library('RColorBrewer')
+library('extrafont')
+library('UpSetR')
+loadfonts(device = "pdf")
+
+sets <-   tibble(
+            tool_set =
+              c( "FamSeq",
+                 "beagle",
+                 "polymutt",
+                 "FamSeq&beagle",
+                 "FamSeq&polymutt",
+                 "beagle&polymutt",
+                 "FamSeq&beagle&polymutt"
+                 )
+          )
+
+current <- tibble()
+
+file_all <- str_c("CCS_1_W.gatk_snps.all_polymutt_beagle_FamSeqMCMC.gtisec")
+all <- as_tibble( read_tsv(file_all, skip = 10, col_names = c("all"), col_types = "i") )
+
+file_cons <- str_c("CCS_1_W.gatk_snps.cons_no_reheader_polymutt_beagle_FamSeqMCMC.gtisec")
+cons <- as_tibble( read_tsv(file_cons, skip = 10, col_names = c("cons"), col_types = "i") )
+
+sets <- sets %>% 
+          add_column(all = all$all, cons = cons$cons) %>%
+          mutate(diff = all - cons)
+
+upset_cons <- fromExpression(
+                sets %>% select("tool_set", "cons") %>%
+                  spread(key = "tool_set", value = "cons")
+                ) %>% add_column(cons = "C")
+
+upset_diff <- fromExpression(
+                sets %>% select("tool_set", "diff") %>%
+                  spread(key = "tool_set", value = "diff")
+                ) %>% add_column(cons = "R")
+
+upset <- bind_rows(upset_cons, upset_diff) %>%
+          mutate(cons = factor(cons),
+                 beagle = as.integer(beagle),
+                 FamSeq = as.integer(FamSeq),
+                 polymutt = as.integer(polymutt)
+                 )
+
+cairo_pdf(file = "Hoell2014_pedigree_consensus_calling.pdf", width = 14, height = 8, onefile = F, family = "Lato")
+upset(upset,
+      queries = list(
+                  list( query = elements,
+                        params = list("cons", levels(upset$cons)[1]),
+                        color = "black",
+                        active = TRUE )
+                ),
+#      query.legend = "top",
+      main.bar.color = gray(0.6),
+      sets.bar.color = gray(0.6),
+      mainbar.y.label = "sites with shared\ngenotype calls",
[…]

### 2020-04-27 manuscript/figs/Lodato.SM.BetaBinomialPlots.R

@@ -1,0 +1,135 @@
+library(VGAM)
+
+# model for heterozygous sites
+al_1 <- function(cov) {
+  return(0.057378844*cov + 0.669733191)
+}
+
+al_2 <- function(cov) {
+  return(0.003233912*cov + 0.399261625)
+}
+
+w <- function(cov) {
+  return(0.000548761*cov + 0.540396786)
+}
+
+het_model <- function(x,cov) {
+  return(w(cov)*dbetabinom.ab(x,cov,al_1(cov),al_1(cov)) + (1-w(cov))*dbetabinom.ab(x,cov,al_2(cov),al_2(cov)))
+}
+
+## generate example value vectors for het model
+for (c in c(5,10,30,60)) {
+  k <- 0:c
+  vals <- sapply(k, het_model, cov = c)
+  print(c)
+  print(format(vals, nsmall = 12, scientific = FALSE))
+}
+
+# model for hom ref sites
+
+alpha <- function(cov) {
+  return(-0.000027183*cov + 0.068567471)
+}
+
+beta <- function(cov) {
+  return(0.007454388*cov + 2.367486659)
+}
+
+hom_ref_model <- function(x,cov) {
+  return( dbetabinom.ab(x,cov,alpha(cov),beta(cov)) )
+}
+
+## generate example value vectors for hom ref model
+for (c in c(5,10,30,60)) {
+  k <- 0:c
+  vals <- sapply(k, hom_ref_model, cov = c)
+  print(c)
+  print(format(vals, nsmall = 12, scientific = FALSE))
+}
+
+# model for hom alt sites (mirror of hom ref model)
+hom_alt_model <- function(x,cov) {
+  return( dbetabinom.ab(x,cov,beta(cov),alpha(cov)) )
+}
+
+## generate example value vectors for hom alt model
+for (c in c(5,10,30,60)) {
+  k <- 0:c
+  vals <- sapply(k, hom_alt_model, cov = c)
+  print(c)
+  print(format(vals, nsmall = 12, scientific = FALSE))
+}
+
+library(ggplot2)
+library(RColorBrewer)
+library(stringr)
+
+# coverages to plot
+covs <- c(20)
+
+# plot het model
+
+base <- ggplot(data.frame(x=c(5,100)), aes(x)) + coord_cartesian(ylim=c(0,7))
+base +
+  stat_function(fun=al_1, linetype=2, size = 1.2) +
+  stat_function(fun=al_2, linetype=3, size = 1.2) + 
[…]

### 2020-04-27 manuscript/figs/SM.BetaBinomialPlots.general.R

@@ -1,0 +1,54 @@
+library(VGAM)
+library(ggplot2)
+library(stringr)
+library(RColorBrewer)
+
+dot_to_minus <- function(x) {
+  str_replace(x, '\\.', '-')
+}
+
+plot_beta_binomial <- function(cov, white, black, shape_scale) {
+  p <- ggplot(data.frame(x=c(0,cov)), aes(x)) +
+    stat_function(fun=dbetabinom.ab, n = cov+1, args=list(size=cov, shape1 = shape_scale * black, shape2 = shape_scale * white),
+      geom="bar", fill = brewer.pal(4, "Greys")[3]) +
+    theme_bw(base_size = 20, base_family = "Helvetica-Narrow") +
+    labs(
+      x = "alternative allele count",
+      y = "probability",
+      subtitle = str_c("w_t = ", white, ", b_t = ", black, ", f = ", shape_scale)
+    )
+  print(p)
+  ggsave(str_c("Beta_binom_cov-", cov,
+               "_w-", dot_to_minus(white),
+               "_b-", dot_to_minus(black),
+               "_f-", dot_to_minus(shape_scale),
+               ".pdf"), device = cairo_pdf, width=9, height=6)
+}
+
+cov <- 20
+
+plot_beta_binomial(cov, 2, 2, 1)
+plot_beta_binomial(cov, 2, 2, .5)
+plot_beta_binomial(cov, 2, 2, .25)
+
+black_error <- 0.00000295 * cov
+white <- 4
+scaling <- .5
+
+plot_beta_binomial(cov, white, black_error, scaling)
+
+
+ggplot(data.frame(x=c(0,cov)), aes(x)) +
+  stat_function(fun=dbetabinom.ab, n = cov+1, args=list(size=cov, shape1 = scaling * black_error, shape2 = scaling * white),
+    geom="bar", fill = brewer.pal(4, "Greys")[3]) +
+  theme_bw(base_size = 20, base_family = "Helvetica-Narrow") +
+  coord_cartesian(ylim = c(0,.0001) ) +
+  labs(
+    x = "",
+    y = ""
+  )
+ggsave(str_c("Beta_binom_cov-", cov,
+             "_w-", dot_to_minus(white),
+             "_b-", dot_to_minus(black_error),
+             "_f-", dot_to_minus(scaling),
+             "_focus_tail.pdf"), device = cairo_pdf, width=9, height=6)

### 2020-04-27 manuscript/figs/common.R

@@ -1,0 +1,97 @@
+library(tidyverse)
+library('RColorBrewer')
+library('ggthemes')
+library('extrafont')
+loadfonts(device = "pdf")
+
+linetype_pal = c(
+  "MonoVar default" = 1,
+  "MonoVar no consensus" = 3,
+  "ProSolo default" = 1,
+  "ProSolo imputation" = 3,
+  "SCAN-SNV sensitive" = 3,
+  "SCcaller default dbsnp" = 1,
+  "SCcaller sensitive dbsnp" = 3,
+  "SCIPhI default" = 1,
+  "SCIPhI sensitive" = 3
+  )
+
+# Based on Okabe and Ito colorblind-safe palette (https://jfly.uni-koeln.de/color/)
+#' Tint a color, i.e. make it lighter by a factor
+#' 
+#' @description Tint a given color by a given factor
+#' 
+#' @param col HEX valued color to lighten up.
+#' @param factor Factor to use for tinting. A higher factor will produce a lighter color.
+tint <- function(col, factor) {
+  rgb_col <- col2rgb(col)
+  rgb(t( ( (255 - rgb_col ) * factor ) + rgb_col), maxColorValue=255)
+}
+
+factor <- 1/3
+
+Black <- colorblind_pal()(8)[[1]]
+light_Black <- tint(Black, factor)
+
+Vermillion_Red <- colorblind_pal()(8)[[7]]
+light_Vermillion_Red <- tint(Vermillion_Red, factor)
+
+Blue <- colorblind_pal()(8)[[6]]
+light_Blue <- tint(Blue, factor)
+
+reddish_Purple <- colorblind_pal()(8)[[8]]
+light_reddish_Purple <- tint(reddish_Purple, factor)
+
+Yellow <- colorblind_pal()(8)[[5]]
+
+# unused Okabe Ito colors
+# Sky_Blue <- colorblind_pal()(8)[[3]]
+# Orange <- colorblind_pal()(8)[[2]]
+# bluish_Green <- colorblind_pal()(8)[[4]]
+
+# a look at the selected colors
+# n_col <- 9
+# dummy_counts <- tibble(color = factor(seq_len(n_col)), count = rep(4, times=n_col))
+# 
+# bp <- ggplot(dummy_counts, aes(color, count)) +
+#         geom_col(aes(fill=color) ) +
+#         scale_fill_manual(
+#           values=c( Black, light_Black,
+#                 Vermillion_Red, light_Vermillion_Red,
[…]

### 2020-04-27 manuscript/figs/plot_prosolo_ado_expected-vs-ground-truth.R

@@ -1,0 +1,142 @@
+library(tidyverse)
+library('RColorBrewer')
+library('extrafont')
+loadfonts(device = "pdf")
+
+
+current <- tibble()
+
+file_l_gt <- str_c("Laehnemann2017/prosolo/single-cells.PNG.alt_sites_only.prosolo.min_sc_cov_1.ado-rates-against-ground-truth.tsv")
+ado_rates <- as_tibble( read_tsv(file_l_gt) ) %>%
+            add_column( dataset = "granulocytes",
+                        type = "ProSolo hom at ground truth het"
+            )
+
+file_d_gt <- str_c("Dong2017/prosolo/single-cells.Clones.alt_sites_only.prosolo.min_sc_cov_1.minGQ_30.ado-rates-against-ground-truth.tsv")
+current <- as_tibble( read_tsv(file_d_gt) ) %>%
+            add_column( dataset = "Dong 2017",
+                        type = "ProSolo hom at ground truth het"
+            )
+ado_rates <- bind_rows(ado_rates, current)
+
+file_known <- str_c("known_ados.tsv")
+current <- as_tibble( read_tsv(file_known) ) %>%
+            mutate( dataset = citation,
+                    ado_rate = ADO) %>%
+            select( dataset, ado_rate ) %>%
+            add_column( cell = "published",
+                        type = "published rate"
+            )
+ado_rates <- bind_rows(ado_rates, current)
+
+cells_l <- c("PAG1", "PAG2", "PAG5", "PAG9", "PAG10")
+for (cell in cells_l) {
+  file <- str_c("Laehnemann2017/prosolo/ado-rate/", cell, ".PNG.alt_sites_only.prosolo.min_sc_cov_1.ground_truth_hets_only.expected-ado-rate.tsv")
+  current <- as_tibble( read_tsv(file) ) %>%
+              add_column( dataset = "granulocytes",
+                          type = "ProSolo expected value ADO events",
+                          cell = cell
+              )
+  ado_rates <- bind_rows(ado_rates, current)
+}
+for (cell in cells_l) {
+  file <- str_c("Laehnemann2017/", cell, ".CCS_1_W.ground_truth.het_snps_only.allele_dropout_rate.tsv")
+  current <- as_tibble( read_tsv(file) ) %>%
+              transmute(ado_rate = ALLELE_DROPOUT_RATE) %>%
+              add_column( dataset = "granulocytes",
+                          type = "no alt/ref read at ground truth het",
+                          cell = cell
+              )
[…]

### 2020-05-13 manuscript/main.tex
minor updates and clarifications introduced after preprint

@@
-\documentclass[fleqn,12pt,inline]{wlscirep}
+\documentclass[12pt,inline]{wlscirep}
 
@@
 
-%\usepackage[left]{lineno}
-%\linenumbers
+\usepackage[left]{lineno}
+\linenumbers
 
@@
 
-\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data}
+\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
 \author[1,2,3,4]{David Lähnemann}
@@
 \author[1,2,+,*]{Alice C. McHardy}
-\author[5,+,*]{Alexander Schönhuth}
+\author[5,6,+,*]{Alexander Schönhuth}
 \affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
@@
 \affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
+\affil[6]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
 \affil[+]{shared last authors}
-\affil[*]{To whom correspondence should be addressed: \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de}, \href{as@cwi.nl}{as@cwi.nl}.}
+\affil[*]{To whom correspondence should be addressed:
+ \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
+ \href{as@cwi.nl}{as@cwi.nl}.}
 
-\begin{abstract}
+\begin{document}
 
+\flushbottom
+\maketitle
+
+\section*{\Large Abstract}
+{\sffamily
 Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
@@
 ProSolo's model is implemented into a flexible framework, encouraging extensions. The source code and usage instructions are available at: \url{https://github.com/prosolo/prosolo}
-\end{abstract}
-
-\begin{document}
-
-\flushbottom
-\maketitle
-
-\thispagestyle{empty}
+}
 
@@
 We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
-Importantly, our model further enables a computationally efficient implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_enhancing_2019}.
+Importantly, our model further enables a computationally efficient implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
 ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.
@@
 At the same time, one of the major features of the core model and its implementation is that it can easily be adapted to flexibly deal with other sampling setups, so it could be extended to further scenarios.
-For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_enhancing_2019}.
+For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_varlociraptor_2020}.
 
@@
 \begin{equation}
-% TODO: center equations
 	\label{eq:BBLodato}
@@
 
-Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_enhancing_2019}.
+Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.
 
@@
 For any such compound event, we can estimate a threshold on the posterior probabilities that controls for a specified false discovery rate. 
-This is based on the approach described by Müller et al.\cite{muller_optimal_2004,muller_fdr_2006}---for further details see the Supplement (Section~\ref{sec:fdr}) and Köster et al\cite{koster_enhancing_2019}.
+This is based on the approach described by Müller et al.\cite{muller_optimal_2004,muller_fdr_2006}---for further details see the Supplement (Section~\ref{sec:fdr}) and Köster et al\cite{koster_varlociraptor_2020}.
 
@@
 ProSolo is an easy-to-use command-line tool---following usability standards\cite{taschuk_ten_2017}---and its source code is available at \url{https://github.com/prosolo/prosolo}, including instructions for an easy installation via Bioconda\cite{gruning_bioconda:_2018}.
-Its main contribution in terms of software is the implementation of its comprehensive statistical model into the Rust variant calling library of Varlociraptor\cite{koster_enhancing_2019}.
+Its main contribution in terms of software is the implementation of its comprehensive statistical model into the Rust variant calling library of Varlociraptor\cite{koster_varlociraptor_2020}.
 See the Supplement (Section~\ref{sec:implementation}) for further implementation details.
@@
 For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
-All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo}.
+All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769116}).
 
@@
  \begin{minipage}{.40\linewidth}
-  \includegraphics[height=45ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
+  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
  \end{minipage}
  \begin{minipage}{.55\linewidth}
-  \includegraphics[height=45ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
+  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
  \end{minipage}
@@
   \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
-  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\\
+  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline \footnotesize
   Threshold parameters:
@@
 
-The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}A and \ref{fig:alt-calling_prec-rec_global}A), at recall rates of 45\% and higher.
+The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}A and \ref{fig:alt-calling_prec-rec_global}A), at recall rates of .45 and higher.
 In comparison to all other tools, ProSolo shows striking increases in recall.
-For example, an increase of nearly 10\% for a precision above 99\%, where its maximum recall is 76.6\%, compared to 70.5\% for MonoVar, 68.7\% for SCIPhI and 61.0\% for SCcaller.
-The only exception with a recall of 0.01\% is SCAN-SNV (at a precision of 99.2\%).
+For example, an increase of nearly 10\% for a precision above .99, where its maximum recall is .766, compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
+The only exception with a recall of .0001 is SCAN-SNV (at a precision of .992).
 This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.
@@
 Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}B).
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above 99\%, with ProSolo reaching a 20\% increase of recall to 17.6\%, compared to SCIPhI's 14.6\%, and SCcaller with 7.2\% (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
-In comparison, MonoVar achieved a maximum precision of only 96.29\%. However, this was at a much higher recall (14.13\%) than for example SCcaller (9.55\% at a precision of 97.23\%).
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .176, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
 SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
 
-On this dataset, SCAN-SNV's recall increased to 0.16\% at a decreased maximum precision of 89.7\%.
+On this dataset, SCAN-SNV's recall increased to .0016 at a decreased maximum precision of .897.
 Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
@@
 We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, \ref{fig:F_1-scores}B and \ref{fig:FDR-ground-truth-vs-theoretical}B for reference).
+
+As almost all tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
+However, it should be noted that SCIPhI took from 3 weeks on the whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
 
@@
 
-Finally, the modular implementation of ProSolo within the context of the Varlociraptor library\cite{koster_enhancing_2019} facilitates the implementation of further features, such as the ones mentioned above, read-backed phasing\cite{bohrson_linked-read_2019,hard_conbase:_2019} or even more variable models of amplification bias\cite{luquette_identification_2019} that could be integrated to exhaustively leverage MDA data information content.
+Finally, the modular implementation of ProSolo within the context of the Varlociraptor library\cite{koster_varlociraptor_2020} facilitates the implementation of further features, such as the ones mentioned above, read-backed phasing\cite{bohrson_linked-read_2019,hard_conbase:_2019} or even more variable models of amplification bias\cite{luquette_identification_2019} that could be integrated to exhaustively leverage MDA data information content.
 Since the Varlociraptor library also provides advanced functionality for the calling of insertions, deletions and multiple nucleotide variants (MNVs), one of the immediately following next steps will be to adapt ProSolo to calling those in single cells.\\

### 2020-05-13 manuscript/supplement.tex
minor updates and clarifications introduced after preprint

@@
 We now define $Z_i[g]$ ($i\in\{1,\dots,l\}$) to be the collection of the observable variables for the alignment of one of $l$ reads $R_i \in \cR$ to a genome position $g$.
-Omitting index $g$, this is the vector ${Z_i = \langle S_{R_i}, Q_{R_i}, M_{R_i} \rangle}$ and drawing on this information, we now introduce hyperparameters analogously to the model for tumor-normal variant calling as outlined by \cite{koster_enhancing_2019} for Varlociraptor.
+Omitting index $g$, this is the vector ${Z_i = \langle S_{R_i}, Q_{R_i}, M_{R_i} \rangle}$ and drawing on this information, we now introduce hyperparameters analogously to the model for tumor-normal variant calling as outlined by \cite{koster_varlociraptor_2020} for Varlociraptor.
 These will be the latent variables $\omega_i\in\{0,1\}$ to model correct placement of an alignment ($\omega_i=1$ if correct and $\omega_i=0$ otherwise) and $\xi_i\in\{0,1\}$ to specify whether an alignment $Z_i$ is associated with the alternative nucleotide ($\xi_i=1$) or not ($\xi_i=0$).
@@
  We will only be looking at single cell samples in the following, as the case of $\rho_b = \theta_b$ is the simpler one and analogous without the whole genome amplification considerations.
-Please also refer to the tumor and normal sample derivations in \cite{koster_enhancing_2019}, but note that we need no purity parameter, here.
+Please also refer to the tumor and normal sample derivations in \cite{koster_varlociraptor_2020}, but note that we need no purity parameter, here.
 }, we get
@@
 The main contribution in terms of software that ProSolo's development provided, is the implementation of its comprehensive statistical model that jointly looks at a single-cell and a bulk sample.
-This was implemented as a part of Varlociraptor, a variant calling library implemented in Rust \citep{koster_enhancing_2019}.
+This was implemented as a part of Varlociraptor, a variant calling library implemented in Rust \citep{koster_varlociraptor_2020}.
 Varlociraptor already makes use of all available sources of information regarding biases of normal sequencing data, induced by ambiguous alignments and sequencing errors, but had to be extended to accommodate this type of sample combination and to allow for accounting for amplification bias.
@@
 At any scale of high throughput sequencing data (from targeted panels over whole exome sequencing to whole genome sequencing), we generate a large amount of event calls and thus have to control for false discoveries, e.g.~by controlling the false discovery rate \citep{benjamini_controlling_1995}.
-Using Varlociraptor's functionality \citep{koster_enhancing_2019}, we can estimate thresholds on our posterior probabilities based on the approach described by \cite{muller_optimal_2004,muller_fdr_2006}.
+Using Varlociraptor's functionality \citep{koster_varlociraptor_2020}, we can estimate thresholds on our posterior probabilities based on the approach described by \cite{muller_optimal_2004,muller_fdr_2006}.
 For ProSolo, we extended Varlociraptor's capabilities from estimating such thresholds for the posterior probability of a single event to estimating thresholds for any combination of the above defined events.
@@
   \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
-  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline \footnotesize
+  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline \footnotesize
   Thresholds used were applied via the following command line options:
@@
 In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
-For example, for a precision above 99\%, its maximum recall is 76.6\% compared to 70.5\% for MonoVar, 68.7\% for SCIPhI and 61.0\% for SCcaller.
-SCAN-SNV achieves up to 99.2\% precision, but only 0.01\% recall.
+For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
+SCAN-SNV achieves up to .992 precision, but only .0001 recall.
 This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.
@@
 See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above 99\%, with ProSolo reaching a 20\% increase of recall to 17.6\%, compared to SCIPhI's 14.6\%, and SCcaller with 7.2\% (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
-In comparison, MonoVar achieved a maximum precision of only 96.29\%. However, this was at a much higher recall (14.13\%) than for example SCcaller (9.55\% at a precision of 97.23\%).
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .176, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
 SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
-On this dataset, SCAN-SNV's recall increased to 0.16\% at a decreased maximum precision of 89.7\%.
+On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
 Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
@@
 We thus excluded all sites where the genotype quality ({\ttfamily GQ} FORMAT field) is below a PHRED score of 30.
-This corresponds to a 0.1\% probability of having erroneously called the current genotype over the second most likely genotype, and thus also filters for a clear distinguishability between heterozygous and homozygous alternative genotypes.
+This corresponds to a 0.001 probability of having erroneously called the current genotype over the second most likely genotype, and thus also filters for a clear distinguishability between heterozygous and homozygous alternative genotypes.
 This is important both for identifying allele dropout and for identifying exact genotypes as in Section~\ref{sec:suppl-gt-calling}.
@@
 
-All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo}.
+All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769116}).
 An overview and citations of software and resources beyond those mentioned in the methods section are provided in Table~\ref{tab:software-overview}.

### 2020-05-13 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -241,21 +241,25 @@ ggplot(avg_per_software_and_parameter %>%
   coord_cartesian( ylim = c(0,1), xlim = c(0,1) ) +
   geom_line( aes( color = software, linetype = software ) ) +
   scale_linetype_manual(
-    values = linetype_pal
+    values = linetype_pal,
+    labels = software_labels
     ) +
   geom_point( aes(shape = filter, color = software, fill = software), size = 4) +
+  scale_colour_manual(
+    values = software_pal,
+    labels = software_labels
+    ) +
+  scale_fill_manual(
+    values = software_pal,
+    labels = software_labels
+    ) +
   scale_shape_manual(
     values = shape_pal
     ) +
-  scale_colour_manual(
-    values = software_pal
-    ) +
-  scale_fill_manual(
-    values = software_pal
-    ) +
   guides(
-    linetype = guide_legend( keywidth = 3),
-    shape = guide_legend( title = "threshold used") ) +
+    linetype = guide_legend( keywidth = 3, keyheight = 1.5),
+    shape = guide_legend( title = "threshold used")
+    ) +
   theme_bw(base_size=24, base_family="Lato")  +
   labs(x = "average recall",
        y = "average precision",
@@ -271,6 +275,7 @@ ggplot(avg_per_software_and_parameter %>%
   geom_line( aes( color = software, linetype = software ) ) +
   scale_linetype_manual(
     values = linetype_pal,
+    labels = software_labels,
     guide = "legend"
     ) +
   geom_point( aes(shape = filter, color = software, fill = software), size = 4) +
@@ -280,14 +285,16 @@ ggplot(avg_per_software_and_parameter %>%
     ) +
   scale_colour_manual(
     guide = "legend",
+    labels = software_labels,
     values = software_pal
     ) +
   scale_fill_manual(
     guide = "legend",
+    labels = software_labels,
     values = software_pal
     ) +
   guides( 
-    linetype = guide_legend( keywidth = 3),
+    linetype = guide_legend( keywidth = 3, keyheight = 1.5),
     shape = guide_legend( title = "threshold used")
     ) +
   theme_bw(base_size=24, base_family="Lato") +

### 2020-05-13 manuscript/figs/common.R

@@ -82,6 +82,18 @@ software_pal <- c(
   "SCIPhI sensitive" = light_reddish_Purple
 )
 
+software_labels <- c(
+   "MonoVar default",
+   "MonoVar no consensus",
+   "ProSolo default",
+   bquote('ProSolo imputation'^-B),
+   "SCAN-SNV sensitive",
+   "SCcaller default dbsnp",
+   "SCcaller sensitive dbsnp",
+   "SCIPhI default",
+   bquote('SCIPhI sensitive'^-B)
+)
+
 shape_pal <-
   c("1e-06" = 17,
     "1e-05" = 4,

### 2021-01-03 analysis_pipelines/Wang2014/somatic_validated_mutations/nature13600-s1-tables_to_ground_truth.py
add Wang2014 pipeline config and validated somatic mutations from original paper, including parsing and filtering script

@@ -1,0 +1,24 @@
+import pandas
+import re
+
+from pyliftover import LiftOver
+liftover = LiftOver('hg18', 'hg19')
+
+clonal = pandas.read_csv("nature13600-s1-table-s6-clonal.tsv", sep = "\t")
+subclonal = pandas.read_csv("nature13600-s1-table-s7-subclonal.tsv", sep = "\t")
+
+all = pandas.concat( [clonal, subclonal] )
+
+# keep only validated variants, i.e. those with a Duplex_P_val smaller 0.01 and not NA
+filtered = all[ all['Duplex_P_val'].str.contains('0,0[123456789]|NA|0,[123456789]') == False ]
+
+with open("Wang2014_ground_truth_non_synonymous_variants.hg18_to_hg19.tsv", mode='w') as out:
+    print( '\t'.join( filtered.columns.values.tolist() ), file=out, end='\n')
+    for index, row in filtered.iterrows():
+        print( row['chrom'], row['pos'], row['REF'], row['VAR'] )
+        lo = liftover.convert_coordinate(row['chrom'],row['pos'] - 1)[0]
+        row['chrom'] = lo[0]
+        row['pos'] = lo[1] + 1
+        print( '\t'.join(map(str, row)) )
+        print( '\t'.join(map(str, row)), file=out, end='\n' )
+

### 2021-01-03 analysis_pipelines/scripts/prosolo.P-N_stats.alt_calls.ground_truth.R
add legacy prosolo positives-negatives calculation R-script

@@ -1,0 +1,51 @@
+#!/usr/bin/env Rscript
+
+library('tidyverse')
+library('argparser')
+
+argv <- arg_parser("True and false positive / negative calculations on ProSolo results FDR filtered by ALT allele presence.") %>%
+        add_argument("--gt",
+		     help="file with ground truth genotypes (incl. path), contents: `CHROM\tPOS\tGT`") %>%
+        add_argument("--calls", help="file with ALT calls for one single cell") %>%
+        add_argument("--out", help="output file") %>%
+        parse_args();
+
+PED_GTs <- as_tibble( read_tsv( argv$gt, col_names = c("CHROM", "POS", "PED_GT") ) ) %>%
+  mutate( PED_GT = factor(PED_GT, levels = c("1/1", "0/1", "0/0", NA) ),
+          CHROM = factor(CHROM) )
+
+probs <- as_tibble(
+  read_tsv(
+    argv$calls,
+    col_types = "cic",
+    col_names = c("CHROM", "POS", "call"),
+    na = c("." )
+    ) ) %>%
+  group_by( CHROM, POS ) %>%
+  summarise( call = if_else( any( call == "ALT" ), "ALT", "REF" ) ) %>%
+  ungroup() %>%
+  right_join(PED_GTs, by = c("CHROM", "POS")) %>%
+  mutate(ped_alt_set_match = case_when(
+           is.na( PED_GT ) ~ "no_ground_truth",
+           is.na( call ) & (PED_GT == "0/1" | PED_GT == "1/1") ~ "P_not_called",
+           is.na( call ) & (PED_GT == "0/0") ~ "N_not_called",
+           call == "ALT" & (PED_GT == "0/1" | PED_GT == "1/1") ~ "TP",
+           call == "ALT" & (PED_GT == "0/0") ~ "FP",
+           call == "REF" & (PED_GT == "0/1" | PED_GT == "1/1") ~ "FN",
+           call == "REF" & (PED_GT == "0/0") ~ "TN" ),
+         ped_alt_set_match = 
+           factor( ped_alt_set_match,
+                   levels=c("TP", "FP", "FN", "TN", "P_not_called","N_not_called", "no_ground_truth") )
+	 ) %>%
+  select( CHROM, POS, ped_alt_set_match )
+
+## ALT presence calculations
+
+mls_pn <- probs %>%
+  count(ped_alt_set_match) %>%
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/multextract.pl
add scVILP scripts in version used for attempts in running scVILP

@@ -1,0 +1,119 @@
+# Written  by Dan Gusfield,  copyright 2009
+#  Permission to use this program comes with no guarantees. You may use it if you don't laugh at it and
+#  #  don't expect the author to understand or explain the code in the future.
+#  #  Permission is not granted for redistribution or distribution of modified versions. Please contact
+#  #  the author concerning distributing any modifications.
+#
+# July 18, 2008 modify multextract5.pl to work for a variable number of states, k, input in the call.
+#multextract.pl
+# Dec. 30, 2007
+#
+#
+#
+$infile = "$ARGV[0]"; # a binary haplotype file produced by ms 
+$prob = "$ARGV[1]"; # this is the probability that a value should be deleted.
+$k = "$ARGV[2]"; # this is the maximum number of states allowed in a character.
+
+open (IN, "$infile");
+open OUTB, ">$ARGV[0]b";  # where the cleaned up binary data is output 
+open (OUTD, '>datalist');  # list of the files for the extracted data, with possible deletions
+open (OOUTD, '>odatalist');  # list of the files for the original extracted data, with no deletions
+#open (STAT, '>>stats');
+$s = 0;
+
+$line = <IN>;
+if ($line =~ /ms (\d+) (\d+) -s (\d+) -r (\d+)/){  #extract the header information
+                                                     #from data in the header line in the ms file
+#   print STAT "$line";
+   $numi = $1;
+   $numreps = $2;
+   $numj = $3;
+   $recrate = $4;
+   $kthnumj = $numj/($k-1);
+   print "$numi, $kthnumj\n";
+#   print STAT "$numreps reps of $numi by $kthnumj with r = $recrate:\n";
+}
+
+@lines = <IN>;
+close(IN);
+
+$i = 0;
+foreach $line (@lines) {
+  if ($line =~ /segsites/) {
+     close OUT;
+     $s++;
+     print OUTB "\n set $s\n";
+
+     $outprefix = "$k" . 'state';
+     $outfile = $outprefix . $numi . '.' . $kthnumj . '.' . $s;
+     $originaloutfile = "o$outfile";  # this will hold the state data for one problem instance
+     print OUTD "$outfile\n";
+     print OOUTD "$originaloutfile\n";
+     open (OUT, ">$outfile");
+     open (OOUT, ">$originaloutfile");
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests.py

@@ -1,0 +1,124 @@
+import sys
+import os
+import subprocess
+
+EXEC = "./perfectphy"
+
+class Result:
+    def __init__(self):
+        self.count = 0
+        self.ms = 0
+    
+    def add(self, usert):
+        self.count += 1
+        
+        # usert = '14.460'
+        part = usert.partition('.')
+        self.ms += 1000 * int(part[0])
+        self.ms += int(part[2])
+    
+    def avg(self):
+        if self.count > 0:
+            return int(self.ms / self.count)
+        else:
+            return -1
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing KW\nk,");
+    
+    outf.write(','.join(str(k) for k in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            if entry.count == TRIALS:
+                outf.write(",%d" % entry.avg() )
+            else:
+                outf.write(",-1")
+    
+    outf.write("\n\n")
+
+
+def runTests():
+    TRIALS = int(sys.argv[1])
+    PARALLEL = int(sys.argv[2])
+    
+    if TRIALS < PARALLEL or TRIALS % PARALLEL != 0:
+        print PARALLEL, "does not divide", TRIALS, "evenly"
+        return
+    
+    for f in (EXEC, "ms", "multextract.pl"):
+        if not os.path.isfile(f):
+            print 'Cannot find file "%s" in current directory' % f
+            return
+    
+    nmRange = [ (x,x) for x in [10, 50, 100, 500, 1000, 2000] ]
+    kRange = [4, 10, 20]
+    
+    # create matrix of results for every (N,M,K) triplet
+    resultmatrix = dict( [ ((n,m,k), Result()) for n,m in nmRange for k in kRange ] )
+    
+    try:
+        for k in kRange:
+            for n,m in nmRange:
+                params = (n,m,k)
+                
+                # generate and interpret datasets
+                gendata = subprocess.Popen("./ms %d %d -s %d -r 0.0 5000 > mybigdata ; perl multextract.pl mybigdata 0 %d > /dev/null ; rm -r o%dstate* > /dev/null" % (n, TRIALS, m * (k-1), k, k), shell = True )
+                
+                print "(%d,%d,%d) Generating..." % params,
+                sys.stdout.flush()
+                
+                entry = resultmatrix[ params ]
+                
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests_dagsize.py

@@ -1,0 +1,112 @@
+import sys
+import os
+import subprocess
+import math
+
+EXEC = "./perfectphy"
+
+class Result:
+    def __init__(self):
+        self.data = []
+        
+    def add(self, size):
+        self.data.append(int(size.split()[0]))
+    
+    def avg(self):
+        if len(self.data) == 0:
+            return -1.0
+        else:
+            return sum(self.data)/float(len(self.data))
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing DAG output file size\nk,");
+    outf.write(','.join(str(k) for k in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            outf.write( ", " + str(entry.data) )
+    outf.write("\n")
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            outf.write( ", " + str(entry.avg()) )
+    outf.write("\n")
+
+def runTests():
+    TRIALS = int(sys.argv[1])
+    PARALLEL = int(sys.argv[2])
+    
+    if TRIALS < PARALLEL or TRIALS % PARALLEL != 0:
+        print PARALLEL, "does not divide", TRIALS, "evenly"
+        return
+    
+    for f in (EXEC, "ms", "multextract.pl"):
+        if not os.path.isfile(f):
+            print 'Cannot find file "%s" in current directory' % f
+            return
+    
+    nmRange = [ (x,x) for x in [10, 50, 100, 500]]#, 1000, 2000] ]
+    kRange = [4, 10, 20]
+    
+    # create matrix of results for every (N,M,K) triplet
+    resultmatrix = dict( [ ((n,m,k), Result()) for n,m in nmRange for k in kRange ] )
+    
+    try:
+        for k in kRange:
+            for n,m in nmRange:
+                params = (n,m,k)
+                
+                # generate and interpret datasets
+                gendata = subprocess.Popen("./ms %d %d -s %d -r 0.0 5000 > mybigdata ; perl multextract.pl mybigdata 0 %d > /dev/null ; rm -r o%dstate* > /dev/null" % (n, TRIALS, m * (k-1), k, k), shell = True )
+                
+                print "(%d,%d,%d) Generating..." % params,
+                sys.stdout.flush()
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests_md.py

@@ -1,0 +1,124 @@
+import sys
+import os
+import subprocess
+
+EXEC = "./perfectphy"
+
+class Result:
+    def __init__(self):
+        self.count = 0
+        self.ms = 0
+    
+    def add(self, usert):
+        self.count += 1
+        
+        # usert = '14.460'
+        part = usert.partition('.')
+        self.ms += 1000 * int(part[0])
+        self.ms += int(part[2])
+    
+    def avg(self):
+        if self.count > 0:
+            return int(self.ms / self.count)
+        else:
+            return -1
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing Missing Data\nk,");
+    
+    outf.write(','.join(str(k) + ' ' + md for k,md in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        
+        for k,md in kRange:
+            entry = resultmatrix[ (n,m,k,md) ]
+            if entry.count == TRIALS:
+                outf.write(",%d" % entry.avg() )
+            else:
+                outf.write(",-1")
+    
+    outf.write("\n\n")
+
+
+def runTests():
+    TRIALS = int(sys.argv[1])
+    PARALLEL = int(sys.argv[2])
+    
+    if TRIALS < PARALLEL or TRIALS % PARALLEL != 0:
+        print PARALLEL, "does not divide", TRIALS, "evenly"
+        return
+    
+    for f in ("perfectphy", "ms", "multextract.pl", "missingdata"):
+        if not os.path.isfile(f):
+            print 'Cannot find file "%s" in current directory' % f
+            return
+    
+    nmRange = [ (x,1000) for x in xrange(10,101,10) ]
+    kRange = [ (4,x) for x in ["0.0", "0.1", "0.2"] ]
+    
+    # create matrix of results for every (N,M,K) triplet
+    resultmatrix = dict( [ ((n,m,k,md), Result()) for n,m in nmRange for k,md in kRange ] )
+    
+    try:
+        for k, md in kRange:
+            for n,m in nmRange:
+                params = (n,m,k,md)
+                
+                # generate and interpret datasets
+                gendata = subprocess.Popen("./ms %d %d -s %d -r 0.0 5000 > mybigdata ; perl multextract.pl mybigdata 0 %d > /dev/null ; rm -r o%dstate* > /dev/null" % (n, TRIALS, m * (k-1), k, k), shell = True )
+                
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests_mintree.py

@@ -1,0 +1,115 @@
+import sys
+import os
+import subprocess
+import math
+
+EXEC = "./perfectphy" + "_mintree"
+
+class Result:
+    def __init__(self):
+        self.numSmallest = 0
+        self.numTrials = 0
+        
+    def add(self, sizes):
+        sizes = map(int,sizes.split())
+        
+        if sizes[0] == sizes[1]:
+            self.numSmallest = self.numSmallest + 1
+        
+        self.numTrials = self.numTrials + 1
+    
+    def val(self):
+        return "%d/%d" % (self.numSmallest, self.numTrials)
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing KW Tree Size vs Smallest Minimal Tree Size\nk,");
+    outf.write(','.join(str(k) for k in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            outf.write( ", " + entry.val() )
+    outf.write("\n")
+
+def runTests():
+    TRIALS = int(sys.argv[1])
+    PARALLEL = int(sys.argv[2])
+    
+    if TRIALS < PARALLEL or TRIALS % PARALLEL != 0:
+        print PARALLEL, "does not divide", TRIALS, "evenly"
+        return
+    
+    for f in (EXEC, "ms", "multextract.pl"):
+        if not os.path.isfile(f):
+            print 'Cannot find file "%s" in current directory' % f
+            return
+    
+    nmRange = [ (x,x) for x in [10, 50, 100, 500, 1000, 2000] ]
+    kRange = [4, 10, 20]
+    
+    # create matrix of results for every (N,M,K) triplet
+    resultmatrix = dict( [ ((n,m,k), Result()) for n,m in nmRange for k in kRange ] )
+    
+    try:
+        for k in kRange:
+            for n,m in nmRange:
+                params = (n,m,k)
+                
+                # generate and interpret datasets
+                gendata = subprocess.Popen("./ms %d %d -s %d -r 0.0 5000 > mybigdata ; perl multextract.pl mybigdata 0 %d > /dev/null ; rm -r o%dstate* > /dev/null" % (n, TRIALS, m * (k-1), k, k), shell = True )
+                
+                print "(%d,%d,%d) Generating..." % params,
+                sys.stdout.flush()
+                
+                entry = resultmatrix[ params ]
+                
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests_rc.py

@@ -1,0 +1,138 @@
+import sys
+import os
+import subprocess
+import time
+
+class Result:
+    def __init__(self):
+        self.count = 0
+        self.ms = 0
+        self.chars = 0
+    
+    def add(self, usert, c):
+        self.count += 1
+        self.chars += int(c)
+        
+        # usert = '14.460'
+        part = usert.partition('.')
+        self.ms += 1000 * int(part[0])
+        self.ms += int(part[2])
+    
+    def avg(self):
+        if self.count > 0:
+            return int(self.ms / self.count), int(self.chars / self.count)
+        else:
+            return -1, -1
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing Character Removal\nk,");
+    
+    outf.write(','.join(str(k) + ' ' + md for k,md in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        
+        for k,md in kRange:
+            entry = resultmatrix[ (n,m,k,md) ]
+            if entry.count == TRIALS:
+                outf.write(",(%d ms, %d ch)" % entry.avg() )
+            else:
+                outf.write(",-")
+    
+    outf.write("\n\n")
+
+
+def runTests():
+    TRIALS = int(sys.argv[1])
+    PARALLEL = int(sys.argv[2])
+    
+    if TRIALS < PARALLEL or TRIALS % PARALLEL != 0:
+        print PARALLEL, "does not divide", TRIALS, "evenly"
+        return
+    
+    for f in (EXEC, "ms", "multextract.pl"):
+        if not os.path.isfile(f):
+            print 'Cannot find file "%s" in current directory' % f
+            return
+    
+    nmRange = [ (50,50) ]#m) for m in xrange(10,41,10) ]
+    kRange = [ (4,r) for r in ["0.0"] ] #["0.01", "0.1", "0.25"] ]
+    
+    # create matrix of results for every (N,M,K) triplet
+    resultmatrix = dict( [ ((n,m,k,r), Result()) for n,m in nmRange for k,r in kRange ] )
+    
+    try:
+        for k, r in kRange:
+            for n,m in nmRange:
+                params = (n,m,k,r)
+                entry = resultmatrix[ params ]
+                
+                #print entry.count
+                #continue
+                
+                while entry.count < TRIALS:
+                    # generate and interpret datasets
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/experiments/run_tests_treecount.py

@@ -1,0 +1,156 @@
+import sys
+import os
+import subprocess
+import math
+
+EXEC = "./perfectphy"
+
+def avg(l):
+    if len(l) == 0:
+        return -1.0
+    return sum(l) / float(len(l))
+
+def sd(l):
+    if len(l) == 0:
+        return -1.0
+    if len(l) == 1:
+        return 0.0
+    
+    a = avg(l)
+    return math.sqrt( sum( (t - a)**2 for t in l ) / len(l) )
+
+def median(l):
+    if len(l) == 0:
+        return -1.0
+    if len(l) == 1:
+        return l[0]
+    
+    l = sorted(l)
+    centerIndex = (len(l) - 1)/2
+    
+    if len(l) % 2 == 0:
+        return (l[centerIndex] + l[centerIndex + 1]) / 2.0
+    else:
+        return float(l[centerIndex])
+
+class Result:
+    def __init__(self):
+        self.trees = []
+        self.times = []
+        
+    def add(self, tcount, usert):
+        self.trees.append( int(tcount) )
+        # usert = '14.460'
+        part = usert.partition('.')
+        ms = 1000 * int(part[0]) + int(part[2])
+        self.times.append(ms)
+    
+    def avg(self):
+        return "(%s, %s ms)" % (avg(self.trees), avg(self.times))
+
+    def sd(self):
+        return "(%s, %s ms)" % (sd(self.trees), sd(self.times))
+    
+    def median(self):
+        return "(%s, %s ms)" % (median(self.trees), median(self.times))
+
+def printResults(outf, resultmatrix, nmRange, kRange, TRIALS):
+    outf.write("Testing KW Tree Count\nk,");
+    outf.write(','.join(str(k) for k in kRange))
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            outf.write( ", " + str(zip(entry.trees, entry.times)) )
+    outf.write("\n")
+    
+    for n,m in nmRange:
+        outf.write( "\n%d %d" % (n,m) )
+        for k in kRange:
+            entry = resultmatrix[ (n,m,k) ]
+            outf.write(", (%s avg, %s sd, %s med)" % (entry.avg(), entry.sd(), entry.median()) )
+    outf.write("\n\n")
+
+
+def runTests():
+    TRIALS = int(sys.argv[1])
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/maxindset.cpp

@@ -1,0 +1,335 @@
+//////// Maximal Independent Set Algorithm by Tsukiyama et al. ////////
+
+void BACKTRACK(vector<vector<int> > &graph, vector<int> &mis, int i, int n, int &count, vector<vector<int> > &output) {
+    int c, x = i+1;
+    bool f;
+    vector<int> bucket;
+    vector<int>::iterator y, z;
+    
+    if (x < n) { // if current scan is not at the end
+        
+        // count the number of neighbors of x which are in mis
+        c = 0;
+        for (y = graph[x].begin(); y != graph[x].end(); ++y) {
+            if (*y < x && mis[*y] == 0) {
+                c++;
+            }
+        }
+        
+        if (c == 0) {
+            // if no neighbors in mis, add it
+            for (y = graph[x].begin(); y != graph[x].end(); ++y) {
+                if (*y < x) {
+                    mis[*y]++;
+                }
+            }
+            
+            // recurse on vertices larger than x
+            BACKTRACK(graph, mis, x, n, count, output);
+            
+            // then remove x
+            for (y = graph[x].begin(); y != graph[x].end(); ++y) {
+                if (*y < x) {
+                    mis[*y]--;
+                }
+            }
+        }
+        else {
+            // x has neighbors, so don't include it
+            mis[x] = c;
+            
+            // recurse on vertices larger than x
+            BACKTRACK(graph, mis, x, n, count, output);
+            
+            mis[x] = 0;
+            
+            f = true;
+            bucket.reserve(graph[x].size());
+            
+            // now attempt to insert x
+            for (y = graph[x].begin(); y != graph[x].end(); ++y) {
+                if (*y < x) {
+                    if (mis[*y] == 0) {
+                        // for each neighbor of x in mis less than x
+                        // remove their neighbors from mis
+                        
+                        bucket.push_back(*y);
+                        for (z = graph[*y].begin(); z != graph[*y].end(); ++z) {
+                            if (*z < x) {
+                                mis[*z]--;
+                                if (mis[*z] == 0) {
+                                    f = false;
+                                }
+                            }
+                        }
+                    }
+                    mis[*y]++;
+                }
+            }
+            
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/ppDAG.cpp

@@ -1,0 +1,1749 @@
+#include "ppDAG.h"
+
+// allows readable 2D indexing with pairs
+#define pair_index(array,p) ( (array)[ (p).first ][ (p).second ] )
+
+//////// ProductNode ////////
+
+PPDAG::ProductNode::ProductNode():rootLabel(-1), count(-1) {}
+PPDAG::ProductNode::ProductNode(const vector<pair<int,int> > &d):rootLabel(-1), count(-1), decomp(d) {}
+PPDAG::ProductNode::ProductNode(const ProductNode &pn):rootLabel(pn.rootLabel), count(pn.count), decomp(pn.decomp) {}
+
+// Pn.count = product of Sn.count below it
+int PPDAG::ProductNode::getCount(PPDAG &dag) {
+    if (count == -1) {
+        count = 1;
+        
+        for (unsigned int i = 0; i < decomp.size(); i++)
+            count *= pair_index(dag.sumNodes, decomp[i]).getCount(dag);
+    }
+    
+    return count;
+}
+
+// returns whether the label of the node is a taxon from the original dataset
+bool PPDAG::ProductNode::hasTaxaRoot(int n) const {
+    return (rootLabel < n);
+}
+
+// get values of i for each sum node in decomp (linear time)
+// example: i =           i3
+//                 + i2 * c3
+//            + i1 * c2 * c3
+//       + i0 * c1 * c2 * c3
+void PPDAG::ProductNode::getIValues(const PPDAG &dag, vector<int> &iValues, int i) const {
+    int product = count;
+    
+    // find i values to give to each sum node in decomp
+    for (unsigned int j = 0; j < decomp.size(); j++) {
+        product /= dag.getSumNode(decomp[j]).count;
+        iValues[j] = i / product;
+        i -= iValues[j] * product;
+    }
+}
+
+// inverse operation of getIValues()
+int PPDAG::ProductNode::collapseIValues(const PPDAG &dag, const vector<int> &iValues) const {
+    int i = 0, product = count;
+    
+    for (unsigned int j = 0; j < decomp.size(); j++) {
+        product /= dag.getSumNode(decomp[j]).count;
+        i += iValues[j] * product;
+    }
+    
+    return i;
+}
+
+
+bool PPDAG::ProductNode::operator<(const ProductNode& rhs) const {
+    return (decomp < rhs.decomp);
+}
+
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/ppDAG.h

@@ -1,0 +1,238 @@
+#ifndef __PPDAG_H__
+#define __PPDAG_H__
+
+class PPDAG {
+public:
+    // forward declarations
+    class ProductNode;
+    class SumNode;
+    class tree_iterator;
+    
+    // maxPC is the pseudo proper cluster which represents all taxa
+    // rootTaxaPC is the taxon which is at the "top" of the DAG
+    // - all Product Nodes under maxPC are rooted by it
+    
+    int maxPC, rootTaxaPC, n;
+    vector<vector<SumNode> > sumNodes;  // for each proper cluster H, for each valid parent y of H
+    vector<vector<int> > nodeLabels;    // stores all possible node labels
+    
+    tree_iterator tree_begin();
+    tree_iterator tree_end();
+    tree_iterator tree_rbegin();
+    tree_iterator tree_rend();
+    
+    // constructors
+    PPDAG(const vector<vector<int> > &propercluster, const Dict &pcdict);
+    PPDAG(const char* filename);
+protected:
+    bool makeFromFile(ifstream &inf);
+public:
+    
+    // construction functions (for sprop4.cpp)
+    void resizeRow(int H, int s);
+    void insertProductNode(ProductNode &pn, int H, int y);
+    void insertManyProductNodes(vector<ProductNode> &mis, int H, int y);
+    bool handleInterior(SumNode *sn, vector<int> &interiorTaxaVp);
+    void finalize(const vector< vector<vector<int> > > &Conn, const vector< vector<vector<int> > > &Vp, const Dict &pcdict, const vector<vector<int> > &propercluster, const vector<vector<int> > &data);
+    
+    // printing functions
+    void makeDOT() const;
+    void makeDOTForTrees() const;
+    void writeProductNode(ofstream &outf, const ProductNode &pn, int H, int y, bool notseen) const;
+    void prettyprint(ostream &outf) const;
+    
+    // accessor functions
+    int getCount();
+    int getCount() const;
+    int getVpSize(int H) const;
+    const SumNode& getSumNode(int H, int y) const;
+    const SumNode& getSumNode(const pair<int,int> Hy) const;
+    const vector<int>& getNodeLabel(int label) const;
+    vector<int> getNodeLabel(int label);
+    
+    bool isSingleton(int H) const;
+    
+    // dynamic programming iteration functions
+    void bottomDP(int &H, int &y) const;
+    void topDP(int &H, int &y) const;
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/qtable.cpp

@@ -1,0 +1,173 @@
+#include"qtable.h"
+
+QTable::QTable() {}
+
+void QTable::construct(const vector<vector<int> > &propercluster) {
+    // propercluster must be sorted lexicographically, NOT also by cluster size
+    int properclustercount = propercluster.size(), n = propercluster[0].size();
+    vector<int> lastone(n);
+    
+    table.resize(properclustercount);
+    indexmap.assign(properclustercount, -1);
+    
+    for(int pc = properclustercount - 1; pc >= 0; pc--) {
+        for(int t = 0; t < n; t++)
+            if (propercluster[pc][t] == 1)
+                lastone[t] = pc;
+        
+        table[pc] = lastone;
+    }
+}
+
+void QTable::setIndexMap(int i, int pc) {
+    indexmap[i] = pc;
+}
+
+int QTable::lookup(const vector<int> &pcluster) const {
+    int pc = 0;
+    unsigned int t, n = pcluster.size();
+    
+    // jump through Q table to find sorted position of pcluster
+    for(t = 0; t < n; t++)
+        if (pcluster[t] == 1)
+            pc = table[pc][t];
+    
+    // ensure that pcluster is actually present by checking that
+    // pcluster[t] == 1 iff Q[pc][t] == pc
+    for(t = 0; t < n; t++)
+        if ((pcluster[t] == 1) != (table[pc][t] == pc))
+            return -1;
+    
+    // return true index
+    return indexmap[pc];
+}
+
+int QTable::lookupComplement(const vector<int> &pcluster, const vector<int> &superpcluster) const {
+    int pc = 0;
+    unsigned int t, n = pcluster.size();
+    
+    #define compHas(i) (pcluster[i] == 0 && superpcluster[i] == 1)
+    
+    // jump through Q table to find sorted position of comp
+    for(t = 0; t < n; t++)
+        if (compHas(t))
+            pc = table[pc][t];
+    
+    // ensure that comp is actually present by checking that
+    // comp[t] == 1 iff Q[pc][t] == pc
+    for(t = 0; t < n; t++)
+        if (compHas(t) != (table[pc][t] == pc))
+            return -1;
+    
+    // return true index
+    return indexmap[pc];
+    
+    #undef compHas
+}
+
+int QTable::lookupComplement(const vector<int> &pcluster) const {
+    int pc = 0;
+    unsigned int t, n = pcluster.size();
+    
+    #define compHas(i) (pcluster[i] == 0)
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/qtable.h

@@ -1,0 +1,44 @@
+#ifndef __QTABLE_H__
+#define __QTABLE_H__
+
+// QTable wraps a L*n table with the following definition:
+// Q[p][i] = the smallest x >= p such that T[x][i] == 1
+
+/*
+??? bucket sort output lists of pc's ???
+*/
+
+class QTable {
+private:
+    vector<vector<int> > table; // Q
+    vector<int> indexmap;       // index mapping from lexicographic order to cluster size then lexicographic order
+    
+public:
+    QTable();
+    
+    // constructs table from lexicographically-ordered proper cluster list
+    void construct(const vector<vector<int> > &propercluster);
+    
+    // externally initialize index mapping
+    // used to handle the fact that propercluster is resorted after construction
+    void setIndexMap(int i, int pc);
+    
+    // find proper cluster index, O(n)
+    int lookup(const vector<int> &pcluster) const;
+    
+    // finds complement of pcluster in superpcluster if it exists, O(n)
+    int lookupComplement(const vector<int> &pcluster, const vector<int> &superpcluster) const;
+    
+    // finds full complement of pcluster if it exists in O(n)
+    int lookupComplement(const vector<int> &pcluster) const;
+    
+    // find all proper cluster indeces in decomp, O(n)
+    // returns whether a non-proper cluster is found
+    bool lookupParallel(const vector<int> &classes, vector<int> &decomp, int n, int maxC, const vector<int> &clustersize) const;
+    
+    // find all proper cluster indeces in decomp, O(n)
+    // returns whether a non-good cluster is found
+    bool lookupParallel(const vector<int> &classes, const vector<int> &good, vector<int> &decomp, int G, int G1, int n, int maxC, const vector<vector<int> > &propercluster) const;
+};
+
+#endif

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/sprop4.cpp

@@ -1,0 +1,3306 @@
+/*
+    Michael Coulombe 2012
+    
+    sprop4 changes:
+        - Extended to C++ (fixed memory bugs)
+        - Implemented full Kannan-Warnow improvements to DP algorithm to determine/construct perfect phylogenies
+            > Extended the above to determine uniqueness (minimal)
+        - Implemented KW algorithm for enumeration of minimal perfect phylogenies in a DAG
+            > Created algorithms, a tree iterator, and simplified Dynamic Program method for the DAG
+*/
+
+
+/*
+ * Written by Dan Gusfield copyright 2008
+ * August 11, 2008 Permission is granted to use but not distribute this code or extensions or modifications of it.
+ *
+ *
+ * Aug 2 increased the defined value of S, which is an upper bound on the number of proper clusters found.
+ * July 30 create tracebackadj that produces an adjacency list of the PP created, where there is one.
+ * July 28 create statetraceback to include the state labels for the nodes of the PP.
+ * July 27 create taxatraceback to print out the proper cluster at each node in the PP.
+ * July 26 begin programming the traceback to actually describe a PP if there is one. This  only shows
+ *  the index of each proper cluster.
+ * July 17, changed sdt[K] to sdt[M] in findpropclusters, which seems to have fixed a bug where some of the
+ * data entries were changed (on homer but not imap) at each iteration in findpropclusters.
+ * July 15, 2008 fixed a bug in findpropclusters - k had been used where statecount[j] was needed
+ * DG June 7, not done - need to find the proper nesting of the tests of the connected components of the
+ * complement of G' in G, and test each one to see if it is good
+
+ * June 4, 2008 this modifies sprop1.c in the unfinished section for case 2 - building and searching the
+relation when G - G' is not a proper cluster. In the unfinished section of sprop1.c I tried to be thoughtful
+about making that code efficient in practice. But it became too messy. Here I do the most straightforward thing
+of building the relation and recording it in a relation matrix, and then I take transitive closer of the
+relation matrix. Later I can implement algorithm, not programming, ideas to make this part more efficient *
+
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/src/sprop4.h

@@ -1,0 +1,148 @@
+#include<stdio.h>
+#include<vector>
+#include<set>
+#include<map>
+#include<fstream>
+#include<iostream>
+#include<string>
+#include<sstream>
+#include<algorithm>
+#include<numeric>
+#include<limits>
+using namespace std;
+
+// prints information about progress and computed data
+#ifndef DEBUGTRACE
+    #define DEBUGTRACE gbl::PRINTDATA
+#endif
+
+#ifndef RUNTESTMINTREE
+    #define RUNTESTMINTREE false
+#endif
+
+#ifndef MISEXT
+    #define MISEXT true
+#endif
+
+#ifndef MISLIMITK
+    #define MISLIMITK true
+#endif
+
+// global configuration variable container
+class gbl {
+public:
+    static bool  PRINTDATA,       // prints important computed data
+                 CHECKUNIQUE,     // checks whether there is a unique minimal perfect phylogeny
+                 ENUMERATE,       // runs KW enumeration algorithm
+                 DAGSTATISTICS,   // runs DAG algorithms for statistics
+                 ALLOWINTERNAL,   // lets taxon in the dataset be internal nodes in minimal trees
+                 NEWICK;          // Prints the initial phylogeny in the Newick tree format
+    
+    static int OUTGROUPTAXON;     // In the initial phylogeny, this taxon is at the top
+    
+    static const char *DAGPRETTYF,// output file where the dag will be prettyprinted, or input if doing statistics on existing DAG
+                      *DAGDOTOUTF;// output dot file of the dag
+                    //*DPDOTOUTF; // output dot file for the tree generated by perf
+};
+
+// old debug flags
+#define VERBOSE 0
+#define VERBOSE1 0
+#define VERBOSEDATA 0
+#define PRINTREAD 0
+#define CLUSTER 0
+#define DEBUG 0
+#define DEBUG1 0
+#define DEBUG2 0
+
+#define DEBUGOLD
+
+enum PrettyPrintFormat {
+    PPAF_NONE,  // normal printing
+    PPAF_DUMMY, // all negative numbers are printed as "*"
+    PPAF_PC,    // zeros are printed as "-", ones printed as their index
+};
+
+void prettyprintarray(const vector<int> &array, const char name[], PrettyPrintFormat format = PPAF_NONE);
+void prettyprintarray(const vector<vector<int> > &array, const char name[], PrettyPrintFormat format = PPAF_NONE);
+
+// utility functions
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/tools/charremoval.py

@@ -1,0 +1,50 @@
+import sys
+import subprocess
+import itertools
+
+EXEC = "./perfectphy"
+
+def charremoval(inf):
+    s = inf.read().split()
+    
+    INDEX, N, M = s[0], s[1], s[2]
+    n = int(N)
+    m = int(M)
+    matrix = [ s[3 + i*m : 3 + i*m + m] for i in xrange(n) ]
+    
+    # loop over number of characters to remove
+    for rem in xrange(0,m):
+        count = m - rem
+        
+        # check every subset
+        for p in itertools.combinations(range(m), count): #subsets(m, count):
+            kw = subprocess.Popen([EXEC], stdin = subprocess.PIPE, stdout = subprocess.PIPE, stderr=subprocess.PIPE)
+            pipe = kw.stdin
+            w = pipe.write
+            
+            w("%s\n%s %d" % (INDEX, N, count) )
+            
+            for row in matrix:
+                w('\n')
+                for i in p:
+                    w( row[i] )
+                    w(' ')
+            
+            output, err = kw.communicate()
+            
+            if output[0] == "1": # return column indices with a pp
+                return p
+    
+    return [] # return no columns
+
+if __name__ == '__main__':
+    try:
+        result = charremoval(sys.stdin)
+        
+        if len(result) > 0:
+            print "1 %d characters: %s" % (len(result), ' '.join(str(i) for i in result))
+        else:
+            print "0 No Perfect Phylogeny for any subset"
+        
+    except Exception as e:
+        print type(e), e

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/tools/convert.cpp

@@ -1,0 +1,239 @@
+#include<vector>
+#include<fstream>
+#include<iostream>
+#include<cctype>
+using namespace std;
+
+enum ArgError { ERROR = -1, QUIT = 0, NONE = 1 };
+
+ArgError handleArguments(int argc, char** argv, char &dummychar, string &index, string &mapfile ) {
+    const char* helpmessage = 
+"Usage: ./convert.out [option] \n"
+"\n"
+"Options:\n"
+"-h, --help       Displays this message then quits.\n"
+"-dummy CHAR      Set CHAR as a dummy state (outputs -1).\n"
+"-index I         Set index of dataset as I (default is 1).\n"
+"-mapfile FILE    Writes the state mapping to FILE\n";
+    
+    string arg;
+    for (int i = 1; i < argc; i++) {
+        if (argv[i][0] != '-') {
+            cerr << "Bad argument: " << argv[i] << " is not a flag" << endl;
+            return ERROR;
+        }
+        
+        arg = argv[i];
+        
+        if (arg.compare("-dummy") == 0) {
+            if (argc > i + 1) {
+                dummychar = argv[i+1][0];
+                
+                if (!isgraph(dummychar)) {
+                    cerr << "Bad argument: dummy character must be graphical" << endl;
+                    return ERROR;
+                }
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede a character" << endl;
+                
+                return ERROR;
+            }
+        }
+        else if (arg.compare("-index") == 0) {
+            if (argc > i + 1) {
+                index = argv[i+1];
+                
+                for (unsigned int j = 0; j < index.size(); j++)
+                    if ( !( isdigit(index[j]) || (j == 0 && index[j] == '-' && index.size() > 1) ) ) {
+                        cerr << "Bad argument: index must be an integer, not " << index << endl;
+                        return ERROR;
+                    }
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede an integer" << endl;
+                
+                return ERROR;
+            }
+        }
+        else if (arg.compare("-mapfile") == 0) {
+            if (argc > i + 1) {
+                mapfile = argv[i+1];
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede a filename" << endl;
+                
+                return ERROR;
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/tools/deconvert.cpp

@@ -1,0 +1,157 @@
+#include<vector>
+#include<fstream>
+#include<iostream>
+#include<cctype>
+using namespace std;
+
+enum ArgError { ERROR = -1, QUIT = 0, NONE = 1 };
+
+ArgError handleArguments(int argc, char** argv, string &mapfile ) {
+    const char* helpmessage = 
+"Usage: ./deconvert.out [option] \n"
+"\n"
+"Options:\n"
+"-h, --help       Displays this message then quits.\n"
+"-mapfile FILE    Read the state mapping from FILE (REQUIRED).\n";
+    
+    string arg;
+    for (int i = 1; i < argc; i++) {
+        if (argv[i][0] != '-') {
+            cerr << "Bad argument: " << argv[i] << " is not a flag." << endl;
+            return ERROR;
+        }
+        
+        arg = argv[i];
+        
+        if (arg.compare("-mapfile") == 0) {
+            if (argc > i + 1) {
+                mapfile = argv[i+1];
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede a filename." << endl;
+                
+                return ERROR;
+            }
+        }
+        else if (arg.compare("-h") == 0 || arg.compare("--help") == 0) {
+            cout << helpmessage;
+            
+            return QUIT;
+        }
+        else {
+            cerr << "Bad argument: " << arg << " is not a valid flag." << endl;
+            
+            return ERROR;
+        }
+    }
+    
+    if (mapfile.empty()) {
+        cerr << "Bad argument: " << "no mapfile given." << endl;
+        
+        return ERROR;
+    }
+    
+    return NONE;
+}
+
+int main(int argc, char** argv) {
+    char dummychar = '\0';
+    string mapfile;
+    
+    #define NAMEWIDTH 10
+    
+    // improve io speed since not using C streams
+    ios_base::sync_with_stdio(false);
+    
+    ArgError argerr = handleArguments(argc, argv, mapfile);
+    if (argerr != NONE)
+        return argerr;
+    
+    ifstream inf(mapfile.c_str());
+    
+    // read number of characters
+    int index, m;
+    inf >> index >> m;
+    
+    // check for a dummy character
+    inf >> noskipws;
+    dummychar = inf.peek();
+    inf >> skipws;
+    
+    if (dummychar != '\n') {
+        inf >> dummychar; // consume actual dummy char
+        
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/tools/missingdata.cpp

@@ -1,0 +1,185 @@
+#include<vector>
+#include<iostream>
+#include<sstream>
+#include<cstdlib>
+#include<ctime>
+using namespace std;
+
+#define ADDMISSING 1
+
+enum ArgError { ERROR = -1, QUIT = 0, NONE = 1 };
+
+ArgError handleArguments(int argc, char** argv, bool &fillmissing, double& prob) {
+    const char* helpmessage =
+"Usage: ./missingdata.out [option] \n"
+"\n"
+"Options:\n"
+"-h, --help       Displays this message then quits.\n"
+#if ADDMISSING
+"-add PERCENT     Randomly add missing data with PERCENT probability per cell\n"
+"                 then fills them in with new states\n"
+#endif
+"-fill            Replace instances of -1 with new states\n";
+    
+    if (argc == 1) {
+        cerr << "This program cannot be executed with no arguments." << endl;
+        return ERROR;
+    }
+    
+    string arg;
+    for (int i = 1; i < argc; i++) {
+        if (argv[i][0] != '-') {
+            cerr << "Bad argument: " << argv[i] << " is not a flag" << endl;
+            return ERROR;
+        }
+        
+        arg = argv[i];
+        
+        if (arg.compare("-fill") == 0) {
+            fillmissing = true;
+        }
+        #if ADDMISSING
+        else if (arg.compare("-add") == 0) {
+            if (argc > i + 1) {
+                stringstream ss(argv[i+1]);
+                ss >> prob;
+                
+                if ( !ss || prob < 0.0 || prob > 1.0) {
+                    cerr << "Bad argument: " << argv[i+1] << " must be a valid double in range [0.0, 1.0]" << endl;
+                    return ERROR;
+                }
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede a double" << endl;
+                return ERROR;
+            }
+        }
+        #endif
+        else if (arg.compare("-h") == 0 || arg.compare("--help") == 0) {
+            cout << helpmessage;
+            
+            return QUIT;
+        }
+        else {
+            cerr << "Bad argument: " << arg << " is not a valid flag" << endl;
+            
+            return ERROR;
+        }
+    }
+    
+    return NONE;
+}
+
+#define readinto(var) do {\
+        cin >> var;\
+        if (cin.bad()) {\
+            cerr << "Input error: " << #var << " could not be read" << endl;\
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/PerfectPhy/tools/newicktodot.cpp

@@ -1,0 +1,196 @@
+#include<vector>
+#include<iostream>
+#include<cstdlib>
+using namespace std;
+
+
+enum ArgError { ERROR = -1, QUIT = 0, NONE = 1 };
+
+ArgError handleArguments(int argc, char** argv, string &name) {
+    const char* helpmessage = "\
+Usage: ./newicktodot.out [option] \n\
+\n\
+Options:\n\
+-h, --help       Displays this message then quits.\n\
+-name NAME       Gives the graph the name NAME (default is \"PerfectPhylogeny\")\n";
+    
+    string arg;
+    for (int i = 1; i < argc; i++) {
+        if (argv[i][0] != '-') {
+            cerr << "Bad argument: " << argv[i] << " is not a flag" << endl;
+            return ERROR;
+        }
+        
+        arg = argv[i];
+        
+        if (arg.compare("-name") == 0) {
+            if (argc > i + 1) {
+                name = argv[i+1];
+                
+                i++;
+            }
+            else {
+                cerr << "Bad argument: " << arg << " must precede a string" << endl;
+                
+                return ERROR;
+            }
+        }
+        else if (arg.compare("-h") == 0 || arg.compare("--help") == 0) {
+            cout << helpmessage;
+            
+            return QUIT;
+        }
+        else {
+            cerr << "Bad argument: " << arg << " is not a valid flag" << endl;
+            
+            return ERROR;
+        }
+    }
+    
+    return NONE;
+}
+
+void writenode(int depth);
+void writesubtree(int depth);
+void writechildren(int depth, vector<int> &children);
+
+int nodeid;
+vector<int> support;
+
+void setSupport(int i, int s) {
+    if (i >= (int) support.size())
+        support.resize(i+1);
+    
+    support[i] = s;
+}
+
+void writeEdges(int parent, vector<int>& children) {
+    for(unsigned int i = 0; i < children.size(); i++) {
+        cout << "    " << nodeid << " -- " << children[i];
+        
+        if (! support.empty())
+            cout << " [ label = \"" << support[ children[i] ] << "\" ];" << endl;
+        else
+            cout << ';' << endl;
+    }
+}
+
+void iogood() {
+    if (! cin.good() ) {
+        cerr << "IO error! cin.rdstate() == " << cin.rdstate() << endl;
+        
+        exit(1);
+    }
+}
+
+void mustbe(char c) {
+    iogood();
+    
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/Phylo_module.py

@@ -1,0 +1,136 @@
+import numpy as np
+import copy
+import sys
+import os
+from ete3 import Tree
+import time 
+import subprocess
+
+def duplicates(lst, item):
+	return [i for i, x in enumerate(lst) if x == item]
+
+def gen_Newick(genotype, PerfectPhy_path, out_dir_path, names_):
+	###########################################################################################
+	###################### Remove the columns with only one state #############################
+	###########################################################################################
+	R = genotype
+	R = np.array(R)
+	l = R.shape[1]
+	removal_R = []
+	num_of_char = 2 ## This is a binary case
+	for j in range(l):	
+		result_count = len(set(R[:,j]))
+		if result_count<num_of_char:
+			removal_R.append(j)
+
+	print(str(len(set(removal_R)))+" characters have been removed from the result")
+	R = np.delete(R, removal_R, 1)
+	#### The number of columns has changed
+	n = len(R)
+	l_R_removal = len(R[0])
+	############################ Second Post Processing Step ###############################
+	#### Build the input matrix of PerfectPhy without columns containing only 0s or 1s #####
+	########################################################################################
+	if not PerfectPhy_path.endswith("/"):
+		PerfectPhy_path+="/"
+	result_name = PerfectPhy_path+"src/tree_result.txt"
+	if not out_dir_path.endswith("/"):
+		out_dir_path+="/"
+	result_newick = out_dir_path+"phylogeny.nex"
+	cmd = 'rm '+result_name
+	os.system(cmd)
+	cmd ='rm '+result_newick
+	os.system(cmd)	
+	tree_result = open(result_name, "w")
+	tree_result.write("0"+"\n")
+	tree_result.write(str(n)+"\t"+str(l_R_removal)+"\n")
+
+
+	####################### Third Post Processing Step ###############################
+	################## Find the duplicated taxa in matrix/matrices ###################
+	##################################################################################
+	R_taxa = []
+	for i in range(n):
+		temp2 = ""
+		for j in range(l_R_removal):
+			temp2+=str(int(R[i][j]))
+			
+		R_taxa.append(temp2)
+
+	R_taxa=np.array(R_taxa)
+
+	##########################################################
+	####### Prepare the input file/files of PerfectPhy #######
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/VCF.py

@@ -1,0 +1,53 @@
+import numpy as np
+from datetime import date
+
+def gen_VCF(out_dir, genotype_mat, read_count_mat_, chrs, posits, alt_counts, rfs, ids, dps):
+	# print dps
+	read_count_mat_ = np.array(read_count_mat_)
+	n = read_count_mat_.shape[0]
+	l = read_count_mat_.shape[1]
+	now = date.today().strftime("%Y%m%d")
+
+	out_f = open(out_dir+"snv.vcf", "w")
+	header = "##fileformat=VCF\n##fileDate="+now+"\n##FILTER=<ID=LowQual,Description=\"Low quality\">\n##INFO=<ID=DP,Number=1,Type=Integer,Description=\"Approximate read depth; some reads may have been filtered\">\n##FORMAT=<ID=GT,Number=1,Type=String,Description=\"Genotype\">\n##FORMAT=<ID=DP,Number=1,Type=Integer,Description=\"Read Depth\">\n##FORMAT=<ID=DPR,Number=1,Type=Integer,Description=\"Number of observation for each allele\">\n##FORMAT=<ID=RO,Number=1,Type=Integer,Description=\"Reference allele observed count\">\n##FORMAT=<ID=AO,Number=1,Type=Integer,Description=\"Alternate allele observed count\">\n#CHROM\tPOS\tID\tREF\tALT\tFILTER\tINFO\tFORMAT"
+	for cell in ids:
+		header+="\t"+cell
+	out_f.write(header)
+	for pos in range(l):
+		total_depth = 0
+		tmp_str = str(chrs[pos])+"\t"+str(posits[pos])+"\t*\t"+rfs[pos]+"\t"
+		for id_indx in range(n):
+			total_depth=total_depth+dps[id_indx][pos]
+			tmp_set = []
+			alt_string = alt_counts[id_indx][pos]
+			for ch in alt_string:
+				tmp_set.append(ch)
+		alt_arr_ = list(set(tmp_set))
+		if len(alt_arr_)!=0:
+			for c in range(len(alt_arr_)):
+				if c==len(alt_string)-1:
+					tmp_str+=alt_arr_[c]
+				else:
+					tmp_str+=alt_string[c]+","
+		else:
+			tmp_str+="*"
+		out_f.write(tmp_str+"\t")
+		out_f.write("PASS\t")
+		out_f.write("DP:"+str(total_depth)+"\t")
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/loci_identify.py

@@ -1,0 +1,104 @@
+import numpy as np
+import copy
+import argparse 
+import sys
+
+def match(str_):
+	tmp = 0
+	tmp+=str_.count(",")
+	tmp+=str_.count(".")
+	return tmp
+def mismatch(str_):
+	tmp=0
+	tmp+=str_.count("A")
+	tmp+=str_.count("C")
+	tmp+=str_.count("G")
+	tmp+=str_.count("T")
+	tmp+=str_.count("N")
+	tmp+=str_.count("a")
+	tmp+=str_.count("c")
+	tmp+=str_.count("g")
+	tmp+=str_.count("t")
+	tmp+=str_.count("n")
+	return tmp
+
+if __name__ == "__main__":
+	num_cells = 0
+	out_path = ""
+	in_path = ""
+	ms = 0
+	nmc = 0
+
+	ap = argparse.ArgumentParser()
+	ap.add_argument("-names","--cell names", required=False, help="file containing the cell names")
+	ap.add_argument("-n","--number of cells",required=False, help="the number of cells ")
+	# ap.add_argument("-lst","--list of the cell names",required=False,
+		# help="a file containing the names of the cells at each row, each row must be of the form \'cell CELL_ID\', for example \'cell 200\' denotes the cell with the ID 200")
+	ap.add_argument("-out","--output file name",required=False, help="path of the summary mpileup file")
+	ap.add_argument("-in","--input mpileup file",required=True, help="path to the input file")
+	ap.add_argument("-ms","--minimum coverage",required=False, help="The minimum number of reads required to support the alternative, default value 3")
+	ap.add_argument("-nmc","--minimum cells",required=False,
+		help="The number of cells having a minimum number of alternative reads needed to consider a genomic location a mutation candidate locus, default value 2")
+	args = vars(ap.parse_args())
+
+	if args['number of cells']==None and args['cell names']==None:
+		print("Please either provide the list of cell names or enter the number of cells \nUsage: python -in <input mpileup file> -n <number of cells> | -names <txt file containing the cell names>")
+		sys.exit()
+	elif args['number of cells']!=None:
+		num_cells = int(args['number of cells'])
+		cell_names = open(out_path+'cellNames.txt', 'w')
+		for cell_indx in range(num_cells):
+			cell_names.write('cell'+str(cell_indx)+'\n')
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/parser.py

@@ -1,0 +1,107 @@
+import numpy as np
+import sys
+import os
+import gc
+
+def chr_extract(chr_):
+	if chr_[-1].endswith("X") or chr_[-1].endswith("x"):
+		return 23
+	elif chr_[-1].endswith("Y") or chr_[-1].endswith("y"):
+		return 24
+	elif chr_[-1].endswith("M") or chr_[-1].endswith("m"):
+		return 25
+	else:
+		chrString = ""
+		for i in chr_:
+			if i.isdigit():
+				chrString+=i
+		return int(chrString)
+
+''' These two functions are used for reading the mpileup files '''
+def match(str_):
+	tmp = 0
+	tmp+=str_.count(",")
+	tmp+=str_.count(".")
+	return tmp
+def mismatch(str_):
+	alternates = ""
+	tmp=0
+	tmp+=str_.count("A")
+	tmp+=str_.count("C")
+	tmp+=str_.count("G")
+	tmp+=str_.count("T")
+	tmp+=str_.count("N")
+	tmp+=str_.count("a")
+	tmp+=str_.count("c")
+	tmp+=str_.count("g")
+	tmp+=str_.count("t")
+	tmp+=str_.count("n")
+	if str_.count("A")!=0 or str_.count("a")!=0:
+		alternates+="A"
+	if str_.count("C")!=0 or str_.count("c")!=0:
+		alternates+="C"
+	if str_.count("T")!=0 or str_.count("t")!=0:
+		alternates+="T"
+	if str_.count("G")!=0 or str_.count("g")!=0:
+		alternates+="G"
+	if str_.count("N")!=0 or str_.count("n")!=0:
+		alternates+="N"
+	if alternates=="":
+		alternates="*"
+	return (tmp, alternates)
+
+def Parse(cell_names_file, mpileup_file):
+	names_file= open(cell_names_file,"r")
+	names_ = []
+	arr_names = names_file.readlines()
+	names_file.close()
+	for line in arr_names:
+		if len(line.strip())!=0:
+			tmp = line.strip().split("\t")
+			names_.append(tmp[0])
+
+	pileup = open(mpileup_file,"r")
+
+	read_counts_ = []
+	chroms_ = []
+	positions_ = []
+	refs_ = []
+	alts_ = []
+	depth_ = []
+	num_cells = len(names_)
+	for i in range(num_cells):
+		read_counts_.append([])
+		alts_.append([])
+		depth_.append([])
+
+	with open(mpileup_file, "r") as infile:
[…]

### 2021-01-03 analysis_pipelines/scripts/scVILP/scripts/scVILP_main.py

@@ -1,0 +1,289 @@
+#!/usr/bin/env python
+import numpy as np 
+import argparse
+from gurobipy import *
+import copy
+import sys
+import os
+import matplotlib.pyplot as plt
+# plt.switch_backend('agg')
+import seaborn as sns 
+# sns.set_style("whitegrid")
+import math
+import scipy
+import subprocess
+from scipy.stats import *
+from scipy.special import comb
+from scipy.special import gammaln
+from ete3 import Tree
+import time 
+from decimal import Decimal
+import scipy.cluster.hierarchy as hier
+import scipy.spatial.distance as dist
+import parser
+import VCF
+import Phylo_module
+
+
+def Binomial_pmf(k,n,p):
+	''' calculates the pmf of binomial distribution '''
+	k_decimal = Decimal(k)
+	n_decimal = Decimal(n)
+	p_decimal = Decimal(p)
+	tmp = Decimal(gammaln(n+1)-gammaln(k+1)-gammaln(n-k+1))+Decimal(k_decimal*p_decimal.ln()+(n_decimal-k_decimal)*Decimal(1-p_decimal).ln())
+	return tmp.exp()
+
+def check_InCompatibility(Matrix):
+
+	'''Count the number of character pairs which violate
+	 the infinite-sites assumption '''
+	num_incompatibles = 0
+	pairs = []
+	Matrix = np.array(Matrix)
+	n = Matrix.shape[0]
+	l = Matrix.shape[1]
+	B_prime = [[[0 for k in range(4)] for j in range(l)] for i in range(l)]
+	for p in range(l):
+		q = p+1
+		while q<l:
+			count01=0
+			count10=0
+			count11=0
+			for cell in range(n):
+				if count01+count10+count11==3:
+					break  
+				# if Matrix[cell][p]==0 and Matrix[cell][q]==0:
+				# 	B_prime[p][q][0]=1
+				if Matrix[cell][p]==0 and Matrix[cell][q]==1:
+					B_prime[p][q][1]=1
+				elif Matrix[cell][p]==1 and Matrix[cell][q]==0:
+					B_prime[p][q][2]=1
+				elif Matrix[cell][p]==1 and Matrix[cell][q]==1:
+					B_prime[p][q][3]=1
+			q+=1
+	for p in range(l):
+		q=p+1
+		while q<l:
+			s = sum(B_prime[p][q])
+			if s==3:
+				num_incompatibles+=1
[…]

### 2021-01-03 manuscript/main.tex
manuscript: update to latest text, scripts, figures and results data

@@
 
-\title{ProSolo: Accurate Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
+\title{ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
 \author[1,2,3,4]{David Lähnemann}
@@
 However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
-As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which when applied to single cells generate significant numbers of false positives and false negatives.
+As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which---when applied to single cells---generate significant numbers of false positives and false negatives.
 Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.
 
-We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified single cell DNA sequencing data.
-It introduces a mechanistically motivated empirical model of amplification bias that improves the quantification of genotyping uncertainty.
-To account for amplification errors, it jointly models the single cell sample with a bulk sequencing sample from the same cell population---also enabling a biologically relevant imputation of missing genotypes for the single cell.
-Through these innovations, ProSolo achieves substantially higher performance in calling and genotyping single nucleotide variants in single cells in comparison to all state-of-the-art tools. 
+We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
+The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
+ProSolo models single cells jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered polymorphic sites.
+As a consequence, ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
 Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.
@@
 With the size of the human (reference) genome at approximately $3.2 \cdot 10^9$ base pairs, these numbers indicate that even during healthy development, most cells harbour cell-specific point mutations.
-This enables retrospective monitoring of lineages involved in normal organism development, merely by sampling some cells\cite{kester_single-cell_2018}, without having to interfere with its general development, or kill the individual.
+This enables retrospective monitoring of lineages involved in normal organism development, merely by sampling some cells\cite{kester_single-cell_2018}, without having to interfere with its general development or having to kill the individual.
 In other words, this establishes a universally applicable methodology for \textit{in vivo} lineage tracing.
@@
 Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
-Improving on that point, tools exist that account for local variation in allelic amplification bias.
-SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019} both estimate the minor allele frequency from nearby germline heterozygous sites.
+Two tools that more variably model alleleic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
+Both estimate the minor allele frequency from nearby germline heterozygous sites. 
 However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
@@
 We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
-Importantly, our model further enables a computationally efficient implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
+Importantly, our model enables a scalable---because computationally efficient---implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
 ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.
@@
    This is formalised with the model in D.
-   \textbf{(D)} Definition of single cell events based on true underlying alternative allele frequencies in the single cell ($\theta_s$) and the bulk ($\theta_b$) (assuming that the bulk sample has a deep coverage that captures somatic variants).
-   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).\small\newline
+   \textbf{(D)} Definition of single cell events based on ProSolo's likelihood density estimates for the spectrum of true underlying alternative allele frequencies in the single cell ($\tilde{\theta}_s$) and the bulk ($\tilde{\theta}_b$).
+   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
+   The model further assumes that the bulk sample has sufficient coverage to capture somatic variants.\small\newline
    ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
@@
 the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
-We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele.
+We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele, to calculate likelihoods for each of these possible true allele frequencies.
 
@@
 
-For our model, we derive likelihoods for all possible alternative allele frequencies in the background bulk sample.
+For our model, we derive likelihood density estimates for all possible alternative allele frequencies in the background bulk sample.
 Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:
@@
 With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
-Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihoods works for calling genotypes in the single cell.
+Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
 However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:
@@
   \label{eq:event-space}
-  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\theta_s} \times [0, 1]_{\theta_b}.
+  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\tilde{\theta}_s} \times [0, 1]_{\tilde{\theta}_b}.
 \end{equation}
 
-Within this space, we define single cell events as mutually exclusive subspaces, for example an error-free homozygous alternative site is defined by allele frequencies $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\theta_s} \times (\frac{1}{2}, 1]_{\theta_b}$  or the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
+For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel).
+At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
+At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
+Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
+For example, an error-free homozygous alternative site is defined by allele frequency likelihood estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
+And the dropout of an alternative allele is defined across allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
+For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
+We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
+This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.
+
 We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):
@@
 For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
-All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769116}).
+All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3824152}).
 
@@
 \begin{figure}[!tpb]
-  \centering
-  \begin{minipage}{.26\linewidth}
-    \includegraphics[height=32ex]{figs/Dong2017_data_overview.pdf} \\
+  \begin{minipage}{.3\linewidth}
+    \includegraphics[height=38ex]{figs/Dong2017_data_overview.pdf} \\
   \end{minipage}
-  \hspace{4ex}
-  \begin{minipage}{.4\linewidth}
-    \includegraphics[height=32ex]{figs/Laehnemann2017_data_overview.pdf} \\
+  \hspace{6ex}
+  \begin{minipage}{.5\linewidth}
+    \includegraphics[height=38ex]{figs/Laehnemann2017_data_overview.pdf} \\
+  \end{minipage}\\
+  \begin{minipage}{.5\linewidth}
+    \includegraphics[height=38ex]{figs/Wang2014_data_overview.pdf} \\
   \end{minipage}
-  \hspace{4ex}
-  \begin{minipage}{.22\linewidth}
-    \includegraphics[height=32ex]{figs/data_overview_color_legend.pdf} \\
+  \hspace{6ex}
+  \begin{minipage}{.3\linewidth}
+    \includegraphics[height=38ex]{figs/data_overview_color_legend.pdf} \\
   \end{minipage}
@@
    Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
+   \textbf{C} Whole exome sequencing (WXS) dataset from a triple negative breast cancer (TNBC) patient\cite{wang_clonal_2014}.
+   For tumor cells (a1 to a8 and h1 to h8), the cell population (pop) sequencing of the normal sample was taken as the ground truth, including known clonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
+   For normal cells (n1 to n16), the cell population (pop) sequencing of the tumor sample was taken as the ground truth, after removing known clonal and subclonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
    }
@@
 
-We benchmarked ProSolo on two experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\
+We benchmarked ProSolo on three experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\
 
@@
 Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
-% refs for MACS and FACS? Ute said, this is usually not required.
 Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
 Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
-% ref for q-rt-PCR? Ute said, this is usually not required.
 For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
@@
 
-\subsubsection{ProSolo achieves highest alternative allele calling accuracy}
+\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC)\cite{wang_clonal_2014}.}
+The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available (Figure~\ref{fig:datasets}C, dataset available via the sequence read archive, project id \mbox{PRJNA168068} or accession \mbox{SRA053195}).
+For each, 16 single cell nuclei and a bulk sample were whole exome sequenced.
+Further, a number of candidate sites for clonal and subclonal tumor mutations was confirmed by targeted duplex deep sequencing in a separate experiment\cite{wang_clonal_2014}.
+We analyzed the tumor and normal cells separately, ensuring that normal cells were called with the normal bulk background sample and tumor cells with the tumor bulk background sample.
+As the ground truth for the tumor cells, we then used the normal bulk sample augmented with the clonal tumor mutations confirmed by targeted duplex sequencing.
+As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations.
+Note that this experimental setup accounts for utmost fairness for all competitors.
+Variant calling for the ground truths was performed as for cell line data above.
+
+\subsubsection{ProSolo achieves consistently high alternative allele calling accuracy}
 
@@
 \begin{figure}[!tpb]
- \begin{minipage}{.40\linewidth}
+ \begin{minipage}{.48\linewidth}
   \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf} \\
  \end{minipage}
- \begin{minipage}{.55\linewidth}
+ \begin{minipage}{.48\linewidth}
   \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \\
+ \end{minipage}\newline
+ \begin{minipage}{.9\linewidth}
+  \includegraphics[height=40ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
  \end{minipage}
@@
   Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
-  Note that both panels are strong zoom-ins, focusing on (different) areas of interest.
+  Note that all panels are strong zoom-ins, focusing on (different) areas of interest.
   Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
   \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
-  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline \footnotesize
+  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
+  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.
+\newline \footnotesize
   Threshold parameters:
@@
   SCcaller {\ttfamily -a cutoff};
-  SCIPhI {\ttfamily none available}.
+  SCIPhI {\ttfamily prosolo --fdr}.
   Software modes:
@@
 Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}B).
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .176, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
 In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
@@
 
-As almost all tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
-However, it should be noted that SCIPhI took from 3 weeks on the whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
+For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
+On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
+On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
+Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a drastically reduced precision.
+This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
+Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
+While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
+As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
+However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
+And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, both will further increase SCIPhI's wall time.
+In addition, adding more cells exponentially grows the space of possible tree topologies that SCIPhI explores\cite{lahnemann_eleven_2020}, which will further increase its runtime.
 
@@
 As can be seen in Figure~\ref{fig:alt-calling_prec-rec} (and Figure~\ref{fig:alt-calling_prec-rec_global}), ProSolo provides flexible control over precision vs.
-recall via specifying a false discovery rate of interest.
+recall via specifying a false discovery rate of interest and is the only tool to achieve a precision of over .99 on the single tumor cell datasets from Wang et al\cite{wang_clonal_2014}.
 While no other tool provides a formal control over the easily interpretable false discovery rate, several of the tools provide other types of thresholds that we varied in attempts to achieve higher precision or recall.
 However, none of them provide control over similar ranges of precision and recall.
+In addition, we ran the standard false discovery rate control implemented in ProSolo on the posterior probabilities provided by SCIPhI, but this did not provide any substantial control over the false discovery rate (Figure~\ref{fig:alt-calling_prec-rec} and \ref{fig:FDR-ground-truth-vs-theoretical}).
 The only limit to that range with ProSolo's current model, is that it becomes less accurate when controlling for very small false discovery rates (below 0.01\% for alternative allele calling in the whole genome dataset, see Supplement Section~\ref{sec:fdr-of-alt-calling}).
@@
 
-ProSolo is the first method for SNV calling from MDA single cell sequencing data to comprehensively model both amplification bias and amplification errors in a way that allows for site-specific variation (Figure \ref{fig:prosolo_alt-calling}).
-We achieve this by combining a data-driven model of amplification errors that incorporates a bulk background sample alongside the single cell sample with an empirical model of amplification bias, based on a statistical understanding of the MDA process that is mechanistically motivated and dependent upon a site's coverage.
-The underlying model calculates posterior probabilities for fine-grained single cell event definitions, whose false discovery rate can be controlled for and that can pass more information about data uncertainties to probabilistic models in downstream analyses.
-Using a whole genome and a whole exome dataset---each with a different type of ground truth (Figure~\ref{fig:datasets})---we demonstrate that these two innovations of ProSolo combined to achieve better accuracy (Figure~\ref{fig:alt-calling_prec-rec}).
+ProSolo is the first method for SNV calling from MDA single cell sequencing data to comprehensively model both amplification bias and amplification errors in a way that enables both
+\begin{enumerate*}
+ \item the integration of site-specific differences (Figure \ref{fig:prosolo_alt-calling}), and
+ \item a computationally efficient---hence practically scalable---evaluation of all relevant data properties.
+\end{enumerate*}
+We achieve this by combining a data-driven model of amplification errors (that incorporates a bulk background sample alongside the single cell sample) with an empirical model of amplification bias which is dependent upon a site's coverage (and is based on a statistical understanding of the MDA process that is mechanistically motivated).
+The underlying model calculates posterior probabilities for fine-grained single cell event definitions, whose false discovery rate can be controlled for and that can pass more information about data uncertainties to probabilistic models in downstream analyses; including methods that can then use this reduced, but highly accurate, data representation to integrate information across cells, for example for the phylogenetic reconstruction of cell relationships.
+Importantly, the calculation of posterior probabilities scales linearly in the coverage of variant sites, which is the fastest theoretically conceivable, documented by runtimes that have substantial advantages over the state of the art in practice.
+Using one whole genome dataset and two whole exome datasets---each with a different type of ground truth (Figure~\ref{fig:datasets})---we demonstrate that these two innovations of ProSolo combined to achieve competing or better accuracy (Figure~\ref{fig:alt-calling_prec-rec}).
 Moreover, the model allows to accurately control the false discovery rate of variant calls, another novelty of great practical value, and both the model and its modular implementation are easy to adapt and extend.
 
-Especially the joint modeling of a single cell and a bulk sample is favorable from a statistical point of view: in terms of MDA induced bias, the bulk acts as an unbiased sample of the population from which the single cell was drawn.
-This provides a drastically more sensitive approach compared to, for example, the consensus rule of calling alternative alleles only when there is evidence in at least two (or even three) single cells (as implemented in MonoVar\cite{zafar_monovar:_2016}).
+Especially the joint modeling of a single cell and a bulk sample is favorable from a statistical point of view: in terms of MDA induced bias and errors, the bulk acts as an unbiased sample of the population from which the single cell was drawn.
+Given enough bulk coverage (Figure~\ref{fig:somatic-recall}), this provides a drastically more sensitive approach compared to, for example, the consensus rule of calling alternative alleles only when there is evidence in at least two (or even three) single cells (as implemented in MonoVar\cite{zafar_monovar:_2016}).
 A more systematic and biologically relevant model of sharing information is implemented in SCIPhI, where the phylogenetic relationship of cells is part of the estimated parameters.
@@
 However, both the consensus rule and the sharing of information via an inferred phylogeny requires that more than one single cell exhibit sufficient coverage of an alternative allele to call it reliably.
-This will rarely be achieved in single cell MDA experiments, which are often limited to a few dozen cells.
+This will rarely be achieved in single cell MDA experiments, which are still often limited to a few dozen cells.
 In contrast, bulk sequencing can easily be scaled to a much larger sampling of a cell population---representing hundreds or thousands of cells of a cell population---simply by adding a single higher coverage sample.
-Our benchmarking demonstrated that ProSolo achieves a substantially higher recall than any other tool for a precision above 99\% in identifying the presence of an alternative allele.
+Our benchmarking experiments demonstrated that ProSolo, by jointly modelling a single cell sample with a bulk sample, is the only tool to consistently achieve a precision above .99.
+In almost all datasets this precision is provided in concert with a substantially higher recall than any other tool.
+The only exception is the normal cells dataset from the single nucleus exome sequencing data, for which SCIPhI attains a higher recall above .99 precision---but at the cost of false positive calls of subclonal heterozygous variants (Figure~\ref{fig:somatic-recall}), while lacking the ability to call sites with a homozygous alternative genotype (Figure~\ref{fig:somatic-recall}).
+Interestingly, a bulk coverage as low as 4X already provides drastic performance improvements in ProSolo's model and allows for a meaningful control over the false discovery rate (Supplementary Figure~\ref{fig:downsampling}, but higher bulk coverage will be necessary to detect low-frequency somatic variants (Supplementary Section~\ref{sec:somatic-recall}).
 To summarize our benchmarking of alternative allele calling: SCcaller seems optimized for precision, MonoVar achieved higher recall than SCcaller, and SCAN-SNV does not seem suitable for general variant calling on MDA single cell sequencing data, but is only applicable when restricting interest to somatic variants.
-ProSolo clearly achieves the best accuracy, with particularly striking increases in recall. 
-The only other tool that shows consistent performance across both datasets, but with a recall 10 to 20\% below ProSolo, is SCIPhI.
-However, SCIPhI takes several weeks on a single core without any possibility of parallelization, compared to a runtime of only a few days on a cluster for all the other tools.
+ProSolo clearly achieves the best precision.
+In the two out of three datasets with fewer cells, this precision comes with a higher recall than in any other tool.
+This demonstrates, that ProSolo's model captures data from individual single cells better than any of the existing models.
+Only in the dataset with more cells and a bulk sample with a lower coverage, SCIPhI can reach a higher recall than ProSolo by integrating information across cells (but at the cost false positive somatic mutation calls).
+However, with default configuration SCIPhI takes weeks on a single core without any possibility of parallelization, compared to a runtime of only days on a cluster for all the other tools.
+This is the result of the general approach, integrating information across cells and across genomic sites all at once.
+Thus, any addition of genomic coverage and cells will further increase SCIPhI's wall clock runtime, where adding cells could prove especially troubling: adding cells will exponentially grow the space of possible tree topologies to explore.
+Other tools, including ProSolo, can simply add more parallel processes per site and cell and can thus leverage cluster computing for larger datasets, something that becomes more urgent with the advent of scalable droplet-based single cell whole genome amplification methods\cite{lahnemann_eleven_2020}.
 
@@
 If genotype profiles without missing values are for example required by downstream tools, ProSolo can provide a cell-population-specific imputation instead of resorting to a majority assignment (limited by the number of single cells sequenced) or even imputation based on external databases (e.g.~dbsnp).
-As our results further demonstrate, ProSolo’s empirical model of amplification bias is robust also for whole exome sequencing data. 
-In contrast, SCcaller---that estimates local amplification bias based on minor allele frequencies at known germline genotypes in the vicinity of an examined site---suffers from a scarcity of neighbouring sites in whole exome data.
+However, any imputation will usually favour germline genotypes over low-frequency somatic genotypes.
+Thus, we suggest to not impute zero-coverage single cell sites whenever possible and instead recommend using and developing downstream software that can deal with both these missing values and the event probabilities that ProSolo provides.
+With this approach, uncertainties in the probabilities and information about missing data are passed on and will allow for more accurate statistical modeling in those downstream analyses.
 
-Finally, ProSolo is the only method that provides a clearly interpretable false discovery rate parameter that can actually control the trade-off between precision and recall.
+Finally, ProSolo is the only method that provides a clearly interpretable false discovery rate parameter that can actually control the trade-off between precision and recall (and this false discovery rate is the only parameter any user will have to specify).
 This can be used on alternative allele calls (Figure~\ref{fig:alt-calling_prec-rec}, Supplement Section~\ref{sec:fdr-of-alt-calling}), the main focus of our benchmarking, but also on any of the events pertinent to single cell analysis (such as allele dropout or amplification errors) or combinations thereof (see Figure~\ref{fig:prosolo_alt-calling}D for all events).
@@
 
-In addition to such global rates, ProSolo determines reliable site-specific posterior probabilities for all single cell events that might be of interest, including allele-resolved genotypes, allele dropout or amplification errors.
-We anticipate that such fine-grained probabilities---and the uncertainties they capture---will be informative for improving probabilistic modelling in downstream analyses.
-They could e.g.~prove useful in models for phylogenetic reconstruction of the lineage relationship of sequenced single cells\cite{singer_single-cell_2018,zafar_siclonefit:_2018,koptagel_scuphr:_2018}, while keeping them computationally tractable\cite{lahnemann_eleven_2020}.
+ProSolo determines such global rates from its reliable site-specific posterior probabilities for all single cell events that might be of interest, including allele-resolved genotypes, allele dropout or amplification errors.
+We anticipate that such fine-grained site-specific probabilities---and the uncertainties they capture---will be informative for improving probabilistic modelling in downstream analyses.
+They could e.g.~prove useful in models for phylogenetic reconstruction of the lineage relationship of sequenced single cells\cite{singer_single-cell_2018,zafar_siclonefit:_2018,koptagel_scuphr:_2018}, thus achieving results similar to or even better than SCIPhI, while keeping the analyses modular and thus computationally tractable\cite{lahnemann_eleven_2020}.
 
@@
 However, the allele dropout rate estimations were slightly off in different directions for the different benchmarking datasets, suggesting that the empirical distributions we currently make use of (based on Lodato et al.\cite{lodato_somatic_2015}) may not suit all datasets.
-For example, whole genome amplification may introduce further variability so far not captured by our model.  
+For example, whole genome amplification may introduce further dataset-specific variability so far not captured by our model.
 This conclusion is further supported by the naively calculated allele dropout rate, which is much lower in the high coverage cells of the two datasets (IL-11 for the Dong et al.
@@
 If samples PAG1 and PAG10 were doublets as well, this would indicate that our use of empirical distributions in ProSolo provides for more robust event probabilities in the presence of doublets, while heuristic thresholding (as in our naive allele dropout estimate) is very sensitive to such perturbations.
-In general, the use of a fixed empirical model for MDA allelic bias does not seem to impede ProSolo's performance in alternative allele calling compared to the other tools, but has a noticeable effect on genotyping (Supplement, Section~\ref{sec:suppl-gt-calling}) and might be responsible for slight imprecisions when controlling for very small false discovery rates (Supplement Section~\ref{sec:fdr-of-alt-calling}).
+In general, the use of a fixed empirical model for MDA allelic bias does not seem to impede ProSolo's performance in alternative allele calling compared to the other tools, but has a noticeable effect on genotyping (Supplement, Section~\ref{sec:suppl-gt-calling} and Figure~\ref{fig:somatic-recall}) and might be responsible for imprecisions when controlling for very small false discovery rates (Supplement Section~\ref{sec:fdr-of-alt-calling}).
 When future datasets are generated based on improved MDA protocols\cite{lahnemann_eleven_2020}, these effects might be exacerbated.
@@
 
-Overall, ProSolo provides an accurate and easy-to-use variant caller for single cell MDA sequencing data, which will empower more research using single cell sequencing data.
+Overall, ProSolo provides an accurate and easy-to-use variant caller for single cell MDA sequencing data, which will easily scale to calling variation on more cells and broader genomic coverage.
+It will thus empower more research using single cell DNA sequencing data.
 

### 2021-01-03 manuscript/supplement.tex
manuscript: update to latest text, scripts, figures and results data

@@
 \usepackage{graphicx}
+\usepackage{subfig}
 \usepackage{amsthm}
@@
  \caption{
-  It depicts single cell events at a particular genomic site, defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
-  The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
+  It depicts single cell events at a particular genomic site, defined by combinations of ProSolo's maximum likelihood estimates for possible true alternative nucleotide frequencies in single cell sequencing ($\tilde{\theta}_s$) and bulk sequencing samples ($\tilde{\theta}_b$) in our probabilistic model.
+  The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\tilde{\theta}_b$ (bottom panel).
   A tabular representation of the same event definitions is given in Table~\ref{tab:Events} below (this is a larger copy of Figure~\ref{fig:prosolo_alt-calling}D in the main manuscript, reproduced  here for easier cross-reference with that table).\footnotesize\newline
@@
   \caption{
-   Single cell events at a particular genomic site are defined by combinations of the possible true alternative nucleotide frequencies in single cell sequencing ($\theta_s$) and bulk sequencing samples ($\theta_b$) in our probabilistic model.
+   Single cell events at a particular genomic site are defined by combinations of ProSolo's likelihood density estimates of the possible true alternative nucleotide frequency ranges in single cell sequencing ($\tilde{\theta}_s$) and bulk sequencing samples ($\tilde{\theta}_b$) in our probabilistic model.
    A graphical representation of the same event definitions is given in Figure~\ref{fig:event-space}.
@@
    \toprule
-   single cell event & $\theta_s(x)$   & $\theta_b(x)$ & implicit single cell genotype \\
+   single cell event & $\tilde{\theta}_s(x)$   & $\tilde{\theta}_b(x)$ & implicit single cell genotype \\
    \toprule
@@
 Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
-All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequencies $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
-We then define mutually exclusive subspaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}$, providing a set of events $\mathfrak{E}$.
-Assuming that there is no informative prior information about different allele frequencies (i.e.
+All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequency likelihood density estimates $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
+We then define mutually exclusive sub-spaces of $\boldsymbol{E}$ as single cell events, for example the dropout of an alternative allele across allele frequencies $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$, providing a set of events $\mathfrak{E}$.
+For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:event-space}, lower panel).
+At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
+At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
+For the above example ($\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$), this means that we can resolve a contradiction between the likelihood estimate at $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and the likelihood density estimate across $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
+Based on our assumption, we decide to trust the bulk sample over the single cell sample, meaning that the bulk is a mixture of a maximum of two sub-populations, and can only contain heterozygous and homozygous alternative cells, rendering a homozygous reference single cell impossible and thus the result of an allele dropout.
+
+Further assuming that there is no informative prior information about different allele frequencies (i.e.
 assuming a flat prior across all possible allele frequencies in both samples), we can then calculate the likelihood of such an event as the product of the likelihoods of the allele frequency ranges in the two samples.
@@
 \begin{figure}[!tpb]
- \begin{minipage}{.99\linewidth}
-  \includegraphics[width=.99\linewidth]{figs/Dong2017/Dong2017_coverage_dist.pdf}
+ \begin{minipage}{\linewidth}
+  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_coverage_dist.pdf}
  \end{minipage} \\
- \begin{minipage}{.99\linewidth}
-  \includegraphics[width=.99\linewidth]{figs/Laehnemann2017/Laehnemann2017_coverage_dist.pdf}
- \end{minipage}
- \caption{
- Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples.
+ \begin{minipage}{\linewidth}
+  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_coverage_dist.pdf}
+ \end{minipage}
+ \caption{
+ Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples (continued on following page).
  Panel \textbf{A} shows the coverage of the single cells (IL-11, IL-12), the kindred ground truth clone (IL-1c) and the more distant background bulk (Clones C1, C2, and C3 joined into one sample) of the dataset from \cite{dong_accurate_2017}.
  Panel \textbf{B} shows the coverage of the single cells (PAG1, PAG2, PAG5, PAG9, PAG10) and the background bulk (PNG) of the granulocyte data generated for this evaluation.
+ }
+ \label{fig:coverage-dists}
+\end{figure}
+
+\begin{figure}[!tpb]
+ \ContinuedFloat
+ \begin{minipage}{\linewidth}
+  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_coverage_dist.pdf}
+ \end{minipage}
+ \caption{
+ Sequencing coverage of targeted genomic sites for multiple displacement amplified (MDA) single cells and respective bulk samples (continued from previous page).
+ Panel \textbf{C} shows the coverage of the single cells (a1 to a8, h1 to h8, and TNBC-n1 to TNBC-n16), and the bulk samples (TNBC-Pop-Tumor and TNBC-Pop-Normal) that were used as ground truth for the normal single cells (with validated somatic variants added) and as ground truth for the tumor single cells (with validated somatic variants removed), respectively \cite{wang_clonal_2014}.
  }
@@
 \end{figure}
+
+\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
+The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
+For each, 16 cells and a bulk sample were whole exome sequenced.
+For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
+In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
+Depending on the single cell, we used two different ground truths for two separate evaluations:
+\begin{enumerate}
+    \item For the tumor single cells, we constructed a ground truth from normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
+    Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
+    Thus, similar caveats as for the germline ground truth in the granulocyte dataset apply.
+    To mitigate this to a certain degree, we added the 373 clonal validated somatic variants, as these are expected to be present in all tumor single cells.
+    \item For the normal single cells, we constructed a ground truth from the tumor bulk sample (as the normal bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
+    To reduce the presence of false-positive somatic variants in this ground truth, we removed all 467 validated somatic variants (clonal and subclonal) from the ground truth.
+\end{enumerate}
+Further, we determined the of the validated somatic variants for all software run on this dataset.
 
@@
 }.
-The different filter thresholds used represent the {\ttfamily --fdr} parameter, which---according to the authors---does not formally control the false discovery rate \citep{luquette_identification_2019}.\\
+The different filter thresholds used represent the {\ttfamily --fdr} parameter, which---according to the authors---does not formally control the false discovery rate \citep{luquette_identification_2019}.
+For the third dataset, added following a suggestion from peer-review, we had to recreate the conda environment for SCAN-SNV, following newer advice in the issues of the repository\footnote{see: \url{https://github.com/parklab/scan-snv/issues/4##issuecomment-596296645}}.
+While this allowed us to run the pipeline, the custom rule for ground truth comparison ran into unsolvable version conflicts.
+After multiple days of debugging attempts, we decided to exclude SCAN-SNV from the analysis of this dataset, also given its performance on the other datasets.\\
 
@@
 A rerun on a server with more performant CPUs led to the sensitive mode finishing after 5 weeks and the default mode after 7.5 weeks.
+On the third dataset, added after a suggestion during peer-review, we only ran SCIPhI with default parameters and reduced the iterations to {\ttfamily -l 80000} (and the number of iterations for the mutation to cell assignment to {\ttfamily --ls 40000}) to finish analysis in a reasonable time-frame.
+No guidance is given for those parameters, but the optimization score seemed to have mostly settled with this amount iterations.
+With these parameters, this analysis finished within a week for both the tumor cell and the normal cell datasets of \cite{wang_clonal_2014}.
 As no option for filtering or thresholding based on sensitivity vs.
-specificity is available, only two unfiltered data points are given in the respective plots (Figure~\ref{fig:alt-calling_prec-rec_global}).
+specificity is available from SCIPhI itself, we parsed output with custom scripts to extract what must be a posterior probability of the presence of a variant at a particular site into a format that ProSolo's false discovery rate control mechanism can handle.
+This is the value in the last (fourth) FORMAT field, which according to the header annotation is either the GQ or the PL field (five FORMAT fields are specified, but only four values per sample are provided).
+Judging from the source code\footnote{For the full relevant source code, see here: \url{https://github.com/cbg-ethz/SCIPhI/blob/34975f7050f86325f361c58e8146883af4a2d289/src/output.h##L202-L216}}, this field contains a PHRED scaled posterior probability of the site not being a variant\footnote{This is 1 minus a value that i is used for determining genotypes by SCIPhI with a fixed cutoff of 0.95 a few lines above: \url{https://github.com/cbg-ethz/SCIPhI/blob/34975f7050f86325f361c58e8146883af4a2d289/src/output.h##L205}}.
+We parsed this into the probability of a variant being present and passed it into {\ttfamily prosolo control-fdr} to attempt controlling the false discovery rate (Figures~\ref{fig:alt-calling_prec-rec_global} and \ref{fig:FDR-ground-truth-vs-theoretical}).
+
+For scVILP (has no versioning\footnote{We used the repository at this commit: \url{https://github.com/mae6/scVILP/tree/ef38bc422d44daafe043c7dcb28112a1f1ee2fe6}}; install instructions had to be amended\footnote{Pull request to amend install instructions: https://github.com/mae6/scVILP/pull/4} and manually fixing parsing and dependency issues\footnote{Pull request to allow handling of mitochondrial chromosome: \url{https://github.com/mae6/scVILP/pull/1}; Pull request to fix impossible import of {\ttfamily cdecimal} by upgrading to {\ttfamily decimal} \url{https://github.com/mae6/scVILP/pull/2}}), we tried two modes:
+\begin{enumerate}
+    \item Default parameters as suggested in the README of the software: \url{https://github.com/mae6/scVILP#identify-the-candidate-loci}.
+    \item A more sensitive identification of candidates by turning off heuristic filters ({\ttfamily -ms 1 -nmc 1}) during the execution of {\ttfamily loci\_identify.py}.
+\end{enumerate}
+However, scVILP was not able to handle any of our three benchmarking datasets, with the Gurobi solver that it uses always crashing with an out of memory error, even when provided with more than a Tebibyte of RAM.
+We documented details of runs and crashes for the developers of the tool in an issue in their code repository: \url{https://github.com/mae6/scVILP/issues/5}.
 
@@
 \begin{figure}[!tpb]
- \begin{minipage}{.36\linewidth}
+ \begin{minipage}{.48\linewidth}
   \includegraphics[height=42ex]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf} \newline
  \end{minipage}
- \begin{minipage}{.55\linewidth}
+ \begin{minipage}{.48\linewidth}
   \includegraphics[height=42ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf} \newline
+ \end{minipage}
+ \begin{minipage}{.9\linewidth}
+  \includegraphics[height=42ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot.pdf} \newline
  \end{minipage}
@@
   \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline \footnotesize
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
+  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
+  Thresholds used were applied via the following command line options:
+  MonoVar {\ttfamily --t};
+  ProSolo {\ttfamily --fdr};
+  SCAN-SNV {\ttfamily --fdr};
+  SCcaller {\ttfamily -a cutoff};
+  SCIPhI {\ttfamily prosolo --fdr}.
+  Software modes:
+  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
+  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
+  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
+  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
+  }
+ \label{fig:alt-calling_prec-rec_global}
+\end{figure}
+
+The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher.
+In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
+For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
+SCAN-SNV achieves up to .992 precision, but only .0001 recall.
+This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants\footnote{E.g.
+see this comment in the issues of the respective repository: \url{https://github.com/parklab/scan-snv/issues/5##issuecomment-628107370}}.
+
+Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
+This dataset is considerably more challenging than the first one.
+See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
+SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
+On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
+Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
+At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
+Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
+The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
+Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
+While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
+As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.
+
+\begin{figure}[!tpb]
+  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf}
+ \caption{
+ Cell-specific $F_1$ scores for different filtering threshold levels for the whole genome dataset.\newline \footnotesize
   Thresholds used were applied via the following command line options:
@@
  }
- \label{fig:alt-calling_prec-rec_global}
-\end{figure}
-
-The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher.
-In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
-For example, for a precision above .99, its maximum recall is .766 compared to .705 for MonoVar, .687 for SCIPhI and .610 for SCcaller.
-SCAN-SNV achieves up to .992 precision, but only .0001 recall.
-This can be explained by it aiming at somatic mutations, while the vast majority of SNVs in a genome will be germline variants.
-
-Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset.
-This dataset is considerably more challenging than the first one.
-See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .176, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
-In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
-SCcaller's decreased recall on this dataset might be due to its estimation of local allelic bias by also taking biases at neighboring sites into account---in whole exome data the number of neighboring sites available for this estimation will be limited and might lead to less reliable estimates.
-On this dataset, SCAN-SNV's recall increased to .0016\% at a decreased maximum precision of .897.
-Most likely, this decreased precision is an artefact of using the germline genotype as ground truth.
-At the sites with somatic mutations in single cells, which SCAN-SNV focuses on, this ground truth will instead contain the germline genotype and will incorrectly classify alternative alleles as false positives.
-Due to this effect, we also expect the calculated precision of all the other tools to be an underestimate, but against the backdrop of large numbers of sites where the single cells harbour the germline genotype, the effect will be much smaller.
-The same ground truth caveat also applies---inversely---for the recall of SCIPhI's sensitive mode and ProSolo's imputation mode.
-Whenever coverage of a site is missing in a single cell, SCIPhI may impute the genotype with the last common ancestor genotype of the most closely related cells, while ProSolo will impute to the majority genotype in the bulk sample.
-While both strategies provide a biologically meaningful imputation that will be more useful than post-hoc modes of imputation, we expect that a number of imputed sites were incorrectly classified as true positives with this germline ground truth.
-As this will lead to an overestimation of the recall, we have excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of this dataset---however, their results are nevertheless displayed in all respective figures.
-
-\begin{figure}[!tpb]
- \begin{minipage}[t]{.35\linewidth}
- \vspace{0pt}
-  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf} \newline
- \end{minipage}
- \begin{minipage}[t]{.64\linewidth}
- \vspace{0pt}
-  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf} \newline
- \end{minipage}
- \caption{
- Cell-specific $F_1$ scores for different filtering threshold levels for \textbf{A} the whole genome dataset and \textbf{B} the whole exome dataset.\newline \footnotesize
+ \label{fig:F_1-scores_Dong2017}
+\end{figure}
+
+\begin{figure}[!tpb]
+  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf}
+ \caption{
+ Cell-specific $F_1$ scores for different filtering threshold levels for the whole exome dataset.\newline \footnotesize
   Thresholds used were applied via the following command line options:
@@
  }
- \label{fig:F_1-scores}
+ \label{fig:F_1-scores_Laehnemann2017}
 \end{figure}
@@
 
-The per-cell $F_1$ score clearly demonstrates ProSolo's performance advantage (Figure~\ref{fig:F_1-scores}), even when restricting our comparison to ProSolo's (and SCIPhI's) default mode for a fairer comparison on the whole exome dataset with a germline ground truth (see caveat discussed above).
-
-\begin{figure}[!tpb]
- \begin{minipage}[t]{.4\linewidth}
+The per-cell $F_1$ score clearly demonstrates ProSolo's performance advantage (Figures~\ref{fig:F_1-scores_Dong2017} and \ref{fig:F_1-scores_Laehnemann2017}), even when restricting our comparison to ProSolo's (and SCIPhI's) default mode for a fairer comparison on the whole exome dataset with a germline ground truth (see caveat discussed above).
+
+\begin{figure}[!tpb]
+ \begin{minipage}{\linewidth}
+  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_per-cell_precision-recall-plot_focus-tools_normal.pdf} \newline
+ \end{minipage}
+ \caption{
+  Zoomed-in precision and recall per normal cell of the single nucleus exome sequencing dataset \cite{wang_clonal_2014}.
+  Note cells TNBC-n13 and TNBC-n14 as clear outliers.\newline \footnotesize
+  For the thresholds used, see caption of Figure~\ref{fig:alt-calling_prec-rec_global}.
+  }
+ \label{fig:wang_alt-calling_prec-rec_per-cell_normal}
+\end{figure}
+
+\begin{figure}[!tpb]
+ \begin{minipage}{\linewidth}
+  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_per-cell_precision-recall-plot_focus-tools_tumor.pdf} \newline
+ \end{minipage}
+ \caption{
+  Zoomed-in precision and recall per tumor cell of the single nucleus exome sequencing dataset \cite{wang_clonal_2014}.
+  Note cell h1 as a clear outlier.\newline \footnotesize
+  For the thresholds used, see caption of Figure~\ref{fig:alt-calling_prec-rec_global}.
+  }
+ \label{fig:wang_alt-calling_prec-rec_per-cell_tumor}
+\end{figure}
+
+For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
+On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
+On the normal cells, both ProSolo and SCIPhI achieve a precision above .99.
+Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a drastically precision.
+This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
+Where ProSolo models each cell and each genomic site separately, SCIPhI's model integrates information across all sites in all cells at once.
+While this can clearly help recover recall, this can also lead to false positive somatic variant calls (Figure~\ref{fig:somatic-recall}) and has clear implications for model complexity and software runtime.
+As all other tools, including ProSolo, are parallelizable over genomic regions and were thus able to process both datasets within days on a multicore machine, we do not report more detailed runtimes for them.
+However, it should be noted that SCIPhI took from 1 week (with iterations reduced below software defaults) on the single nucleus whole exome dataset up to 7.5 weeks on the whole genome dataset, running on a single core without any possibility of parallelization.
+And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, this will further increase SCIPhI's wall time.
+In addition, with more cells added, the space of possible tree topologies that SCIPhI explores grows exponentially, further increasing runtimes.
+
+In contrast to the whole genome dataset and the granulocyte whole exome dataset, this dataset showed more variation across cells.
+To document this, without further conflating the general precision-recall plots, we include per-cell precision-recall plots, here.
+This shows two outlier cells among the single normal cells, TNBC-n13 and TNBC-n14 (Figure~\ref{fig:wang_alt-calling_prec-rec_per-cell_normal}), and one drastic outlier among the single tumor cells, h1 (Figure~\ref{fig:wang_alt-calling_prec-rec_per-cell_normal}).
+With h1 at a clearly reduced overall coverage (Figure~\ref{fig:coverage-dists}C), these outliers---and further more subtle variation across cells---probably reflect variation in the data generation processes rather than biological differences in genomic variation across cells.
+
+\subsection{Estimating the bulk contribution to ProSolo results via downsampling} \label{sec:downsampling}
+
+To estimate the importance of bulk data in ProSolo's single nucleotide variant calling model, we performed a downsampling analysis with the whole exome sequencing dataset of five granulocytes.
+In addition to the full bulk coverage of around 40X at 80\% of exome sites (compare Figure~\ref{fig:coverage-dists}B), we downsampled the bulk sample to around 28X, 16X, 4X, and 0X (Figure~\ref{fig:downsampling}).
+From this analysis, we can see that even a very minimal coverage of 4X will result in very precise variation calls for any false discovery rate control at 0.1 or below.
+Only removing the bulk entirely (0X), will significantly impact ProSolo's precision.
+This is to be expected, as the model does not have any data available to compensate for the uncertainty of amplification bias that the empirical bias distributions account for in the single cell data.
+In line with this increased uncertainty in variation calls, false discovery rates below 0.2 will remove all calls (Figure~\ref{fig:downsampling}), while the recall remains very high in unfiltered calls.
+
+\begin{figure}[!tpb]
+  \includegraphics[width=.99\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-downsampling_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \newline
+ \caption{
+  Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth, including the results of ProSolo after downsampling from the maximum coverage of 50X to 35X, 20X, 5X or 0X.\newline
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline \footnotesize
+  Thresholds used were applied via the following command line options:
+  MonoVar {\ttfamily --t};
+  ProSolo {\ttfamily --fdr};
+  SCAN-SNV {\ttfamily --fdr};
+  SCcaller {\ttfamily -a cutoff};
+  SCIPhI {\ttfamily none available}.
+  Software modes:
+  MonoVar with consensus filtering ({\itshape default}) or without ({\itshape no consensus});
+  ProSolo with minimum coverage 1 in single cell ({\itshape default}), or imputing zero coverage sites based on bulk sample ({\itshape imputation});
+  SCcaller with recommended settings ({\itshape default}) or with a more {\itshape sensitive} calling;
+  SCIPhI with default parameters ({\itshape default}) or all heuristics off ({\itshape sensitive}).
+  }
+ \label{fig:downsampling}
+\end{figure}
+
+\subsection{Recall of validated somatic variants from \cite{wang_clonal_2014}} \label{sec:somatic-recall}
+
+\begin{figure}[!tpb]
+  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_validated_somatic_genotype_recall.pdf}
+ \caption{
+ Recall of validated non-synonymous somatic mutations in the single tumor cells of \cite{wang_clonal_2014} stratified across genotypes ("0/1" for heterozygous and "1/1" for homozygous alternative) and clonality (clonal vs.
+subclonal).
+ The expected number of variants with a somatic variant present in the respective number of single tumor cells is given in grey.
+ Clonal somatic variants are always expected to be present in all 16 single tumor cells.
+ }
+ \label{fig:somatic-recall}
+\end{figure}
+For the tumor cells from the \cite{wang_clonal_2014} dataset, we could further determine the recall of non-synonymous somatic variants they had validated via deep targeted duplex sequencing and provided in their Supplementary Tables 6 and 7 (specifying clonal and subclonal variants, respectively).
+Clonal variants were assumed to be present in all 16 tumor cells, with the number of single cells with the subclonal somatic variant present provided by \cite{wang_clonal_2014}.
+While SCIPhI and SCcaller have a better recall of heterozygous genotype calls, neither of them can call homozygous alternative genotypes (Figure~\ref{fig:somatic-recall}).
+On the clonal variants, MonoVar has a higher homozygous alternative recall than ProSolo, but ProSolo has a higher heterozygous recall.
+On the subclonal variants, two things are notable:
+\begin{enumerate}
+  \item ProSolo has lower recall than the other tools, indicating that the depth of coverage in the tumor bulk sample (around 35X for 80\% of the exome, see Figure~\ref{fig:coverage-dists}C) is not sufficient to pick up low-frequency variants and the bulk sample thus further decreases the likelihood of such low frequency variants in the single cells.
+  Thus, while the downsampling analysis above suggests that a very low bulk coverage is sufficient for ProSolo's model to perform well, this seems to be restricted to higher frequency mutations.
+  Assuming an evenly mixed cell population (thus ignoring any population structure or in turn assuming an even sampling of all relevant subclones), the bulk coverage of 35X should allow for the discovery of variant alleles present at a frequency of at least 2.86\% of sampled alleles.
+  Judging by the frequency detected in the targeted duplex sequencing of \cite{wang_clonal_2014}, this should make the detection of 52 of the 94 subclonal variants, which corresponds to the number of missed subclonal variants in Figure~\ref{fig:somatic-recall}.
+  As a consequence, if the aim is to detect lower frequency mutations, we simply recommend a deeper sequencing of the bulk sample.
+  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells, while the maximum frequency for any of the heterozygous subclonal mutation detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
+  This suggests that SCIPhI's model over-imputes mutations by aggregating information across cells.
+\end{enumerate}
+
+\subsection{False Discovery Rate Control of Alternative Allele Calling} \label{sec:fdr-of-alt-calling}
+
+\begin{figure}[!tpb]
+ \begin{minipage}[t]{.49\linewidth}
  \vspace{0pt}
-  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo_FDR_ground_truth_vs_theoretical.pdf} \newline
- \end{minipage}
- \begin{minipage}[t]{.59\linewidth}
+  \includegraphics[height=40ex]{figs/Dong2017/Dong2017_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf} \newline
+ \end{minipage}
+ \begin{minipage}[t]{.49\linewidth}
  \vspace{0pt}
-  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo__FDR_ground_truth_vs_theoretical.pdf} \newline
+  \includegraphics[height=40ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf} \newline
+ \end{minipage}
+ \begin{minipage}[t]{.99\linewidth}
+ \vspace{0pt}
+  \includegraphics[height=40ex]{figs/Wang2014/Wang2014_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf} \newline
  \end{minipage}
@@
 
-\subsection{False Discovery Rate Control of Alternative Allele Calling} \label{sec:fdr-of-alt-calling}
-
 Finally, a feature where ProSolo clearly stands out, is the control over the false discovery rate.
@@
 However, none of them provide control over similar ranges of precision and recall.
+In addition, we ran the standard false discovery rate control implemented in ProSolo on the posterior probabilities provided by SCIPhI, but this did not provide any substantial control over the false discovery rate (Figures~\ref{fig:alt-calling_prec-rec} and \ref{fig:FDR-ground-truth-vs-theoretical}).
 Thus, ProSolo is the only tool that provides the user with the choice of either aiming for more discoveries at the cost of a higher rate of false discoveries, or at aiming for a more limited number of discoveries with higher confidence in each of them.
@@
 This indicates, that ProSolo's model still needs refinement in its resolution for sites with a very high likelihood of an alternative allele present.
-A similar trend can be seen on the whole exome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B), but for any attempt to control for a false discovery rate below 0.02.
+The same trend can be seen on the other datasets (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B and C), but for any attempt to control for a false discovery rate below 0.02 or 0.005.
 This is within the common range of values often chosen as a false discovery rate to control for in practice and not being able to actually achieve the false discovery rate control at such values is usually unacceptable.
-However, as also discussed above, we generally expect the precision (of all tools) to be underestimated on this dataset: using the germline genotypes as ground truth results in misclassifying actual somatic variants in single cells as false positives.
+However, as also discussed above, we generally expect the precision (of all tools) to be underestimated on the whole exome granulocyte dataset: using the germline genotypes as ground truth results in misclassifying actual somatic variants in single cells as false positives.
 This underestimation of precision on this dataset in turn means an overestimation of the false discovery rate ($1 - p$recision), and is thus to be expected.
-This artefact of using a germline ground truth also shows in the discrepancy in the ground-truth based false discovery rate between ProSolo's default and imputation mode when controlling for higher false discovery rates.
+A similar effect is to be expected for the tumor cells in the TNBC dataset, as these will very likely contain more than the clonal somatic variants validated in the original publication \citep{wang_clonal_2014}.
+Further, the artefact of using a germline ground truth on the granulocyte dataset also shows in the discrepancy in the ground-truth based false discovery rate between ProSolo's default and imputation mode when controlling for higher false discovery rates.
 As expected, imputation turned out less precise in the whole genome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}A).
@@
 
-All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769116}).
+All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3824152}).
 An overview and citations of software and resources beyond those mentioned in the methods section are provided in Table~\ref{tab:software-overview}.

### 2021-01-03 manuscript/figs/Dong2017/Dong2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -61,13 +61,13 @@ for (cell in cells) {
 
 # Read in ProSolo stats
 
-fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
+fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "0-5", "1")
 modes <- c( "", ".min_sc_cov_1" )
 
 for (cell in cells) {
   for ( f in fdrs ) {
     for ( m in modes ) {
-      file <- str_c( "prosolo/fdr_alt-presence/", cell, ".Clones.alt_sites_only.fdr_", f, "_alt-presence.prosolo", m , ".positives_negatives.ML_set.tsv" )
+      file <- str_c( "prosolo/fdr_alt-presence/", cell, ".Clones.alt_sites_only.fdr_", f, "_alt-presence.prosolo", m , ".alt_calls.awk_positives_negatives.ML_set.tsv" )
       
       if ( m == ".min_sc_cov_1" ) {
         mode = "ProSolo default"
@@ -78,7 +78,8 @@ for (cell in cells) {
       current <- as_tibble( read_tsv(file) ) %>%
                   add_column( software = mode,
                               cell = cell,
-                              filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ) ) )
+                              filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ) )
+                  )
       
       positives_negatives <- bind_rows(positives_negatives, current)
     }
@@ -108,23 +109,27 @@ for (cell in cells) {
 # Read in SciPhi stats
 
 modes <- c("default", "sensitive")
+fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "0-5", "1")
 
 for (cell in cells) {
   for (m in modes) {
-    if (m == "default") {
-      file <- str_c("sciphi/", cell, ".single_cells.default.sciphi.positives_negatives.tsv")
-      mode = "SCIPhI default"
-    } else if (m == "sensitive") {
-      file <- str_c("sciphi/", cell, ".single_cells.400000_iterations.sensitive.sciphi.positives_negatives.tsv")
-      mode = "SCIPhI sensitive"
+    for ( f in fdrs ) {
+      if (m == "default") {
[…]

### 2021-01-03 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -5,7 +5,6 @@ cells <- c("PAG1", "PAG2", "PAG5", "PAG9", "PAG10")
 
 # Read in MonoVar stats
 
-# TODO: add "0-1" once computed
 ps <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
 cs <- c("0", "1") # indicator whether consensus filtering was on in MonoVar
 
@@ -61,7 +60,7 @@ for (cell in cells) {
 
 # Read in ProSolo stats
 
-fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
+fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "0-5", "1")
 modes <- c( "", ".min_sc_cov_1" )
 
 for (cell in cells) {
@@ -79,6 +78,34 @@ for (cell in cells) {
                   add_column( software = mode,
                               cell = cell,
                               filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ) ) )
+      
+      positives_negatives <- bind_rows(positives_negatives, current)
+    }
+  }
+}
+
+fdrs <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1", "0-2", "0-5", "1")
+downsampling <- c(".PNG", ".PNG_70-percent_seed-67", ".PNG_40-percent_seed-54", ".PNG_10-percent_seed-32", ".PNG_empty")
+
+for (cell in cells) {
+  for ( f in fdrs ) {
+    for (d in downsampling) {
+      if ( d == ".PNG_empty" & f %in% c("0-000001", "0-00001", "0-0001", "0-001", "0-005", "0-01", "0-02", "0-05", "0-1") ) {
+        next
+      }
+      file <- str_c( "prosolo/fdr_alt-presence/", cell, d, ".alt_sites_only.fdr_", f, "_alt-presence.prosolo.min_sc_cov_1.positives_negatives.ML_set.tsv" )
+      
+      current <- as_tibble( read_tsv(file) ) %>%
+                  add_column( software = case_when( 
+                                d == ".PNG" ~ "ProSolo def 40X bulk",
[…]

### 2021-01-03 manuscript/figs/Wang2014/Wang2014_plot_coverage_dist.R

@@ -1,0 +1,125 @@
+library(tidyverse)
+library('RColorBrewer')
+library('extrafont')
+loadfonts(device = "pdf")
+
+cells <- tibble(
+  name = c(
+    "a1",
+    "a2",
+    "a3",
+    "a4",
+    "a5",
+    "a6",
+    "a7",
+    "a8",
+    "h1",
+    "h2",
+    "h3",
+    "h4",
+    "h5",
+    "h6",
+    "h7",
+    "h8",
+    "TNBC-n10",
+    "TNBC-n11",
+    "TNBC-n12",
+    "TNBC-n13",
+    "TNBC-n14",
+    "TNBC-n15",
+    "TNBC-n16",
+    "TNBC-n1",
+    "TNBC-n2",
+    "TNBC-n3",
+    "TNBC-n4",
+    "TNBC-n5",
+    "TNBC-n6",
+    "TNBC-n7",
+    "TNBC-n8",
+    "TNBC-n9",
+    "TNBC-Pop-Tumor",
+    "TNBC-Pop-Normal"
+  ),
+  batch = c(
+    rep("tumor single cell", times = 16),
+    rep("normal single cell", times = 16),
+    rep("bulk", times = 2)
+  )
+)
+
+# Read in data
+
+current <- tibble()
+coverage_dists <- tibble()
+for (sample in cells$name) {
+  file <- str_c( sample, ".bps.TruSeq_exome_targeted_regions.mosdepth.region.dist.txt")
+      
+  batch = cells %>% filter(name == sample) %>% pull(batch)
+  current <- as_tibble(
+      read_tsv(
+        file,
+        col_names = c("region", "coverage", "fraction"),
+        col_types = "cid"
+      )
+    ) %>%
+    add_column(
+      sample = sample,
+      batch = batch
+    ) %>%
+    mutate(
+      sample = factor(sample, levels = cells$name, ordered = TRUE)
+    )
+  
+  coverage_dists <- bind_rows(coverage_dists, current)
+}
+
+max_power <- 2
+powers <- 10**(0:max_power)
+majors <- c(1,3,5)
+minors <- 1:9
+major_ticks <- rep(majors, times = max_power+1)*rep(powers, each = length(majors))
+minor_ticks <- rep(minors, times = max_power+1)*rep(powers, each = length(minors))
+
+# we need 32 somewhat distinguishable colors, so we go for four 9-value
+# sequential palettes without their lightest color -- palettes are only
+# colorblind-safe within themselves, but together with linetypes this
[…]

### 2021-01-03 manuscript/figs/Wang2014/Wang2014_plot_prosolo-monovar-scansnv-sccaller-scvilp.precision-recall.R

@@ -1,0 +1,443 @@
+# packages, colors, palettes and stuff
+source("../common.R")
+
+cells <- tibble(
+  name = c(
+    "a1",
+    "a2",
+    "a3",
+    "a4",
+    "a5",
+    "a6",
+    "a7",
+    "a8",
+    "h1",
+    "h2",
+    "h3",
+    "h4",
+    "h5",
+    "h6",
+    "h7",
+    "h8",
+    "TNBC-n10",
+    "TNBC-n11",
+    "TNBC-n12",
+    "TNBC-n13",
+    "TNBC-n14",
+    "TNBC-n15",
+    "TNBC-n16",
+    "TNBC-n1",
+    "TNBC-n2",
+    "TNBC-n3",
+    "TNBC-n4",
+    "TNBC-n5",
+    "TNBC-n6",
+    "TNBC-n7",
+    "TNBC-n8",
+    "TNBC-n9"
+  ),
+  batch = c(
+    rep("tumor", times = 16),
+    rep("normal", times = 16)
+  )
+)
+
+get_cells_batch <- function(cells, cell) {
+  cells %>% filter(name == cell) %>% pull(batch)
+}
+
+get_batch_ground_truth <- function(batch) {
+  if (batch == "normal") {
+    ground_truth = ".tumor_ground_truth."
+  } else if (batch == "tumor") {
+    ground_truth = ".normal_ground_truth."
+  } else {
+    stop("Unknown batch type.")
+  }
+  ground_truth
+}
+
+# Read in MonoVar stats
+
+# TODO: add "0-1" once computed
+ps <- c( "0-000001", "0-00001", "0-0001", "0-001", "0-01", "0-02", "0-05", "0-1", "0-2", "1")
+cs <- c("0", "1") # indicator whether consensus filtering was on in MonoVar
+
+current <- tibble()
+positives_negatives <- tibble()
+for (cell in cells[["name"]]) {
+  for ( c in cs ) {
+    for ( p in ps ) {
+      batch <- get_cells_batch(cells, cell)
+      ground_truth = get_batch_ground_truth(batch)
+      file <- str_c("monovar/", cell, ".", batch, "_cells", ground_truth, p, ".c", c, ".monovar.alt_calls.awk_positives_negatives.tsv")
+      
+      if ( c == 1 ) {
+        mode = "MonoVar default"
+      } else {
+        mode  = "MonoVar no consensus"
+      }
+       
+      current <- as_tibble( read_tsv(file) ) %>%
+                  add_column( software = mode,
[…]

### 2021-01-03 manuscript/figs/Wang2014/Wang2014_plot_prosolo-monovar-sccaller-sciphi_som_gt_recall.R

@@ -1,0 +1,202 @@
+# packages, colors, palettes and stuff
+source("../common.R")
+
+tumor_cells <- c(
+    "a1",
+    "a2",
+    "a3",
+    "a4",
+    "a5",
+    "a6",
+    "a7",
+    "a8",
+    "h1",
+    "h2",
+    "h3",
+    "h4",
+    "h5",
+    "h6",
+    "h7",
+    "h8"
+  )
+
+fdr <- c( "1" )
+
+current <- tibble()
+recall_per_cell <- tibble()
+
+# read in ProSolo data
+for ( sc in tumor_cells ) {
+  for (f in fdr ) {
+    file <- str_c(
+      "prosolo/fdr_alt-presence/som_gt_recall/",
+      sc,
+      ".TNBC-Pop-Tumor.alt_sites_only.fdr_",
+      f,
+      "_alt-presence.prosolo.min_sc_cov_1.gt_calls.ML_set.somatic_genotype_recall.tsv"
+    )
+    current <- as_tibble(
+        read_tsv(
+          file,
+          col_types = "cicccccc"
+          )
+      ) %>%
+      add_column(
+        cell = sc, 
+        software = "ProSolo default",
+        filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ))
+      )
+    recall_per_cell <- bind_rows(recall_per_cell,current)
+  }
+}
+
+# read in SCIPhI data
+for ( sc in tumor_cells ) {
+  for (f in fdr ) {
+    file <- str_c(
+      "sciphi/fdr_alt-presence/som_gt_recall/",
+      sc,
+      ".tumor_cells.80000-40000_iterations.default.sciphi.alt_prob.fdr_",
+      f,
+      "_alt-presence.gt_calls.ML_set.somatic_genotype_recall.tsv"
+    )
+    current <- as_tibble(
+        read_tsv(
+          file,
+          col_types = "cicccccc"
+          )
+      ) %>%
+      add_column(
+        cell = sc, 
+        software = "SCIPhI default",
+        filter = if_else(f == "1", 1.0, as.numeric( str_replace(f, '-', '.') ))
+      )
+    recall_per_cell <- bind_rows(recall_per_cell,current)
+  }
+}
+
+# Read in Monovar data
+for ( sc in tumor_cells ) {
+  for (f in fdr ) {
+    file <- str_c(
+      "monovar/som_gt_recall/",
+      sc,
+      ".tumor_cells.",
+      f,
[…]

### 2021-01-03 manuscript/figs/common.R

@@ -8,12 +8,31 @@ linetype_pal = c(
   "MonoVar default" = 1,
   "MonoVar no consensus" = 3,
   "ProSolo default" = 1,
+  "ProSolo def 40X bulk" = 1,
+  "ProSolo def 28X bulk" = 5,
+  "ProSolo def 16X bulk" = 6,
+  "ProSolo def  4X bulk" = 2,
+  "ProSolo def  0X bulk" = 4,
   "ProSolo imputation" = 3,
   "SCAN-SNV sensitive" = 3,
   "SCcaller default dbsnp" = 1,
   "SCcaller sensitive dbsnp" = 3,
   "SCIPhI default" = 1,
-  "SCIPhI sensitive" = 3
+  "SCIPhI sensitive" = 3,
+  "scVILP default" = 1,
+  "scVILP sensitive" = 3
+  )
+
+basic_software_levels = c(
+  "MonoVar default",
+  "MonoVar no consensus",
+  "ProSolo default",
+  "ProSolo imputation",
+  "SCAN-SNV sensitive",
+  "SCcaller default dbsnp",
+  "SCcaller sensitive dbsnp",
+  "SCIPhI default",
+  "SCIPhI sensitive"
   )
 
 # Based on Okabe and Ito colorblind-safe palette (https://jfly.uni-koeln.de/color/)
@@ -32,6 +51,7 @@ factor <- 1/3
 
 Black <- colorblind_pal()(8)[[1]]
 light_Black <- tint(Black, factor)
+very_light_Black <- tint(Black, 3/4)
 
 Vermillion_Red <- colorblind_pal()(8)[[7]]
 light_Vermillion_Red <- tint(Vermillion_Red, factor)
@@ -71,27 +91,43 @@ Yellow <- colorblind_pal()(8)[[5]]
 # cvd_grid(bp)
 
 software_pal <- c(
+  "expected" = very_light_Black,
   "MonoVar default" = Black,
   "MonoVar no consensus" = light_Black,
   "ProSolo default" = Vermillion_Red,
+  "ProSolo def 40X bulk" = Vermillion_Red,
+  "ProSolo def 28X bulk" = Vermillion_Red,
+  "ProSolo def 16X bulk" = Vermillion_Red,
+  "ProSolo def  4X bulk" = Vermillion_Red,
+  "ProSolo def  0X bulk" = Vermillion_Red,
   "ProSolo imputation" = light_Vermillion_Red,
   "SCAN-SNV sensitive" = Yellow,
[…]

### 2021-02-02 manuscript/main.tex
last updates to text, e.g. citing the general Zenodo DOI that points to latest pipeline and results data version

@@
 \newcommand{\cB}{{\mathbf{B}}}
-
 \usepackage{euscript}
@@
 Interestingly, SCIPhI\cite{singer_single-cell_2018} additionally models allelic amplification bias to be governed by one global beta-binomial distribution (to apply for all cells in a dataset), thereby accounting for allelic dropout a second time (as the extreme values at 0 and 1 of that distribution).
-Two tools that more variably model alleleic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
+Two tools that more variably model allelic amplification bias are SCcaller \cite{dong_accurate_2017} and SCAN-SNV\cite{luquette_identification_2019}.
 Both estimate the minor allele frequency from nearby germline heterozygous sites. 
@@
 
+\begin{table}[tbp]
+ \caption{Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
+Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.}
+ \label{tab:Lodato}
+ \renewcommand{\arraystretch}{1.15}
+
+ \begin{center}
+  \begin{tabular}{lrr}
+   parameter  & slope        & intersect   \\
+   \toprule
+   $\alpha$   & -0.000027183 & 0.068567471 \\
+   $\beta$    & 0.007454388  & 2.367486659 \\
+   \midrule
+   $w$        & 0.000548761  & 0.540396786 \\
+   $\alpha_1$ & 0.057378844  & 0.669733191 \\
+   $\alpha_2$ & 0.003233912  & 0.399261625 \\
+   \bottomrule
+  \end{tabular}
+ \end{center}
+\end{table}
+
 \paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
@@
 With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
-Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
+Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
 However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:
@@
 For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
-All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3824152}).
+All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).
 
@@
 \paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
+\begin{sloppypar}
 For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
@@
 From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.
+\end{sloppypar}
 
@@
 As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations.
-Note that this experimental setup accounts for utmost fairness for all competitors.
+This experimental setup aims for fairness across all competitors. 
 Variant calling for the ground truths was performed as for cell line data above.
@@
   Precision-recall plots for alternative allele calls of ProSolo, MonoVar, SCAN-SNV, SCcaller and SCIPhI.
-  Note that all panels are strong zoom-ins, focusing on (different) areas of interest.
+  All panels are strong zoom-ins, focusing on (different) areas of interest.
   Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
@@
 While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
-We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, \ref{fig:F_1-scores}B and \ref{fig:FDR-ground-truth-vs-theoretical}B for reference).
+We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:F_1-scores_Laehnemann2017} for reference).
 
@@
 Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
-While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a drastically reduced precision.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at reduced precision.
 This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
@@
 And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, both will further increase SCIPhI's wall time.
-In addition, adding more cells exponentially grows the space of possible tree topologies that SCIPhI explores\cite{lahnemann_eleven_2020}, which will further increase its runtime.
+In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.
 
@@
 
-ProSolo is the first method for SNV calling from MDA single cell sequencing data to comprehensively model both amplification bias and amplification errors in a way that enables both
-\begin{enumerate*}
- \item the integration of site-specific differences (Figure \ref{fig:prosolo_alt-calling}), and
- \item a computationally efficient---hence practically scalable---evaluation of all relevant data properties.
-\end{enumerate*}
+ProSolo is the first method for SNV calling from MDA single cell sequencing data to comprehensively model both amplification bias and amplification errors in a way that enables both the integration of site-specific differences (Figure \ref{fig:prosolo_alt-calling}), and a computationally efficient---hence practically scalable---evaluation of all relevant data properties.
 We achieve this by combining a data-driven model of amplification errors (that incorporates a bulk background sample alongside the single cell sample) with an empirical model of amplification bias which is dependent upon a site's coverage (and is based on a statistical understanding of the MDA process that is mechanistically motivated).

### 2021-02-02 manuscript/supplement.tex
last updates to text, e.g. citing the general Zenodo DOI that points to latest pipeline and results data version

@@
 In addition, the exact distribution used could be further refined with the below considerations regarding the suitability of beta-binomial distributions in this case.\\
-
-\begin{table}[tbp]
- \caption{Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
-Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.}
- \label{tab:Lodato}
- \renewcommand{\arraystretch}{1.15}
-
- \begin{center}
-  \begin{tabular}{lrr}
-   parameter  & slope        & intersect   \\
-   \toprule
-   $\alpha$   & -0.000027183 & 0.068567471 \\
-   $\beta$    & 0.007454388  & 2.367486659 \\
-   \midrule
-   $w$        & 0.000548761  & 0.540396786 \\
-   $\alpha_1$ & 0.057378844  & 0.669733191 \\
-   $\alpha_2$ & 0.003233912  & 0.399261625 \\
-   \bottomrule
-  \end{tabular}
- \end{center}
-\end{table}
 
@@
 
-All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3824152}).
+All analysis code for both the main processing pipeline and the manuscript figures is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).
 An overview and citations of software and resources beyond those mentioned in the methods section are provided in Table~\ref{tab:software-overview}.
@@
     bcftools            & 1.6       & handle VCF / BCF files        & \citep{danecek_variant_2011} \\
+                        & 1.10      & some rule-specific environments & \citep{danecek_variant_2011} \\
     \midrule
@@
     \midrule
-    ggplot2             & 3.2.1     & result plots                  & \citep{wickham_ggplot2_2016} \\
+    ggplot2             & 3.3.0     & result plots                  & \citep{wickham_ggplot2_2016} \\
     \midrule

### 2021-02-05 manuscript/supplement.tex
add FACS sorting strategy figure to supplement

@@
  }
- \label{fig:coverage-dists}
 \end{figure}
@@
 \textbf{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
-For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figure~\ref{fig:datasets}B).
+For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figures~\ref{fig:datasets}B and \ref{fig:FACS-scheme}).
 Single cells from that cell population were isolated using a Fluidigm C1 microfluidics chip and subjected to multiple displacement amplification (MDA) using the GE Healthcare GenomiPhi Kit in its version adapted to that microfluidic chip.
@@
 From the remaining sorted cell population, we extracted bulk DNA and submitted it to the same whole exome capture and DNA sequencing without MDA.\\
+
+\begin{figure}[!tpb]
+ \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_Granulozyten_P.pdf}
+ \caption{
+ 59 CD66b+ granulocytes were sorted out from a total population of 35779 CD3-T cell depleted peripheral blood cells (gate C in the right-most panel) and from these, five single cells were chosen where whole genome amplification had yielded DNA for at least 15 out of 16 control loci.
+ }
+ \label{fig:FACS-scheme}
+\end{figure}
 
@@
     bcftools            & 1.6       & handle VCF / BCF files        & \citep{danecek_variant_2011} \\
-                        & 1.10      & some rule-specific environments & \citep{danecek_variant_2011} \\
+                        & 1.10      & rule-specific environment     &  \\
     \midrule

### 2021-06-29 manuscript/main.tex
update manuscript to 2nd revision

@@
 The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
-ProSolo models single cells jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered polymorphic sites.
-As a consequence, ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
-Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to identify, such as allele dropout.
+ProSolo models a single cell jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered sites, when downstream tools cannot handle missing data.
+ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
+Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to quantify, such as allele dropout.
 
@@
 However, SCcaller employs a fixed global false positive error rate for the calling of alternative alleles\cite{dong_accurate_2017}, and SCAN-SNV makes use of heuristics for filtering candidate variants\cite{luquette_identification_2019}.
-Thus, to the best of our knowledge, there is no statistical model that allows for local variation of bias and errors due to amplification, and for statistically sound false discovery control when calling and genotyping SNVs in single cells.
+Thus, to the best of our knowledge, there is no statistical model that allows for both local variation of bias and errors due to amplification, and for statistically sound false discovery rate control when calling and genotyping SNVs in single cells.
 
@@
 \begin{enumerate*}[label={(\roman*)}]
-  \item The theoretical underlying allele frequency at a site in a single cell: $\theta_s$.
+  \item The true (but usually unknown) underlying allele frequency at a site in a single cell: $\theta_s$.
    This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
@@
 \end{enumerate*}
+
+The goal is to estimate the likelihood density across the three possible underlying allele frequencies in the single cell (we denote $\tilde{\theta}_s$ as the density estimate across all $\theta_s\in\{0.0,0.5,1.0\}$).
 To accurately quantify the uncertainty introduced by amplification bias, we consider
@@
 \begin{table}[tbp]
- \caption{Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
-Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.}
+ \caption{
+ Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
+ The full precision as used in the implementation is given in Supplementary Table~\ref{tab:Lodato-full}
+ }
  \label{tab:Lodato}
@@
    \toprule
-   $\alpha$   & -0.000027183 & 0.068567471 \\
-   $\beta$    & 0.007454388  & 2.367486659 \\
+   $\alpha$   & -0.00003 & 0.06857 \\
+   $\beta$    & 0.00745  & 2.36749 \\
    \midrule
-   $w$        & 0.000548761  & 0.540396786 \\
-   $\alpha_1$ & 0.057378844  & 0.669733191 \\
-   $\alpha_2$ & 0.003233912  & 0.399261625 \\
+   $w$        & 0.00055  & 0.54040 \\
+   $\alpha_1$ & 0.05738  & 0.66973 \\
+   $\alpha_2$ & 0.00323  & 0.39926 \\
    \bottomrule
@@
 Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
-This establishes a formal, statistical argument to why one should consider bulk experiments in single cell sequencing whenever possible and why the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
+This establishes a formal, statistical argument why one should consider bulk experiments in single cell sequencing whenever possible: the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell, in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
+At the same time, the absence of a single cell candidate mutation in a bulk sample (with sufficient sequencing depth) increases the probability of an amplification error.
 As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.
@@
 Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.
+Further, as for the single cell sample, when referring to the likelihood density estimates across all possible allele frequencies (as opposed to a particular allele frequency), we denote this with $\tilde{\theta}_b$.
 
@@
 With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
-Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across allele frequency $\tilde{\theta}_s,\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
+Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across all possible allele frequencies $\tilde{\theta}_s$ and $\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
 However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:
@@
 
-For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel).
+For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel; for discussion on the validity of this assumption, see Supplementary Section \ref{sec:event-space-def}).
 At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
@@
 Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
-For example, an error-free homozygous alternative site is defined by allele frequency likelihood estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
-And the dropout of an alternative allele is defined across allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
+For example, an error-free homozygous alternative site is defined by allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
+And the dropout of an alternative allele is defined across allele frequency likelihood density estimates $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
 For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
@@
 Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
-However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will usually favor germline genotypes over any existing (lower frequency) somatic genotypes at a site.
+However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will favor germline genotypes over any existing (lower frequency) somatic genotypes at a site, unless such a somatic genotype is present in a majority of cells.
+Thus, such an imputation carries the potential to introduce erroneous calls (especially when looking at subclonal somatic mutations), and we recommend to instead use downstream tools that can accommodate for missing data and data uncertainty wherever they are available.
 
@@
 We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}.
-We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 5.4.0}) to implement the benchmarking workflows.
+We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 6.3.0}) to implement the benchmarking workflows.
 For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
@@
 \paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
-The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
+The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from SRA project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
 A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
@@
 \newline \footnotesize
-  Threshold parameters:
+  Threshold parameters (not comparable across tools):
   MonoVar {\ttfamily --t};
@@
 While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
-We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:F_1-scores_Laehnemann2017} for reference).
+We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:per-cell_prec-rec_Laehnemann2017} for reference).
 
@@
 And where adding breadth of coverage (i.e. more genomic sites) or more cells simply means adding more parallel processes in most tools, both will further increase SCIPhI's wall time.
-In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.
+In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores super-exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.
 
@@
 Importantly, the calculation of posterior probabilities scales linearly in the coverage of variant sites, which is the fastest theoretically conceivable, documented by runtimes that have substantial advantages over the state of the art in practice.
-Using one whole genome dataset and two whole exome datasets---each with a different type of ground truth (Figure~\ref{fig:datasets})---we demonstrate that these two innovations of ProSolo combined to achieve competing or better accuracy (Figure~\ref{fig:alt-calling_prec-rec}).
+Using one whole genome dataset and two whole exome datasets---each with a different type of ground truth (Figure~\ref{fig:datasets})---we demonstrate that these two innovations of ProSolo combine to achieve competing or better accuracy (Figure~\ref{fig:alt-calling_prec-rec}).
 Moreover, the model allows to accurately control the false discovery rate of variant calls, another novelty of great practical value, and both the model and its modular implementation are easy to adapt and extend.
@@
 The only exception is the normal cells dataset from the single nucleus exome sequencing data, for which SCIPhI attains a higher recall above .99 precision---but at the cost of false positive calls of subclonal heterozygous variants (Figure~\ref{fig:somatic-recall}), while lacking the ability to call sites with a homozygous alternative genotype (Figure~\ref{fig:somatic-recall}).
-Interestingly, a bulk coverage as low as 4X already provides drastic performance improvements in ProSolo's model and allows for a meaningful control over the false discovery rate (Supplementary Figure~\ref{fig:downsampling}, but higher bulk coverage will be necessary to detect low-frequency somatic variants (Supplementary Section~\ref{sec:somatic-recall}).
+Interestingly, a bulk coverage as low as 4X already provides drastic performance improvements in ProSolo's model and allows for a meaningful control over the false discovery rate (Supplementary Figure~\ref{fig:downsampling}; but higher bulk coverage will be necessary to detect low-frequency somatic variants, Supplementary Section~\ref{sec:somatic-recall}).
 To summarize our benchmarking of alternative allele calling: SCcaller seems optimized for precision, MonoVar achieved higher recall than SCcaller, and SCAN-SNV does not seem suitable for general variant calling on MDA single cell sequencing data, but is only applicable when restricting interest to somatic variants.
@@
 This demonstrates, that ProSolo's model captures data from individual single cells better than any of the existing models.
-Only in the dataset with more cells and a bulk sample with a lower coverage, SCIPhI can reach a higher recall than ProSolo by integrating information across cells (but at the cost false positive somatic mutation calls).
+Only in the dataset with more cells and a bulk sample with a lower coverage, SCIPhI can reach a higher recall than ProSolo by integrating information across cells (but at the cost of false positive somatic mutation calls).
 However, with default configuration SCIPhI takes weeks on a single core without any possibility of parallelization, compared to a runtime of only days on a cluster for all the other tools.
 This is the result of the general approach, integrating information across cells and across genomic sites all at once.
-Thus, any addition of genomic coverage and cells will further increase SCIPhI's wall clock runtime, where adding cells could prove especially troubling: adding cells will exponentially grow the space of possible tree topologies to explore.
+Thus, any addition of genomic coverage and cells will further increase SCIPhI's wall clock runtime, where adding cells could prove especially troubling: adding cells will super-exponentially grow the space of possible tree topologies to explore.
 Other tools, including ProSolo, can simply add more parallel processes per site and cell and can thus leverage cluster computing for larger datasets, something that becomes more urgent with the advent of scalable droplet-based single cell whole genome amplification methods\cite{lahnemann_eleven_2020}.
@@
 If genotype profiles without missing values are for example required by downstream tools, ProSolo can provide a cell-population-specific imputation instead of resorting to a majority assignment (limited by the number of single cells sequenced) or even imputation based on external databases (e.g.~dbsnp).
-However, any imputation will usually favour germline genotypes over low-frequency somatic genotypes.
+However, any imputation will usually favour germline genotypes over low-frequency somatic genotypes (even though the opposite case also happens for SCIPhI, where some heterozygous subclonal variants are called in a lot more cells than the ground truth suggests, Figure~\ref{fig:somatic-recall}).
 Thus, we suggest to not impute zero-coverage single cell sites whenever possible and instead recommend using and developing downstream software that can deal with both these missing values and the event probabilities that ProSolo provides.
@@
 We anticipate that such fine-grained site-specific probabilities---and the uncertainties they capture---will be informative for improving probabilistic modelling in downstream analyses.
-They could e.g.~prove useful in models for phylogenetic reconstruction of the lineage relationship of sequenced single cells\cite{singer_single-cell_2018,zafar_siclonefit:_2018,koptagel_scuphr:_2018}, thus achieving results similar to or even better than SCIPhI, while keeping the analyses modular and thus computationally tractable\cite{lahnemann_eleven_2020}.
+They could e.g.~prove useful in models for phylogenetic reconstruction of the lineage relationship of sequenced single cells\cite{singer_single-cell_2018,zafar_siclonefit:_2018,koptagel_scuphr:_2018}, thus achieving reconstruction results similar to or better than SCIPhI, while keeping the analyses modular and thus computationally tractable\cite{lahnemann_eleven_2020}.
 
@@
 However, the allele dropout rate estimations were slightly off in different directions for the different benchmarking datasets, suggesting that the empirical distributions we currently make use of (based on Lodato et al.\cite{lodato_somatic_2015}) may not suit all datasets.
-For example, whole genome amplification may introduce further dataset-specific variability so far not captured by our model.
+For example, the degree to which whole genome amplification introduces variability may be dataset-specific, something not currently captured in our model.
 This conclusion is further supported by the naively calculated allele dropout rate, which is much lower in the high coverage cells of the two datasets (IL-11 for the Dong et al.
@@
 If samples PAG1 and PAG10 were doublets as well, this would indicate that our use of empirical distributions in ProSolo provides for more robust event probabilities in the presence of doublets, while heuristic thresholding (as in our naive allele dropout estimate) is very sensitive to such perturbations.
-In general, the use of a fixed empirical model for MDA allelic bias does not seem to impede ProSolo's performance in alternative allele calling compared to the other tools, but has a noticeable effect on genotyping (Supplement, Section~\ref{sec:suppl-gt-calling} and Figure~\ref{fig:somatic-recall}) and might be responsible for imprecisions when controlling for very small false discovery rates (Supplement Section~\ref{sec:fdr-of-alt-calling}).
+
+In general, the use of a fixed empirical model for MDA allelic bias does not seem to impede ProSolo's performance in alternative allele calling compared to the other tools, but has a noticeable effect on genotyping (Supplement, Section~\ref{sec:suppl-gt-calling} and Figure~\ref{fig:somatic-recall}) and might be responsible for imprecisions when controlling for very small false discovery rates (Supplement, Section~\ref{sec:fdr-of-alt-calling}).
 When future datasets are generated based on improved MDA protocols\cite{lahnemann_eleven_2020}, these effects might be exacerbated.
@@
 For example, the parameters of the mechanistically motivated combination of beta-binomial distributions for modelling heterozygous genotypes could be learned per single cell sample at germline heterozygous sites---similar to the approaches of SCcaller and SCAN-SNV, but globally per cell with their local variation modelled by their dependence on a site's coverage.
+
+In addition, the current model prevents the calling of subclonal somatic mutations with a bulk sample that has insufficient coverage to sample the respective allele, which can easily be remedied by sequencing the bulk sample to a greater depth.
+In addition, including a prior for the somatic mutation rate in the bulk sample further improve recall.
+Newer versions of the Varlociraptor library\cite{koster_varlociraptor_2020} already allow using such a prior, so this is immediate future work.
+When bulk coverage is low and an alternative allele is not sampled by any read, the prior will result in non-zero alternative allele frequencies in the bulk being assigned non-zero likelihoods, which is the desired behaviour for this constellation.
+When bulk coverage is high, the increased amount of data will progressively overrule the prior.
+
 Further room for improvement also remains for the modeling of the homozygous genotypes: while the current distributions account for both amplification and sequencing errors, sequencing errors are already safely accounted for elsewhere in our latent variable model.
-An amplification error profile that is not compounded with sequencing errors, would need to be based on the $\Phi$29 polymerase error rate and a better understanding of the statistical distribution that these errors generate (Supplement, Section~\ref{sec:better_beta-binom}).
-Studying and implementing the corresponding changes in the future has the potential to further improve the accurate site-specific event probabilities that ProSolo already provides through the joint modelling with a bulk background sample.
+Beyond low coverage in the bulk background sample, this might be another reason why some subclonal somatic heterozygous variants are mis-classified as homozygous by ProSolo (Supplement, Figure~\ref{fig:somatic-recall}).
+Here, an amplification error profile that is not compounded with sequencing errors would need to be based on the $\Phi$29 polymerase error rate and a better understanding of the statistical distribution that these errors generate (Supplement, Section~\ref{sec:better_beta-binom}).
+Studying and implementing the corresponding changes in the future has the potential to further improve the accurate site-specific event probabilities that ProSolo already provides through the joint modelling with a (sufficiently deep) bulk background sample.
 
-Finally, the modular implementation of ProSolo within the context of the Varlociraptor library\cite{koster_varlociraptor_2020} facilitates the implementation of further features, such as the ones mentioned above, read-backed phasing\cite{bohrson_linked-read_2019,hard_conbase:_2019} or even more variable models of amplification bias\cite{luquette_identification_2019} that could be integrated to exhaustively leverage MDA data information content.
-Since the Varlociraptor library also provides advanced functionality for the calling of insertions, deletions and multiple nucleotide variants (MNVs), one of the immediately following next steps will be to adapt ProSolo to calling those in single cells.\\
+Finally, the modular implementation of ProSolo within the context of the Varlociraptor library\cite{koster_varlociraptor_2020} facilitates the implementation of further features, such as read-backed phasing\cite{bohrson_linked-read_2019,hard_conbase:_2019} or even more variable models of amplification bias\cite{luquette_identification_2019} that could be integrated to exhaustively leverage MDA data information content.
+While no standardized software is available, it seems to be possible to obtain copy number profiles from single cells \cite{baslan_genome-wide_2012,satas_scarlet_2020}.
+Extending our model to make use of such profiles, it should also be possible to salvage cancer-related variant cases that our current model setup does not capture (Supplement, Section~\ref{sec:event-space-def}).
+And since the Varlociraptor library also provides advanced functionality for the calling of insertions, deletions and multiple nucleotide variants (MNVs), one of the immediately following next steps will be to adapt ProSolo to calling those in single cells.\\
 

### 2021-06-29 manuscript/supplement.tex
update manuscript to 2nd revision

@@
 \end{equation}
-as the (true, but usually unknown) frequency of alternative nucleotide $x$ in the bulk and single cell sample, respectively.
+as the true (but usually unknown) frequency of alternative nucleotide $x$ in the bulk and single cell sample, respectively.
 These frequencies translate into probabilities of sampling a DNA fragment containing the respective nucleotide at a site:
@@
 or indirectly, as with the sequencing of whole genome amplified DNA of a single cell $s$.
-It is our eventual goal to work back from bulk and single cell data to derive likelihoods for possible values of $\theta_{b}(x)$ and $\theta_{s}(x)$, respectively, allowing us to place likelihoods on possible underlying genotypes and events like allele dropouts in single cells.\\
-
-For a single diploid cell $s$, the possible alternative nucleotide frequencies are
+It is our eventual goal to work back from bulk and single cell data to derive likelihoods for all possible values of $\theta_{b}(x)$ (denoted as $\tilde{\theta_b}(x)$)\footnote{We use $\theta$ notation for the true underlying alternate allele frequency at a site, and $\tilde{\theta}$ notation when referring to the likelihood density estimates derived for all possible underlying alternate allele frequencies.
+The latter notation is thus mostly used for the definition of single cell events in section \ref{sec:event-space-def}.} and $\theta_{s}(x)$ (denoted as $\tilde{\theta_s}(x)$), respectively, allowing us to place likelihoods on possible underlying genotypes and events like allele dropouts in single cells.\\
+
+For a single diploid cell $s$, the true alternative nucleotide frequency can be
 \begin{equation}
@@
 
+\begin{table}[tbp]
+ \caption{
+ Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
+ Each of these scales linearly up to a site coverage of 60, with the slopes and intercepts as given here.
+ }
+ \label{tab:Lodato-full}
+ \renewcommand{\arraystretch}{1.15}
+
+ \begin{center}
+  \begin{tabular}{lrr}
+   parameter  & slope        & intersect   \\
+   \toprule
+   $\alpha$   & -0.000027183 & 0.068567471 \\
+   $\beta$    & 0.007454388  & 2.367486659 \\
+   \midrule
+   $w$        & 0.000548761  & 0.540396786 \\
+   $\alpha_1$ & 0.057378844  & 0.669733191 \\
+   $\alpha_2$ & 0.003233912  & 0.399261625 \\
+   \bottomrule
+  \end{tabular}
+ \end{center}
+\end{table}
+
 With the resulting likelihoods for $\rho_s \in [0,1]$, we can calculate the likelihood of a particular $\theta_s$:
@@
 
-For the current implementation, we use the empirical values for $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ as deduced by \cite{lodato_somatic_2015} (Table~\ref{tab:Lodato}).
+For the current implementation, we use the empirical values for $\alpha(l), \beta(l), w(l), \alpha_1(l), \alpha_2(l)$ as deduced by \cite{lodato_somatic_2015} (Table~\ref{tab:Lodato-full}).
 But as mentioned above, these could be learned from a dataset at hand whenever high-confidence sets of heterozygous and homozygous sites are available (e.g.~by calling variants on a background bulk and cross-referencing with dbSNP).\\
@@
 
-Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell and the bulk sequencing sample, given the respective data.
+\subsubsection{Defining the Event Space}\label{sec:event-space-def}
+
+Using the above model for a genome position $g$, we can compute the likelihood of any combination of possible underlying alternative nucleotide frequencies in the single cell ($\tilde{\theta_s}$) and the bulk sequencing sample ($\tilde{\theta_b}$), given the respective data.
 All the possible combinations of allele frequencies in the two samples define a two-dimensional space of allele frequency likelihood density estimates $\boldsymbol{E}$ (Equation~\ref{eq:event-space}, Figure~\ref{fig:event-space}, Table~\ref{tab:Events}).
@@
 Based on our assumption, we decide to trust the bulk sample over the single cell sample, meaning that the bulk is a mixture of a maximum of two sub-populations, and can only contain heterozygous and homozygous alternative cells, rendering a homozygous reference single cell impossible and thus the result of an allele dropout.
+
+To evaluate whether this assumption of only two sub-populations in the bulk is safe to make, we can examine all the possible ways in which this assumption may be violated.
+This requires that we have a sequence of two mutational events affecting the same genomic site, each generating a new sub-population.
+To know which types of mutational event sequences to consider, we note that the same point mutation has been observed to occur independently in different single cells from the same individual \citep{kuipers_single-cell_2017}, possibly due to selection for an advantageous effect of that mutation.
+However, the net effect of such parallel evolution still restricts the cell bulk to two sub-populations with distinct genotypes\footnote{As we are looking at different single cells separately, this also means that we do not have to worry about this kind of violation of the infinite sites assumption.
+Such parallel evolution only becomes problematic for phylogenetic reconstruction approaches that assume infinite sites, which would assume that each mutation only happens once in the tree.}.
+At the same time, we are not aware of cases where different cells from the same population acquire different point mutations at the exact same locus, and expect this to be highly unlikely.
+Instead, a third sub-population will usually be generated by a loss of heterozygosity event on a larger scale, covering a mutated site \citep{kuipers_single-cell_2017,satas_scarlet_2020}.
+
+Thus, if we start with a homozygous reference bulk, a first event would introduce an alternative allele mutation and a second event in that sub-population would lead to the loss of either the reference allele or the alternative allele.
+Losing the recently generated alternative allele leads to an allele frequency in that sub-population that is similar to the homozygous reference genotype (it only has coverage of the reference allele), and this sub-population is thus indistinguishable from the original homozygous reference population.
+ProSolo will correctly identify any single cells with the loss event as homozygous reference (Figure~\ref{fig:event-space}).
+
+However, losing the reference allele leads to a sub-population that only has coverage of the alternative allele and is thus similar to a homozygous alternative genotype and clearly distinguishable from the other two genotypes.
+While this type of event was not observed in a systematic survey of multiple mutational hits per site \citep{kuipers_single-cell_2017}, it has coincidentally been observed in a recent study that explicitly modeled copy number alterations in single cells for their phylogenetic reconstruction of cell lineages (mutation LINGO2:2 in \cite{satas_scarlet_2020}).
+Thus, we expect this type of event to be rare, but possible.
+In this case, as the alternative allele was only introduced in a sub-population and the loss of the reference allele is only present in an even smaller sub-population of that, the bulk frequency of this alternative allele will be well below $0.5$.
+Here, the ProSolo model assumes the homozygous alternative genotype of this sub-population impossible and instead calls cells from that sub-population as heterozygous (Figure~\ref{fig:event-space}).
+However, this only concerns the genotype, while the presence of the mutation will be identified correctly.
+As MonoVar is the only other variant calling software that can even genotype beyond calling the presence or absence of a mutation, this could only be considered a drawback compared to that tool.
+
+As germline homozygous alternative sites are very rare compared to germline homozygous reference sites (due to the reference genome being biased towards more common alleles), we expect the symmetric case\footnote{A mutation towards the reference with a subsequent complete loss of the originally homozygous alternative allele.} to be much less likely to be observed.
+
+Finally, starting from a heterozygous germline genotype in the bulk will be more likely than starting from a homozygous alternative genotype, but a lot less likely than starting from a homozygous reference genotype.
+From this starting point, three sub-populations regarding the same alternative allele can only be generated by two events in opposite directions in distinct cells from that population:
+one point mutation or loss event towards the reference and one point mutation or loss event towards the alternative allele, with the loss events much more likely to affect any site than the point mutations.
+Whether the resulting bulk alternative allele frequency then ends up above or below $0.5$, depends on which event happens earlier and/or grows faster within the population.
+As a result, the homozygous genotype (reference or alternative) with the lower frequency will erroneously be considered a heterozygous single cell genotype, but ProSolo will correctly call the more frequent (and thus probably more relevant) homozygous genotype.
+However, we are not aware of any such events being reported in the literature and expect them to be vary rare.
+
+We conclude that the assumption is reasonable to make and we only expect it to lead to the misclassification of homozygous alternative single cell sites as heterozygous when they have been generated by the introduction of an alternative allele via a point mutation, followed by an allelic loss of the reference allele in some daughter cells.
+We also discuss possible respective cases in the context of the tumor single cells from the data of \citet{wang_clonal_2014} (Supplement Section~\ref{sec:somatic-recall}).
+
+It should be possible to mitigate such cases with further extensions of the model to account for copy number changes.
+As in \citet{satas_scarlet_2020}, this would require separately obtaining copy number profiles that are then provided as input to ProSolo, but we do not yet see a standardized way of obtaining such profiles from single cell DNA-sequencing data.
+So far, such copy number profiles seem to have usually been obtained with custom scripts tailored to specific datasets (for example \cite{baslan_genome-wide_2012,satas_scarlet_2020,kuipers_single-cell_2020}), but when more generalized approaches for obtaining such profiles become available, they could then be used to adjust the exact model setup for each site based on the local copy number profile.
+Newer versions of varlociraptor, whose variant calling functionality ProSolo uses in the background, already allow for a chromosome-specific specification of copy numbers\footnote{See the "ploidy" definition in this documentation section: \url{https://varlociraptor.github.io/docs/calling/\#configuring-the-joint-prior-distribution}}.
+Thus, it should be possible to extend this to site-specific copy numbers in the future and to extend the ProSolo model to arbitrary ploidies.
+
+\subsubsection{Posterior Probabilities of Events}\label{sec:event-posteriors}
 
@@
 
+\begin{figure}[!tpb]
+  \includegraphics[width=\linewidth]{figs/Wang2014/Wang2014_validated_somatic_clonal_zygosity_misclassifications.pdf}
+ \caption{
+   The number of clonal variants with a respective alternative allele frequency, stratified by the genotype (hom for homozygous alternative and het for heterozygous) provided by \cite{wang_clonal_2014} for their somatic ground truth list.
+   We consider the peak around $0.4$ to represent genuine heterozygous variants and the peak around $0.7$ to represent genuine homozygous alternative variants, with a tumor sample purity in the duplex sequencing experiment of somewhere around those $0.7$.
+ }
+\label{fig:wang-clonals-zygosity}
+\end{figure}
+
 \textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
 The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
-For each, 16 cells and a bulk sample were whole exome sequenced.
+For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}C).
 For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
 In addition, \cite{wang_clonal_2014} provide a list of 373 non-synonymous clonal and 94 non-synonymous subclonal somatic variants (Supplementary Tables 6 and 7 in \cite{wang_clonal_2014}) whose presence they validated with deep targeted duplex sequencing.
-Depending on the single cell, we used two different ground truths for two separate evaluations:
+As they provide a duplex sequencing based alternative allele frequency, we examined the alternative allele frequency distribution of the non-synonymous clonal variants, stratified by the specified zygosity (Figure~\ref{fig:wang-clonals-zygosity}).
+This clearly shows two alternative allele frequency peaks, one at around $0.4$ and the other at roughly double the alternative allele frequency around $0.7$.
+These two peaks should correspond to heterozygous ($0.4$) and homozygous alternative variants ($0.7$), respectively.
+As these are supposed to be clonal variants in the tumor, the remaining fraction of $0.3$ beyond the homozygous alternative variants peak (from $0.7$ to $1$) should be contamination with normal cells (without these somatic variants).
+This indicates a duplex sample tumor purity of around $0.7$.
+Unexpectedly, we also see the $0.7$ peak in variants that \cite{wang_clonal_2014} classified as heterozygous and the $0.4$ peak in variants that \cite{wang_clonal_2014} classified as homozygous alternative (Figure~\ref{fig:wang-clonals-zygosity}).
+It is unclear from \cite{wang_clonal_2014} and its supplementary material how the original variant classifications were generated, we assume that this is based on the genotypes obtained for the single cells.
+But as the single cells were genotyped with standard variant calling software (GATK UnifiedGenotyper), they very likely suffer from considerable allelic dropout and we thus reclassified the clonal variants' genotypes based on the duplex sequencing.
+If the alternative allele frequency determined by the ultra-deep duplex sequencing was below $0.6$, we classified the variant as heterozygous, and as homozygous alternative if this frequency was above (or equal to) $0.6$ (Figure~\ref{fig:wang-clonals-zygosity}).
+Then, depending on the single cell, we used two different ground truths for two separate evaluations:
 \begin{enumerate}
-    \item For the tumor single cells, we constructed a ground truth from normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
+    \item For the tumor single cells, we constructed a ground truth from the normal bulk sample (as the tumor bulk sample was used for calling with ProSolo) using the same methodology as used for the cell line data (GATK HaplotypeCaller for variant sites, bcftools mpileup for non-variant sites).
     Please bear in mind that this ground truth will mostly only contain germline variants and only few somatic mutations that appeared early enough in the patient's life to rise to high somatic frequencies.
@@
   \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
-  Thresholds used were applied via the following command line options:
+  Used thresholds are not comparable between tools.
+  They were applied via the following command line options:
   MonoVar {\ttfamily --t};
@@
 \begin{figure}[!tpb]
-  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf}
- \caption{
- Cell-specific $F_1$ scores for different filtering threshold levels for the whole genome dataset.\newline \footnotesize
-  Thresholds used were applied via the following command line options:
+  \includegraphics[width=\linewidth]{figs/Dong2017/Dong2017_prosolo-monovar-scansnv-sccaller_per-cell_prec-rec.pdf}
+ \caption{
+ Cell-specific precision-recall plots across different filtering threshold levels for the whole genome dataset.\newline \footnotesize
+  Used thresholds are not comparable between tools.
+  They were applied via the following command line options:
   MonoVar {\ttfamily --t};
@@
  }
- \label{fig:F_1-scores_Dong2017}
-\end{figure}
-
-\begin{figure}[!tpb]
-  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf}
- \caption{
- Cell-specific $F_1$ scores for different filtering threshold levels for the whole exome dataset.\newline \footnotesize
-  Thresholds used were applied via the following command line options:
+ \label{fig:per-cell_prec-rec_Dong2017}
+\end{figure}
+
+\begin{figure}[!tpb]
+  \includegraphics[width=\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_per-cell_prec-rec.pdf}
+ \caption{
+ Cell-specific precision-recall plots across different filtering threshold levels for the whole exome dataset.\newline \footnotesize
+  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
+  Used thresholds are not comparable between tools.
+  They were applied via the following command line options:
   MonoVar {\ttfamily --t};
@@
  }
- \label{fig:F_1-scores_Laehnemann2017}
-\end{figure}
-
-ProSolo's improvements in accuracy can be further formalised using the $F_1$ score, the harmonic mean of precision $p$ and recall $r$, with $\beta = 1$ for the general $F$ score defined as \citep{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}:\\
-
-\begin{equation} \label{eq:F_1_score}
-  F_{\beta} = (\beta^2 + 1) \times \frac{p r}{\beta^2 p + r}
-\end{equation}
-
-The per-cell $F_1$ score clearly demonstrates ProSolo's performance advantage (Figures~\ref{fig:F_1-scores_Dong2017} and \ref{fig:F_1-scores_Laehnemann2017}), even when restricting our comparison to ProSolo's (and SCIPhI's) default mode for a fairer comparison on the whole exome dataset with a germline ground truth (see caveat discussed above).
+ \label{fig:per-cell_prec-rec_Laehnemann2017}
+\end{figure}
 
@@
 Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
-While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a drastically precision.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a reduced precision.
 This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
@@
 \begin{figure}[!tpb]
-  \includegraphics[width=.99\linewidth]{figs/Laehnemann2017/Laehnemann2017_prosolo-downsampling_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf} \newline
- \caption{
-  Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth, including the results of ProSolo after downsampling from the maximum coverage of 50X to 35X, 20X, 5X or 0X.\newline
+ \begin{minipage}[t]{.4\linewidth}
+  \includegraphics[height=70ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-downsampling_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf}
+ \end{minipage}
+ \begin{minipage}[t]{.58\linewidth}
+  \includegraphics[height=70ex]{figs/Laehnemann2017/Laehnemann2017_prosolo-downsampling_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left_non-zero-bulk.pdf}
+ \end{minipage}\newline
+ \caption{
+  Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth, including the results of ProSolo after downsampling from the maximum coverage of 50X to 35X, 20X, 5X or 0X.
+  The left panel includes all bulk coverages, the right panel focuses on the differences in non-zero bulk coverages.\newline 
   $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline \footnotesize
-  Thresholds used were applied via the following command line options:
+  Used thresholds are not comparable between tools.
+  They were applied via the following command line options:
   MonoVar {\ttfamily --t};
@@
 \end{figure}
+
 For the tumor cells from the \cite{wang_clonal_2014} dataset, we could further determine the recall of non-synonymous somatic variants they had validated via deep targeted duplex sequencing and provided in their Supplementary Tables 6 and 7 (specifying clonal and subclonal variants, respectively).
@@
 On the clonal variants, MonoVar has a higher homozygous alternative recall than ProSolo, but ProSolo has a higher heterozygous recall.
-On the subclonal variants, two things are notable:
+On the subclonal variants, three things are notable:
 \begin{enumerate}
@@
   Assuming an evenly mixed cell population (thus ignoring any population structure or in turn assuming an even sampling of all relevant subclones), the bulk coverage of 35X should allow for the discovery of variant alleles present at a frequency of at least 2.86\% of sampled alleles.
-  Judging by the frequency detected in the targeted duplex sequencing of \cite{wang_clonal_2014}, this should make the detection of 52 of the 94 subclonal variants, which corresponds to the number of missed subclonal variants in Figure~\ref{fig:somatic-recall}.
+  Judging by the frequency detected in the targeted duplex sequencing of \cite{wang_clonal_2014}, this should render the detection of 52 of the 94 subclonal variants impossible, which corresponds to the number of missed subclonal variants in Figure~\ref{fig:somatic-recall}.
   As a consequence, if the aim is to detect lower frequency mutations, we simply recommend a deeper sequencing of the bulk sample.
-  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells, while the maximum frequency for any of the heterozygous subclonal mutation detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
+  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells, while the maximum frequency for any of the heterozygous subclonal mutations detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
   This suggests that SCIPhI's model over-imputes mutations by aggregating information across cells.
+  \item None of the tools calls any of the six homozygous alternative variants.
+  SCcaller and SCIPhI do not genotype and thus can only determine the presence or absence of an alternative allele.
+  We chose to represent this as a heterozygous alternative genotype, as is this will be the correct choice in the majority of cases.
+  For MonoVar, this is surprising, as all these homozygous alternative variants should be present in three cells, which should allow MonoVar to recover them even when enforcing an occurrence in a minimum of three independent single cells.
+  For ProSolo, this seems to be due to the low bulk coverage of these variants.
+  The low coverage of these variants in the duplex sequencing means that the alternative allele is probably not covered in the low coverage bulk tumor sample at all, and ProSolo accordingly assigns a homozygous reference genotype.
+  However, if the bulk coverage were higher, ProSolo might still suffer from its assumption of a maximum of two different genotypes at a site in the bulk sample.
+  This can lead to erroneously mis-calling homozygous alternative sites as heterozygous, when they are the result of a point mutation followed by a loss of the reference allele in a daughter cell (Supplementary Section~\ref{sec:event-space-def}).
+  However, an even simpler explanation (for all tools) could be that the ground truth genotype was called with GATK UnifiedGenotyper, a variant caller that is not adapted to single cell DNA-sequencing data, and may have thus suffered from allele dropout of the reference allele to generate these homozygous alternative genotypes.
+  This could mean that these homozygous alternative ground truth are not correct to begin with.
 \end{enumerate}
@@
  Any points underneath it indicate overly conservative FDR control, any points above it indicate an FDR control that is too lenient.\newline \footnotesize
-  Thresholds used were applied via the following command line options:
+  Used thresholds are not comparable between tools.
+  They were applied via the following command line options:
   MonoVar {\ttfamily --t};
@@
 
-Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
+Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C), with the $F_1$ score defined as the harmonic mean of precision $p$ and recall $r$ (with $\beta = 1$ for the general $F$ score, see \cite{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}):\\
+
+\begin{equation} \label{eq:F_1_score}
+  F_{\beta} = (\beta^2 + 1) \times \frac{p r}{\beta^2 p + r}
+\end{equation}.
+
 For ProSolo, this amounts to the default mode without imputation (which also ensures a fair comparison against the germline-based ground truth) and controlling for a false discovery rate of $0.2$.

### 2021-06-29 manuscript/figs/Dong2017/Dong2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -211,13 +211,10 @@ ggsave("Dong2017_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf", device = c
   
 ggplot(metrics %>% 
         filter(software != "SCcaller default bulk", software != "SCcaller sensitive bulk"),
-  aes(x = filter, y = F1) ) +
-  coord_cartesian( ylim = c(0,1) ) +
-  geom_line( aes(group = interaction(cell, software), color = software, linetype = software)) +
+  aes(x = recall, y = precision) ) +
+  coord_cartesian( ylim = c(0.95,1), xlim = c(0.35,0.85) ) +
+  geom_line( aes(color = software, linetype = software)) +
   geom_point( aes(shape = filter_factor, color = software, fill = software), size = 4) +
-  scale_x_continuous(
-    trans = 'log10'
-  ) +
   scale_shape_manual(
     values = shape_pal
   ) +
@@ -246,11 +243,11 @@ ggplot(metrics %>%
   ) +
   facet_wrap(~cell, ncol = 2) +
   labs(
-    x = "threshold used",
-    y = "F_1 score",
+    x = "recall",
+    y = "precision",
     title = "IL-11 and IL-12"
   )
-ggsave("Dong2017_prosolo-monovar-scansnv-sccaller_F1_plot.pdf", device = cairo_pdf, width=21, height=10)
+ggsave("Dong2017_prosolo-monovar-scansnv-sccaller_per-cell_prec-rec.pdf", device = cairo_pdf, width=21, height=10)
 
 ggplot(avg_per_software_and_parameter %>%
     filter(software != "SCcaller default bulk", software != "SCcaller sensitive bulk"),

### 2021-06-29 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -245,13 +245,10 @@ ggplot(metrics %>%
 #          software != "ProSolo imputation",
 #          software != "SCIPhI sensitive"
           ),
-  aes(x = filter, y = F1) ) +
-  coord_cartesian( ylim = c(0,1) ) +
+  aes(x = recall, y = precision) ) +
+  coord_cartesian( ylim = c(0.875,1), xlim = c(0,0.35) ) +
   geom_point( aes(shape = filter_factor, color = software, fill = software), size = 4) +
   geom_line( aes(group = interaction(cell, software), color = software, linetype = software)) +
-  scale_x_continuous(
-    trans = 'log10'
-  ) +
   scale_shape_manual(
     values = shape_pal
   ) +
@@ -291,10 +288,10 @@ ggplot(metrics %>%
     )
   ) +
   facet_wrap(~cell, ncol=2) +
-  labs(x = "threshold used",
-       y = "F_1 score",
+  labs(x = "recall",
+       y = "precision",
        title = "5 granulocytes")
-ggsave("Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_F1_plot.pdf", device = cairo_pdf, width=21, height=24)
+ggsave("Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_per-cell_prec-rec.pdf", device = cairo_pdf, width=21, height=24)
 
 ggplot(avg_per_software_and_parameter %>%
     filter(
@@ -391,6 +388,53 @@ ggplot(avg_per_software_and_parameter %>%
   scale_linetype_manual(
     values = linetype_pal,
     labels = software_labels,
+    guide = "none"
+    ) +
+  geom_point( aes(shape = filter, color = software, fill = software), size = 4) +
+  scale_shape_manual(
+    guide = "none",
+    values = shape_pal
+    ) +
+  scale_colour_manual(
+    guide = "none",
+    labels = software_labels,
+    values = software_pal
+    ) +
+  scale_fill_manual(
+    guide = "none",
+    labels = software_labels,
+    values = software_pal
+    ) +
+#  guides( 
+#    linetype = guide_legend( keywidth = 3, keyheight = 1.5),
+#    shape = guide_legend( title = "threshold used")
+#    ) +
+  theme_bw(base_size=24, base_family="Lato") +
+  labs(x = "average recall",
+       y = "average precision",
+       title = "Low bulk coverage sufficient",
[…]

### 2021-06-29 manuscript/figs/Wang2014/Wang2014_motivation_reclassify_zygosity_clonal_somatics.R

@@ -1,0 +1,44 @@
+# packages, colors, palettes and stuff
+source("../common.R")
+
+# Read in validated somatics annotations
+validated_clonals <- read_tsv(
+  "nature13600-s1-table-s6-clonal.tsv"
+) %>%
+  mutate(
+    Duplex_Freq = as.numeric( str_replace(Duplex_Freq, ",", ".") ),
+    Duplex_P_val = as.numeric(
+      str_replace(
+        str_replace(
+          str_replace(Duplex_P_val, "-[^\\d]+", "-"),
+          "<", ""),
+        ",", ".")
+      )
+    )
+
+reclassification_motivation <- ggplot(validated_clonals) +
+  geom_histogram(
+    aes(
+      Duplex_Freq
+    ),
+    bins = 50
+  ) +
+  scale_x_continuous(
+    limits = c(0.0, 1.0),
+    n.breaks = 11
+  ) +
+  facet_grid(
+    cols = vars(class),
+    rows = vars(zygosity)
+  ) +
+  theme_bw() +
+  xlab("alternative allele frequency (duplex sequencing)") +
+  ylab("clonal variant calls") 
+
+
+ggsave(
+  filename="Wang2014_validated_somatic_clonal_zygosity_misclassifications.pdf",
+  plot=reclassification_motivation,
+  width=10,
+  height=5
+)

### 2021-06-29 manuscript/figs/Wang2014/Wang2014_plot_prosolo-monovar-sccaller-sciphi_som_gt_recall.R

@@ -128,19 +128,26 @@ for ( sc in tumor_cells ) {
 
 validated_somatics <- read_tsv(
   "Wang2014_ground_truth_non_synonymous_variants.hg18_to_hg19.tsv"
-)
+) %>%
+  mutate(
+    Duplex_Freq = as.numeric( str_replace(Duplex_Freq, ",", ".") ),
+    Duplex_P_val = as.numeric(
+      str_replace(
+        str_replace(
+          str_replace(Duplex_P_val, "-[^\\d]+", "-"),
+          "<", ""),
+        ",", ".")
+      )
+    )
 
 max_subclonal_het <- validated_somatics %>%
   filter(class == "subclonal") %>%
   filter(zygosity == "het") %>%
-  mutate(
-    Duplex_Freq = as.double(str_replace(Duplex_Freq, ',', '.'))
-  ) %>%
   summarize(max(Duplex_Freq))
 write_tsv(max_subclonal_het, path = "Wang2014_max_subclonal_duplex_frequency.tsv")  
 
 expected_numbers <- validated_somatics %>%
-  select(class, chrom, pos, REF, VAR, cells, zygosity) %>%
+  select(class, chrom, pos, REF, VAR, cells, zygosity, Duplex_Freq) %>%
   rename(
     CHROM = chrom,
     POS = pos,
@@ -148,25 +155,51 @@ expected_numbers <- validated_somatics %>%
     ALT = VAR,
     number_of_cells = cells,
     CLONALITY = class,
-    GT = zygosity
   ) %>%
   mutate(
     number_of_cells = if_else(number_of_cells == "NaN", 16, number_of_cells),
     GT = case_when(
-      GT == "het" ~ "0/1",
-      GT == "hom" ~ "1/1",
-      TRUE ~ "unknown"
-    )
+      zygosity == "het" ~ "0/1",
+      zygosity == "hom" ~ "1/1"
+    ),
+    expected_GT = GT
   ) %>%
   add_column(
     software = "expected"
   )
-  
+ 
 per_software <- recall_per_cell %>%
-  filter(GT_CALL == "TP") %>%
   count(software, CHROM, POS, REF, ALT, CLONALITY, GT, name = "number_of_cells") %>%
-  full_join(expected_numbers) %>%
-  complete(software, nesting(CHROM, POS, REF, ALT, CLONALITY, GT), fill = list(number_of_cells = 0))
+  # add the expected genotype into software records
+  full_join(
+    expected_numbers %>%
+      select(
+        -software,
[…]

### 2021-06-29 analysis_pipelines/Wang2014/somatic_validated_mutations/nature13600-s1-tables_to_ground_truth.py
update analysis_pipeline to reclassification of clonal somatic mutations

@@ -1,3 +1,15 @@
+# install dependencies with mamba:
+# mamba create -n pyliftover-pandas -c conda-forge -c bioconda pyliftover=0.4 pandas=1.1.5
+# 
+# activate the new environment before running this script:
+# conda activate pyliftover-pandas
+#
+# then run with:
+# python3 nature13600-s1-tables_to_ground_truth.py >Wang2014_ground_truth_non_synonymous_variants.hg18_to_hg19.tsv
+#
+# and deactivate the environment:
+# conda deactivate
+
 import pandas
 import re
 
@@ -19,6 +31,13 @@ with open("Wang2014_ground_truth_non_synonymous_variants.hg18_to_hg19.tsv", mode
         lo = liftover.convert_coordinate(row['chrom'],row['pos'] - 1)[0]
         row['chrom'] = lo[0]
         row['pos'] = lo[1] + 1
+        # reclassify clonals' zygosity based on the Duplex_Freq
+        if row['class'] == 'clonal':
+            freq = float(row['Duplex_Freq'].replace(',', '.'))
+            if freq >= 0.6:
+                row['zygosity'] = 'hom'
+            else:
+                row['zygosity'] = 'het'
         print( '\t'.join(map(str, row)) )
         print( '\t'.join(map(str, row)), file=out, end='\n' )
 

### 2021-09-09 manuscript/main.tex
update manuscript, scripts and figures to final submission (last reviewer comment addressed, editorial requests addressed)

@@
 \usepackage{xr-hyper}
+\usepackage{hyperref}
 
@@
 
-\usepackage[left]{lineno}
-\linenumbers
+%\usepackage[left]{lineno}
+%\linenumbers
 
@@
 \title{ProSolo: Accurate and Scalable Variant Calling from Single Cell DNA Sequencing Data\vspace{3ex}}
-\author[1,2,3,4]{David Lähnemann}
-\author[4,5]{Johannes Köster}
-\author[3]{Ute Fischer}
-\author[3]{Arndt Borkhardt}
-\author[1,2,+,*]{Alice C. McHardy}
-\author[5,6,+,*]{Alexander Schönhuth}
+\author[1,2,3,4,5]{David Lähnemann}
+\author[5,6]{Johannes Köster}
+\author[4]{Ute Fischer}
+\author[4]{Arndt Borkhardt}
+\author[1,2,3,*]{Alice C. McHardy}
+\author[6,7,*]{Alexander Schönhuth}
 \affil[1]{Department for Computational Biology of Infection Research, Helmholtz Centre for Infection Research, 38124 Braunschweig, Germany.}
-\affil[2]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
-\affil[3]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
-\affil[4]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
-\affil[5]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
-\affil[6]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
-\affil[+]{shared last authors}
-\affil[*]{To whom correspondence should be addressed:
+\affil[2]{Braunschweig Integrated Centre of Systems Biology (BRICS), Technische Universität Braunschweig, Braunschweig, Germany.}
+\affil[3]{Algorithmic Bioinformatics, Faculty of Mathematics and Natural Sciences, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
+\affil[4]{Department of Paediatric Oncology, Haematology and Immunology, University Hospital, Medical Faculty, Heinrich Heine University Düsseldorf, 40225 Düsseldorf, Germany.}
+\affil[5]{Algorithms for Reproducible Bioinformatics, Institute of Human Genetics, University of Duisburg-Essen, 45147 Essen, Germany.}
+\affil[6]{Genome Data Science, Life Sciences Group, Centrum Wiskunde \& Informatica, 1098 XG Amsterdam, The Netherlands.}
+\affil[7]{Genome Data Science, Faculty of Technology, Bielefeld University, 33615 Bielefeld, Germany.}
+\affil[*]{These authors jointly supervised this work:
  \href{alice.mchardy@helmholtz-hzi.de}{alice.mchardy@helmholtz-hzi.de},
@@
 {\sffamily
-Obtaining accurate mutational profiles from single cell DNA is essential for the analysis of genomic cell-to-cell heterogeneity at the finest level of resolution.
+Accurate single cell mutational profiles can reveal genomic cell-to-cell heterogeneity.
 However, sequencing libraries suitable for genotyping require whole genome amplification, which introduces allelic bias and copy errors.
-As a result, single cell DNA sequencing data violates the assumptions of variant callers developed for bulk sequencing, which---when applied to single cells---generate significant numbers of false positives and false negatives.
-Only dedicated models accounting for amplification bias and errors will be able to provide more accurate calls.
+The resulting data violates assumptions of variant callers developed for bulk sequencing.
+Thus, only dedicated models accounting for amplification bias and errors can provide accurate calls.
 
-We present ProSolo, a probabilistic model for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
-The model encompassingly integrates all relevant biases emerging during MDA in a site-specific and scalable---because computationally efficient---manner.
-ProSolo models a single cell jointly with a bulk sequencing sample, which enhances variant calling performance, enables identifying MDA induced errors, and supports imputation of genotypes in single cells at insufficiently covered sites, when downstream tools cannot handle missing data.
-ProSolo achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools. 
-Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly; not only for single nucleotide variant calls, but also for artefacts of single cell methodology that one may wish to quantify, such as allele dropout.
+We present ProSolo for calling single nucleotide variants from multiple displacement amplified (MDA) single cell DNA sequencing data.
+ProSolo probabilistically models a single cell jointly with a bulk sequencing sample and integrates all relevant MDA biases in a site-specific and scalable---because computationally efficient---manner.
+This achieves a higher accuracy in calling and genotyping single nucleotide variants in single cells in comparison to state-of-the-art tools and supports imputation of insufficiently covered genotypes, when downstream tools cannot handle missing data.
+Moreover, ProSolo implements the first approach to control the false discovery rate reliably and flexibly.
 
-ProSolo's model is implemented into a flexible framework, encouraging extensions. The source code and usage instructions are available at: \url{https://github.com/prosolo/prosolo}
+ProSolo is implemented in an extendable framework, with code and usage at: \url{https://github.com/prosolo/prosolo}
 }
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-We describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
+In this work we describe ProSolo, a variant caller using a unifying statistical framework that takes into account all relevant MDA related biases and errors, allowing for them to vary locally.
 Importantly, our model enables a scalable---because computationally efficient---implementation, which is challenging even in bulk variant calling when considering local effects due to statistical uncertainties affecting the data\cite{koster_varlociraptor_2020}.
 ProSolo's statistical rigor allows for an accurate control of the false discovery rate when calling alternative alleles or identifying other relevant effects, such as allele dropout.
+It achieves a higher variant calling accuracy compared to state-of-the-art tools.
 
-\section{Results}
+\section*{Results}
 
-\subsection{Single Cell Sequencing Model}
+\subsection*{Comprehensive and Flexible Single Cell Sequencing Model}
 
-We describe a novel probabilistic model that addresses the genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
-In the following, we introduce the central innovations of our model and demonstrate its advantages in comparison to existing approaches.
-More details and a detailed derivation of all model elements can be found in the Supplement, including a summary of the core model.
+We describe a probabilistic model that addresses the genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
+Here, we name the central innovations of our model and demonstrate its advantages in comparison to existing approaches.
+A more detailed description of the innovations is available in the Methods section.
 
-Our model addresses the two major issues of MDA: (i) the differential amplification of the two alleles present in a diploid cell ("amplification bias" in the following); (ii) MDA induced errors ("amplification errors" in the following) which are copy errors introduced by the $\Phi29$ polymerase used in MDA.
+Briefly, our model addresses the two major issues of MDA: (i) the differential amplification of the two alleles present in a diploid cell ("amplification bias" in the following); (ii) MDA induced errors ("amplification errors" in the following) which are copy errors introduced by the $\Phi29$ polymerase used in MDA.
 To address amplification bias, we leverage a mechanistically motivated, empirically derived model of differential amplification of alleles.
@@
 At the same time, one of the major features of the core model and its implementation is that it can easily be adapted to flexibly deal with other sampling setups, so it could be extended to further scenarios.
-For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_varlociraptor_2020}.
 
-\begin{figure}[!tpb]%figure1
-  \begin{minipage}[t]{.99\linewidth}
-    \begin{minipage}{.47\linewidth}
-      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
-    \end{minipage}
-    \hspace{2ex}
-    \begin{minipage}{.47\linewidth}
-      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
-    \end{minipage}
-  \end{minipage}
-  \vspace{2ex}
-  \hspace{1ex}
-  \begin{minipage}[t]{.99\linewidth}
-    \begin{minipage}{.47\linewidth}
-      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
-    \end{minipage}
-    \hspace{2ex}
-    \begin{minipage}{.49\linewidth}
-      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
-    \end{minipage}
-  \end{minipage}
-  \caption{
-   \textbf{A,B} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
-   Homozygous reference sites in \textbf{(A)} are assumed to follow a beta-binomial distribution;
-   sites heterozygous for the alternative allele in \textbf{(B)} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
-   \textbf{(C)} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs.
-$\theta_s = 0.5$).
-   Alternative nucleotide T (orange) is an amplification error.
-   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
-   This is formalised with the model in D.
-   \textbf{(D)} Definition of single cell events based on ProSolo's likelihood density estimates for the spectrum of true underlying alternative allele frequencies in the single cell ($\tilde{\theta}_s$) and the bulk ($\tilde{\theta}_b$).
-   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
-   The model further assumes that the bulk sample has sufficient coverage to capture somatic variants.\small\newline
-   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
-  \label{fig:prosolo_alt-calling}
-\end{figure}
-
-\paragraph{A mechanistically motivated model of amplification bias, trained on data, gives realistic coverage-specific single-cell (genotype) likelihoods.}
-To account for MDA amplification bias up to the complete dropout of individual alleles, we distinguish between two alternative allele frequencies:
-\begin{enumerate*}[label={(\roman*)}]
-  \item The true (but usually unknown) underlying allele frequency at a site in a single cell: $\theta_s$.
-   This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
-   However, the ratio of reads harboring the different alleles from a single cell sequencing experiment does not reflect the true allele frequency, because of the biases induced by amplification. 
-  Instead, the ratio of reads reflects
-  \item the allele frequency after its distortion through amplification bias.
-   For a site with total coverage of $l$ reads, of which $k$ reads bear the alternative allele, the formal definition of this measurable frequency is $\rho_s=\frac{k}{l}$.
-\end{enumerate*}
-
-The goal is to estimate the likelihood density across the three possible underlying allele frequencies in the single cell (we denote $\tilde{\theta}_s$ as the density estimate across all $\theta_s\in\{0.0,0.5,1.0\}$).
-To accurately quantify the uncertainty introduced by amplification bias, we consider
-
-\begin{equation}
- \mathbf{P}(\rho_s\mid\theta_s)\qquad \text{for}  \quad\theta_s\in\{0,0.5,1\},
-\end{equation}
-
-the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}C).
-We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele, to calculate likelihoods for each of these possible true allele frequencies.
-
-To do this, we follow the considerations of Lodato et al.\cite{lodato_somatic_2015} (see Figure S5 and the section ``Modeling MDA-derived alternative read counts'' of the respective supplement for the details), who fitted well-studied probability distributions to empirical distributions they obtained for $\mathbf{P}(\rho_s\mid\theta_s)$ of $\theta_s\in\{0,0.5\}$.
-For sites that are homozygous for the reference allele, amplification bias cannot initially happen.
-However, once an amplification error creates an alternative allele, this can be amplified to large frequencies due to amplification bias.
-Lodato et al.\cite{lodato_somatic_2015} thus consider the empirical distribution $\mathbf{P}(\rho_s\mid\theta_s=0)$ to follow a beta-binomial.
-Effectively, this means that the probability of a non-zero alternative read count  ($\rho_s>0$, which is any alternative read count above $0$ in Figure~\ref{fig:prosolo_alt-calling}A) will be non-zero ($\mathbf{P}(\rho_s\mid\theta_s=0)>0$), merely because of sequencing and amplification errors.
-In contrast, at heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B), the distribution is dominated by amplification bias.
-Thus, for $\mathbf{P}(\rho_s\mid\theta_s=0.5)$ they found a mixture of two beta-binomial distributions to appropriately fit the empirically observed distributions (Figure~\ref{fig:prosolo_alt-calling}B).
-We further motivate the choice of the beta-binomial distribution mechanistically by an analogy to its generative urn model named after Pólya\cite{eggenberger_uber_1923}:
-Take an urn with two white and two black balls, where each ball represents one strand of each (double-stranded) allele at a heterozygous site.
-In the Pólya urn model, drawing a ball leads to replacement of that ball with two balls of the same color, analogous to a strand copy by the $\Phi29$ polymerase (additional discussion in the Supplement, Section~\ref{sec:polya-urn}).
-Finally, the necessity for a mixture of beta-binomials (Figure~\ref{fig:prosolo_alt-calling}B) becomes evident when contrasting it with the binomial distribution observed at heterozygous sites in bulk experiments.
-Namely, the bulk distribution for reads supporting the alternative allele would narrowly peak at a count of $k=\frac{l}{2}$, which is not the case for the corresponding $k=10$ in Figure~\ref{fig:prosolo_alt-calling}B.
-Instead, the mixture of beta-binomials peaks at the extreme read counts of $k=0$ and $k=l$, highlighting that the dropout of the alternative or reference allele is quite likely to occur.
-Most likely, this mixture arises from differences in the accessibility of different pieces of the genomic DNA for amplification\cite{picher_trueprime_2016} (Supplement, Section~\ref{sec:polya-urn}).
-Finally, for homozygous alternative sites (i.e. $\mathbf{P}(\rho_s\mid\theta_s=1)$), we rely on the symmetry of the cases $\theta_s=0$ and $\theta_s=1$.
-In summary, we obtain the following equations for $\mathbf{P}(\rho_s\mid\theta_s)$:
-
-\begin{equation}
-	\label{eq:BBLodato}
-	\begin{split}
-		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 0 ) 			=~ &\cB\cB(~ k,~ l;~ \alpha(l),~ \beta(l)~ ),\\
-		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = \frac12) =~ &w(l) \times \cB\cB(~ k,~ l;~ \alpha_1(l),~ \alpha_1(l)~ )~ +\\
-                                                       ~&~(~ 1-w(l)~ ) \times \cB\cB(~ k,~ l;~ \alpha_2(l),~ \alpha_2(l)~ ),\\
-		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 1 ) 			=~ &\cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ).
-	\end{split}
-\end{equation}
-
-Here, $\cB\cB$ represents the beta-binomial probability mass function where $k\in\{0,...,l\}$ with $l$ being the total read coverage of the site that is considered.
-All of parameters $\alpha, \alpha_1,\alpha_2$, $\beta$, and $w$ scale linearly in $l$.
-Allowing to vary distributions through these parameters allows amplification bias to depend on the total coverage of a site, and thereby to vary locally.
-Symmetry of distributions for the homozygous sites ($\theta_s = 0$ in Figure~\ref{fig:prosolo_alt-calling}A, and $\theta_s = 1$) is established by swapping the shape parameters $\alpha(l)$ and $\beta(l)$.
-The distribution for heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}B) corresponds to a mixture of two beta-binomials with shape parameters $\alpha_1(l)=\beta_2(l)$ and $\alpha_2(l)=\beta_2(l)$, where equality of $\alpha(l)$ and $\beta(l)$ yields symmetry of the distributions relative to $k = \frac{l}{2}$ (the expected value for alternative read counts at heterozygous sites).
-Slopes and intersects for the scaling of all parameter values across different choices of $l$ are given in Table~\ref{tab:Lodato}.
-
-\begin{table}[tbp]
- \caption{
- Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
- The full precision as used in the implementation is given in Supplementary Table~\ref{tab:Lodato-full}
- }
- \label{tab:Lodato}
- \renewcommand{\arraystretch}{1.15}
-
- \begin{center}
-  \begin{tabular}{lrr}
-   parameter  & slope        & intersect   \\
-   \toprule
-   $\alpha$   & -0.00003 & 0.06857 \\
-   $\beta$    & 0.00745  & 2.36749 \\
-   \midrule
-   $w$        & 0.00055  & 0.54040 \\
-   $\alpha_1$ & 0.05738  & 0.66973 \\
-   $\alpha_2$ & 0.00323  & 0.39926 \\
-   \bottomrule
-  \end{tabular}
- \end{center}
-\end{table}
-
-\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
-A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
-Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
-Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
-This establishes a formal, statistical argument why one should consider bulk experiments in single cell sequencing whenever possible: the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell, in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}C).
-At the same time, the absence of a single cell candidate mutation in a bulk sample (with sufficient sequencing depth) increases the probability of an amplification error.
-As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.
-
-For our model, we derive likelihood density estimates for all possible alternative allele frequencies in the background bulk sample.
-Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:
-
-\begin{equation}
- \label{eq:bulk-prob}
- L\left(\theta_b = \frac{m}{n} \mid \boldsymbol{Z}^b\right) \propto
- P\left(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n}\right) =
- \prod_{j=1}^n P\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)
-\end{equation}
-
-Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.
-Further, as for the single cell sample, when referring to the likelihood density estimates across all possible allele frequencies (as opposed to a particular allele frequency), we denote this with $\tilde{\theta}_b$.
-
-\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
-With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
-Figure~\ref{fig:prosolo_alt-calling}C gives a simplified illustration of how the combination of likelihood density estimates across all possible allele frequencies $\tilde{\theta}_s$ and $\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
-However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:
-
-\begin{equation}
-  \label{eq:event-space}
-  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\tilde{\theta}_s} \times [0, 1]_{\tilde{\theta}_b}.
-\end{equation}
-
-For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}D, lower panel; for discussion on the validity of this assumption, see Supplementary Section \ref{sec:event-space-def}).
-At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
-At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
-Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
-For example, an error-free homozygous alternative site is defined by allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
-And the dropout of an alternative allele is defined across allele frequency likelihood density estimates $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}D, Table~\ref{tab:Events}).
-For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
-We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
-This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.
-
-We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}D):
-
-\begin{equation}
-  \label{eq:event-set}
-  \begin{split}
-    \mathfrak{E} &= \{ \boldsymbol{E}_{\text{hom ref}}, \boldsymbol{E}_{\text{err alt}}, \boldsymbol{E}_{\text{ADO to alt}}, \boldsymbol{E}_{\text{het}}, \boldsymbol{E}_{\text{ADO to ref}}, \boldsymbol{E}_{\text{err ref}}, \boldsymbol{E}_{\text{hom alt}} \}
-  \end{split}
-\end{equation}
-
-Assuming a flat prior across the possible underlying allele frequencies for both the bulk and the single cell, we can compute likelihoods for all those single cell events (e.g.~Equation~\ref{eq:ado-to-ref-likelihood-data}).
-The sum of the likelihoods of all these (mutually exclusive) events yields the marginal probability (Equation \ref{eq:marginal-prob}).
-Using the marginal probability, we can calculate the posterior probability for any of these events.
-For example, the posterior probability of event $\boldsymbol{E}_{\text{hom alt}}$ (Figure~\ref{fig:prosolo_alt-calling}D) can be calculated with:
-
-\begin{equation}
- \label{eq:hom-alt-posterior-samples}
- \begin{split}
-  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
-    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
-      \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{hom alt}}) \\
-    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
-      \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) \times \Prob(\boldsymbol{Z}^b \mid \theta_b \in (\frac12,1])\\
-  \end{split}    
-\end{equation}
-
-Accounting for the sample likelihoods based on Equation \ref{eq:theta-likelihood} (assuming $\rho_b = \theta_b$ for the bulk that has no amplification step, Equation \ref{eq:rho-b}), and evaluating only point estimates of these likelihoods at possible alternative allele frequencies, this gives:
-
-\begin{equation}
- \label{eq:hom-alt-posterior-likelihoods}
- \begin{split}   
- \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
-   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
-     &\sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \Prob(\rho_s = \frac{k}{l} \mid \theta_s =  1)\right\} \times \\
-     &\sum_{m=\frac{n}{2}}^{n} \Prob(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n})\\
-  \end{split}    
-\end{equation}
-
-Accounting for amplification bias with Equation \ref{eq:BBLodato} and computing the likelihood of the sample-specific allele frequency ranges with Equations \ref{eq:read-product} and \ref{eq:bulk-prob}, this becomes:
-
-\begin{equation}
- \label{eq:hom-alt-posterior-reads}
- \begin{split}   
- \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
-   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
-      &\sum_{k=0}^{l} \left\{ \prod_{i=1}^l \Prob(Z_i^s \mid \rho_s = \frac{k}{l}) \times \cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ) \right\} \times\\
-      &\sum_{m=\frac{n}{2}}^{n} \prod_{j=1}^l \Prob\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)\\
-  \end{split}    
-\end{equation}
-
-With read probabilities calculated with Equation \ref{eq:read_probability}, we can calculate the posterior probability of the $\boldsymbol{E}_{\text{hom alt}}$ event (for an analogous and more detailed derivation for $\boldsymbol{E}_{\text{ADO to ref}}$, see Supplement, Section~\ref{sec:sc-model}).
-To obtain posterior probabilities for compound events, we sum up the posterior probabilities of the events it comprises, effectively joining up their respective allele frequency range combinations into compound range combinations.
-For the site-specific single cell probability for the presence of an alternative allele in the single cell, we thus get the compound event (blue events in Figure~\ref{fig:prosolo_alt-calling}D; Table~\ref{tab:Events}):
-
-\begin{equation}
- \label{eq:compound-event-alt-presence}
- \begin{split}
-  \boldsymbol{E}_{\text{alt presence}_s}
-  =~&\boldsymbol{E}_\text{ADO to ref}~~\cup~~
-    \boldsymbol{E}_\text{err ref}~~\cup~~
-    \boldsymbol{E}_\text{het}~~\cup~~
-    \boldsymbol{E}_\text{ADO to alt}~~\cup~~
-    \boldsymbol{E}_\text{hom alt} \\
-  =~&\{0\}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}~~\cup~~
-    \{0, \frac{1}{2}\}_{\theta_s} \times [1]_{\theta_b}~~\cup~~
-    \{\frac{1}{2}\}_{\theta_s} \times (0,1)_{\theta_b}~~\cup~~\\
-    &\{1\}_{\theta_s} \times (0,\frac{1}{2}]_{\theta_b}~~\cup~~
-    \{1\}_{\theta_s} \times (\frac{1}{2},1]_{\theta_b} \\
-  =~&\{0\}_{\theta_s} \times [\frac{1}{2},1]_{\theta_b}~~\cup~~
-    \{\frac{1}{2},1\}_{\theta_s} \times (0,1]_{\theta_b}
- \end{split}
-\end{equation}
-
-whose posterior probability we can obtain from this sum:
-
-\begin{equation}
- \label{eq:alt-posterior-prob}
- \begin{split}
-  \Prob(\text{alt presence}_s \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =~&
-    \Prob(\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~ 
-    \Prob(\boldsymbol{E}_{\text{err ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
-  &\Prob(\boldsymbol{E}_{\text{het}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
-  \Prob(\boldsymbol{E}_{\text{ADO to alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
-  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
- \end{split}
-\end{equation}
-
-To genotype, we calculate the posterior probability of all three possible single-cell genotypes and choose the genotype with the maximum posterior probability. 
-In Figure~\ref{fig:prosolo_alt-calling}D, events are colored purple when implying the homozygous reference genotype, light blue when implying the heterozygous genotype and dark blue when implying the homozygous alternative genotype (Table \ref{tab:Events}).
-Similarly, to calculate the posterior probability of an allele dropout at a particular site, we sum up the posterior probabilities of the two {\ttfamily ADO} events.
-
-For any such compound event, we can estimate a threshold on the posterior probabilities that controls for a specified false discovery rate. 
-This is based on the approach described by Müller et al.\cite{muller_optimal_2004,muller_fdr_2006}---for further details see the Supplement (Section~\ref{sec:fdr}) and Köster et al\cite{koster_varlociraptor_2020}.
-
-\paragraph{Biologically relevant imputation based on the bulk sample.}
-As argued above, a large enough sampling of the bulk cell population that the single cell comes from should contain the single cell's genotype at a particular site, unless this cell is genuinely the first cell to harbour a mutation at that site.
-This bulk background sample can thereby render credibility to single cell variants with low coverage, while at the same time eliminating amplification errors in the single cell sample, as these will not exist in the bulk sample.
-Interestingly, the bulk sample also provides a mechanism of biologically meaningful imputation at sites that have no read coverage in the single cell.
-If imputation is desired for sites with no read coverage in the single cell, we set $\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) = \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac{1}{2}) = \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) = 1$, rendering all (unknown underlying) single cell genotypes equally likely.
-Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
-However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will favor germline genotypes over any existing (lower frequency) somatic genotypes at a site, unless such a somatic genotype is present in a majority of cells.
-Thus, such an imputation carries the potential to introduce erroneous calls (especially when looking at subclonal somatic mutations), and we recommend to instead use downstream tools that can accommodate for missing data and data uncertainty wherever they are available.
-
-\paragraph{ProSolo is an easy-to-use command-line tool, based on a modular framework.}
-
-ProSolo is an easy-to-use command-line tool---following usability standards\cite{taschuk_ten_2017}---and its source code is available at \url{https://github.com/prosolo/prosolo}, including instructions for an easy installation via Bioconda\cite{gruning_bioconda:_2018}.
-Its main contribution in terms of software is the implementation of its comprehensive statistical model into the Rust variant calling library of Varlociraptor\cite{koster_varlociraptor_2020}.
-See the Supplement (Section~\ref{sec:implementation}) for further implementation details.
-
-\subsection{Benchmarking}
-
-We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}.
-We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 6.3.0}) to implement the benchmarking workflows.
-For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
-All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).
-
-\subsubsection{Datasets and Ground Truths}
-
-\begin{figure}[!tpb]
-  \begin{minipage}{.3\linewidth}
-    \includegraphics[height=38ex]{figs/Dong2017_data_overview.pdf} \\
-  \end{minipage}
-  \hspace{6ex}
-  \begin{minipage}{.5\linewidth}
-    \includegraphics[height=38ex]{figs/Laehnemann2017_data_overview.pdf} \\
-  \end{minipage}\\
-  \begin{minipage}{.5\linewidth}
-    \includegraphics[height=38ex]{figs/Wang2014_data_overview.pdf} \\
-  \end{minipage}
-  \hspace{6ex}
-  \begin{minipage}{.3\linewidth}
-    \includegraphics[height=38ex]{figs/data_overview_color_legend.pdf} \\
-  \end{minipage}
-  \caption{
-   Benchmark Datasets.
-   For single cells, DNA was multiple displacement amplified (MDA).
-   \textbf{A} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
-   \textbf{B} Newly generated whole exome sequencing (WXS) dataset of blood cells.
-   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
-   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
-   \textbf{C} Whole exome sequencing (WXS) dataset from a triple negative breast cancer (TNBC) patient\cite{wang_clonal_2014}.
-   For tumor cells (a1 to a8 and h1 to h8), the cell population (pop) sequencing of the normal sample was taken as the ground truth, including known clonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
-   For normal cells (n1 to n16), the cell population (pop) sequencing of the tumor sample was taken as the ground truth, after removing known clonal and subclonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
-   }
-  \label{fig:datasets}
-\end{figure}
-
-We benchmarked ProSolo on three experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\
-
-\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
-The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}A, dataset available from SRA project \mbox{PRJNA305211}, accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
-A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
-After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
-From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
-The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
-IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
-The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
-IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
-Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
-Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\
-
-\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
-\begin{sloppypar}
-For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
-Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}B, dataset available as \mbox{EGAD00001005929} in EGA study \mbox{EGAS00001004123}).
-Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
-Using a panel of 16 primer pairs covering different genes across chromosomes for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
-For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
-From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.
-\end{sloppypar}
-
-To generate the ground truth of this dataset, we could leverage previously sequenced bulk whole exome data from the same person, their parents and three siblings \cite{hoell_constitutional_2014} (Figure~\ref{fig:datasets}B).
-To create ground truth germline alternative allele calls, we ran three pedigree-aware variant callers (BEAGLE4.0\cite{browning_improving_2013}, polymutt\cite{li_likelihood-based_2012} and FamSeq\cite{peng_rare_2013,peng_famseq:_2014}) and created a consensus by including only calls where all callers agree at a site and where a maximum of one caller not calling the site was allowed (Figure~\ref{fig:granulocytes-ground-truth}).
-
-\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC)\cite{wang_clonal_2014}.}
-The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available (Figure~\ref{fig:datasets}C, dataset available via the sequence read archive, project id \mbox{PRJNA168068} or accession \mbox{SRA053195}).
-For each, 16 single cell nuclei and a bulk sample were whole exome sequenced.
-Further, a number of candidate sites for clonal and subclonal tumor mutations was confirmed by targeted duplex deep sequencing in a separate experiment\cite{wang_clonal_2014}.
-We analyzed the tumor and normal cells separately, ensuring that normal cells were called with the normal bulk background sample and tumor cells with the tumor bulk background sample.
-As the ground truth for the tumor cells, we then used the normal bulk sample augmented with the clonal tumor mutations confirmed by targeted duplex sequencing.
-As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations.
-This experimental setup aims for fairness across all competitors. 
-Variant calling for the ground truths was performed as for cell line data above.
-
-\subsubsection{ProSolo achieves consistently high alternative allele calling accuracy}
+\subsection*{Consistently High Alternative Allele Calling Accuracy}
 
@@
 the absence of an alternative allele (i.e.
-the heterozygous and the homozygous alternative genotypes called jointly, the joint probability of the blue event areas in Figure~\ref{fig:prosolo_alt-calling}D).
+the heterozygous and the homozygous alternative genotypes called jointly, the joint probability of the blue event areas in Figure~\ref{fig:prosolo_alt-calling}d).
 We thus focused on this for the main benchmarking.
@@
  \end{minipage}\newline
- \begin{minipage}{.9\linewidth}
-  \includegraphics[height=40ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
+ \begin{minipage}{.95\linewidth}
+  \includegraphics[height=45ex]{figs/Wang2014/Wang2014_prosolo-monovar-scansnv-sccaller-scvilp_precision-recall-plot_focus-tools.pdf} \\
  \end{minipage}
@@
   Global views of these panels are provided in Figure~\ref{fig:alt-calling_prec-rec_global}.
-  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
-  \textbf{B} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
-  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.
+  \textbf{a} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1C as ground truth genotypes.
+  \textbf{b} Precision and recall average of the five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.
+  \textbf{c} Precision and recall average of 16 tumor and 16 normal single cells sequenced at the whole exome level. \newline
+  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel b.
 \newline \footnotesize
@@
 
-The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}A and \ref{fig:alt-calling_prec-rec_global}A), at recall rates of .45 and higher.
+The whole genome cell line dataset (Figure~\ref{fig:datasets}a, Methods section) seems much less challenging than the other dataset, as all methods achieved very high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec}a and \ref{fig:alt-calling_prec-rec_global}a), at recall rates of .45 and higher.
 In comparison to all other tools, ProSolo shows striking increases in recall.
@@
 
-Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}B).
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+Although a relative increase in recall of about 10\% at utmost precision is certainly remarkable, ProSolo demonstrates its power on the second (whole exome) dataset (Figure \ref{fig:datasets}b, Methods section).
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b).
 In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
@@
 While the underestimation of precision equally affects all tools and generally means that benchmarking results are more conservative than with a more accurate (somatic) ground truth, this overestimation of recall in only these modes of two tools does not allow for a fair comparison.
-We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}B, \ref{fig:alt-calling_prec-rec_global}B, and \ref{fig:per-cell_prec-rec_Laehnemann2017} for reference).
+We have thus excluded both SCIPhI's sensitive mode and ProSolo's imputation mode from the discussion of the whole exome dataset with its germline ground truth (but their results are nevertheless displayed in Figures~\ref{fig:alt-calling_prec-rec}b, \ref{fig:alt-calling_prec-rec_global}b, and \ref{fig:per-cell_prec-rec_Laehnemann2017} for reference).
 
-For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
+For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}c, Methods section).
 On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
@@
 Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
-While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at reduced precision.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}c), this comes at reduced precision.
 This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
@@
 In addition, adding more cells grows the space of possible tree topologies that SCIPhI explores super-exponentially\cite{lahnemann_eleven_2020}, which will further increase its runtime.
+
+\subsection*{Flexible Control Over The False Discovery Rate}
 
@@
 
-\subsubsection{Estimates of allele dropout rate validate ProSolo's model}
-
-Leveraging our ground truths and using three different ways to calculate the allele dropout rate, we can confirm the general validity of our single-cell event definitions and also explore limitations of the current model.
-For the allele dropout rate, we will focus on the set of sites where the respective ground truth call is heterozygous, as these are the sites where the dropout of one of the alleles can be identified in a non-ambiguous manner.
-
-More details for the three ways in which we calculate allele dropout rates are given in the Supplement (Section~\ref{sec:suppl-adr-calc}).
-Here, we give a short explanation:
+\subsection*{Valid Model Estimates of Allele Dropout Rate}
 
@@
 
-\begin{enumerate*}
- \item At each ground truth heterozygous site, we sum the posterior probabilities of the two allele dropout events defined in ProSolo ("ADO to alt" and "ADO to ref" in Figure~\ref{fig:prosolo_alt-calling}D) to obtain a total allele dropout probability (Equation \ref{eq:ado-posterior-prob}) and use these to compute the expected value of allele dropout across all sites.
- This gives us an expected allele dropout rate ("expected value" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.expected-allele-dropout-rate}).
- \item We genotype all the ground truth heterozygous sites with ProSolo, take the most likely genotype and then compare against the ground truth: heterozygous sites that ProSolo calls as homozygous are counted as dropout sites.
-The number of such sites is divided by the total number of ground truth heterozygous sites where the respective single cell had coverage ("hom at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.hom-at-ground-truth-het-allele-dropout-rate}).
- \item We identify all heterozygous ground truth sites with a coverage of at least 7, where without amplification bias we could be reasonably sure to sample both alleles (for the reasoning see Supplement, Section~\ref{sec:suppl-adr-calc}).
- We then count a site as an allele dropout if there is one allele (reference or alternative) with no read coverage at all, and again divide by the total number of ground truth heterozygous sites where the respective single cell had coverage ("no alt/ref read at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.no-alt-ref-read-at-ground-truth-het-adr}).
-\end{enumerate*}
-
+Leveraging our ground truths and using three different ways to calculate the allele dropout rate, we can confirm the general validity of our single-cell event definitions and also explore limitations of the current model (Methods section).
 The expected allele dropout rates based on the ProSolo probabilities for allele dropout clearly fall into the range of previously published allele dropout rates\cite{wang_clonal_2014,hou_single-cell_2012,xu_single-cell_2012,ling_evaluation_2009,spits_whole-genome_2006,spits_optimization_2006,renwick_proof_2006,lodato_somatic_2015} ("published" in Figure~\ref{fig:allele-dropout}).
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 Finally, ProSolo is the only method that provides a clearly interpretable false discovery rate parameter that can actually control the trade-off between precision and recall (and this false discovery rate is the only parameter any user will have to specify).
-This can be used on alternative allele calls (Figure~\ref{fig:alt-calling_prec-rec}, Supplement Section~\ref{sec:fdr-of-alt-calling}), the main focus of our benchmarking, but also on any of the events pertinent to single cell analysis (such as allele dropout or amplification errors) or combinations thereof (see Figure~\ref{fig:prosolo_alt-calling}D for all events).
+This can be used on alternative allele calls (Figure~\ref{fig:alt-calling_prec-rec}, Supplement Section~\ref{sec:fdr-of-alt-calling}), the main focus of our benchmarking, but also on any of the events pertinent to single cell analysis (such as allele dropout or amplification errors) or combinations thereof (see Figure~\ref{fig:prosolo_alt-calling}d for all events).
 Thereby, beyond just calling alternative allele presence in a statistically reliable way, ProSolo can, for example, also compute the expected allele dropout rate across the entire genome of a particular cell in a robust manner (Equation \ref{eq:ado-posterior-prob}, Figure~\ref{fig:allele-dropout}).
@@
 
-\bibliography{Benchmarking_ProSolo}
+\section*{Methods}
+
+\subsection*{Single Cell Sequencing Model}
+
+Here, we introduce the central innovations of our probabilistic model for genotyping of diploid single cells whose DNA has been subject to whole genome multiple displacement amplification (MDA)\cite{dean_comprehensive_2002}.
+More details and a detailed derivation of all model elements can be found in the Supplement, including a summary of the core model.
+
+We address amplification bias by leveraging a mechanistically motivated, empirically derived model of differential amplification of alleles and assess amplification errors by evaluating single cell samples together with a bulk sample from which the single cell is supposed to stem.
+Regarding the latter, we argue that a bulk sample should be added to single cell sequencing experiments wherever possible: it samples from the same cell population without requiring amplification, and is therefore unaffected by amplification bias and errors and thus makes a particularly useful background sample to address the statistical uncertainties and biases induced by MDA.
+For related work on flexible bulk sequencing sample composition, see Köster et al.\cite{koster_varlociraptor_2020}.
+
+\begin{figure}[!tpb]%figure1
+  \begin{minipage}[t]{.99\linewidth}
+    \begin{minipage}{.47\linewidth}
+      \includegraphics[width=.95\linewidth]{figs/Lodato_hom-ref-model_cov-20.pdf}
+    \end{minipage}
+    \hspace{2ex}
+    \begin{minipage}{.47\linewidth}
+      \includegraphics[width=.95\linewidth]{figs/Lodato_het-model_cov-20.pdf}
+    \end{minipage}
+  \end{minipage}
+  \vspace{2ex}
+  \hspace{1ex}
+  \begin{minipage}[t]{.99\linewidth}
+    \begin{minipage}{.47\linewidth}
+      \includegraphics[width=.95\linewidth]{figs/prosolo_alt-calling_overview.pdf}
+    \end{minipage}
+    \hspace{2ex}
+    \begin{minipage}{.49\linewidth}
+      \includegraphics[width=.95\linewidth]{figs/Event_space.pdf}
+    \end{minipage}
+  \end{minipage}
+  \caption{
+   \textbf{a,b} Exemplary alternative allele read count distributions for sites covered by 20 reads, as derived by Lodato et al.\cite{lodato_somatic_2015}
+   Homozygous reference sites in \textbf{a} are assumed to follow a beta-binomial distribution;
+   sites heterozygous for the alternative allele in \textbf{b} are assumed to follow the linear combination of two symmetrical beta-binomial distributions (dotted and dashed lines).
+   \textbf{c} Toy example of calling the same genomic site in two single cells from the same population that differ in their true underlying allele frequencies for alternative allele C (blue, $\theta_s = 0$ vs.
+$\theta_s = 0.5$).
+   Alternative nucleotide T (orange) is an amplification error.
+   Empirical distributions in A and B account for the  amplification bias, and likelihoods for the alternative allele candidates from the bulk reduce the likelihoods of amplification errors, thereby correctly identifying both the error and the original true mutation.
+   This is formalised with the model in D.
+   \textbf{d} Definition of single cell events based on ProSolo's likelihood density estimates for the spectrum of true underlying alternative allele frequencies in the single cell ($\tilde{\theta}_s$) and the bulk ($\tilde{\theta}_b$).
+   The bulk is always assumed to be a combination of a maximum of two genotypes at a particular site, generating all possible $\theta_b$ (bottom panel).
+   The model further assumes that the bulk sample has sufficient coverage to capture somatic variants.\small\newline
+   ADO -- allele dropout, alt -- alternative, err -- error, het -- heterozygous, hom -- homozygous, ref -- reference.}
+  \label{fig:prosolo_alt-calling}
+\end{figure}
+
+\paragraph{A mechanistically motivated model of amplification bias, trained on data, gives realistic coverage-specific single-cell (genotype) likelihoods.}
+To account for MDA amplification bias up to the complete dropout of individual alleles, we distinguish between two alternative allele frequencies:
+\begin{enumerate*}[label={(\roman*)}]
+  \item The true (but usually unknown) underlying allele frequency at a site in a single cell: $\theta_s$.
+   This can be assigned one of three possible values, namely $\theta_s\in\{0,0.5,1\}$, where 0 and 1 represent the homozygous reference and alternative genotype and $0.5$ a heterozygous genotype.
+   However, the ratio of reads harboring the different alleles from a single cell sequencing experiment does not reflect the true allele frequency, because of the biases induced by amplification. 
+  Instead, the ratio of reads reflects
+  \item the allele frequency after its distortion through amplification bias.
+   For a site with total coverage of $l$ reads, of which $k$ reads bear the alternative allele, the formal definition of this measurable frequency is $\rho_s=\frac{k}{l}$.
+\end{enumerate*}
+
+The goal is to estimate the likelihood density across the three possible underlying allele frequencies in the single cell (we denote $\tilde{\theta}_s$ as the density estimate across all $\theta_s\in\{0.0,0.5,1.0\}$).
+To accurately quantify the uncertainty introduced by amplification bias, we consider
+
+\begin{equation}
+ \mathbf{P}(\rho_s\mid\theta_s)\qquad \text{for}  \quad\theta_s\in\{0,0.5,1\},
+\end{equation}
+
+the probability distributions that reflect the shift from the true allele frequency $\theta_s$ to the distorted allele frequency $\rho_s$, as induced by MDA (Figure \ref{fig:prosolo_alt-calling}c).
+We thus formally describe the statistics of read counts skewed by MDA at all sites, encompassing sites that are homozygous for the reference allele, heterozygous, or homozygous for the alternative allele, to calculate likelihoods for each of these possible true allele frequencies.
+
+To do this, we follow the considerations of Lodato et al.\cite{lodato_somatic_2015} (see Figure S5 and the section ``Modeling MDA-derived alternative read counts'' of the respective supplement for the details), who fitted well-studied probability distributions to empirical distributions they obtained for $\mathbf{P}(\rho_s\mid\theta_s)$ of $\theta_s\in\{0,0.5\}$.
+For sites that are homozygous for the reference allele, amplification bias cannot initially happen.
+However, once an amplification error creates an alternative allele, this can be amplified to large frequencies due to amplification bias.
+Lodato et al.\cite{lodato_somatic_2015} thus consider the empirical distribution $\mathbf{P}(\rho_s\mid\theta_s=0)$ to follow a beta-binomial.
+Effectively, this means that the probability of a non-zero alternative read count  ($\rho_s>0$, which is any alternative read count above $0$ in Figure~\ref{fig:prosolo_alt-calling}a) will be non-zero ($\mathbf{P}(\rho_s\mid\theta_s=0)>0$), merely because of sequencing and amplification errors.
+In contrast, at heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}b), the distribution is dominated by amplification bias.
+Thus, for $\mathbf{P}(\rho_s\mid\theta_s=0.5)$ they found a mixture of two beta-binomial distributions to appropriately fit the empirically observed distributions (Figure~\ref{fig:prosolo_alt-calling}b).
+We further motivate the choice of the beta-binomial distribution mechanistically by an analogy to its generative urn model named after Pólya\cite{eggenberger_uber_1923}:
+Take an urn with two white and two black balls, where each ball represents one strand of each (double-stranded) allele at a heterozygous site.
+In the Pólya urn model, drawing a ball leads to replacement of that ball with two balls of the same color, analogous to a strand copy by the $\Phi29$ polymerase (additional discussion in the Supplement, Section~\ref{sec:polya-urn}).
+Finally, the necessity for a mixture of beta-binomials (Figure~\ref{fig:prosolo_alt-calling}b) becomes evident when contrasting it with the binomial distribution observed at heterozygous sites in bulk experiments.
+Namely, the bulk distribution for reads supporting the alternative allele would narrowly peak at a count of $k=\frac{l}{2}$, which is not the case for the corresponding $k=10$ in Figure~\ref{fig:prosolo_alt-calling}b.
+Instead, the mixture of beta-binomials peaks at the extreme read counts of $k=0$ and $k=l$, highlighting that the dropout of the alternative or reference allele is quite likely to occur.
+Most likely, this mixture arises from differences in the accessibility of different pieces of the genomic DNA for amplification\cite{picher_trueprime_2016} (Supplement, Section~\ref{sec:polya-urn}).
+Finally, for homozygous alternative sites (i.e. $\mathbf{P}(\rho_s\mid\theta_s=1)$), we rely on the symmetry of the cases $\theta_s=0$ and $\theta_s=1$.
+In summary, we obtain the following equations for $\mathbf{P}(\rho_s\mid\theta_s)$:
+
+\begin{equation}
+	\label{eq:BBLodato}
+	\begin{split}
+		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 0 ) 			=~ &\cB\cB(~ k,~ l;~ \alpha(l),~ \beta(l)~ ),\\
+		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = \frac12) =~ &w(l) \times \cB\cB(~ k,~ l;~ \alpha_1(l),~ \alpha_1(l)~ )~ +\\
+                                                       ~&~(~ 1-w(l)~ ) \times \cB\cB(~ k,~ l;~ \alpha_2(l),~ \alpha_2(l)~ ),\\
+		\Prob(\rho_s = \frac{k}{l} \mid \theta_s = 1 ) 			=~ &\cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ).
+	\end{split}
+\end{equation}
+
+Here, $\cB\cB$ represents the beta-binomial probability mass function where $k\in\{0,...,l\}$ with $l$ being the total read coverage of the site that is considered.
+All of parameters $\alpha, \alpha_1,\alpha_2$, $\beta$, and $w$ scale linearly in $l$.
+Allowing to vary distributions through these parameters allows amplification bias to depend on the total coverage of a site, and thereby to vary locally.
+Symmetry of distributions for the homozygous sites ($\theta_s = 0$ in Figure~\ref{fig:prosolo_alt-calling}a, and $\theta_s = 1$) is established by swapping the shape parameters $\alpha(l)$ and $\beta(l)$.
+The distribution for heterozygous sites (Figure~\ref{fig:prosolo_alt-calling}b) corresponds to a mixture of two beta-binomials with shape parameters $\alpha_1(l)=\beta_2(l)$ and $\alpha_2(l)=\beta_2(l)$, where equality of $\alpha(l)$ and $\beta(l)$ yields symmetry of the distributions relative to $k = \frac{l}{2}$ (the expected value for alternative read counts at heterozygous sites).
+Slopes and intersects for the scaling of all parameter values across different choices of $l$ are given in Table~\ref{tab:Lodato}.
+
+\begin{table}[tbp]
+ \caption{
+ Parameters of the multiple displacement amplification (MDA) amplification bias model as presented in Supplementary Figure S5, panels C and D of \cite{lodato_somatic_2015}.
+ The full precision as used in the implementation is given in Supplementary Table~\ref{tab:Lodato-full}
+ }
+ \label{tab:Lodato}
+ \renewcommand{\arraystretch}{1.15}
+
+ \begin{center}
+  \begin{tabular}{lrr}
+   parameter  & slope        & intersect   \\
+   \toprule
+   $\alpha$   & -0.00003 & 0.06857 \\
+   $\beta$    & 0.00745  & 2.36749 \\
+   \midrule
+   $w$        & 0.00055  & 0.54040 \\
+   $\alpha_1$ & 0.05738  & 0.66973 \\
+   $\alpha_2$ & 0.00323  & 0.39926 \\
+   \bottomrule
+  \end{tabular}
+ \end{center}
+\end{table}
+
+\paragraph{Using bulk evidence of alternative alleles to increase the accuracy of variant calls.}
+A bulk sequencing sample of the same cell type from the same organism is a much larger sampling of that cell population than the sequencing of dozens of single cells.
+Unless a single cell is the only one to harbour a particular mutation, a deep enough sequencing of the bulk sample from which the single cell was drawn should contain reads from cells that share the particular mutation with the single cell.
+Thus, considering bulk samples corresponds to drawing unbiased samples of the entire population (of genome copies), in contrast to single cells that correspond to heavily distorted measurements on pairs of copies.
+This establishes a formal, statistical argument why one should consider bulk experiments in single cell sequencing whenever possible: the identification of a mutation in an accompanying bulk sample lends further credibility to the mutation in the single cell, in a data-driven way, without assuming any fixed error rate (Figure~\ref{fig:prosolo_alt-calling}c).
+At the same time, the absence of a single cell candidate mutation in a bulk sample (with sufficient sequencing depth) increases the probability of an amplification error.
+As a consequence, bulk samples can be employed to improve both the sensitivity and specificity of variant calls in the single cell, where increasing depth of coverage of the bulk sample increases the accuracy of the calls.
+
+For our model, we derive likelihood density estimates for all possible alternative allele frequencies in the background bulk sample.
+Given a set of $n$ reads from the bulk ($b$) read data $\boldsymbol{Z}^b  = \{ \boldsymbol{Z}_1^b, \boldsymbol{Z}_2^b, \dots, \boldsymbol{Z}_n^b \}$, and discrete possible allele frequencies $\frac{m}{n}$ ($m \in 0, 1, \dots, n$), we compute the probability of the data given a particular allele frequency as the product of the probabilities of all the reads:
+
+\begin{equation}
+ \label{eq:bulk-prob}
+ L\left(\theta_b = \frac{m}{n} \mid \boldsymbol{Z}^b\right) \propto
+ P\left(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n}\right) =
+ \prod_{j=1}^n P\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)
+\end{equation}
+
+Here, the probability of an individual read, given a particular allele frequency, $\Prob(Z_j^b \mid \theta_b = \frac{m}{n})$, is defined as in Equation \ref{eq:read_probability}, based on the model described in Köster et al.\cite{koster_varlociraptor_2020}.
+Further, as for the single cell sample, when referring to the likelihood density estimates across all possible allele frequencies (as opposed to a particular allele frequency), we denote this with $\tilde{\theta}_b$.
+
+\paragraph{Calculating posterior probabilities for events at single cell sites, including a bulk background sample.}
+With single cell genotype likelihoods adjusted by our empirical amplification bias model (Equation \ref{eq:BBLodato}) and auxiliary evidence on alternative alleles from a bulk sample (Equation \ref{eq:bulk-prob}), we define mutually exclusive single cell events.
+Figure~\ref{fig:prosolo_alt-calling}c gives a simplified illustration of how the combination of likelihood density estimates across all possible allele frequencies $\tilde{\theta}_s$ and $\tilde{\theta}_b$ works for calling a heterozygous and a homozygous genotype in the single cell.
+However, our model fully defines the two-dimensional space of possible underlying alternative allele frequencies in the two samples as:
+
+\begin{equation}
+  \label{eq:event-space}
+  \boldsymbol{E} =\{0, \frac{1}{2}, 1\}_{\tilde{\theta}_s} \times [0, 1]_{\tilde{\theta}_b}.
+\end{equation}
+
+For these definitions, we always assume that---regarding a particular genomic site---the bulk cell population can only consist of a maximum of two sub-populations that are exactly one mutated allele apart from each other (Figure~\ref{fig:prosolo_alt-calling}d, lower panel; for discussion on the validity of this assumption, see Supplementary Section \ref{sec:event-space-def}).
+At allele frequencies $\theta_b \in \{0, 0.5, 1\}$, we assume a homogeneous population of only homozygous reference, heterozygous, or homozygous alternative cells, respectively.
+At frequencies in between, we assume a mixture of heterozygous cells with homozygous reference cells ($\theta_b \in (0,0.5)$), or with homozygous alternative cells ($\theta_b \in (0.5,1)$).
+Bearing this assumption in mind, we define single cell events as mutually exclusive two-dimensional sub-spaces.
+For example, an error-free homozygous alternative site is defined by allele frequency likelihood density estimates across $\boldsymbol{E}_{\text{hom alt}} = \{ 1 \}_{\tilde{\theta}_s} \times (\frac{1}{2}, 1]_{\tilde{\theta}_b}$.
+And the dropout of an alternative allele is defined across allele frequency likelihood density estimates $\boldsymbol{E}_{\text{ADO to ref}} = \{ 0 \}_{\tilde{\theta}_s} \times [\frac{1}{2}, 1)_{\tilde{\theta}_b}$ (Figure~\ref{fig:prosolo_alt-calling}d, Table~\ref{tab:Events}).
+For the latter, please note that we here resolve a contradiction between $\tilde{\theta}_s = 0$, which indicates a homozygous reference single cell, and $\tilde{\theta}_b \in [0.5,1)$, which indicates that the bulk does not contain any homozygous reference cells.
+We decide to trust the bulk sample over the single cell sample, and with our above assumption that the bulk is a mixture of a maximum of two sub-populations, the bulk can only contain heterozygous and homozygous alternative cells.
+This renders a homozygous reference single cell impossible, and we classify this event as evidence for an allele dropout of the alternative allele.
+
+We thus obtain a set of mutually exclusive single cell events (Figure~\ref{fig:prosolo_alt-calling}d):
+
+\begin{equation}
+  \label{eq:event-set}
+  \begin{split}
+    \mathfrak{E} &= \{ \boldsymbol{E}_{\text{hom ref}}, \boldsymbol{E}_{\text{err alt}}, \boldsymbol{E}_{\text{ADO to alt}}, \boldsymbol{E}_{\text{het}}, \boldsymbol{E}_{\text{ADO to ref}}, \boldsymbol{E}_{\text{err ref}}, \boldsymbol{E}_{\text{hom alt}} \}
+  \end{split}
+\end{equation}
+
+Assuming a flat prior across the possible underlying allele frequencies for both the bulk and the single cell, we can compute likelihoods for all those single cell events (e.g.~Equation~\ref{eq:ado-to-ref-likelihood-data}).
+The sum of the likelihoods of all these (mutually exclusive) events yields the marginal probability (Equation \ref{eq:marginal-prob}).
+Using the marginal probability, we can calculate the posterior probability for any of these events.
+For example, the posterior probability of event $\boldsymbol{E}_{\text{hom alt}}$ (Figure~\ref{fig:prosolo_alt-calling}d) can be calculated with:
+
+\begin{equation}
+ \label{eq:hom-alt-posterior-samples}
+ \begin{split}
+  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
+    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
+      \Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_{\text{hom alt}}) \\
+    =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times
+      \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) \times \Prob(\boldsymbol{Z}^b \mid \theta_b \in (\frac12,1])\\
+  \end{split}    
+\end{equation}
+
+Accounting for the sample likelihoods based on Equation \ref{eq:theta-likelihood} (assuming $\rho_b = \theta_b$ for the bulk that has no amplification step, Equation \ref{eq:rho-b}), and evaluating only point estimates of these likelihoods at possible alternative allele frequencies, this gives:
+
+\begin{equation}
+ \label{eq:hom-alt-posterior-likelihoods}
+ \begin{split}   
+ \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
+   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
+     &\sum_{k=0}^{l} \left\{ \Prob(\boldsymbol{Z}^s \mid \rho_s = \frac{k}{l}) \times \Prob(\rho_s = \frac{k}{l} \mid \theta_s =  1)\right\} \times \\
+     &\sum_{m=\frac{n}{2}}^{n} \Prob(\boldsymbol{Z}^b \mid \theta_b = \frac{m}{n})\\
+  \end{split}    
+\end{equation}
+
+Accounting for amplification bias with Equation \ref{eq:BBLodato} and computing the likelihood of the sample-specific allele frequency ranges with Equations \ref{eq:read-product} and \ref{eq:bulk-prob}, this becomes:
+
+\begin{equation}
+ \label{eq:hom-alt-posterior-reads}
+ \begin{split}   
+ \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
+   =~&\frac{1}{\sum_{\boldsymbol{E}_e \in \mathfrak{E}}{\Prob(\boldsymbol{Z}^s,\boldsymbol{Z}^b \mid \boldsymbol{E}_e )}} \times \\
+      &\sum_{k=0}^{l} \left\{ \prod_{i=1}^l \Prob(Z_i^s \mid \rho_s = \frac{k}{l}) \times \cB\cB(~ k,~ l;~ \beta(l),~ \alpha(l)~ ) \right\} \times\\
+      &\sum_{m=\frac{n}{2}}^{n} \prod_{j=1}^l \Prob\left(Z_j^b \mid \theta_b = \frac{m}{n} \right)\\
+  \end{split}    
+\end{equation}
+
+With read probabilities calculated with Equation \ref{eq:read_probability}, we can calculate the posterior probability of the $\boldsymbol{E}_{\text{hom alt}}$ event (for an analogous and more detailed derivation for $\boldsymbol{E}_{\text{ADO to ref}}$, see Supplement, Section~\ref{sec:sc-model}).
+To obtain posterior probabilities for compound events, we sum up the posterior probabilities of the events it comprises, effectively joining up their respective allele frequency range combinations into compound range combinations.
+For the site-specific single cell probability for the presence of an alternative allele in the single cell, we thus get the compound event (blue events in Figure~\ref{fig:prosolo_alt-calling}d; Table~\ref{tab:Events}):
+
+\begin{equation}
+ \label{eq:compound-event-alt-presence}
+ \begin{split}
+  \boldsymbol{E}_{\text{alt presence}_s}
+  =~&\boldsymbol{E}_\text{ADO to ref}~~\cup~~
+    \boldsymbol{E}_\text{err ref}~~\cup~~
+    \boldsymbol{E}_\text{het}~~\cup~~
+    \boldsymbol{E}_\text{ADO to alt}~~\cup~~
+    \boldsymbol{E}_\text{hom alt} \\
+  =~&\{0\}_{\theta_s} \times [\frac{1}{2}, 1)_{\theta_b}~~\cup~~
+    \{0, \frac{1}{2}\}_{\theta_s} \times [1]_{\theta_b}~~\cup~~
+    \{\frac{1}{2}\}_{\theta_s} \times (0,1)_{\theta_b}~~\cup~~\\
+    &\{1\}_{\theta_s} \times (0,\frac{1}{2}]_{\theta_b}~~\cup~~
+    \{1\}_{\theta_s} \times (\frac{1}{2},1]_{\theta_b} \\
+  =~&\{0\}_{\theta_s} \times [\frac{1}{2},1]_{\theta_b}~~\cup~~
+    \{\frac{1}{2},1\}_{\theta_s} \times (0,1]_{\theta_b}
+ \end{split}
+\end{equation}
+
+whose posterior probability we can obtain from this sum:
+
+\begin{equation}
+ \label{eq:alt-posterior-prob}
+ \begin{split}
+  \Prob(\text{alt presence}_s \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b) =~&
+    \Prob(\boldsymbol{E}_{\text{ADO to ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~ 
+    \Prob(\boldsymbol{E}_{\text{err ref}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+ \\
+  &\Prob(\boldsymbol{E}_{\text{het}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
+  \Prob(\boldsymbol{E}_{\text{ADO to alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)~+~
+  \Prob(\boldsymbol{E}_{\text{hom alt}} \mid \boldsymbol{Z}^s,\boldsymbol{Z}^b)
+ \end{split}
+\end{equation}
+
+To genotype, we calculate the posterior probability of all three possible single-cell genotypes and choose the genotype with the maximum posterior probability. 
+In Figure~\ref{fig:prosolo_alt-calling}d, events are colored purple when implying the homozygous reference genotype, light blue when implying the heterozygous genotype and dark blue when implying the homozygous alternative genotype (Table \ref{tab:Events}).
+Similarly, to calculate the posterior probability of an allele dropout at a particular site, we sum up the posterior probabilities of the two {\ttfamily ADO} events.
+
+For any such compound event, we can estimate a threshold on the posterior probabilities that controls for a specified false discovery rate. 
+This is based on the approach described by Müller et al.\cite{muller_optimal_2004,muller_fdr_2006}---for further details see the Supplement (Section~\ref{sec:fdr}) and Köster et al\cite{koster_varlociraptor_2020}.
+
+\paragraph{Biologically relevant imputation based on the bulk sample.}
+As argued above, a large enough sampling of the bulk cell population that the single cell comes from should contain the single cell's genotype at a particular site, unless this cell is genuinely the first cell to harbour a mutation at that site.
+This bulk background sample can thereby render credibility to single cell variants with low coverage, while at the same time eliminating amplification errors in the single cell sample, as these will not exist in the bulk sample.
+Interestingly, the bulk sample also provides a mechanism of biologically meaningful imputation at sites that have no read coverage in the single cell.
+If imputation is desired for sites with no read coverage in the single cell, we set $\Prob(\boldsymbol{Z}^s \mid \theta_s = 0) = \Prob(\boldsymbol{Z}^s \mid \theta_s = \frac{1}{2}) = \Prob(\boldsymbol{Z}^s \mid \theta_s = 1) = 1$, rendering all (unknown underlying) single cell genotypes equally likely.
+Thus, the posterior probabilities of events at sites with no read coverage become solely dependent on the read data from the bulk sample, providing the most common genotype in the bulk.
+However, while this is a biologically meaningful way of imputation at the vast majority of genomic sites, it should be noted that this imputation will favor germline genotypes over any existing (lower frequency) somatic genotypes at a site, unless such a somatic genotype is present in a majority of cells.
+Thus, such an imputation carries the potential to introduce erroneous calls (especially when looking at subclonal somatic mutations), and we recommend to instead use downstream tools that can accommodate for missing data and data uncertainty wherever they are available.
+
+\paragraph{ProSolo is an easy-to-use command-line tool, based on a modular framework.}
+
+ProSolo is an easy-to-use command-line tool---following usability standards\cite{taschuk_ten_2017}---and its source code is available at \url{https://github.com/prosolo/prosolo}, including instructions for an easy installation via Bioconda\cite{gruning_bioconda:_2018}.
+Its main contribution in terms of software is the implementation of its comprehensive statistical model into the Rust variant calling library of Varlociraptor\cite{koster_varlociraptor_2020}.
+See the Supplement (Section~\ref{sec:implementation}) for further implementation details.
+
+\subsection*{Benchmarking}
+
+We compare ProSolo to the state-of-the-art for SNV calling from single cell sequencing data of multiple displacement amplified (MDA) DNA: MonoVar\cite{zafar_monovar:_2016}, SCAN-SNV\cite{luquette_identification_2019}, SCcaller\cite{dong_accurate_2017} and SCIPhI\cite{singer_single-cell_2018}.
+We used Snakemake\cite{koster_snakemakescalable_2012} (version {\ttfamily 6.3.0}) to implement the benchmarking workflows.
+For detailed information on benchmarking setup and results, see Supplementary Section~\ref{sec:suppl-benchmarking}.
+All code used for benchmarking is available at: \url{https://github.com/prosolo/benchmarking_prosolo} (or as preserved by Zenodo at: \url{https://doi.org/10.5281/zenodo.3769115}).
+
+\subsubsection*{Datasets and Ground Truths}
+
+\begin{figure}[!tpb]
+  \begin{minipage}{.3\linewidth}
+    \includegraphics[height=38ex]{figs/Dong2017_data_overview.pdf} \\
+  \end{minipage}
+  \hspace{6ex}
+  \begin{minipage}{.5\linewidth}
+    \includegraphics[height=38ex]{figs/Laehnemann2017_data_overview.pdf} \\
+  \end{minipage}\\
+  \begin{minipage}{.5\linewidth}
+    \includegraphics[height=38ex]{figs/Wang2014_data_overview.pdf} \\
+  \end{minipage}
+  \hspace{6ex}
+  \begin{minipage}{.3\linewidth}
+    \includegraphics[height=38ex]{figs/data_overview_color_legend.pdf} \\
+  \end{minipage}
+  \caption{
+   Benchmark Datasets.
+   For single cells, DNA was multiple displacement amplified (MDA).
+   \textbf{a} Whole genome sequencing (WGS) dataset, generated from a clonal population started with an individual cell\cite{dong_accurate_2017}, and expanded further downstream to generate different bulk samples (C1, C2, C3, IL-1C).
+   \textbf{b} Newly generated whole exome sequencing (WXS) dataset of blood cells.
+   Ground truth genotypes of patient CCS1 were determined from sequencing data of family members (boxes male; circles female).
+   Granulocytes were isolated from blood using magnetic and fluorescence activated cell sorting (MACS and FACS).
+   \textbf{c} Whole exome sequencing (WXS) dataset from a triple negative breast cancer (TNBC) patient\cite{wang_clonal_2014}.
+   For tumor cells (a1 to a8 and h1 to h8), the cell population (pop) sequencing of the normal sample was taken as the ground truth, including known clonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
+   For normal cells (n1 to n16), the cell population (pop) sequencing of the tumor sample was taken as the ground truth, after removing known clonal and subclonal tumor-specific somatic variants previously validated by deep targeted duplex (td, dx) sequencing\cite{wang_clonal_2014}.
+   }
+  \label{fig:datasets}
+\end{figure}
+
+We benchmarked ProSolo on three experimental datasets (Figure~\ref{fig:datasets}), each with a different kind of ground truth:\\
+
+\paragraph{Whole genome sequencing of almost identical kin-cells from a cell line \cite{dong_accurate_2017}.}
+The first dataset comes from the publication of the SCcaller software\cite{dong_accurate_2017} (Figure~\ref{fig:datasets}a, dataset available from SRA project \mbox{PRJNA305211} [\url{https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA305211}], accessions \mbox{SRR2976561} to \mbox{SRR2976566}).
+A single starting cell was grown in two steps (Figure 1 of the original paper\cite{dong_accurate_2017}):
+After an initial mini-expansion, a single cell was selected as the founder for the secondary IL expansion into 20-30 cells.
+From this, two cells were extracted (IL-11 and IL-12) and sequenced following MDA.
+The remaining kindred cells from that clone were used as a bulk sequencing sample without amplification (IL-1C).
+IL-1C serves as the ground truth, as these cells are only very few cell divisions away from IL-11 and IL-12, and thus have almost no difference in the somatic mutations acquired.
+The ground truth genotype was generated using GATK HaplotypeCaller to call variant sites and bcftools mpileup to identify homozygous reference sites (with read coverage above 25 but no alternative allele present).
+IL-1C was only used as a ground truth and not provided as input to any of the software compared here.
+Three more distant clones (C1, C2, C3), generated from cells after the first mini-expansion, were merged into a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
+Unlike other callers (all of which finished in less than 5 days), SCIPhI took five weeks to finish on this dataset in sensitive mode and 7.5 weeks in default mode.\\
+
+\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
+\begin{sloppypar}
+For the second benchmarking dataset, blood was taken from a patient with a constitutional mismatch repair-deficiency\cite{hoell_constitutional_2014} after informed consent.
+Data acquisition protocols were approved by the Ethics Committee of the Medical Faculty of the Heinrich Heine University Düsseldorf.
+Granulocytes were selected via Magnetic-Activated Cell Sorting (MACS) and Fluorescence Activated Cell Sorting (FACS, Figure~\ref{fig:datasets}b, dataset available as \mbox{EGAD00001005929} [\url{https://ega-archive.org/datasets/EGAD00001005929}] in EGA study \mbox{EGAS00001004123}).
+Individual cells were isolated using a microfluidics device and subjected to multiple displacement amplification (MDA).
+Using a panel of 16 primer pairs covering different genes across chromosomes (Table~\ref{tab:selection-primers}) for quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
+For those cells, we performed whole exome capture, sequencing library preparation and paired-end Illumina sequencing.
+From the remaining sorted cell population, we also extracted bulk DNA and submitted it to whole exome capture and paired-end Illumina sequencing without MDA, to generate a bulk background sample for ProSolo and SCcaller.
+\end{sloppypar}
+
+To generate the ground truth of this dataset, we could leverage previously sequenced bulk whole exome data from the same person, their parents and three siblings \cite{hoell_constitutional_2014} (Figure~\ref{fig:datasets}b).
+To create ground truth germline alternative allele calls, we ran three pedigree-aware variant callers (BEAGLE4.0\cite{browning_improving_2013}, polymutt\cite{li_likelihood-based_2012} and FamSeq\cite{peng_rare_2013,peng_famseq:_2014}) and created a consensus by including only calls where all callers agree at a site and where a maximum of one caller not calling the site was allowed (Figure~\ref{fig:granulocytes-ground-truth}).
+
+\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC)\cite{wang_clonal_2014}.}
+The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available (Figure~\ref{fig:datasets}c, dataset available via the sequence read archive, project id \mbox{PRJNA168068} [\url{https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA168068}] or accession \mbox{SRA053195}).
+For each, 16 single cell nuclei and a bulk sample were whole exome sequenced.
+Further, a number of candidate sites for clonal and subclonal tumor mutations was confirmed by targeted duplex deep sequencing in a separate experiment\cite{wang_clonal_2014}.
+We analyzed the tumor and normal cells separately, ensuring that normal cells were called with the normal bulk background sample and tumor cells with the tumor bulk background sample.
+As the ground truth for the tumor cells, we then used the normal bulk sample augmented with the clonal tumor mutations confirmed by targeted duplex sequencing.
+As the ground truth for the normal cells, we used the tumor bulk sample, removing the confirmed clonal and subclonal tumor mutations.
+This experimental setup aims for fairness across all competitors. 
+Variant calling for the ground truths was performed as for cell line data above.
+
+\subsubsection*{Estimates of allele dropout rate validate ProSolo's model}
+
+For the allele dropout rate, we will focus on the set of sites where the respective ground truth call is heterozygous, as these are the sites where the dropout of one of the alleles can be identified in a non-ambiguous manner.
+More details for the three ways in which we calculate allele dropout rates are given in the Supplement (Section~\ref{sec:suppl-adr-calc}).
+Here, we give a short explanation:
+
+\begin{enumerate*}
+ \item At each ground truth heterozygous site, we sum the posterior probabilities of the two allele dropout events defined in ProSolo ("ADO to alt" and "ADO to ref" in Figure~\ref{fig:prosolo_alt-calling}d) to obtain a total allele dropout probability (Equation \ref{eq:ado-posterior-prob}) and use these to compute the expected value of allele dropout across all sites.
+ This gives us an expected allele dropout rate ("expected value" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.expected-allele-dropout-rate}).
+ \item We genotype all the ground truth heterozygous sites with ProSolo, take the most likely genotype and then compare against the ground truth: heterozygous sites that ProSolo calls as homozygous are counted as dropout sites.
+The number of such sites is divided by the total number of ground truth heterozygous sites where the respective single cell had coverage ("hom at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.hom-at-ground-truth-het-allele-dropout-rate}).
+ \item We identify all heterozygous ground truth sites with a coverage of at least 7, where without amplification bias we could be reasonably sure to sample both alleles (for the reasoning see Supplement, Section~\ref{sec:suppl-adr-calc}).
+ We then count a site as an allele dropout if there is one allele (reference or alternative) with no read coverage at all, and again divide by the total number of ground truth heterozygous sites where the respective single cell had coverage ("no alt/ref read at ground truth het" in Figure~\ref{fig:allele-dropout}, Equation \ref{eq.no-alt-ref-read-at-ground-truth-het-adr}).
+\end{enumerate*}
+
+\section*{Data Availability}
+
+Lists of accessions for all samples used are given in the accompanying GitHub repository, namely at: \url{https://github.com/ProSolo/benchmarking_prosolo/tree/v5.0/analysis_pipelines#dataset}.
+This repository is also archived via Zenodo: \url{https://doi.org/10.5281/zenodo.3769115}.
+
+Dong et al.
+(2017)\cite{dong_accurate_2017} provide the first dataset  (Figure~\ref{fig:datasets}a) via the Sequence Read Archive as SRA project \mbox{PRJNA305211} (accessions \mbox{SRR2976561} to \mbox{SRR2976566}): \url{https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA305211}.
+
+We provide the second dataset alongside this publication (Figure~\ref{fig:datasets}b) via the European Genome-Phenome Archive as EGA dataset \mbox{EGAD00001005929}: \url{https://ega-archive.org/datasets/EGAD00001005929}.
+As this is patient data, interested researchers need to contact the data access committee of this dataset via the EGA platform and sign a license agreement to gain data access: \url{https://ega-archive.org/dacs/EGAC00001001468}.
+
+Wang et al.
+(2014)\cite{wang_clonal_2014} provide the third dataset (Figure~\ref{fig:datasets}c) via the Sequence Read Archive as SRA project \mbox{PRJNA168068} (accession \mbox{SRA053195}): \url{https://www.ncbi.nlm.nih.gov/sra/?term=PRJNA168068}.
+
+\section*{Code Availability}
+
+All code used in data analysis and preparation of the manuscript, alongside a description of necessary steps for reproducing results, can be found in a GitHub repository accompanying this manuscript: \url{https://github.com/prosolo/benchmarking_prosolo}.
+To ensure its long-term availability and make it citable, we have created a release of this repository and archived it on Zenodo: \url{https://doi.org/10.5281/zenodo.3769115}.
+
+Further, the code of ProSolo itself is in the following GitHub repository: \url{https://github.com/prosolo/prosolo}.
+The version of the code used in this study has been archived under: \url{https://doi.org/10.5281/zenodo.5366555}.
+
+\begin{thebibliography}{10}
+\expandafter\ifx\csname url\endcsname\relax
+  \def\url#1{\texttt{#1}}\fi
+\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi
+\expandafter\ifx\csname doiprefix\endcsname\relax\def\doiprefix{DOI }\fi
+\providecommand{\bibinfo}[2]{#2}
+\providecommand{\eprint}[2][]{\url{#2}}
+
+\bibitem{forsberg_mosaicism_2017}
+\bibinfo{author}{Forsberg, L.~A.}, \bibinfo{author}{Gisselsson, D.} \&
+  \bibinfo{author}{Dumanski, J.~P.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Mosaicism in health and disease
+  {\textemdash} clones picking up speed}}.
+\newblock {\emph{\JournalTitle{Nature Reviews Genetics}}}
+  \textbf{\bibinfo{volume}{18}}, \bibinfo{pages}{128--142}
+  (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1038/nrg.2016.145.
+
+\bibitem{lynch_rate_2010}
+\bibinfo{author}{Lynch, M.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Rate, molecular spectrum, and
+  consequences of human mutation}}.
+\newblock {\emph{\JournalTitle{Proceedings of the National Academy of
+  Sciences}}} \textbf{\bibinfo{volume}{107}}, \bibinfo{pages}{961--968}
+  (\bibinfo{year}{2010}).
+\newblock \doiprefix 10.1073/pnas.0912629107.
+
+\bibitem{lynch_evolution_2010}
+\bibinfo{author}{Lynch, M.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Evolution of the mutation rate}}.
+\newblock {\emph{\JournalTitle{Trends in Genetics}}}
+  \textbf{\bibinfo{volume}{26}}, \bibinfo{pages}{345--352}
+  (\bibinfo{year}{2010}).
+\newblock \doiprefix 10.1016/j.tig.2010.05.003.
+
+\bibitem{tomasetti_half_2013}
+\bibinfo{author}{Tomasetti, C.}, \bibinfo{author}{Vogelstein, B.} \&
+  \bibinfo{author}{Parmigiani, G.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Half or more of the somatic
+  mutations in cancers of self-renewing tissues originate prior to tumor
+  initiation}}.
+\newblock {\emph{\JournalTitle{Proceedings of the National Academy of
+  Sciences}}} \textbf{\bibinfo{volume}{110}}, \bibinfo{pages}{1999--2004}
+  (\bibinfo{year}{2013}).
+\newblock \doiprefix 10.1073/pnas.1221068110.
+
+\bibitem{li_somatic_2013}
+\bibinfo{author}{Li, R.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Somatic point mutations occurring
+  early in development: a monozygotic twin study}}.
+\newblock {\emph{\JournalTitle{Journal of Medical Genetics}}}
+  \bibinfo{pages}{jmedgenet--2013--101712} (\bibinfo{year}{2013}).
+\newblock \doiprefix 10.1136/jmedgenet-2013-101712.
+
+\bibitem{milholland_differences_2017}
+\bibinfo{author}{Milholland, B.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Differences between germline and
+  somatic mutation rates in humans and mice}}.
+\newblock {\emph{\JournalTitle{Nature Communications}}}
+  \textbf{\bibinfo{volume}{8}}, \bibinfo{pages}{15183} (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1038/ncomms15183.
+
+\bibitem{kester_single-cell_2018}
+\bibinfo{author}{Kester, L.} \& \bibinfo{author}{van Oudenaarden, A.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Single-{Cell} {Transcriptomics}
+  {Meets} {Lineage} {Tracing}}}.
+\newblock {\emph{\JournalTitle{Cell Stem Cell}}} \textbf{\bibinfo{volume}{23}},
+  \bibinfo{pages}{166--179} (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1016/j.stem.2018.04.014.
+
+\bibitem{ross_onconem:_2016}
+\bibinfo{author}{Ross, E.~M.} \& \bibinfo{author}{Markowetz, F.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{OncoNEM}: inferring tumor
+  evolution from single-cell sequencing data}}.
+\newblock {\emph{\JournalTitle{Genome Biology}}} \textbf{\bibinfo{volume}{17}},
+  \bibinfo{pages}{69} (\bibinfo{year}{2016}).
+\newblock \doiprefix 10.1186/s13059-016-0929-9.
+
+\bibitem{zafar_sifit:_2017}
+\bibinfo{author}{Zafar, H.}, \bibinfo{author}{Tzen, A.},
+  \bibinfo{author}{Navin, N.}, \bibinfo{author}{Chen, K.} \&
+  \bibinfo{author}{Nakhleh, L.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{SiFit}: inferring tumor trees from
+  single-cell sequencing data under finite-sites models}}.
+\newblock {\emph{\JournalTitle{Genome Biology}}} \textbf{\bibinfo{volume}{18}},
+  \bibinfo{pages}{178} (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1186/s13059-017-1311-2.
+
+\bibitem{malikic_integrative_2019}
+\bibinfo{author}{Malikic, S.}, \bibinfo{author}{Jahn, K.},
+  \bibinfo{author}{Kuipers, J.}, \bibinfo{author}{Sahinalp, S.~C.} \&
+  \bibinfo{author}{Beerenwinkel, N.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Integrative inference of subclonal
+  tumour evolution from single-cell and bulk sequencing data}}.
+\newblock {\emph{\JournalTitle{Nature Communications}}}
+  \textbf{\bibinfo{volume}{10}}, \bibinfo{pages}{2750} (\bibinfo{year}{2019}).
+\newblock \doiprefix 10.1038/s41467-019-10737-5.
+
+\bibitem{kuipers_advances_2017}
+\bibinfo{author}{Kuipers, J.}, \bibinfo{author}{Jahn, K.} \&
+  \bibinfo{author}{Beerenwinkel, N.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Advances in understanding tumour
+  evolution through single-cell~sequencing}}.
+\newblock {\emph{\JournalTitle{Biochimica et Biophysica Acta (BBA) - Reviews on
+  Cancer}}} \textbf{\bibinfo{volume}{1867}}, \bibinfo{pages}{127--138}
+  (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1016/j.bbcan.2017.02.001.
+
+\bibitem{skums_inference_2019}
+\bibinfo{author}{Skums, P.}, \bibinfo{author}{Tsyvina, V.} \&
+  \bibinfo{author}{Zelikovsky, A.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Inference of clonal selection in
+  cancer populations using single-cell sequencing data}}.
+\newblock {\emph{\JournalTitle{Bioinformatics}}} \textbf{\bibinfo{volume}{35}},
+  \bibinfo{pages}{i398--i407} (\bibinfo{year}{2019}).
+\newblock \doiprefix 10.1093/bioinformatics/btz392.
+
+\bibitem{lahnemann_eleven_2020}
+\bibinfo{author}{L{\"a}hnemann, D.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Eleven grand challenges in
+  single-cell data science}}.
+\newblock {\emph{\JournalTitle{Genome Biology}}} \textbf{\bibinfo{volume}{21}},
+  \bibinfo{pages}{31} (\bibinfo{year}{2020}).
+\newblock \doiprefix 10.1186/s13059-020-1926-6.
+
+\bibitem{dean_comprehensive_2002}
+\bibinfo{author}{Dean, F.~B.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Comprehensive human genome
+  amplification using multiple displacement amplification}}.
+\newblock {\emph{\JournalTitle{Proceedings of the National Academy of
+  Sciences}}} \textbf{\bibinfo{volume}{99}}, \bibinfo{pages}{5261--5266}
+  (\bibinfo{year}{2002}).
+\newblock \doiprefix 10.1073/pnas.082089499.
+
+\bibitem{de_bourcy_quantitative_2014}
+\bibinfo{author}{de~Bourcy, C. F.~A.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{A {Quantitative} {Comparison} of
+  {Single}-{Cell} {Whole} {Genome} {Amplification} {Methods}}}.
+\newblock {\emph{\JournalTitle{PLoS ONE}}} \textbf{\bibinfo{volume}{9}},
+  \bibinfo{pages}{e105585} (\bibinfo{year}{2014}).
+\newblock \doiprefix 10.1371/journal.pone.0105585.
+
+\bibitem{hou_comparison_2015}
+\bibinfo{author}{Hou, Y.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Comparison of variations detection
+  between whole-genome amplification methods used in single-cell
+  resequencing}}.
+\newblock {\emph{\JournalTitle{GigaScience}}} \textbf{\bibinfo{volume}{4}},
+  \bibinfo{pages}{37} (\bibinfo{year}{2015}).
+\newblock \doiprefix 10.1186/s13742-015-0068-3.
+
+\bibitem{huang_single-cell_2015}
+\bibinfo{author}{Huang, L.}, \bibinfo{author}{Ma, F.},
+  \bibinfo{author}{Chapman, A.}, \bibinfo{author}{Lu, S.} \&
+  \bibinfo{author}{Xie, X.~S.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Single-{Cell} {Whole}-{Genome}
+  {Amplification} and {Sequencing}: {Methodology} and {Applications}}}.
+\newblock {\emph{\JournalTitle{Annual Review of Genomics and Human Genetics}}}
+  \textbf{\bibinfo{volume}{16}}, \bibinfo{pages}{79--102}
+  (\bibinfo{year}{2015}).
+\newblock \doiprefix 10.1146/annurev-genom-090413-025352.
+
+\bibitem{estevez-gomez_comparison_2018}
+\bibinfo{author}{Est{\'e}vez-G{\'o}mez, N.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Comparison of single-cell
+  whole-genome amplification strategies}}.
+\newblock {\emph{\JournalTitle{bioRxiv}}} \bibinfo{pages}{443754}
+  (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1101/443754.
+
+\bibitem{paez_genome_2004}
+\bibinfo{author}{Paez, J.~G.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Genome coverage and sequence
+  fidelity of $\phi$29 polymerase-based multiple strand displacement whole
+  genome amplification}}.
+\newblock {\emph{\JournalTitle{Nucleic Acids Research}}}
+  \textbf{\bibinfo{volume}{32}}, \bibinfo{pages}{e71--e71}
+  (\bibinfo{year}{2004}).
+\newblock \doiprefix 10.1093/nar/gnh069.
+
+\bibitem{wang_genome-wide_2012}
+\bibinfo{author}{Wang, J.}, \bibinfo{author}{Fan, H.~C.},
+  \bibinfo{author}{Behr, B.} \& \bibinfo{author}{Quake, S.~R.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Genome-wide {Single}-{Cell}
+  {Analysis} of {Recombination} {Activity} and {De} {Novo} {Mutation} {Rates}
+  in {Human} {Sperm}}}.
+\newblock {\emph{\JournalTitle{Cell}}} \textbf{\bibinfo{volume}{150}},
+  \bibinfo{pages}{402--412} (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1016/j.cell.2012.06.030.
+
+\bibitem{wang_clonal_2014}
+\bibinfo{author}{Wang, Y.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Clonal evolution in breast cancer
+  revealed by single nucleus genome sequencing}}.
+\newblock {\emph{\JournalTitle{Nature}}} \textbf{\bibinfo{volume}{512}},
+  \bibinfo{pages}{155--160} (\bibinfo{year}{2014}).
+\newblock \doiprefix 10.1038/nature13600.
+
+\bibitem{lodato_somatic_2015}
+\bibinfo{author}{Lodato, M.~A.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Somatic mutation in single human
+  neurons tracks developmental and transcriptional history}}.
+\newblock {\emph{\JournalTitle{Science}}} \textbf{\bibinfo{volume}{350}},
+  \bibinfo{pages}{94--98} (\bibinfo{year}{2015}).
+\newblock \doiprefix 10.1126/science.aab1785.
+
+\bibitem{de_paz_high-resolution_2018}
+\bibinfo{author}{de~Paz, A.~M.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{High-resolution mapping of {DNA}
+  polymerase fidelity using nucleotide imbalances and next-generation
+  sequencing}}.
+\newblock {\emph{\JournalTitle{Nucleic Acids Research}}}
+  \textbf{\bibinfo{volume}{46}}, \bibinfo{pages}{e78--e78}
+  (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1093/nar/gky296.
+
+\bibitem{baumer_exploring_2018}
+\bibinfo{author}{B{\"a}umer, C.}, \bibinfo{author}{Fisch, E.},
+  \bibinfo{author}{Wedler, H.}, \bibinfo{author}{Reinecke, F.} \&
+  \bibinfo{author}{Korfhage, C.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Exploring {DNA} quality of single
+  cells for genome analysis with simultaneous whole-genome amplification}}.
+\newblock {\emph{\JournalTitle{Scientific Reports}}}
+  \textbf{\bibinfo{volume}{8}}, \bibinfo{pages}{1--10} (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1038/s41598-018-25895-7.
+
+\bibitem{picher_trueprime_2016}
+\bibinfo{author}{Picher, {\'A}.~J.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{TruePrime} is a novel method for
+  whole-genome amplification from single cells based on
+  \textit{{Tth}}{PrimPol}}}.
+\newblock {\emph{\JournalTitle{Nature Communications}}}
+  \textbf{\bibinfo{volume}{7}}, \bibinfo{pages}{13296} (\bibinfo{year}{2016}).
+\newblock \doiprefix 10.1038/ncomms13296.
+
+\bibitem{spits_optimization_2006}
+\bibinfo{author}{Spits, C.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Optimization and evaluation of
+  single-cell whole-genome multiple displacement amplification}}.
+\newblock {\emph{\JournalTitle{Human Mutation}}} \textbf{\bibinfo{volume}{27}},
+  \bibinfo{pages}{496--503} (\bibinfo{year}{2006}).
+\newblock \doiprefix 10.1002/humu.20324.
+
+\bibitem{zafar_monovar:_2016}
+\bibinfo{author}{Zafar, H.}, \bibinfo{author}{Wang, Y.},
+  \bibinfo{author}{Nakhleh, L.}, \bibinfo{author}{Navin, N.} \&
+  \bibinfo{author}{Chen, K.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Monovar: single-nucleotide variant
+  detection in single cells}}.
+\newblock {\emph{\JournalTitle{Nature Methods}}} \textbf{\bibinfo{volume}{13}},
+  \bibinfo{pages}{505--507} (\bibinfo{year}{2016}).
+\newblock \doiprefix 10.1038/nmeth.3835.
+
+\bibitem{dong_accurate_2017}
+\bibinfo{author}{Dong, X.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Accurate identification of
+  single-nucleotide variants in whole-genome-amplified single cells}}.
+\newblock {\emph{\JournalTitle{Nature Methods}}} \textbf{\bibinfo{volume}{14}},
+  \bibinfo{pages}{491--493} (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1038/nmeth.4227.
+
+\bibitem{singer_single-cell_2018}
+\bibinfo{author}{Singer, J.}, \bibinfo{author}{Kuipers, J.},
+  \bibinfo{author}{Jahn, K.} \& \bibinfo{author}{Beerenwinkel, N.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Single-cell mutation identification
+  via phylogenetic inference.}}
+\newblock {\emph{\JournalTitle{Nature communications}}}
+  \textbf{\bibinfo{volume}{9}}, \bibinfo{pages}{5144--5144}
+  (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1038/s41467-018-07627-7.
+
+\bibitem{luquette_identification_2019}
+\bibinfo{author}{Luquette, L.~J.}, \bibinfo{author}{Bohrson, C.~L.},
+  \bibinfo{author}{Sherman, M.~A.} \& \bibinfo{author}{Park, P.~J.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Identification of somatic mutations
+  in single cell {DNA}-seq using a spatial model of allelic imbalance}}.
+\newblock {\emph{\JournalTitle{Nature Communications}}}
+  \textbf{\bibinfo{volume}{10}}, \bibinfo{pages}{1--14} (\bibinfo{year}{2019}).
+\newblock \doiprefix 10.1038/s41467-019-11857-8.
+
+\bibitem{koster_varlociraptor_2020}
+\bibinfo{author}{Köster, J.}, \bibinfo{author}{Dijkstra, L.~J.},
+  \bibinfo{author}{Marschall, T.} \& \bibinfo{author}{Schönhuth, A.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Varlociraptor: enhancing
+  sensitivity and controlling false discovery rate in somatic indel
+  discovery}}.
+\newblock {\emph{\JournalTitle{Genome Biology}}} \textbf{\bibinfo{volume}{21}},
+  \bibinfo{pages}{98} (\bibinfo{year}{2020}).
+\newblock \doiprefix 10.1186/s13059-020-01993-6.
+
+\bibitem{hou_single-cell_2012}
+\bibinfo{author}{Hou, Y.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Single-{Cell} {Exome} {Sequencing}
+  and {Monoclonal} {Evolution} of a {JAK2}{\textbackslash}mbox-{Negative}
+  {Myeloproliferative} {Neoplasm}}}.
+\newblock {\emph{\JournalTitle{Cell}}} \textbf{\bibinfo{volume}{148}},
+  \bibinfo{pages}{873--885} (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1016/j.cell.2012.02.028.
+\newblock \bibinfo{note}{Bibtex: Hou2012}.
+
+\bibitem{xu_single-cell_2012}
+\bibinfo{author}{Xu, X.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Single-{Cell} {Exome} {Sequencing}
+  {Reveals} {Single}-{Nucleotide} {Mutation} {Characteristics} of a {Kidney}
+  {Tumor}}}.
+\newblock {\emph{\JournalTitle{Cell}}} \textbf{\bibinfo{volume}{148}},
+  \bibinfo{pages}{886--895} (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1016/j.cell.2012.02.025.
+\newblock \bibinfo{note}{Bibtex: Xu2012}.
+
+\bibitem{ling_evaluation_2009}
+\bibinfo{author}{Ling, J.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Evaluation of genome coverage and
+  fidelity of multiple displacement amplification from single cells by {SNP}
+  array}}.
+\newblock {\emph{\JournalTitle{MHR: Basic science of reproductive medicine}}}
+  \textbf{\bibinfo{volume}{15}}, \bibinfo{pages}{739--747}
+  (\bibinfo{year}{2009}).
+\newblock \doiprefix 10.1093/molehr/gap066.
+
+\bibitem{spits_whole-genome_2006}
+\bibinfo{author}{Spits, C.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Whole-genome multiple displacement
+  amplification from single cells}}.
+\newblock {\emph{\JournalTitle{Nature Protocols}}}
+  \textbf{\bibinfo{volume}{1}}, \bibinfo{pages}{1965--1970}
+  (\bibinfo{year}{2006}).
+\newblock \doiprefix 10.1038/nprot.2006.326.
+
+\bibitem{renwick_proof_2006}
+\bibinfo{author}{Renwick, P.~J.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Proof of principle and first cases
+  using preimplantation genetic haplotyping {\textendash} a paradigm shift for
+  embryo diagnosis}}.
+\newblock {\emph{\JournalTitle{Reproductive BioMedicine Online}}}
+  \textbf{\bibinfo{volume}{13}}, \bibinfo{pages}{110--119}
+  (\bibinfo{year}{2006}).
+\newblock \doiprefix 10.1016/S1472-6483(10)62024-X.
+
+\bibitem{zafar_siclonefit:_2018}
+\bibinfo{author}{Zafar, H.}, \bibinfo{author}{Navin, N.},
+  \bibinfo{author}{Chen, K.} \& \bibinfo{author}{Nakhleh, L.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{SiCloneFit}: {Bayesian} inference
+  of population structure, genotype, and phylogeny of tumor clones from
+  single-cell genome sequencing data}}.
+\newblock {\emph{\JournalTitle{bioRxiv}}} \bibinfo{pages}{394262}
+  (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1101/394262.
+
+\bibitem{koptagel_scuphr:_2018}
+\bibinfo{author}{Koptagel, H.}, \bibinfo{author}{Jun, S.-H.} \&
+  \bibinfo{author}{Lagergren, J.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{SCuPhr}: {A} {Probabilistic}
+  {Framework} for {Cell} {Lineage} {Tree} {Reconstruction}}}.
+\newblock {\emph{\JournalTitle{bioRxiv}}} \bibinfo{pages}{357442}
+  (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1101/357442.
+
+\bibitem{bohrson_linked-read_2019}
+\bibinfo{author}{Bohrson, C.~L.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Linked-read analysis identifies
+  mutations in single-cell {DNA}-sequencing data}}.
+\newblock {\emph{\JournalTitle{Nature Genetics}}} \bibinfo{pages}{1}
+  (\bibinfo{year}{2019}).
+\newblock \doiprefix 10.1038/s41588-019-0366-2.
+
+\bibitem{hard_conbase:_2019}
+\bibinfo{author}{H{\r a}rd, J.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Conbase: a software for
+  unsupervised discovery of clonal somatic mutations in single cells through
+  read phasing}}.
+\newblock {\emph{\JournalTitle{Genome Biology}}} \textbf{\bibinfo{volume}{20}},
+  \bibinfo{pages}{68} (\bibinfo{year}{2019}).
+\newblock \doiprefix 10.1186/s13059-019-1673-8.
+
+\bibitem{baslan_genome-wide_2012}
+\bibinfo{author}{Baslan, T.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Genome-wide copy number analysis of
+  single cells}}.
+\newblock {\emph{\JournalTitle{Nature Protocols}}}
+  \textbf{\bibinfo{volume}{7}}, \bibinfo{pages}{1024--1041}
+  (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1038/nprot.2012.039.
+
+\bibitem{satas_scarlet_2020}
+\bibinfo{author}{Satas, G.}, \bibinfo{author}{Zaccaria, S.},
+  \bibinfo{author}{Mon, G.} \& \bibinfo{author}{Raphael, B.~J.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{SCARLET}: {Single}-cell tumor
+  phylogeny inference with copy-number constrained mutation losses}}.
+\newblock {\emph{\JournalTitle{Cell Systems}}} \textbf{\bibinfo{volume}{10}},
+  \bibinfo{pages}{323--332.e8} (\bibinfo{year}{2020}).
+\newblock \doiprefix 10.1016/j.cels.2020.04.001.
+
+\bibitem{eggenberger_uber_1923}
+\bibinfo{author}{Eggenberger, F.} \& \bibinfo{author}{P{\'o}lya, G.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{\"U}ber die {Statistik}
+  verketteter {Vorg{\"a}nge}}}.
+\newblock {\emph{\JournalTitle{ZAMM - Journal of Applied Mathematics and
+  Mechanics / Zeitschrift f{\"u}r Angewandte Mathematik und Mechanik}}}
+  \textbf{\bibinfo{volume}{3}}, \bibinfo{pages}{279--289}
+  (\bibinfo{year}{1923}).
+\newblock \doiprefix 10.1002/zamm.19230030407.
+
+\bibitem{muller_optimal_2004}
+\bibinfo{author}{M{\"u}ller, P.}, \bibinfo{author}{Parmigiani, G.},
+  \bibinfo{author}{Robert, C.} \& \bibinfo{author}{Rousseau, J.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Optimal {Sample} {Size} for
+  {Multiple} {Testing}: {The} {Case} of {Gene} {Expression} {Microarrays}}}.
+\newblock {\emph{\JournalTitle{Journal of the American Statistical
+  Association}}} \textbf{\bibinfo{volume}{99}}, \bibinfo{pages}{990--1001}
+  (\bibinfo{year}{2004}).
+\newblock \doiprefix 10.1198/016214504000001646.
+
+\bibitem{muller_fdr_2006}
+\bibinfo{author}{M{\"u}ller, P.}, \bibinfo{author}{Parmigiani, G.} \&
+  \bibinfo{author}{Rice, K.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{FDR} and {Bayesian} multiple
+  comparisons rules}}.
+\newblock {\emph{\JournalTitle{Proceedings of the ISBA 8th World Meeting on
+  Bayesian Statistics (Valencia)}}}  (\bibinfo{year}{2006}).
+
+\bibitem{taschuk_ten_2017}
+\bibinfo{author}{Taschuk, M.} \& \bibinfo{author}{Wilson, G.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Ten simple rules for making
+  research software more robust}}.
+\newblock {\emph{\JournalTitle{PLOS Computational Biology}}}
+  \textbf{\bibinfo{volume}{13}}, \bibinfo{pages}{e1005412}
+  (\bibinfo{year}{2017}).
+\newblock \doiprefix 10.1371/journal.pcbi.1005412.
+
+\bibitem{gruning_bioconda:_2018}
+\bibinfo{author}{Gr{\"u}ning, B.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Bioconda: sustainable and
+  comprehensive software distribution for the life sciences}}.
+\newblock {\emph{\JournalTitle{Nature Methods}}} \textbf{\bibinfo{volume}{15}},
+  \bibinfo{pages}{475--476} (\bibinfo{year}{2018}).
+\newblock \doiprefix 10.1038/s41592-018-0046-7.
+
+\bibitem{koster_snakemakescalable_2012}
+\bibinfo{author}{K{\"o}ster, J.} \& \bibinfo{author}{Rahmann, S.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Snakemake{\textemdash}a scalable
+  bioinformatics workflow engine}}.
+\newblock {\emph{\JournalTitle{Bioinformatics}}} \textbf{\bibinfo{volume}{28}},
+  \bibinfo{pages}{2520--2522} (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1093/bioinformatics/bts480.
+
+\bibitem{hoell_constitutional_2014}
+\bibinfo{author}{Hoell, J.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Constitutional {Mismatch}
+  {Repair}-deficiency and {Whole}-exome {Sequencing} as the {Means} of the
+  {Rapid} {Detection} of the {Causative} {MSH6} {Defect}}}.
+\newblock {\emph{\JournalTitle{Klinische P{\"a}diatrie}}}
+  \textbf{\bibinfo{volume}{226}}, \bibinfo{pages}{357--361}
+  (\bibinfo{year}{2014}).
+\newblock \doiprefix 10.1055/s-0034-1389905.
+
+\bibitem{browning_improving_2013}
+\bibinfo{author}{Browning, B.~L.} \& \bibinfo{author}{Browning, S.~R.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Improving the {Accuracy} and
+  {Efficiency} of {Identity}-by-{Descent} {Detection} in {Population} {Data}}}.
+\newblock {\emph{\JournalTitle{Genetics}}} \textbf{\bibinfo{volume}{194}},
+  \bibinfo{pages}{459--471} (\bibinfo{year}{2013}).
+\newblock \doiprefix 10.1534/genetics.113.150029.
+
+\bibitem{li_likelihood-based_2012}
+\bibinfo{author}{Li, B.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{A {Likelihood}-{Based} {Framework}
+  for {Variant} {Calling} and {De} {Novo} {Mutation} {Detection} in
+  {Families}}}.
+\newblock {\emph{\JournalTitle{PLoS Genet}}} \textbf{\bibinfo{volume}{8}},
+  \bibinfo{pages}{e1002944} (\bibinfo{year}{2012}).
+\newblock \doiprefix 10.1371/journal.pgen.1002944.
+
+\bibitem{peng_rare_2013}
+\bibinfo{author}{Peng, G.} \emph{et~al.}
+\newblock \bibinfo{journal}{\bibinfo{title}{Rare variant detection using
+  family-based sequencing analysis}}.
+\newblock {\emph{\JournalTitle{Proceedings of the National Academy of
+  Sciences}}} \textbf{\bibinfo{volume}{110}}, \bibinfo{pages}{3985--3990}
+  (\bibinfo{year}{2013}).
+\newblock \doiprefix 10.1073/pnas.1222158110.
+
+\bibitem{peng_famseq:_2014}
+\bibinfo{author}{Peng, G.}, \bibinfo{author}{Fan, Y.} \& \bibinfo{author}{Wang,
+  W.}
+\newblock \bibinfo{journal}{\bibinfo{title}{{FamSeq}: {A} {Variant} {Calling}
+  {Program} for {Family}-{Based} {Sequencing} {Data} {Using} {Graphics}
+  {Processing} {Units}}}.
+\newblock {\emph{\JournalTitle{PLoS Comput Biol}}}
+  \textbf{\bibinfo{volume}{10}}, \bibinfo{pages}{e1003880}
+  (\bibinfo{year}{2014}).
+\newblock \doiprefix 10.1371/journal.pcbi.1003880.
+
+\end{thebibliography}
 
@@
 
-\section*{Funding}
 This work has been supported by the Helmholtz Association, in particular through a Helmholtz Incubator grant (Sparse2Big ZT-I-0007), by the compute cluster at the Helmholtz Institute for Infection Research, and by the Katharina Hardt Stiftung.
 Alexander Schönhuth was supported by the Netherlands Organisation for Scientific Research (NWO: Vidi grant 639.072.309).
-Arndt Borkhardt and Ute Fischer were further supported by the German Federal Office for Radiation Protection (BfS) grant nos.
-3618S32274 and 3618S32275.
+Arndt Borkhardt and Ute Fischer were further supported by the German Federal Office for Radiation Protection (BfS) grant nos. 3618S32274 and 3618S32275.
 
-\section{Online Methods}
-See Supplement PDF.
+\section*{Author contributions}
+
+A.C.M, A.B. and U.F. conceived the original project to study single immune cells.
+A.C.M., A.B., U.F. and D.L. created the experimental design.
+U.F. obtained and processed samples.
+A.S. and D.L. formulated the statistical model and wrote the manuscript, with comments from A.C.M.
+The model was implemented by J.K. and D.L. in Rust.
+D.L. conducted all computational analyses, under the supervision of A.C.M. and A.S.
+Results were interpreted by D.L., A.C.M. and A.S.
+All authors edited the manuscript.
+
+\section*{Competing interests}
+The authors declare no competing interests.
 

### 2021-09-09 manuscript/supplement.tex
update manuscript, scripts and figures to final submission (last reviewer comment addressed, editorial requests addressed)

@@
 
+\usepackage{xr-hyper}
 \usepackage{hyperref}
@@
 
-\usepackage{xr-hyper}
-
 % xr helper code to generate dependency definition command \myexternaldocument
@@
 
-\author{David Lähnemann, Johannes Köster, Ute Fischer, \\ Arndt Borkhardt, Alice McHardy, Alexander Schönhuth}
+\author{David Lähnemann, Johannes Köster, Ute Fischer, \\ Arndt Borkhardt, Alice C. McHardy, Alexander Schönhuth}
 
@@
 
-\textbf{Whole genome sequencing of almost identical kin-cells from a cell-line \citep{dong_accurate_2017}.}
-The first dataset is from the publication of a previous variant caller for single cell sequencing data generated using MDA, SCcaller (\cite{dong_accurate_2017}; Figure~\ref{fig:datasets}A).
+\paragraph{Whole genome sequencing of almost identical kin-cells from a cell-line \citep{dong_accurate_2017}.}
+The first dataset is from the publication of a previous variant caller for single cell sequencing data generated using MDA, SCcaller (\cite{dong_accurate_2017}; Figure~\ref{fig:datasets}a).
 They started with one initial cell from a cell line and grew it in several steps (a complete overview of their setup is given in Figure 1 of \cite{dong_accurate_2017}).
@@
 This bulk sample (IL-1C) was only used as a ground truth and was not provided as input to any of the software compared here.
-Three more distant clones (C1, C2, C3 in Figure~\ref{fig:datasets}A), grown from the cells after the initial mini-expansion, were jointly used as a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
+Three more distant clones (C1, C2, C3 in Figure~\ref{fig:datasets}a), grown from the cells after the initial mini-expansion, were jointly used as a further bulk sample for SCcaller and ProSolo (see Software and Parameters below).
 With three separate bulk sequencing runs, their joint coverage of about 50 for 80\% of genomic sites is much higher than for the MDA single cells and the unamplified IL-1C bulk (Clones in Figure~\ref{fig:coverage-dists}A).\\
@@
 
-\textbf{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
-For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figures~\ref{fig:datasets}B and \ref{fig:FACS-scheme}).
+\paragraph{Whole exome sequencing of five human granulocytes with a pedigree ground truth.}
+For the second benchmarking dataset, Blood was taken from a patient after informed consent and granulocytes were sorted out in several steps: Initially, CD3- cells were captured via Magnetic-Activated Cell Sorting (MACS); among them, CD66b+ granulocytes were filtered out via Fluorescence Activated Cell Sorting (FACS) (Figures~\ref{fig:datasets}b and \ref{fig:FACS-scheme}).
 Single cells from that cell population were isolated using a Fluidigm C1 microfluidics chip and subjected to multiple displacement amplification (MDA) using the GE Healthcare GenomiPhi Kit in its version adapted to that microfluidic chip.
-With a panel of 16 custom primer pairs covering different genes across chromosomes used in a quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
+With a panel of 16 custom primer pairs covering different genes across chromosomes (Table~\ref{tab:selection-primers}) used in a quantitative real-time PCR, we selected granulocytes where at least 15 of these loci were properly amplified.
 For those cells, we performed whole exome capture using the Agilent SureSelectXT Human All Exon V5+UTRs Kit.
@@
 
+\begin{table}
+    \centering
+    \caption{
+    Primers used for selection of well amplified single cell samples via quantitative real-time PCR.
+    These were taken from Table S3 in \cite{zong_genome-wide_2012}.
+    }
+    \small
+    \begin{tabular}{lll}
+      \toprule
+      chromosome  &  forward primer & reverse primer \\
+      \toprule
+      chr1  & AGGAAAGGCATACTGGAGGGACAT & TTAGGGATGGCACCACACTCTTGA \\
+      \midrule
+      chr2  & TCCCAGAGAAGCATCCTCCATGTT & CACCACACTGCCTCAAATGTTGCT \\
+      \midrule
+      chr3  & TCAAGTTGCCAGCTGTGGCTGTAT & AGAAGGGCATTTCCTGTCAGTGGA \\
+      \midrule
+      chr4  & ATGGGCAAATCCAGAAGAGTCCAG & CCATTCACTTCCTTGGAAAGGTAGCC \\
+      \midrule
+      chr5  & AATAGCGTGCAGTTCTGGGTAGCA & TTCACATCCTGGGAGGAACAGCAT \\
+      \midrule
+      chr6  & TGAATGCCAGGGTGAGACCTTTGA & TGTTCATTATCCCACGCCAGGACT \\
+      \midrule
+      chr7  & ACCAAAGGAAAGCCAGCCAGTCTA & ACTCCACAGCTCCCAAGCATACAA \\
+      \midrule
+      chr9  & TCCCAGCTCTCTCTCTTGCATCTT & AGTGAAGCTGGTGTATGCAGAGGT \\
+      \midrule
+      chr12 & AGAGGGCTGCTTTATGCAGGTG   & CTACATTTGGGTCTTTGCTGCCATG \\
+      \midrule
+      chr13 & AGCAGCCCCAGGCAGAT        & CGGAGAGGACGGTCACGTTTAC \\
+      \midrule
+      chr14 & CGGAGAGGACGGTCACGTTTAC   & CGTGGGAGTTTTGAAATGCGATGT \\
+      \midrule
+      chr15 & CCTGTCTCTGCTCCTGCG       & TGCACACATGCACAGTGGAG \\
+      \midrule
+      chr16 & CTCCAAGGTTCTGCAGCCTC     & GGTATGACTACACATTCAGGCTGG \\
+      \midrule
+      chr17 & GTGGTACATAGTGCATGGTCCG   & GGCGACATACCCCAACTTCATAAG \\
+      \midrule
+      chr18 & CGTTCTTAGGACCAAAGGGCTG   & CCAGCATCCATGTCTCTGCAC \\
+      \midrule
+      chr19 & GCCCAGAGCGCCTGA          & CCAGCCCCTGGACCACT \\
+      \bottomrule
+    \end{tabular}
+    \label{tab:selection-primers}
+\end{table}
+
 \begin{figure}[!tpb]
@@
 
-For the ground truth of this dataset, we could leverage previously published bulk whole exome sequencing data from the same person, their parents and three siblings (\cite{hoell_constitutional_2014}; Figure~\ref{fig:datasets}B).
+For the ground truth of this dataset, we could leverage previously published bulk whole exome sequencing data from the same person, their parents and three siblings (\cite{hoell_constitutional_2014}; Figure~\ref{fig:datasets}b).
 For ground truth alternative allele calls, we ran three pedigree-aware variant callers on the family whole exome data.
@@
 
-\textbf{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
+\paragraph{Single nucleus exome sequencing of 16 normal and 16 tumor cells from a triple negative breast cancer (TNBC) \citep{wang_clonal_2014}.}
 The third dataset is a published dataset from a TNBC patient, with both tumor and matched normal cells available.
-For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}C).
+For each, 16 cells and a bulk sample were whole exome sequenced (Figure~\ref{fig:datasets}c).
 For the 32 single cells, library preparation was performed on isolated single nuclei \citep{wang_clonal_2014}.
@@
   Global view of the precision-recall plots for alternative allele calling across different software modes and filtering thresholds (for zoomed in view on the most interesting areas of the plot, see main manuscript Figure~\ref{fig:alt-calling_prec-rec}).
-  \textbf{A} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
-  \textbf{B} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel B.\newline
-  \textbf{C} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
+  \textbf{a} Precision and recall average of two whole genome sequenced single cells IL-11 and IL-12 against their kindred clone IL-1c as ground truth genotypes.
+  \textbf{b} Precision and recall average of five whole exome sequenced single granulocytes against their pedigree-based germline genotype ground truth.\newline 
+  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset in panel b.\newline
+  \textbf{c} Precision and recall average of 16 tumor and 16 normal single cells sequence at the whole exome level. \newline \footnotesize
   Used thresholds are not comparable between tools.
@@
 
-The whole genome cell line dataset (Figure~\ref{fig:datasets}A) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}A and \ref{fig:alt-calling_prec-rec}A), at recall rates of 45\% and higher.
+The whole genome cell line dataset (Figure~\ref{fig:datasets}a) seems much less challenging than the other dataset: all methods achieve high precision in alternative allele calling (Figures~\ref{fig:alt-calling_prec-rec_global}a and \ref{fig:alt-calling_prec-rec}a), at recall rates of 45\% and higher.
 In comparison with all other tools, ProSolo achieves the most striking increases in recall of nearly 10\%.
@@
 This dataset is considerably more challenging than the first one.
-See Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B for respective results.
-For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}B and \ref{fig:alt-calling_prec-rec_global}B).
+See Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b for respective results.
+For this dataset, only SCcaller, SCIPhI and ProSolo achieved a precision above .99, with ProSolo reaching a 20\% increase of recall to .178, compared to SCIPhI's .146, and SCcaller with .072 (Figures~\ref{fig:alt-calling_prec-rec}b and \ref{fig:alt-calling_prec-rec_global}b).
 In comparison, MonoVar achieved a maximum precision of only .962. However, this was at a much higher recall (.141) than for example SCcaller (.095 at a precision of .972).
@@
  Cell-specific precision-recall plots across different filtering threshold levels for the whole exome dataset.\newline \footnotesize
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
+  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline
   Used thresholds are not comparable between tools.
@@
 
-For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}C).
+For the third dataset of single nucleus whole exome sequencing data of 32 cells from a TNBC patient (16 tumor cells, 16 normal cells), we analysed tumor and normal cells separately (Figure~\ref{fig:alt-calling_prec-rec}c).
 On the tumor cells, ProSolo is the only tool to achieve a precision above .99 (at a recall of .319).
@@
 Here, SCIPhI outcompetes ProSolo with a maximum recall of .650 compared to ProSolo's .548.
-While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}C), this comes at a reduced precision.
+While ProSolo can achieve a similar recall (.625, see {\ttfamily --fdr} threshold of 0.2 in Figure~\ref{fig:alt-calling_prec-rec_global}c), this comes at a reduced precision.
 This reflects the inherent uncertainty of the single cell data and showcases a key difference in the approaches of ProSolo and SCIPhI.
@@
   The left panel includes all bulk coverages, the right panel focuses on the differences in non-zero bulk coverages.\newline 
-  $^\text{\sffamily-B}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline \footnotesize
+  $^\text{\sffamily-b}$ The germline ground truth induces an artificial increase of recall for SCIPhI's sensitive and ProSolo's imputation mode; these modes should thus be disregarded for a fair comparison on the granulocyte dataset.\newline \footnotesize
   Used thresholds are not comparable between tools.
@@
 
+\begin{table}
+  \caption{
+  Genomic sites where any tool calls the presence of a variant in at least three more cells than the number reported in \cite{wang_clonal_2014}.
+  "Wang duplex freq" gives the alternative allele frequency from the targeted duplex bulk sequencing (deep coverage and high accuracy) in \cite{wang_clonal_2014}.
+  All variants in this table are reported as heterozygous by \cite{wang_clonal_2014}.
+  Results are from each software's best performing software mode without any threshold-based filtering.
+  \footnotesize\newline$^*$ MonoVar called 5 cells as heterozygous and 4 cells as homozygous alternative}
+  \begin{tabular}{p{9ex}rp{9ex}rrrrr}
+    \toprule
+    & & & \multicolumn{5}{c}{number of cells with variant (out of 16)} \\
+    \cmidrule(l){4-8}
+    chromosome & position & Wang duplex freq & Wang & MonoVar & ProSolo & SCcaller & SCIPhI \\
+    \toprule
+    chr16 &  1537917  & 0.102820 & 3 &     3 & 0 & 4 & 12\\
+    \midrule
+    chr16 & 15149761  & 0.057895 & 2 &     3 & 0 & 3 & 12\\
+    \midrule
+    chr16 & 66643819  & 0.189566 & 5 &     6 & 6 & 6 & 15\\
+    \midrule
+    chr17 & 77809062  & 0.001294 & 2 &     3 & 0 & 3 & 15\\
+    \midrule
+    chr2  & 74435818  & 0.091089 & 3 &     4 & 0 & 3 & 12\\
+    \midrule
+    chr5  & 55204134  & 0.233289 & 6 & $^*$9 & 11 & 10 & 15\\
+    \midrule
+    chr9  & 19298104  & 0.131816 & 4 &     6 & 0 & 6 & 15\\
+    \midrule
+    chr9  & 80046280  & 0.000269 & 2 &     2 & 0 & 2 & 15\\
+    \midrule
+    chr9  & 117853210 & 0.001309 & 4 &     3 & 3 & 4 & 15\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:sciphi-overimputed}
+\end{table}
+
 For the tumor cells from the \cite{wang_clonal_2014} dataset, we could further determine the recall of non-synonymous somatic variants they had validated via deep targeted duplex sequencing and provided in their Supplementary Tables 6 and 7 (specifying clonal and subclonal variants, respectively).
@@
   As a consequence, if the aim is to detect lower frequency mutations, we simply recommend a deeper sequencing of the bulk sample.
-  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells, while the maximum frequency for any of the heterozygous subclonal mutations detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
-  This suggests that SCIPhI's model over-imputes mutations by aggregating information across cells.
+  \item SCIPhI calls nine heterozygous mutations in 12 or 15 cells (Table~\ref{tab:sciphi-overimputed} and Figure~\ref{fig:somatic-recall}), while the maximum frequency for any of the heterozygous subclonal mutations detected in the deeply sequenced tumor bulk was 0.233, more in the range of the maximum number of six out of 16 cells that \cite{wang_clonal_2014} found any of the subclonal variants in.
+  Also, when able to detect the variant, all other tools report a maximum excess of two cells with the mutation compared to \cite{wang_clonal_2014}.
+  The only exception is a mutation on chromosome five, where all tools agree on a much larger number of cells with the mutation (9-11) than originally reported by Wang et al.
+(6), but SCIPhI still clearly exceeds that consensus (15).
+  Altogether, these numbers suggest that SCIPhI's model over-imputes mutations by aggregating information across cells.
   \item None of the tools calls any of the six homozygous alternative variants.
@@
  \caption{
- False discovery rate (FDR, $1 - p$recision) as calculated based on the ground truth, compared to the theoretical FDR we controlled for using ProSolo, for (A) the whole genome data set and (B) the whole exome data set.
+ False discovery rate (FDR, $1 - p$recision) as calculated based on the ground truth, compared to the theoretical FDR we controlled for using ProSolo, for \textbf{a} the whole genome cell-line data set, \textbf{b} the whole exome data set with the pedigree ground truth and \textbf{c} the whole exome cancer data set.
  The diagonal indicates identity of the ground truth FDR with the FDR that was controlled for.
@@
 
-Further, Figure~\ref{fig:FDR-ground-truth-vs-theoretical}A demonstrates, that ProSolo provides a rigorous control over the false discovery rate on the whole genome dataset.
+Further, Figure~\ref{fig:FDR-ground-truth-vs-theoretical}a demonstrates, that ProSolo provides a rigorous control over the false discovery rate on the whole genome dataset.
 The only exception---in both default and imputation mode---is when attempting to control for a false discovery rate below 0.0001.
 This indicates, that ProSolo's model still needs refinement in its resolution for sites with a very high likelihood of an alternative allele present.
-The same trend can be seen on the other datasets (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B and C), but for any attempt to control for a false discovery rate below 0.02 or 0.005.
+The same trend can be seen on the other datasets (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}b and c), but for any attempt to control for a false discovery rate below 0.02 or 0.005.
 This is within the common range of values often chosen as a false discovery rate to control for in practice and not being able to actually achieve the false discovery rate control at such values is usually unacceptable.
@@
 Further, the artefact of using a germline ground truth on the granulocyte dataset also shows in the discrepancy in the ground-truth based false discovery rate between ProSolo's default and imputation mode when controlling for higher false discovery rates.
-As expected, imputation turned out less precise in the whole genome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}A).
-In contrast, imputation seems to have improved precision for the whole exome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}B).
+As expected, imputation turned out less precise in the whole genome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}a).
+In contrast, imputation seems to have improved precision for the whole exome dataset (Figure~\ref{fig:FDR-ground-truth-vs-theoretical}b).
 This would suggest that the larger call set from ProSolo's imputation mode (with a higher recall) is also more precise than the smaller call set from the default mode.
@@
 
-Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C), with the $F_1$ score defined as the harmonic mean of precision $p$ and recall $r$ (with $\beta = 1$ for the general $F$ score, see \cite{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}):\\
+Comparisons provided here are based on the software modes that achieved the highest $F_1$ score across most single cells for alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}c), with the $F_1$ score defined as the harmonic mean of precision $p$ and recall $r$ (with $\beta = 1$ for the general $F$ score, see \cite{van_rijsbergen_evaluation_1979,chinchor_muc-4_1992}):\\
 
@@
   Genotype calling precision and recall in comparison against the respective ground truth genotypes (from IL-1c in the left plots; from PNG in the plots on the right).
-  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
+  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}c).
   Facet labels at the top of plots indicate the single cell.
@@
   Genotype calling comparison against the respective ground truth genotypes (from IL-1c in the left plots; from PNG in the plots on the right).
-  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}C).
+  MonoVar and ProSolo were run with the parameters that maximised the $F_1$ score alternative allele calling (Figure~\ref{fig:alt-calling_prec-rec}c).
   Panels A and B correspond to the respective panels in Figure~\ref{fig:gt-calling-true-false-colouring}, but with a finer resolution of the false genotypes.
@@
 
-\subsection{Further Software and Resources used}
+\subsection{Software and Resources used}
 
@@
  \begin{table}[tbp]
-  \caption{Software and related resources used in this manuscript. All software versions as installed with conda.}
+  \caption{Software and related resources used in this manuscript.
+All software versions as installed with conda, except: Kaluza and Summit, which were pre-installed on the fluorescence activated cell sorting machine; Brewer and Okabe Ito, which are color palettes; bioconda, which is a package repository for the conda (and mamba) package manager.}
   \label{tab:software-overview}
+  \small
   \begin{tabular}{llll}
     \toprule
-    software / resource & version   & purpose                       & citation \\
+    resource            & version   & purpose                       & citation \\
     \toprule
-    cutadapt            & 1.14      & adapter trimming              & \citep{martin_cutadapt_2011} \\
-    \midrule
-    trimmomatic         & 0.36      & quality trimming              & \citep{bolger_trimmomatic_2014} \\
-    \midrule
-    bwa mem             & 0.7.15    & read alignment / mapping      & \citep{li_aligning_2013} \\
-    \midrule
-    samtools            & 1.4.1     & handle SAM / BAM files        & \citep{li_sequence_2009} \\
-    \midrule
-    picardtools         & 2.9.2     & mark duplicates               &  \\
-    \midrule
     bcftools            & 1.6       & handle VCF / BCF files        & \citep{danecek_variant_2011} \\
@@
     \midrule
+    BEAGLE 4.0          & 4.0       & pedigree-aware variant calling  & \citep{browning_improving_2013} \\
+    \midrule
+    bioconda            &           & software package management   &  \citep{gruning_bioconda:_2018}\\
+    \midrule
+    Brewer              &           & colorblind-friendly palettes  & \citep{brewer_colorbrewer_2006} \\
+    \midrule
+    conda               & 4.9.0     & software package management   &  \\
+    \midrule
+    cutadapt            & 1.14      & adapter trimming              & \citep{martin_cutadapt_2011} \\
+    \midrule
+    bwa mem             & 0.7.15    & read alignment / mapping      & \citep{li_aligning_2013} \\
+    \midrule
+    FamSeq              & 1.0.3     & pedigree-aware variant calling & \citep{peng_rare_2013,peng_famseq:_2014} \\
+    \midrule
+    ggplot2             & 3.3.0     & result plots                  & \citep{wickham_ggplot2_2016} \\
+    \midrule
+    ggthemes            & 0.2.5     & Okabe Ito palette             & \citep{arnold_ggthemes_2019} \\
+    \midrule
+    Inkscape            & 0.92.4    & info-graphics, improved labels & \\
+    \midrule
+    Kaluza              & 2.1       & FACS data analysis   &  \\
+    \midrule
+    mamba               & 0.6.4     & software package management   &  \\
+    \midrule
+    MonoVar             & 0.0.1     & single cell variant calling   & \citep{zafar_monovar:_2016} \\
+    \midrule
     mosdepth            & 0.2.5     & coverage calculations         & \citep{pedersen_mosdepth_2018} \\
     \midrule
-    Okabe Ito palette   &           & colorblind-safe               & \citep{okabe_color_2008} \\
-    \midrule
-    ggthemes            & 0.2.5     & Okabe Ito palette             & \citep{arnold_ggthemes_2019} \\
-    \midrule
-    Brewer palettes     &           & colorblind-friendly           & \citep{brewer_colorbrewer_2006} \\
+    Okabe Ito           &           & colorblind-safe palette       & \citep{okabe_color_2008} \\
+    \midrule
+    picardtools         & 2.9.2     & mark duplicates               &  \\
+    \midrule
+    polymutt           & 0.18      & pedigree-aware variant calling & \citep{li_likelihood-based_2012} \\
+    \midrule
+    ProSolo             & 0.6.1     & single cell variant calling        &  \\
     \midrule
     RColorBrewer        & 1.1\_2    & Brewer palettes              & \citep{neuwirth_rcolorbrewer_2014} \\
+    \midrule
+    samtools            & 1.4.1     & handle SAM / BAM files        & \citep{li_sequence_2009} \\
+    \midrule
+    SCAN-SNV            & 19-10-30 & single cell variant calling  & \citep{luquette_identification_2019} \\
+    \midrule
+    SCcaller            & 1.2       & single cell variant calling  & \citep{dong_accurate_2017} \\
+    \midrule
+    SCIPhI              & 0.1.4     & single cell variant calling  & \citep{singer_single-cell_2018} \\
+    \midrule
+    Snakemake           & 5.4.0     & workflow management           & \citep{koster_snakemakescalable_2012} \\
+    \midrule
+    Summit              & 5.1.0     & FACS data acquisition         & \\
     \midrule
@@
     \midrule
-    ggplot2             & 3.3.0     & result plots                  & \citep{wickham_ggplot2_2016} \\
-    \midrule
-    Inkscape            & 0.92.4    & info-graphics, improved labels & \\
+    trimmomatic         & 0.36      & quality trimming              & \citep{bolger_trimmomatic_2014} \\
+    \midrule
+    UpsetR              &           & set intersection plot \ref{fig:granulocytes-ground-truth} & \citep{conway_upsetr_2017} \\
     \midrule
     VGAM                & 1.1\_1     & distributions in R            & \citep{yee_vgam_2008,yee_vector_2015} \\
-    \midrule
-    UpsetR              &           & set intersection plot \ref{fig:granulocytes-ground-truth} & \citep{conway_upsetr_2017} \\
     \bottomrule

### 2021-09-09 manuscript/figs/Dong2017/Dong2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -182,7 +182,7 @@ ggplot(avg_per_software_and_parameter %>%
         mutate(num_filter = as.numeric(levels(filter)[filter]) )
   ) + 
   geom_point(aes(x = num_filter, y = avg_FDR, color = software, shape = filter), size = 4) +
-  geom_abline(intercept = 0) +
+  geom_segment(aes(x = 0.000001, y = 0.000001, xend = 1, yend = 1)) +
   scale_x_continuous(
     trans = "log10",
     limits=axis_limits
@@ -204,7 +204,7 @@ ggplot(avg_per_software_and_parameter %>%
   labs(
     x = "theoretical FDR",
     y = "ground truth FDR",
-    tag = "A",
+    tag = "a",
     title = "IL-11 and IL-12"
   )
 ggsave("Dong2017_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf", device = cairo_pdf, width=8, height=7.5)
@@ -277,7 +277,7 @@ ggplot(avg_per_software_and_parameter %>%
   theme_bw(base_size=24, base_family="Lato")  +
   labs(x = "average recall",
        y = "average precision",
-       tag = "A",
+       tag = "a",
        title = "IL-11 and IL-12")
 ggsave("Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot.pdf", device = cairo_pdf, width=8, height=7.5)
 
@@ -310,7 +310,7 @@ ggplot(avg_per_software_and_parameter %>%
   theme_bw(base_size=24, base_family="Lato") +
   labs(x = "average recall",
        y = "average precision",
-       tag = "A",
+       tag = "a",
        title = "IL-11 and IL-12")
 ggsave("Dong2017_prosolo-monovar-scansnv-sccaller_precision-recall-plot_focus-tools.pdf", device = cairo_pdf, width=8, height=7.5)
 

### 2021-09-09 manuscript/figs/Laehnemann2017/Laehnemann2017_plot_prosolo-monovar-scansnv-sccaller-sciphi.precision-recall.R

@@ -209,7 +209,7 @@ ggplot(avg_per_software_and_parameter %>%
          mutate(num_filter = as.numeric(levels(filter)[filter]) )
   ) + 
   geom_point(aes(x = num_filter, y = avg_FDR, color = software, shape = filter), size = 4) +
-  geom_abline(intercept = 0) +
+  geom_segment(aes(x = 0.000001, y = 0.000001, xend = 1, yend = 1)) +
   scale_x_continuous(
     trans = "log10",
     limits=axis_limits
@@ -231,7 +231,7 @@ ggplot(avg_per_software_and_parameter %>%
   labs(
     x = "theoretical FDR",
     y = "ground truth FDR",
-    tag = "B",
+    tag = "b",
     title = "5 granulocytes"
   )
 ggsave("Laehnemann2017_prosolo-sciphi_FDR_ground_truth_vs_theoretical.pdf", device = cairo_pdf, width=8, height=7.5)
@@ -329,7 +329,7 @@ ggplot(avg_per_software_and_parameter %>%
   theme_bw(base_size=24, base_family="Lato")  +
   labs(x = "average recall",
        y = "average precision",
-       tag = "B",
+       tag = "b",
        title = "5 granulocytes")
 ggsave("Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot.pdf", device = cairo_pdf, width=8, height=7.5)
 
@@ -370,7 +370,7 @@ ggplot(avg_per_software_and_parameter %>%
   theme_bw(base_size=24, base_family="Lato") +
   labs(x = "average recall",
        y = "average precision",
-       tag = "B",
+       tag = "b",
        title = "5 granulocytes")
 ggsave("Laehnemann2017_prosolo-monovar-scansnv-sccaller-sciphi_precision-recall-plot_focus-top-left.pdf", device = cairo_pdf, width=8, height=7.5)
 

### 2021-09-09 manuscript/figs/Wang2014/Wang2014_plot_prosolo-monovar-scansnv-sccaller-scvilp.precision-recall.R

@@ -238,6 +238,15 @@ for (b in unique( cells %>% pull(batch) ) ) {
 fils = levels(avg_per_software_batch_and_parameter$filter)
 
 axis_limits <- c(0.0000001,1)
+reduced_software_pal <- software_pal %>%
+  enframe() %>% 
+  filter(
+    name == "ProSolo default" |
+      name == "ProSolo imputation" |
+      name == "SCIPhI default" |
+      name == "SCIPhI sensitive"
+    ) %>%
+  deframe()
 
 ggplot(avg_per_software_batch_and_parameter %>%
          filter(
@@ -253,7 +262,7 @@ ggplot(avg_per_software_batch_and_parameter %>%
          )
   ) + 
   geom_point(aes(x = num_filter, y = avg_FDR, color = software, shape = filter), size = 4) +
-  geom_abline(intercept = 0) +
+  geom_segment(aes(x = 0.000001, y = 0.000001, xend = 1, yend = 1)) +
   scale_x_continuous(
     trans = "log10",
     limits=axis_limits
@@ -263,9 +272,9 @@ ggplot(avg_per_software_batch_and_parameter %>%
     limits=axis_limits
   ) +
   scale_color_manual(
-    values = software_pal,
+    values = reduced_software_pal,
     labels = software_labels,
-    drop = FALSE
+    drop = TRUE
   ) +
   scale_shape_manual(
     name = "theoretical FDR",
@@ -279,7 +288,7 @@ ggplot(avg_per_software_batch_and_parameter %>%
   labs(
     x = "theoretical FDR",
     y = "ground truth FDR",
-    tag = "C",
+    tag = "c",
     title = "TNBC"
   )
 ggsave(
@@ -288,6 +297,20 @@ ggsave(
   width=16,
   height=7.5
 )
+
+basic_software_pal <- software_pal %>%
+  enframe() %>%
+  filter(
+    name %in% basic_software_levels
+  ) %>%
+  deframe()
+
+basic_linetype_pal <- linetype_pal %>%
+  enframe() %>%
+  filter(
+    name %in% basic_software_levels
+  ) %>%
+  deframe()
 
 ggplot(avg_per_software_batch_and_parameter %>%
     filter(
@@ -299,18 +322,18 @@ ggplot(avg_per_software_batch_and_parameter %>%
   coord_cartesian( ylim = c(0,1), xlim = c(0,1) ) +
   geom_line( aes( color = software, linetype = software ) ) +
[…]

### 2021-09-09 manuscript/figs/Wang2014/Wang2014_plot_prosolo-monovar-sccaller-sciphi_som_gt_recall.R

@@ -1,5 +1,6 @@
 # packages, colors, palettes and stuff
 source("../common.R")
+library(knitr)
 
 tumor_cells <- c(
     "a1",
@@ -168,19 +169,44 @@ expected_numbers <- validated_somatics %>%
     software = "expected"
   )
  
-per_software <- recall_per_cell %>%
+counts_per_software <- recall_per_cell %>%
   count(software, CHROM, POS, REF, ALT, CLONALITY, GT, name = "number_of_cells") %>%
   # add the expected genotype into software records
   full_join(
     expected_numbers %>%
+      rename(expected_number_of_cells = number_of_cells) %>%
       select(
         -software,
-        -number_of_cells,
         -zygosity,
         -GT
       ),
     by = c("CHROM", "POS", "REF", "ALT", "CLONALITY")
-    ) %>%
+    )
+  
+overimputed <- counts_per_software %>%
+  semi_join(
+    counts_per_software %>%
+    filter(
+      CLONALITY == "subclonal",
+      software == "SCIPhI default",
+      GT == expected_GT,
+      number_of_cells > expected_number_of_cells + 2
+    ),
+    by = c("CHROM", "POS", "REF", "ALT", "CLONALITY", "expected_number_of_cells", "Duplex_Freq", "expected_GT")
+  ) %>%
+  filter(
+    GT != "0/0"
+  ) %>%
+  pivot_wider(
+    names_from = c(software, GT),
+    names_prefix = "cells_",
+    values_from = number_of_cells,
+    values_fill = 0
+  ) %>%
+  arrange(CHROM, POS) %>%
+  kable(format = "latex")
+ 
+per_software <- counts_per_software %>%
   # remove incorrect calls
   filter(GT == expected_GT) %>%
   # restrict to variants in the truth set

### 2021-09-09 manuscript/figs/common.R

@@ -120,14 +120,14 @@ software_labels <- c(
    "ProSolo def 16X bulk" = "ProSolo def 16X bulk",
    "ProSolo def  4X bulk" = "ProSolo def  4X bulk",
    "ProSolo def  0X bulk" = "ProSolo def  0X bulk",
-   "ProSolo imputation" = bquote('ProSolo imputation'^-B),
+   "ProSolo imputation" = bquote('ProSolo imputation'^-b),
    "SCAN-SNV sensitive" = "SCAN-SNV sensitive",
    "SCcaller default dbsnp" = "SCcaller default dbsnp",
    "SCcaller sensitive dbsnp" = "SCcaller sensitive dbsnp",
    "SCIPhI default" = "SCIPhI default",
-   "SCIPhI sensitive" = bquote('SCIPhI sensitive'^-B),
+   "SCIPhI sensitive" = bquote('SCIPhI sensitive'^-b),
    "scVILP default" = "scVILP default",
-   "scVILP sensitive" = bquote('scVILP sensitive'^-B)
+   "scVILP sensitive" = bquote('scVILP sensitive'^-b)
 )
 
 shape_pal <-
